### 2020-07-17 manuscript_00.Rmd
setup doc for writing

---
title: 'Integrated Ocular Million Single Cell Meta-Atlas for Discovery and Development'
author:
  - Vinay Swamy:
      institute: bg
  - Zachary Batz:
      institute: nnrl
  - Temesgen Fufa:
      institute: mgog
  - Robert Hufnagel:
      institute: mgog
  - David McGaughey:
      institute:
        - bg
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - bg: Bioinformatics Group, Ophthalmic Genetics & Visual Function Branch, National Eye Institute, National Institutes of Health
  - nnrl: Neurobiology, Neurodegeneration & Repair Laboratory, National Eye Institute, National Institutes of Health
  - mgog: Medical Genetics and Ophthalmic Genomics Unit, National Eye Institute, National Institutes of Health

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: massive_integrated_eye_scRNA.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To curate a accessible and reliable ocular cell-type specific transcriptome dataset of over one million normal vertebrate ocular cell types with a matching high performance reactive web application to query gene expression across cell type, species, study, and other factors. METHODS: We queried several public repositories to find over a dozen healthy, non-perturbed, ocular tissue single cell RNA-seq studies As a non-ocular reference, we also downloaded the pan-mouse Tabula Muris raw data. The XX ocular and XX Tabula Muris cells were sent into a Snakemake-based reproducible pipeline we wrote to quantify transcripts, remove cells with poor quality, normalize expression values, creates a batch-corrected low dimensional space, performs XX differential expression tests, and outputs all as a single SQLite database file: the Single Cell Eye in a Disk (scEiaD) dataset. XXX thousand published cell labels assignments were curated and a machine learning  projected onto the non-labelled cells. Furthermore, we wrote the web application PLAE (https://plae.nei.nih.gov) to display scEiaD. RESULTS: The new portal provides quick visualization of vertebrate eye-related transcriptomes published to date by gene/transcript, XX cell types, XX ocular tissues, XX body tissues. As a test of the value of this unified pan-eye dataset, we showed XX. CONCLUSION: The PLAE v1.0 web app serves the pan-ocular and body dataset, scEiaD. This offers the eye community a powerful and quick means to test hypotheses on human gene and transcript expression across 54 body and 19 eye tissues. Futhermore the XXX labelled cells across XX studies, X techologies, and 3 species provides a highly challenging benchmarking resource for single cell algorithm development."
keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
---

```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)

# setup caption-ing
fig_cap <- captioner("Figure")
supFig_cap <- captioner("Supplemental Figure")
tab_cap <- captioner("Table")
supTab_cap <- captioner("Supplemental Table")

```

### 2020-08-04 manuscript_00.Rmd
text update

@@
 abstract: "PURPOSE: To curate a accessible and reliable ocular cell-type specific transcriptome dataset of over one million normal vertebrate ocular cell types with a matching high performance reactive web application to query gene expression across cell type, species, study, and other factors.
-METHODS: We queried several public repositories to find over a dozen healthy, non-perturbed, ocular tissue single cell RNA-seq studies As a non-ocular reference, we also downloaded the pan-mouse Tabula Muris raw data.
-The XX ocular and XX Tabula Muris cells were sent into a Snakemake-based reproducible pipeline we wrote to quantify transcripts, remove cells with poor quality, normalize expression values, creates a batch-corrected low dimensional space, performs XX differential expression tests, and outputs all as a single SQLite database file: the Single Cell Eye in a Disk (scEiaD) dataset.
+METHODS: We queried several public repositories to find eighteen healthy, non-perturbed, ocular tissue single cell RNA-seq studies As a non-ocular reference, we also used the pan-mouse Tabula Muris resource.
+The XX ocular and XX Tabula Muris cells were sent into a Snakemake-based reproducible pipeline we wrote to quantify transcripts, remove cells with poor quality, normalize expression values, create a batch-corrected low dimensional space, performs XX differential expression tests, and outputs all as a single SQLite database file: the Single Cell Eye in a Disk (scEiaD) dataset.
 XXX thousand published cell labels assignments were curated and a machine learning  projected onto the non-labelled cells.
@@
 ```
+
+# Introduction
+## Many published ocular atlases
+
+## Data not easily available for all
+
+## Some have accompanying apps
+
+## But apps are generally slow and have minimal exploration tooling
+
+## Cross comparisons difficult
+
+## If exploring a gene expression pattern, crucial to have reproducibility across disparate groups to enhance confidence that effect is biological
+
+## Single cell datasets getting larger
+Valentine Svensson figure?
+MOCA atlas (mouse 2e6 cells?)
+
+## Thus crucial for algorithms to be highly performant
+Our multidimensional atlas of 1e6 cells will be a useful reference dataset in:
+  - batch effect correction tools
+  - clustering algorithms
+  - cell type prediction / projection
+  - trajectory inference
+
+## What we did
+We have built an integrated meta-atlas that:
+
+  Has a batch corrected lower dimensional space (8 dimensions) from scVI
+
+  Curated previously labelled cell types for >400,000 cells.
+
+  Projected these onto the remaining unlabelled cells.
+
+  Made 77 clusters
+   - subclustered each to create XX total clusters
+
+  Pre-calculated pseudobulk differential expression testing in 9 different testing schemes:
+
+## Will do???
+
+Trajectory
+
+Velocity
+
+# Methods
+
+## Identify ocular scRNA studies
+SRA
+EBI
+Pubmed
+Found 18 (?) studies
+
+## pan mouse as non-ocular reference
+Tabula Muris
+
+## Quantification
+
+kallisto/bustools
+
+## Batch normalization
+Tested:
+  - scVI
+  - CCA
+  - insct
+  - combat
+  - nothing
+  - magic
+  - scanorama
+  - harmony
+  - fastMNN
+
+## Grid search to pick optimal parameters
+
+Varied:
+  - number of Hyper Variable Genes (HVG): 1000, 2000,5000,1000
+  - dims: 4,6,8,10,20,30,50,100
+  - normalization: scran, seurat standard, libSize, sqrt, counts (scVI only)
+  - nneighbors (UMAP viz only)
+  - knn: 5,7,10 (clustering)
+
+benchmarked with:
+  - LISI
+  - ARI
+  - Silhouette
+  - PCA variance explained before/after correction
+
+## Clustering
+Louvain-jaccard from Seurat
+
+##
+
+# Results
+
+## pseudobulk differential expression testing
+  - celltype (published) against remaining
+  - celltype (published) pairwise against celltype (published)
+  - species specific within celltype
+  - same for celltype (projected)
+  - same for cluster
+
+# Conclusion

### 2020-08-06 src/pull_data.R
script started to pull from scEiaD sqlite

@@ -1,0 +1,9 @@
+# Extract data for manuscript figures
+
+library(pool)
+library(RSQLite)
+library(tidyverse)
+
+scEiaD <- dbPool(drv = SQLite(), dbname = "~/data/massive_integrated_eye_scRNA/MOARTABLES__anthology_limmaFALSE___Mus_musculus_Macaca_fascicularis_Homo_sapiens-5000-counts-TabulaDroplet-batch-scVI-8-0.1-15-7.sqlite", idleTimeout = 3600000)
+
+meta_filter <- scEiaD %>% tbl('metadata_filter')

### 2020-11-03 manuscript_00.Rmd
script started to pull from scEiaD sqlite
text updates

--- 2020-07-17 manuscript_00.Rmd
@@
 ---
-title: 'Integrated Ocular Million Single Cell Meta-Atlas for Discovery and Development'
+title: 'An Million Transcriptome Ocular Meta-Atlas for Discovery and Development'
 author:
@@
       institute: bg
-  - Zachary Batz:
-      institute: nnrl
   - Temesgen Fufa:
@@
   - bg: Bioinformatics Group, Ophthalmic Genetics & Visual Function Branch, National Eye Institute, National Institutes of Health
-  - nnrl: Neurobiology, Neurodegeneration & Repair Laboratory, National Eye Institute, National Institutes of Health
   - mgog: Medical Genetics and Ophthalmic Genomics Unit, National Eye Institute, National Institutes of Health
@@
 csl: investigative-ophthalmology-and-visual-science.csl
-abstract: "PURPOSE: To curate a accessible and reliable ocular cell-type specific transcriptome dataset of over one million normal vertebrate ocular cell types with a matching high performance reactive web application to query gene expression across cell type, species, study, and other factors.
-METHODS: We queried several public repositories to find over a dozen healthy, non-perturbed, ocular tissue single cell RNA-seq studies As a non-ocular reference, we also downloaded the pan-mouse Tabula Muris raw data.
-The XX ocular and XX Tabula Muris cells were sent into a Snakemake-based reproducible pipeline we wrote to quantify transcripts, remove cells with poor quality, normalize expression values, creates a batch-corrected low dimensional space, performs XX differential expression tests, and outputs all as a single SQLite database file: the Single Cell Eye in a Disk (scEiaD) dataset.
-XXX thousand published cell labels assignments were curated and a machine learning  projected onto the non-labelled cells.
-Furthermore, we wrote the web application PLAE (https://plae.nei.nih.gov) to display scEiaD.
-RESULTS: The new portal provides quick visualization of vertebrate eye-related transcriptomes published to date by gene/transcript, XX cell types, XX ocular tissues, XX body tissues.
-As a test of the value of this unified pan-eye dataset, we showed XX.
-CONCLUSION: The PLAE v1.0 web app serves the pan-ocular and body dataset, scEiaD.
-This offers the eye community a powerful and quick means to test hypotheses on human gene and transcript expression across 54 body and 19 eye tissues.
-Futhermore the XXX labelled cells across XX studies, X techologies, and 3 species provides a highly challenging benchmarking resource for single cell algorithm development."
-keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
+abstract: "The development of highly scalable single cell transcriptome technology has allowed for a high resolution view into huge numbers of single cell transcriptomes.
+Analyzing the transcriptomes between different projects is highly desirable as this would theoretically allow for better assessment of which effects are consistent across independent studies.
+However it is difficult to compare and contrast data across different projects as there are substantial batch effects from computational processing, single cell technology utilized, and the natural biological variation.
+While many single cell transcriptome specific batch correction methods purport to remove the technical noise it is difficult to ascertain which method works best.
+We developed an R package (scPOP) that can scan the results of multiple methods and key parameter choices and create a ranked list of the best performing methods and parameters.
+We use this package along with a Snakefile based workflow system to demonstrate how to optimally merge over 1.4 million cells from 30 studies and three species to create a massive ocular single cell transcriptome meta-atlas.
+This provides a model how to efficiently create meta-atlases for tissues and cells of interest."
+keywords: "RNA-seq, retina, RPE, ocular, eye, Snakemake, singlecell, scRNA, single-cell, singlecell, scPOP, R"
 ---
@@
 library(captioner)
-
+library(pool)
+library(RSQLite)
+library(formattable)
 # setup caption-ing
@@
 
+scEiaD_2020_v01 <- dbPool(drv = SQLite(), dbname = "~/data/massive_integrated_eye_scRNA/MOARTABLES__anthology_limmaFALSE___Mus_musculus_Macaca_fascicularis_Homo_sapiens-0-2000-counts-TabulaDroplet-batch-scVI-8-0.001-500-0.6.sqlite", idleTimeout = 3600000)
+
+meta_filter <- fst::read_fst('~/git/plaeApp/inst/app/www/meta_filter.fst') %>% as_tibble()
+
+# cells before QC removal (mito / likely doublets)
+load('~/data/massive_integrated_eye_scRNA/cell_info_labelled.Rdata')
+
+# meta data before doublet removal
+load('~/data/massive_integrated_eye_scRNA/Mus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features2000__counts__TabulaDroplet__batch__scVI__dims8__preFilter__mindist0.001__nneighbors500.umap.Rdata')
+
+methods <- factor(c('scArches', 'bbknn','insct','magic', 'scVI','CCA', 'scanorama', 'harmony', 'fastMNN', 'combat', 'none', 'liger'), levels = c('bbknn','CCA','combat','fastMNN','harmony','insct','liger','magic','scanorama','scArches','scVI', 'none'))
 ```
+
+# Introduction
+## Many published ocular atlases
+
+The recent explosion of research using single cell transcriptome technology has led to insights in development, disease progression, etc.
+Some of the seminal work in the single cell transcriptomics field (Macosko et al) have used the retina.
+For example in the ocular community, over a dozen studies have used single cell technology to profile cell type patterns, etc.
+
+DISCUSS RETINA DEVELOPMENT/CELL TYPE stuff
+
+## Data not easily available for all
+
+While these groups has leveraged this new technology to [do stuff], there is a large chasm to cross the enable outside groups to use the data in new ways.
+Several of these papers are accompanied by reactive web apps, which allow for relatively quick checks of gene expression across cell type or cluster assignment.
+However these web apps have minimal to no data exploration tools and are slow to operate, especially with the larger datasets.
+
+## Impossible to compare results across independent studies
+Most crucially, these paper data web portals, should they exist, are silos which only contain the single cell transcriptomes related to their study.
+Thus it is impossible to compare directly across datasets.
+When exploring a gene expression pattern, it is crucial to have reproducibility across disparate groups to enhance confidence that effect is biological as bench experiments require large investments in time.
+
+While lightly processed counts data is often provided, differences in both single cell technology (e.g.
+droplet or well based) and bioinformatic processing choices make it impractical to quickly aggregate the datasets together as differences found may be technical.
+
+## There is a path 
+As the raw sequence data is required to be deposited on publication, it is possible to remove the technical differences due to bioinformatic processing choices by re-processing all of the data in a unified framework, though this still requires a large amount of bioinformatic expertise and compute infrastructure.
+Also there still remain technical (or batch) effects due to the variety in single cell technologies available and variation in tissue handling and processing across different scientific groups.
+Fortunately the single cell community has recognized that removal of technical (also referred to as batch) effects is a crucial issue and have independently developed many tools.
+
+## The meta-atlas
+
+We propose that by re-processing publicly available raw single cell transcriptome data in a consistent bioinformatic framework and optimally applying batch correction methodology we can create a meta-atlas, which combines individual datasets into one unified (or meta-) atlas.
+As there are thousands of possible permutations of single cell tools, references, and parameter choices, we create our meta-atlas (which we will refer to as the single cell eye in a disk or scEiaD) by benchmarking across multiple important integration parameters (integration method, number of hyper-variable genes, etc).
+The benchmarking system we developed uses a wide variety of metrics and we make available to all via an R package scPOP (single-cell Pick Optimal Parameters).
+The scEiaD will be of utility to two communities: the ocular community who can search scEiaD for gene expression across many dimensions (e.g.
+cluster, cell type, study) and the single cell informatic community who can use this very large, well-curated dataset to test algorithms for compute efficiency and performance in a diverse environment.
+
+<!-- ## As Single cell datasets getting larger, crucial for algorithms to be highly performant -->
+<!--   - Valentine Svensson figure? -->
+<!--   - MOCA atlas (mouse 2e6 cells?) -->
+
+<!-- Our multidimensional atlas of 1e6 cells will be a useful reference dataset in: -->
+<!--   - batch effect correction tools -->
+<!--   - clustering algorithms -->
+<!--   - cell type prediction / projection -->
+<!--   - trajectory inference -->
+
+# Results
+
+## We identify `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` ocular scRNA datasets across 3 species
+
+We identified ocular single cell RNA seq (scRNA) studies by querying PubMed, the Sequence Read Archive (SRA), and the European Nucleotide Archive (ENA) for the inclusive terms "retina", "single cell", "scRNA", "ocular", "eye", "transcriptome" and hand filtering the results to only keep normal (non-perturbed or mutagenized) data from single cell RNA-seq technology.
+We identified `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` deposited datasets that have been published in `r meta_filter %>% pull(Citation) %>% unique() %>% length()` publications (`r supTab_cap(name = 'study_cell_count', display = 'cite')`).
+  To provide a non-ocular reference we also downloaded the raw sequence data for the Tabula Muris project.
+In cases where the fastq file from the SRA was not processed properly, we acquired the original bam files and re-extracted the fastq.
+After downloading all the data we had 5.7 TB across `r scEiaD_2020_v01 %>% tbl('metadata_filter') %>% pull(sample_accession) %>% unique() %>% length()` fastq file sets.
+
+```{r fig1, fig.width=13, fig.height=5, fig.cap=fig1_cap, echo=FALSE, message = FALSE}
+fig1_cap <- fig_cap(name='fig1', caption =  'a.
+Counts of (study) accession, published papers, and batches for each scRNA technlogy, split by organism.
+b.
+Simplified directed workflow of major steps in scEiaD creation from raw counts to gene counts, benchmarking optimal integration methods to produce batch corrected latent dimensions (LDims), then downstream analysis outputs like clustering, differential gene testing (DiffT), and 2D UMAP visualization.
+c.
+Cell type counts extracted from published studies for select cell types, split by species.
+Count of study accessions for each species overlaid on bar plot.')
+
+source('figs_and_tables/fig1.R')
+plot_grid(plot_grid(a, c, align = 'v',  axis = 'lr', ncol = 1, labels = c('','c')), b,  nrow = 1, rel_widths = c(1.1,1), labels = c('a','b'))
+```
+
+## `r cell_info_labels %>% nrow() %>% as.integer() %>% as.integer() %>% format(., big.mark=",", scientific=FALSE)` cells before quality control
+
+  Gene-level counts were quantified with the kallisto bustool pseudo-aligner.
+For the droplet-based technology data, the inflection point between empty and non-empty droplets was automatically determined using the DropletUtils barcodeRank function in R.
+After empty droplet removal, we had `r format((umap %>% nrow() %>% as.numeric()) - (meta_filter %>% nrow() %>% as.numeric()), big.mark=",", scientific=FALSE)` cells.
+We then created a Seurat v3 object, calculated the percent mitochondrial genes counts, and removed cells which had more than 10% mitochondrial reads across all gene counts.
+For the droplet-based data, we also removed cells which had more than 3000 unique genes detected as these are likely to be doublets.
+After these standard quality control steps we were left with `r format(umap %>% nrow, big.mark = ",", scientific=FALSE)` cells.
+
+## Extraction and curation of cell type labels
+
+  A core objective of many scRNA based studies is labeling distinct cell types.
+As this information is crucial to assess dataset integration and provide an accurate reference for user querying, we extracted individual cell labels with a combination of hand-searching the Gene Expression Omnibus, supplemental information from the publication, web resources (if an web app was made available), and personal correspondence.
+After normalizing cell type names, we obtained labels for `r meta_filter$CellType %>% table() %>% sum()` cells across `r meta_filter$CellType %>% table() %>% length()` cell types.
+
+## Integration benchmarking of `r methods %>% length()` tools
+
+  A key step in extracting biological information out of the unified dataset is to remove the technical effects from the data.
+A wide variety of tools have been designed for this function.
+As we were uncertain which would perform the best and it is difficult to manually assess performance we designed a R-based benchmarking system to evaluate performance (scPOP).
+There are two key metrics which have to be balanced in order to optimize performance: batch mixing (e.g.
+different studies with the same cell type should be similar) and cell type or cluster purity (where different cell types or clusters should be distinct).
+While these can be visually assessed by looking at marker gene expression across the 2D t-SNE or UMAP projection, it is more rigorous and scalable to quantify this process.
+  We define batch as being each biological sample.
+We assume each study is at least one unique sample.
+We also studied the metadata and methods of each study to determine if the particular study contained multiple samples.
+In the end we identified `r meta_filter$batch %>% table() %>% length()` batches to be integrated.
+  The Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, were used to assess integration performance.
+We also use the Adjusted Rand Index (ARI), Normalized Mutual Information (NMI), and PCR metrics as implemented by scIB.
+The LISI and ASW were used to measure batch mixing (where lower is better), celltype mixing (higher is better), and cluster mixing (higher and better).
+NMI and ARI were used to assess the consistency of celltype to cluster assignment (where 1 is perfect correspondence between cluster and cell type).
+The PCR metric (higher is better) runs PCA before and after integration to assess how much variance is explained by the batch corrected dimensional space.
+  To visualize the interplay between batch mixing and cell type distinction we plot in FIG XXXXX the batch mixing LISI score (which has been multiplied by -1) was plotted on the y-axis (higher is better) against the cell type LISI on the x-axis (higher is better).
+The best performer on both metrics will be in the top right corner.
+
+## scVI has highest integration performance
+
+  To rank the integration algorithms across the various benchmarks, each score was Z scale normalized so that each score has an equal weight.
+The scores were then summed which allows us to rank the performance of each integration tool across the benchmarks.
+scVI has the strongest performance, with a sum score of SOMETHING.
+The next highest performers are Harmony, fastMNN, and Scanorama.
+
+## Further optimization of scVI with grid search
+
+  To find the best set of parameters for integration and clustering we did a grid search across key parameters: hyper variable genes (HVG), latent dimensions, and k nearest neighbors.
+Using scPOP we found the optimal parameters to be 2000 HVG, 8 latent dimensions, and a resolution of 0.6 with the leiden algorithm.
+We also varied the UMAP projection values of nearest neighbors and minimum distance to qualitatively pick a 2D projection, selecting a minimum distance of 0.001 and 500 nearest neighbors.
+    While we removed droplet-based cells with more than 3,000 transcripts as these are highly likely to be doublets (more than one cell per droplet) there are likely many more doublets.
+We ran DoubletDetect and scrublet and calculated the distribution of DoubletDetect and scrublet scores across all clusters and removed clusters with a DoubletDetect and scrublet score more than 4 standard deviations above the mean.
+This removed another `r format((umap %>% nrow() %>% as.integer()) - (meta_filter %>% nrow() %>% as.integer()), big.mark=",", scientific=FALSE)` cells, leaving `r format((meta_filter %>% nrow()) %>% as.integer(), big.mark=",", scientific=FALSE)` in total.
+
+## xgboost model built to label unknown cell types
+  To further study cell type specific expression patterning we needed to label the `r format(umap %>% nrow() - meta_filter %>% filter(!is.na(CellType)) %>% nrow(), big.mark=",", scientific=FALSE)` unlabelled cells.
+Traditionally this is done by clustering the cells, then using cell type specific markers to label the clusters.
+However, as we had hundreds of thousands of pre-labelled cells we decided to leverage the expert labels by building a xgboost-based machine learning model that used the half of the labeled cells as a training set (see methods for more details).
+The model was used to predict the cell type assignments for all cells in scEiaD.
+In this manner we both label most cells (cells which cannot be assigned a cell type assignment with a confidence above 0.5 were left unlabelled) and correct the small number of mislabels in the truth set.
+
+  ~~ALLUVIAL PLOT FIGURE~~
+
+## scEiaD UMAP projection blends batches and separates cell types
+  The 2D UMAP projection of the 8 scVI-calculated latent dimensional blends the `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` studies together while also maintaining distinct space for the `r meta_filter$CellType_predict %>% table() %>% length()` unique cell types.
+We see the neurogenic population from which the retinal cell types are derived near the center of the UMAP visualization.
+Cones and rods, which are the photoreceptors of the retina which are responsible for color and low-light vision, respectively, are near each other in the UMAP.
+The amacrine cells ....say more about cell type in space
+
+## diff testing reveals....
+pseudo and haystack
+
+## Pan mouse as non-ocular reference
+
+# Methods
+
+## Quantification
+
+kallisto/bustools
+
+## Batch normalization
+Tested:
+- scVI
+- CCA
+- insct
+- combat
+- nothing
+- magic
+- scanorama
+- harmony
+- fastMNN
+- scArches
+
+## Grid search to pick optimal parameters
+
+Varied:
+- number of Hyper Variable Genes (HVG): 1000, 2000,5000,1000
+- dims: 4,6,8,10,20,30,50,100
+- normalization: scran, seurat standard, libSize, sqrt, counts (scVI only)
+- nneighbors (UMAP viz only)
+- knn: 5,7,10 (clustering)
+
+benchmarked with:
+- LISI
+- ARI
+- Silhouette
+- PCA variance explained before/after correction
+
+## Clustering
+Louvain-jaccard as implemented by Seurat
+
+## pseudobulk differential expression testing
+- celltype (published) against remaining
+- celltype (published) pairwise against celltype (published)
+- species specific within celltype
+- same for celltype (projected)
+- same for cluster
+
+# Conclusion

### 2020-11-05 manuscript_00.Rmd
text updates

@@
 While many single cell transcriptome specific batch correction methods purport to remove the technical noise it is difficult to ascertain which method works best.
-We developed an R package (scPOP) that can scan the results of multiple methods and key parameter choices and create a ranked list of the best performing methods and parameters.
-We use this package along with a Snakefile based workflow system to demonstrate how to optimally merge over 1.4 million cells from 30 studies and three species to create a massive ocular single cell transcriptome meta-atlas.
+We developed an R package (scPOP) that can scan the results of multiple methods and key parameter choices and create a ranked list of the best performing method / parameter combinations.
+We use this package along with a Snakefile based workflow system to demonstrate how to optimally merge over 1.4 million cells from 30 datsets and three species to create a massive ocular single cell transcriptome meta-atlas.
 This provides a model how to efficiently create meta-atlases for tissues and cells of interest."
@@
 methods <- factor(c('scArches', 'bbknn','insct','magic', 'scVI','CCA', 'scanorama', 'harmony', 'fastMNN', 'combat', 'none', 'liger'), levels = c('bbknn','CCA','combat','fastMNN','harmony','insct','liger','magic','scanorama','scArches','scVI', 'none'))
+
+qc <- readr::read_tsv('~/data/massive_integrated_eye_scRNA/QC.tsv.gz')
 ```
@@
 # Introduction
-## Many published ocular atlases
+## Explosion of single cell transcriptomic atlases
+The  recent introduction of relatively low cost and high throughput single cell sequencing technlogy has led to an explosion of research across many fields.
+GIVE SOME EXAMPLES.
+As of 2020, over 30 million cells have been sequenced across n studies (Svennson Pachter http://www.nxn.se/single-cell-studies)
 
-The recent explosion of research using single cell transcriptome technology has led to insights in development, disease progression, etc.
-Some of the seminal work in the single cell transcriptomics field (Macosko et al) have used the retina.
-For example in the ocular community, over a dozen studies have used single cell technology to profile cell type patterns, etc.
+## Many in the eye 
+  Some of the seminal work in the high throughput single cell transcriptomics field (Macosko et al) have used the retina.
+As of late 2020, over a dozen published studies have used single cell technology to profile cell type specific gene expression patterns, developmental trends, species specific differentiation, and disease alterations.
 
-DISCUSS RETINA DEVELOPMENT/CELL TYPE stuff
+## Impossible to easily compare results across independent studies
+Study specific web portals, should they exist, are silos which only contain the single cell transcriptomes related to their study.
+There are notable efforts to create web portals (UCSC Single Cell Browser, EBI Single Cell Expression Atlas) that allow for GUI-based viewing of data for most publicly available datasets.
+However, these again have the shortcoming as the study-specific web portals discussed above in that you still cannot directly compare between independent studies.
+In these situations it is impossible to compare directly across datasets.
+When exploring a gene expression pattern, it is crucial to have reproducibility across disparate groups to enhance confidence that effect is biological as bench experiments require large investments in time.
 
-## Data not easily available for all
+<!-- ## Aggregation of results to create meta-atlas important -->
 
-While these groups has leveraged this new technology to [do stuff], there is a large chasm to cross the enable outside groups to use the data in new ways.
+<!-- While these groups has leveraged this new technology to [do stuff], there is a large chasm to cross the enable outside groups to use the data in new ways.
 Several of these papers are accompanied by reactive web apps, which allow for relatively quick checks of gene expression across cell type or cluster assignment.
 However these web apps have minimal to no data exploration tools and are slow to operate, especially with the larger datasets.
+-->
 
-## Impossible to compare results across independent studies
-Most crucially, these paper data web portals, should they exist, are silos which only contain the single cell transcriptomes related to their study.
-Thus it is impossible to compare directly across datasets.
-When exploring a gene expression pattern, it is crucial to have reproducibility across disparate groups to enhance confidence that effect is biological as bench experiments require large investments in time.
-
-While lightly processed counts data is often provided, differences in both single cell technology (e.g.
-droplet or well based) and bioinformatic processing choices make it impractical to quickly aggregate the datasets together as differences found may be technical.
+<!-- While lightly processed counts data is often provided, differences in both single cell technology (e.g.
+droplet or well based) and bioinformatic processing choices make it impossible to simply concatenate the datasets together as differences found will likely be technical.
+-->
 
@@
 Fortunately the single cell community has recognized that removal of technical (also referred to as batch) effects is a crucial issue and have independently developed many tools.
+The challenge is still to determine which tool works best.
+
+## scRNA datasets are rapidly increasing in size
+  - Valentine Svensson figure?
+  - http://www.nxn.se/single-cell-studies
+  - MOCA atlas (mouse 2e6 cells)
 
@@
 
-<!-- ## As Single cell datasets getting larger, crucial for algorithms to be highly performant -->
-<!--   - Valentine Svensson figure? -->
-<!--   - MOCA atlas (mouse 2e6 cells?) -->
-
-<!-- Our multidimensional atlas of 1e6 cells will be a useful reference dataset in: -->
-<!--   - batch effect correction tools -->
-<!--   - clustering algorithms -->
-<!--   - cell type prediction / projection -->
-<!--   - trajectory inference -->
-
 # Results
@@
 
-We identified ocular single cell RNA seq (scRNA) studies by querying PubMed, the Sequence Read Archive (SRA), and the European Nucleotide Archive (ENA) for the inclusive terms "retina", "single cell", "scRNA", "ocular", "eye", "transcriptome" and hand filtering the results to only keep normal (non-perturbed or mutagenized) data from single cell RNA-seq technology.
-We identified `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` deposited datasets that have been published in `r meta_filter %>% pull(Citation) %>% unique() %>% length()` publications (`r supTab_cap(name = 'study_cell_count', display = 'cite')`).
+  The first step in building a meta-atlas is identifying studies to draw the data from.
+We identified ocular single cell RNA seq (scRNA) studies by querying PubMed, the Sequence Read Archive (SRA), and the European Nucleotide Archive (ENA) for the inclusive terms "retina", "single cell", "scRNA", "ocular", "eye", "transcriptome" and then hand filtering the results to only keep normal (non-perturbed or mutagenized) data from single cell RNA-seq technology.
+We identified `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` deposited datasets that have been published in `r meta_filter %>% pull(Citation) %>% unique() %>% length()` publications (`r fig_cap(name = 'fig1', display = 'cite')`).
   To provide a non-ocular reference we also downloaded the raw sequence data for the Tabula Muris project.
@@
 b.
-Simplified directed workflow of major steps in scEiaD creation from raw counts to gene counts, benchmarking optimal integration methods to produce batch corrected latent dimensions (LDims), then downstream analysis outputs like clustering, differential gene testing (DiffT), and 2D UMAP visualization.
+Simplified directed workflow of major steps in scEiaD creation from raw counts to gene counts, benchmarking optimal integration methods (SnakePOP) to produce batch corrected latent dimensions (LatentDims), then downstream analysis outputs like clustering, differential gene testing (DiffTesting), and 2D UMAP visualization.
 c.
-Cell type counts extracted from published studies for select cell types, split by species.
+Cell type counts extracted from published studies for the more common retina cell types, split by species.
 Count of study accessions for each species overlaid on bar plot.')
@@
 
-## `r cell_info_labels %>% nrow() %>% as.integer() %>% as.integer() %>% format(., big.mark=",", scientific=FALSE)` cells before quality control
+## `r cell_info_labels %>% nrow() %>% as.integer() %>% format(., big.mark=",", scientific=FALSE)` cells before quality control
 
-  Gene-level counts were quantified with the kallisto bustool pseudo-aligner.
+  Gene-level counts were quantified with the kallisto bustool pseudo-aligner for both the droplet and well based samples.
 For the droplet-based technology data, the inflection point between empty and non-empty droplets was automatically determined using the DropletUtils barcodeRank function in R.
-After empty droplet removal, we had `r format((umap %>% nrow() %>% as.numeric()) - (meta_filter %>% nrow() %>% as.numeric()), big.mark=",", scientific=FALSE)` cells.
+After empty droplet removal, we had `r format(cell_info_labels %>% nrow() %>% as.integer() , big.mark=",", scientific=FALSE)` cells.
 We then created a Seurat v3 object, calculated the percent mitochondrial genes counts, and removed cells which had more than 10% mitochondrial reads across all gene counts.
@@
 As this information is crucial to assess dataset integration and provide an accurate reference for user querying, we extracted individual cell labels with a combination of hand-searching the Gene Expression Omnibus, supplemental information from the publication, web resources (if an web app was made available), and personal correspondence.
-After normalizing cell type names, we obtained labels for `r meta_filter$CellType %>% table() %>% sum()` cells across `r meta_filter$CellType %>% table() %>% length()` cell types.
+After normalizing cell type names, we obtained labels for `r meta_filter$CellType %>% table() %>% sum()` cells across `r meta_filter$CellType %>% table() %>% length()` cell types (`r supTab_cap(name = 'supTab1', display = 'cite')`).
 
-## Integration benchmarking of `r methods %>% length()` tools
+## Running `r methods %>% length()` tools in a Snakefile-based system
 
@@
 A wide variety of tools have been designed for this function.
-As we were uncertain which would perform the best and it is difficult to manually assess performance we designed a R-based benchmarking system to evaluate performance (scPOP).
+As we were uncertain which would perform the best and it is difficult to manually assess performance we ran `r methods %>% length()` tools with a commonly used set of key parameters that effect integration: 2000 hyper variable genes, the louvain clustering algorithm with 7 k-nearest neighbors
+
   There are two key metrics which have to be balanced in order to optimize performance: batch mixing (e.g.
@@
 In the end we identified `r meta_filter$batch %>% table() %>% length()` batches to be integrated.
-  The Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, were used to assess integration performance.
-We also use the Adjusted Rand Index (ARI), Normalized Mutual Information (NMI), and PCR metrics as implemented by scIB.
+
+# scPOP wraps several different methods at measuring integration performance.
+
+  Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, are used as well as the Adjusted Rand Index (ARI), Normalized Mutual Information (NMI), and PCR metrics as implemented by scIB.
 The LISI and ASW were used to measure batch mixing (where lower is better), celltype mixing (higher is better), and cluster mixing (higher and better).
@@
 
-  To rank the integration algorithms across the various benchmarks, each score was Z scale normalized so that each score has an equal weight.
-The scores were then summed which allows us to rank the performance of each integration tool across the benchmarks.
-scVI has the strongest performance, with a sum score of SOMETHING.
+  To rank the integration algorithms across the various benchmarks, scPOP Z scale normalizes all the metrics so that each score has an equal weight.
+scPOP then sums them all which allows us to rank the performance of each integration tool across the benchmarks.
+For users who have custom needs, the weights of each metrics can be customized to prioritize mixing or sepraration.
+scPOP finds that scVI has the strongest performance, with a sum score of SOMETHING.
 The next highest performers are Harmony, fastMNN, and Scanorama.
+We do not claim that these results are gegeral - rather we emphasize that for each batch correction analysis that scPOP or some similar approach be used to identify the optimal method for each dataset.
 
@@
 Traditionally this is done by clustering the cells, then using cell type specific markers to label the clusters.
-However, as we had hundreds of thousands of pre-labelled cells we decided to leverage the expert labels by building a xgboost-based machine learning model that used the half of the labeled cells as a training set (see methods for more details).
-The model was used to predict the cell type assignments for all cells in scEiaD.
-In this manner we both label most cells (cells which cannot be assigned a cell type assignment with a confidence above 0.5 were left unlabelled) and correct the small number of mislabels in the truth set.
+However, as we had hundreds of thousands of pre-labelled cells we leveraged the expert labels by building a xgboost-based machine learning model that used the half of the labeled cells as a training set (see methods for more details) to train a cell type predictor for scEiaD input.
+The trained model was used to predict the cell type assignments for all cells in scEiaD.
+In this manner we both label most cells (cells which cannot be assigned a cell type assignment with a confidence above 0.5 were left unlabelled) and correct a small number of probable mislabels in the truth set.
 
@@
 # Conclusion
+
+# Supplemental
+```{r supFig1, fig.width=4, fig.height=7, out.width=300, fig.cap=supFig1_cap, echo=FALSE, message = FALSE}
+supFig1_cap <- supFig_cap(name='supFig1_cap', caption =  'The left bar delineates the number of cells for each organism - technology combination.
+The right bar specifies the number of each cells in each post QC category.
+In silico doublets were identified with scrublet and DoubletDetector.')
+
+source('figs_and_tables/sup_fig1__qc_failure.R')
+supFig1_plot
+```
+
+```{r supTab1, fig.width=4, fig.height=7, out.width=300, fig.cap=supTab1_cap, echo=FALSE, message = FALSE}
+supTab1_cap <- supTab_cap(name='supTab1_cap', caption =  'Counts for cell type labels.
+Published are the author created labels from the published datasets.
+Transferred area the cell labels that were transferred by a machine learning model onto the entire scEiaD dataset.')
+
+source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
+ctTable
+```

### 2020-11-05 figs_and_tables/fig1.R

@@ -1,0 +1,72 @@
+library(tidyverse)
+library(tidygraph)
+library(ggdag)
+library(ggraph)
+library(cowplot)
+library(patchwork)
+a <- meta_filter %>% 
+  mutate(Platform = gsub('C1', 'SMARTSeq_v4', Platform)) %>%
+  group_by(organism, Platform) %>% 
+  summarise(accession = length(unique(study_accession)), paper = length(unique(PMID)), batch = length(unique(batch))) %>% 
+  pivot_longer(cols = accession:batch) %>% 
+  mutate(Count = value,
+         Type = name,
+         organism = gsub(' ','\n',organism)) %>%  
+  ggplot(aes(x=Platform, y=Count, fill = Type, group = Count)) + 
+  facet_wrap(~organism) + 
+  geom_bar(stat = 'identity', position = position_dodge2()) + 
+  cowplot::theme_cowplot() + 
+  coord_flip()
+
+dagified <- dagify('counts' ~ 'fastq',
+                   'SnakePOP' ~ 'CellType',
+                   'SnakePOP' ~ 'counts',
+                   'LatentDims' ~ 'SnakePOP',
+                   'cluster' ~ 'LatentDims',
+                   'umap' ~ 'LatentDims' ,
+                   'ML' ~ 'CellType',
+                   'projectedCellType' ~ 'ML',
+                   'ML' ~ 'LatentDims',
+                   'DiffTesting' ~ 'counts',
+                   'DiffTesting' ~ 'projectedCellType',
+                   'DiffTesting' ~ 'CellType',
+                   'DiffTesting' ~ 'cluster',
+                   'DiffTesting' ~ 'LatentDims',
+                   'projectedCellType' ~ 'LatentDims',
+                   'projectedCellType' ~ 'CellType')
+
+tidy_dagitty(dagified)
+set.seed(2534)
+b <- as_tbl_graph(dagified) %>% mutate(bop = name) %>% ggraph(layout = 'graphopt') + 
+  geom_edge_link(arrow = arrow(length = unit(3, 'mm')), 
+                 end_cap = circle(8, 'mm'), start_cap = circle(2, 'mm')) + 
+  #geom_node_point(size = 10) +
+  geom_node_label(aes(label = name), color = 'black', alpha = 1)  + 
+    theme_nothing()
+
+  
+
+# CellType
[…]

### 2020-11-05 figs_and_tables/sup_fig1__qc_failure.R

@@ -1,0 +1,46 @@
+library(ggalluvial)
+counts <- c(qc %>% 
+              mutate(QC = gsub(' QC','', QC),
+                     QC = gsub('Uni', '\nUni', QC), 
+                     QC = gsub('Doubl', '\nDoubl', QC), 
+                     TechType = case_when(Platform %in% c('10xv2','10xv3','DropSeq') ~ 'Droplet',
+                                          TRUE ~ 'Well')) %>% 
+              mutate(OrgTech = paste0(organism, '\n', TechType)) %>% 
+              mutate(OrgTech = gsub('fasci','\nfasci', OrgTech)) %>% 
+              group_by(OrgTech) %>% 
+              summarise(Count = n()) %>% pull(Count) %>% rev(),
+            qc %>% 
+              mutate(QC = gsub(' QC','', QC),
+                     QC = gsub('Uni', '\nUni', QC), 
+                     QC = gsub('Doubl', '\nDoubl', QC), 
+                     TechType = case_when(Platform %in% c('10xv2','10xv3','DropSeq') ~ 'Droplet',
+                                          TRUE ~ 'Well')) %>% 
+              mutate(OrgTech = paste0(organism, '\n', TechType)) %>% 
+              mutate(OrgTech = gsub('fasci','\nfasci', OrgTech)) %>% 
+              group_by(QC) %>% 
+              summarise(Count = n()) %>% pull(Count) %>% rev()
+) 
+
+supFig1_plot <- qc %>% 
+  mutate(QC = gsub(' QC','', QC),
+         QC = gsub('Uni', '\nUni', QC), 
+         QC = gsub('Doubl', '\nDoubl', QC), 
+         TechType = case_when(Platform %in% c('10xv2','10xv3','DropSeq') ~ 'Droplet',
+                              TRUE ~ 'Well')) %>% 
+  mutate(OrgTech = paste0(organism, '\n', TechType)) %>% 
+  mutate(OrgTech = gsub('fasci','\nfasci', OrgTech)) %>% 
+  group_by(OrgTech, QC) %>% 
+  summarise(Count = n()) %>% 
+  #ggplot(aes(y = sqrt(Count), axis1 = organism, axis2 = TechType, axis3 = QC)) +
+  ggplot(aes(y = sqrt(Count), axis1 = OrgTech, axis2 = QC, label = Count)) +
+  geom_alluvium(aes(fill = `QC`), alpha = 0.4) +
[…]

### 2020-11-05 figs_and_tables/sup_fig2__celltype_xgboost_labelling.R

@@ -1,0 +1,45 @@
+ct_processing <-  meta_filter %>% 
+  #filter(CellType %in% common) %>% 
+  filter(is.na(CellType) | CellType != 'Mast') %>% 
+  filter(!(!Platform %in% c('10xv2','10xv3','DropSeq') & CellType_predict == 'Unlabelled'  )) %>% 
+  filter(!(!Platform %in% c('10xv2','10xv3','DropSeq') & CellType_predict == 'Muller Glia Progenitor'  )) %>% 
+  mutate(CellType = case_when(is.na(CellType) ~ 'Unlabelled', 
+                              CellType == 'AC/HC_Precurs' ~ 'AC/HC Precursors',
+                              TRUE ~ CellType)) %>% 
+  mutate(CellType_predict = case_when(CellType_predict == 'RPC' ~ 'RPCs',
+                                      CellType_predict == 'AC/HC_Precurs' ~ 'AC/HC Precursors',
+                                      grepl('Mesenchymal', CellType_predict) ~ 'Endothelial',
+                                      is.na(CellType_predict) ~ 'Unlabelled', 
+                                      TRUE ~ CellType_predict)) 
+
+##########
+predictedCT <- ct_processing %>% 
+  group_by(organism, CellType_predict) %>% 
+  summarise(`Published Count` = n()) %>% pivot_wider(values_from = `Published Count`, names_from = c(organism))
+colnames(predictedCT) <- c('CellType','HS Transferred','MF Transferred','MM Transferred')
+prelabelledCT <- ct_processing %>% 
+  group_by(organism, CellType) %>% 
+  summarise(`Published Count` = n()) %>% pivot_wider(values_from = `Published Count`, names_from = c(organism))
+colnames(prelabelledCT) <- c('CellType','HS Published','MF Published','MM Published')
+joinedCT <- left_join(prelabelledCT, predictedCT)
+joinedCT[is.na(joinedCT)] <- 0
+ctTable <- joinedCT %>% flextable()
+##########
+
+ct_alluvial <- ct_processing %>% 
+  select(CellType, CellType_predict) %>% 
+  group_by(CellType, CellType_predict) %>% 
+  summarise(Count = n()) %>% 
+  ggplot(aes(y = sqrt(Count), axis1 = CellType, axis2 = CellType_predict)) +
+  geom_alluvium(aes(fill = `CellType`), alpha = 0.8) +
+  geom_stratum(alpha = 0) +
+  ggrepel::geom_label_repel(stat = "stratum", 
[…]

### 2020-11-05 figs_and_tables/sup_fig2__publishedCounts.R

@@ -1,0 +1,15 @@
+nxn_sc <- read_tsv('http://www.nxn.se/single-cell-studies/data.tsv')
+sum <- nxn_sc %>% 
+  mutate(Date = as.Date(x = as.character(Date), format = '%Y%m%d'), 
+         Year = format(Date, '%Y')) %>% 
+  group_by(Year) %>% 
+  summarise(Count = sum(`Reported cells total`, na.rm = TRUE)) %>% 
+  mutate(Sum = cumsum(Count)) %>% 
+  ggplot(aes(x=Year, y = Sum)) +
+  geom_bar(stat = 'identity') +
+  ylab("Sum of SC Counts") +
+  cowplot::theme_cowplot() + 
+  coord_flip()
+cowplot::plot_grid(sum, sum + scale_y_log10() + ylab('Log(Sum of SC Counts)'))
+
+

### 2020-11-07 manuscript_00.Rmd
text and fig updates

@@
 The  recent introduction of relatively low cost and high throughput single cell sequencing technlogy has led to an explosion of research across many fields.
-GIVE SOME EXAMPLES.
-As of 2020, over 30 million cells have been sequenced across n studies (Svennson Pachter http://www.nxn.se/single-cell-studies)
+As of 2020, over 30 million cells have been sequenced across over 1,000 studies (Svennson Pachter http://www.nxn.se/single-cell-studies).
+As of November 2020, the average size of each individual dataset is near 100,000 cells.
+Some of the seminal work in the high throughput single cell transcriptomics field (Macosko et al) have used the retina.
+As of late 2020, over a dozen published studies containing over a million cells have used single cell technology to profile cell type specific gene expression patterns, cell fate trajectory, tissue and cell differentiation, and disease perturbation.
 
-## Many in the eye 
-  Some of the seminal work in the high throughput single cell transcriptomics field (Macosko et al) have used the retina.
-As of late 2020, over a dozen published studies have used single cell technology to profile cell type specific gene expression patterns, developmental trends, species specific differentiation, and disease alterations.
+## Many researchers use processed count tables for methods development and hypothesis testing
 
-## Impossible to easily compare results across independent studies
-Study specific web portals, should they exist, are silos which only contain the single cell transcriptomes related to their study.
+While the gene - cell count tables are often made available in repositories like the Gene Expression Omnibus (GEO), there are no requirements to uniformly process the data.
+This means the count tables cannot be used in cross-study comparisons as even small differences in the computational pipeline (aligner, transcriptome reference, etc.) create study-specific effects.
+This issue can be addressed only by re-quantifying the data in a uniform environment.
+Fortunately, due to the continued development of computationally light-weight gene quantification tools into the single-cell space (kbtools, alevin-something), re-quantification does not require massive compute resources.
+
+Unfortunately, there still remain study specific batch effects due to the variety in single cell technologies used and variation in tissue handling and processing across different scientific groups.
+The single cell community has recognized that removal of these technical (also referred to as batch) effects is a critical issue and have independently developed many tools.
+It remains unclear which tool(s) is optimal for a particular dataset.
+
+<!-- Study specific web portals, should they exist, are silos which only contain the single cell transcriptomes related to their study.
 There are notable efforts to create web portals (UCSC Single Cell Browser, EBI Single Cell Expression Atlas) that allow for GUI-based viewing of data for most publicly available datasets.
@@
 When exploring a gene expression pattern, it is crucial to have reproducibility across disparate groups to enhance confidence that effect is biological as bench experiments require large investments in time.
+-->
 
@@
 
-## There is a path 
-As the raw sequence data is required to be deposited on publication, it is possible to remove the technical differences due to bioinformatic processing choices by re-processing all of the data in a unified framework, though this still requires a large amount of bioinformatic expertise and compute infrastructure.
-Also there still remain technical (or batch) effects due to the variety in single cell technologies available and variation in tissue handling and processing across different scientific groups.
-Fortunately the single cell community has recognized that removal of technical (also referred to as batch) effects is a crucial issue and have independently developed many tools.
-The challenge is still to determine which tool works best.
-
-## scRNA datasets are rapidly increasing in size
-  - Valentine Svensson figure?
-  - http://www.nxn.se/single-cell-studies
-  - MOCA atlas (mouse 2e6 cells)
-
 ## The meta-atlas
 
-We propose that by re-processing publicly available raw single cell transcriptome data in a consistent bioinformatic framework and optimally applying batch correction methodology we can create a meta-atlas, which combines individual datasets into one unified (or meta-) atlas.
-As there are thousands of possible permutations of single cell tools, references, and parameter choices, we create our meta-atlas (which we will refer to as the single cell eye in a disk or scEiaD) by benchmarking across multiple important integration parameters (integration method, number of hyper-variable genes, etc).
+We propose that by re-processing publicly available raw single cell transcriptome data in a consistent bioinformatic framework and optimally using batch correction tools we can create a meta-atlas, which combines individual datasets into one unified (or meta-) atlas.
+As there are thousands of possible permutations of single cell tools, references, and parameter choices, we create our meta-atlas (which we refer to as the single cell eye in a disk or scEiaD) by benchmarking integration outcomes across multiple important single cell RNA-seq processing parameters (integration tool, number of hyper-variable genes, clustering resolution, etc).
 The benchmarking system we developed uses a wide variety of metrics and we make available to all via an R package scPOP (single-cell Pick Optimal Parameters).
 The scEiaD will be of utility to two communities: the ocular community who can search scEiaD for gene expression across many dimensions (e.g.
-cluster, cell type, study) and the single cell informatic community who can use this very large, well-curated dataset to test algorithms for compute efficiency and performance in a diverse environment.
+cluster, cell type, study) and the computational community who can use this very large, well-curated dataset to test algorithms for compute efficiency and performance in a diverse environment.
+As we believe data re-use is a powerful and accessible means to drive forward research, we provide our meta-atlas code-base and propose general guidelines to optimally create custom meta-atlases.
 
@@
 We identified ocular single cell RNA seq (scRNA) studies by querying PubMed, the Sequence Read Archive (SRA), and the European Nucleotide Archive (ENA) for the inclusive terms "retina", "single cell", "scRNA", "ocular", "eye", "transcriptome" and then hand filtering the results to only keep normal (non-perturbed or mutagenized) data from single cell RNA-seq technology.
-We identified `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` deposited datasets that have been published in `r meta_filter %>% pull(Citation) %>% unique() %>% length()` publications (`r fig_cap(name = 'fig1', display = 'cite')`).
+As of September 2020 we identified `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` deposited datasets that have been published in `r meta_filter %>% pull(Citation) %>% unique() %>% length()` publications (`r fig_cap(name = 'fig1', display = 'cite')`).
 To provide a non-ocular reference we also downloaded the raw sequence data for the Tabula Muris project.
-In cases where the fastq file from the SRA was not processed properly, we acquired the original bam files and re-extracted the fastq.
+  In cases where the fastq file from the SRA was not processed properly, we acquired the original bam files (personal correspondence or the SRA) and re-extracted the fastq.
 After downloading all the data we had 5.7 TB across `r scEiaD_2020_v01 %>% tbl('metadata_filter') %>% pull(sample_accession) %>% unique() %>% length()` fastq file sets.
 
-```{r fig1, fig.width=13, fig.height=5, fig.cap=fig1_cap, echo=FALSE, message = FALSE}
+```{r fig1, fig.width=6, fig.height=3, fig.cap=fig1_cap, echo=FALSE, message = FALSE}
 fig1_cap <- fig_cap(name='fig1', caption =  'a.
@@
 A wide variety of tools have been designed for this function.
-As we were uncertain which would perform the best and it is difficult to manually assess performance we ran `r methods %>% length()` tools with a commonly used set of key parameters that effect integration: 2000 hyper variable genes, the louvain clustering algorithm with 7 k-nearest neighbors
+As we were uncertain which would perform the best and it is difficult to manually assess performance we ran `r methods %>% length()` tools with a commonly used set of key parameters that affect integration: 2000 hyper variable genes, 8 and 30 latent dimensions returned, and the louvain clustering algorithm with 7 k-nearest neighbors
 
@@
   Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, are used as well as the Adjusted Rand Index (ARI), Normalized Mutual Information (NMI), and PCR metrics as implemented by scIB.
-The LISI and ASW were used to measure batch mixing (where lower is better), celltype mixing (higher is better), and cluster mixing (higher and better).
-NMI and ARI were used to assess the consistency of celltype to cluster assignment (where 1 is perfect correspondence between cluster and cell type).
+The LISI and ASW were used to measure batch mixing (where lower is better), cell type mixing (higher is better), and cluster mixing (higher is better).
+NMI and ARI were used to assess the consistency of cell type to cluster assignment (where 1 is perfect correspondence between cluster and cell type).
 The PCR metric (higher is better) runs PCA before and after integration to assess how much variance is explained by the batch corrected dimensional space.
@@
   To rank the integration algorithms across the various benchmarks, scPOP Z scale normalizes all the metrics so that each score has an equal weight.
-scPOP then sums them all which allows us to rank the performance of each integration tool across the benchmarks.
-For users who have custom needs, the weights of each metrics can be customized to prioritize mixing or sepraration.
-scPOP finds that scVI has the strongest performance, with a sum score of SOMETHING.
+The scores are summed which allows us to rank the performance of each integration tool across the diverse benchmarks.
+For users who have custom needs, the weights of each metrics can be customized to prioritize mixing or separation.
+scPOP finds that for our dataset scVI has the strongest performance, with a sum score of SOMETHING.
 The next highest performers are Harmony, fastMNN, and Scanorama.
-We do not claim that these results are gegeral - rather we emphasize that for each batch correction analysis that scPOP or some similar approach be used to identify the optimal method for each dataset.
+We do not claim that these results are gegnral - rather we emphasize that for each (meta) atlas creation that scPOP or some similar approach be used to identify the optimal method for each dataset.
 
@@
 # Supplemental
+```{r supFig1, fig.width=2, fig.height=3, out.width=300, fig.cap=supFig1_cap, echo=FALSE, message = FALSE}
+supFig1_cap <- supFig_cap(name='supFig1_cap', caption =  'The left bar delineates the number of cells for each organism - technology combination.
+The right bar specifies the number of each cells in each post QC category.
+In silico doublets were identified with scrublet and DoubletDetector.')
+
+source('figs_and_tables/sup_fig1__qc_failure.R')
+supFig1_plot
+```
+
 ```{r supFig1, fig.width=4, fig.height=7, out.width=300, fig.cap=supFig1_cap, echo=FALSE, message = FALSE}
@@
 supFig1_plot
+```
+
+```{r supFig1, fig.width=5, fig.height=5, out.width=300, fig.cap=supFig2_cap, echo=FALSE, message = FALSE}
+supFig2_cap <- supFig_cap(name='supFig2_cap', caption =  '')
+
+source('figs_and_tables/fig2__integrationPerf.R')
+cowplot::plot_grid(
+  cowplot::plot_grid(lisi + theme(legend.position="none"), silhouette, nrow = 1), 
+  cowplot::plot_grid(ari, legend, rel_widths = c(1,0.2), ncol = 2), 
+  nrow = 2)
 ```

### 2020-11-07 figs_and_tables/fig2__integrationPerf.R

@@ -1,0 +1,103 @@
+load('~/data/massive_integrated_eye_scRNA/metrics_2020_10_07.Rdata')
+
+perf <- perf %>% 
+  mutate(knn = case_when(method == 'scArches' ~ 7, TRUE ~ knn))
+cluster_stats <- perf %>% select(clusterN, clusterMedian, dims, nf, knn, method, normalization, set) %>% unique() 
+
+perf_well <- perf %>% unique() %>% filter(set == 'onlyWELL') %>% 
+  select(Score, Group, Value, set, dims:normalization) %>% 
+  filter(Score %in% c('LISI','Silhouette')) 
+
+perf_tabula <- perf %>% unique() %>% filter(set == 'TabulaDroplet') %>% 
+  select(Score, Group, Value, set, dims:normalization) %>% 
+  filter(Score %in% c('LISI','Silhouette', 'ARI', 'PCR', 'NMI')) %>% 
+  filter(method != 'CCA')
+
+lisi <- perf_tabula %>% 
+  filter(Score == 'LISI', nf == 2000, dims %in% c(8,30), knn == 7) %>%
+  pivot_wider(values_from = Value, names_from = c('Group')) %>% 
+  ggplot(aes(y=Batch, x=-Cluster, shape = normalization)) + geom_point(aes(color=method), size = 5) +
+  cowplot::theme_cowplot() + scale_color_manual(values = pals::alphabet() %>% unname()) +
+  ggtitle('LISI')
+
+ari <- perf_tabula %>% 
+  filter(Score == 'ARI', Group == 'CellType-Cluster', nf == 2000, dims %in% c(8, 30), knn == 7) %>%
+  ggplot(aes(y=Value, x=normalization)) + geom_point(aes(color=method, shape = normalization), size = 5) +
+  cowplot::theme_cowplot() + scale_color_manual(values = pals::alphabet() %>% unname()) + ylab('ARI (higher is better)') +
+  ggtitle('ARI') + theme(legend.position="none")
+
+silhouette <- perf_tabula %>% filter(Score == 'Silhouette', nf == 2000, dims %in% c(8,30), knn == 7) %>% 
+  pivot_wider(values_from = Value, names_from = c('Group')) %>% 
+  ggplot(aes(y=-Batch, x=Cluster, shape = normalization)) + geom_point(aes(color=method), size = 5) +
[…]

### 2020-11-07 figs_and_tables/sup_fig2__publishedCounts.R

@@ -12,4 +12,26 @@ sum <- nxn_sc %>%
   coord_flip()
 cowplot::plot_grid(sum, sum + scale_y_log10() + ylab('Log(Sum of SC Counts)'))
 
+median <- nxn_sc %>% 
+  mutate(Date = as.Date(x = as.character(Date), format = '%Y%m%d'), 
+         Year = format(Date, '%Y')) %>% 
+  group_by(Year) %>% 
+  summarise(Count = sum(`Reported cells total`, na.rm = TRUE), Median = median(`Reported cells total`, na.rm = TRUE)) %>% 
+  ggplot(aes(x=Year, y = Median)) +
+  geom_bar(stat = 'identity') +
+  ylab("Median of SC Counts") +
+  cowplot::theme_cowplot() + 
+  coord_flip()
+
+mean <- nxn_sc %>% 
+  mutate(Date = as.Date(x = as.character(Date), format = '%Y%m%d'), 
+         Year = format(Date, '%Y')) %>% 
+  group_by(Year) %>% 
+  summarise(Count = sum(`Reported cells total`, na.rm = TRUE), Mean = mean(`Reported cells total`, na.rm = TRUE)) %>% 
+  ggplot(aes(x=Year, y = Mean)) +
+  geom_bar(stat = 'identity') +
+  ylab("Mean of SC Counts / Dataset") +
+  cowplot::theme_cowplot() + 
+  coord_flip() +
+  theme(plot.margin=unit(c(1,1,1.5,1.2),"cm"))
 

### 2020-11-09 manuscript_00.Rmd
add nmi plot

@@
 
-```{r supFig1, fig.width=4, fig.height=7, out.width=300, fig.cap=supFig1_cap, echo=FALSE, message = FALSE}
-supFig1_cap <- supFig_cap(name='supFig1_cap', caption =  'The left bar delineates the number of cells for each organism - technology combination.
-The right bar specifies the number of each cells in each post QC category.
-In silico doublets were identified with scrublet and DoubletDetector.')
-
-source('figs_and_tables/sup_fig1__qc_failure.R')
-supFig1_plot
-```
-
 ```{r supFig1, fig.width=5, fig.height=5, out.width=300, fig.cap=supFig2_cap, echo=FALSE, message = FALSE}
@@
 cowplot::plot_grid(
-  cowplot::plot_grid(lisi + theme(legend.position="none"), silhouette, nrow = 1), 
-  cowplot::plot_grid(ari, legend, rel_widths = c(1,0.2), ncol = 2), 
+  cowplot::plot_grid(lisi + theme(legend.position="none"), silhouette, NULL, rel_widths = c(1,1,0.2), nrow = 1, labels = c('A' ,'B')), 
+  cowplot::plot_grid(ari, nmi, legend, rel_widths = c(1,1, 0.2), ncol = 3, labels = c('C', 'D')), 
   nrow = 2)

### 2020-11-09 figs_and_tables/fig2__integrationPerf.R

@@ -23,8 +23,14 @@ lisi <- perf_tabula %>%
 ari <- perf_tabula %>% 
   filter(Score == 'ARI', Group == 'CellType-Cluster', nf == 2000, dims %in% c(8, 30), knn == 7) %>%
   ggplot(aes(y=Value, x=normalization)) + geom_point(aes(color=method, shape = normalization), size = 5) +
-  cowplot::theme_cowplot() + scale_color_manual(values = pals::alphabet() %>% unname()) + ylab('ARI (higher is better)') +
+  cowplot::theme_cowplot() + scale_color_manual(values = pals::alphabet() %>% unname()) + ylab('ARI') +
   ggtitle('ARI') + theme(legend.position="none")
+
+nmi <- perf_tabula %>% 
+  filter(Score == 'NMI', Group == 'CellType-Cluster', nf == 2000, dims %in% c(8, 30), knn == 7) %>%
+  ggplot(aes(y=Value, x=normalization)) + geom_point(aes(color=method, shape = normalization), size = 5) +
+  cowplot::theme_cowplot() + scale_color_manual(values = pals::alphabet() %>% unname()) + ylab('NMI') +
+  ggtitle('NMI') + theme(legend.position="none")
 
 silhouette <- perf_tabula %>% filter(Score == 'Silhouette', nf == 2000, dims %in% c(8,30), knn == 7) %>% 
   pivot_wider(values_from = Value, names_from = c('Group')) %>% 

### 2020-11-10 manuscript_00.Rmd
text updates, more figures

@@
 load('~/data/massive_integrated_eye_scRNA/Mus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features2000__counts__TabulaDroplet__batch__scVI__dims8__preFilter__mindist0.001__nneighbors500.umap.Rdata')
+umapRef <- umap
 
@@
 For the droplet-based data, we also removed cells which had more than 3000 unique genes detected as these are likely to be doublets.
-After these standard quality control steps we were left with `r format(umap %>% nrow, big.mark = ",", scientific=FALSE)` cells.
+After these standard quality control steps we were left with `r format(umapRef %>% nrow, big.mark = ",", scientific=FALSE)` cells.
 
@@
 
+```{r figBenchmark, fig.width=6, fig.height=4, fig.cap=figBenchmark_cap, echo=FALSE, message = FALSE}
+figBenchmark_cap <- fig_cap(name='figBenchmark', caption =  'a.
+Example of a method (Combat) which has a high level of batch blending, but poor separation of cell types (colored by cell type).
+b.
+example of an approach (no batch correction) that cleanly separates cell types but does not mix batches (colored by study).
+d.
+Optimal parameters for scVI method.
+sumZScale not directly to c')
+
+source('figs_and_tables/fig_umap_benchmarking.R')
+plot_grid(
+  plot_grid(
+    umap_plot_maker(umapCombat), 
+    umap_plot_maker(umapNone, color_against = 'study_accession'),
+    rel_widths = c(1.3, 1),
+    labels = c('a','b')), 
+  plot_grid(
+    zscore_sum_all_methods, 
+    zscore_droplet_scVI_optimize,
+    labels = c('c','d')),
+  nrow = 2,
+  rel_heights = c(1, 0.4)
+)
+```
+
 ## scVI has highest integration performance
@@
 The scores are summed which allows us to rank the performance of each integration tool across the diverse benchmarks.
-For users who have custom needs, the weights of each metrics can be customized to prioritize mixing or separation.
+For users who have custom needs, the weights of each metrics can be customized to prioritize batch mixing or cluster/cell type separation.
 scPOP finds that for our dataset scVI has the strongest performance, with a sum score of SOMETHING.
-The next highest performers are Harmony, fastMNN, and Scanorama.
+The next highest performers are fastMNN, harmony, and scanorama.
 We do not claim that these results are gegnral - rather we emphasize that for each (meta) atlas creation that scPOP or some similar approach be used to identify the optimal method for each dataset.
@@
 We ran DoubletDetect and scrublet and calculated the distribution of DoubletDetect and scrublet scores across all clusters and removed clusters with a DoubletDetect and scrublet score more than 4 standard deviations above the mean.
-This removed another `r format((umap %>% nrow() %>% as.integer()) - (meta_filter %>% nrow() %>% as.integer()), big.mark=",", scientific=FALSE)` cells, leaving `r format((meta_filter %>% nrow()) %>% as.integer(), big.mark=",", scientific=FALSE)` in total.
+This removed another `r format((umapRef %>% nrow() %>% as.integer()) - (meta_filter %>% nrow() %>% as.integer()), big.mark=",", scientific=FALSE)` cells, leaving `r format((meta_filter %>% nrow()) %>% as.integer(), big.mark=",", scientific=FALSE)` in total.
 
 ## xgboost model built to label unknown cell types
-  To further study cell type specific expression patterning we needed to label the `r format(umap %>% nrow() - meta_filter %>% filter(!is.na(CellType)) %>% nrow(), big.mark=",", scientific=FALSE)` unlabelled cells.
+To further study cell type specific expression patterning we needed to label the `r format(umapRef %>% nrow() - meta_filter %>% filter(!is.na(CellType)) %>% nrow(), big.mark=",", scientific=FALSE)` unlabelled cells.
 Traditionally this is done by clustering the cells, then using cell type specific markers to label the clusters.
@@
 ```{r supFig1, fig.width=5, fig.height=5, out.width=300, fig.cap=supFig2_cap, echo=FALSE, message = FALSE}
-supFig2_cap <- supFig_cap(name='supFig2_cap', caption =  '')
+supFig2_cap <- supFig_cap(name='supFig2_cap', caption =  'Performance of the various batch correction tools across various benchmarking metrics.
+For the LISI and Silhouette plots in A, B higher means better batch mixing and further to the right means better cluster purity.
+For the ARI and NMI metrics (which reflects how well cluster matches with cell type) in C, D, higher means a better score.')
 

### 2020-11-10 figs_and_tables/fig2__integrationPerf.R

@@ -14,25 +14,25 @@ perf_tabula <- perf %>% unique() %>% filter(set == 'TabulaDroplet') %>%
   filter(method != 'CCA')
 
 lisi <- perf_tabula %>% 
-  filter(Score == 'LISI', nf == 2000, dims %in% c(8,30), knn == 7) %>%
+  filter(Score == 'LISI', nf == 2000, dims %in% c(30), knn == 7) %>%
   pivot_wider(values_from = Value, names_from = c('Group')) %>% 
   ggplot(aes(y=Batch, x=-Cluster, shape = normalization)) + geom_point(aes(color=method), size = 5) +
   cowplot::theme_cowplot() + scale_color_manual(values = pals::alphabet() %>% unname()) +
   ggtitle('LISI')
 
 ari <- perf_tabula %>% 
-  filter(Score == 'ARI', Group == 'CellType-Cluster', nf == 2000, dims %in% c(8, 30), knn == 7) %>%
+  filter(Score == 'ARI', Group == 'CellType-Cluster', nf == 2000, dims %in% c(30), knn == 7) %>%
   ggplot(aes(y=Value, x=normalization)) + geom_point(aes(color=method, shape = normalization), size = 5) +
-  cowplot::theme_cowplot() + scale_color_manual(values = pals::alphabet() %>% unname()) + ylab('ARI') +
+  cowplot::theme_cowplot() + scale_color_manual(values = pals::alphabet() %>% unname()) + ylab('ARI (higher is better)') +
   ggtitle('ARI') + theme(legend.position="none")
 
 nmi <- perf_tabula %>% 
-  filter(Score == 'NMI', Group == 'CellType-Cluster', nf == 2000, dims %in% c(8, 30), knn == 7) %>%
+  filter(Score == 'NMI', Group == 'CellType-Cluster', nf == 2000, dims %in% c(30), knn == 7) %>%
   ggplot(aes(y=Value, x=normalization)) + geom_point(aes(color=method, shape = normalization), size = 5) +
   cowplot::theme_cowplot() + scale_color_manual(values = pals::alphabet() %>% unname()) + ylab('NMI') +
   ggtitle('NMI') + theme(legend.position="none")
 
-silhouette <- perf_tabula %>% filter(Score == 'Silhouette', nf == 2000, dims %in% c(8,30), knn == 7) %>% 
[…]

### 2020-11-10 figs_and_tables/fig_umap_benchmarking.R

@@ -1,0 +1,51 @@
+library(scattermore)
+source('figs_and_tables/fig2__integrationPerf.R')
+load('~/data/massive_integrated_eye_scRNA/Mus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features2000__libSize__TabulaDroplet__batch__none__dims30__preFilter__mindist0.3__nneighbors30.umap.Rdata')
+umapNone <- umap
+
+load('~/data/massive_integrated_eye_scRNA/Mus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features2000__libSize__TabulaDroplet__batch__combat__dims30__preFilter__mindist0.3__nneighbors30.umap.Rdata')
+umapCombat <- umap
+
+umap_plot_maker <- function(umap, color_against = 'CellType', red = 'UMAP', ptsize = 2){
+  celltype_col <- 'CellType' 
+  # filter
+  umapFig <- umap %>% 
+    #rename(Stage = integration_group) %>% 
+    mutate(CellType = gsub('Rod Bipolar Cells', 'Bipolar Cells', !!as.symbol(celltype_col))) %>% 
+    filter(!is.na(CellType), 
+           !is.na(study_accession), 
+           !CellType %in% c('Doublet', 'Doublets'),
+           !grepl('Mesenchyme/Lens', CellType))  %>% 
+    mutate(Size = case_when(organism == 'Homo sapiens' ~ 0.015,
+                            TRUE ~ 0.01)) 
+  
+  # attach colors to cell types
+  cell_types <- umapFig %>% 
+    pull(CellType) %>% unique() %>% sort()
+  type_val <- setNames(c(pals::alphabet(), pals::alphabet2())[1:length(cell_types)], cell_types)
+  type_col <- scale_colour_manual(values = type_val)
+  type_fill <- scale_fill_manual(values = type_val)
+  # cell type known
+  if (color_against == 'CellType'){
+    umapFig %>% 
+      ggplot() + 
+      geom_scattermore(aes(x=umapFig[,paste0(red,'_1')] %>% pull(1), 
+                           y = umapFig[,paste0(red,'_2')] %>% pull(1), 
+                           colour = CellType), pointsize = (ptsize/3), alpha = 0.1) + 
+      guides(colour = guide_legend(override.aes = list(size=2, alpha = 1))) + 
+      theme_void() +
+      type_col + 
[…]

### 2020-11-10 manuscript_00.Rmd
updates

@@
 
+gse <- read_tsv('~/git/massive_integrated_eye_scRNA/data/GEO_Study_Level_Metadata.tsv')
+
 qc <- readr::read_tsv('~/data/massive_integrated_eye_scRNA/QC.tsv.gz')
@@
 
-```{r fig1, fig.width=6, fig.height=3, fig.cap=fig1_cap, echo=FALSE, message = FALSE}
+```{r fig1, fig.width=13, fig.height=6, fig.cap=fig1_cap, echo=FALSE, message = FALSE}
 fig1_cap <- fig_cap(name='fig1', caption =  'a.
@@
 
-A key step in extracting biological information out of the unified dataset is to remove the technical effects from the data.
+Removal of the technical effects from the data is a crucial step in assessing the biological patterns out of the a meta-dataset.
 A wide variety of tools have been designed for this function.
@@
 
-There are two key metrics which have to be balanced in order to optimize performance: batch mixing (e.g.
-different studies with the same cell type should be similar) and cell type or cluster purity (where different cell types or clusters should be distinct).
+There are two key metrics which have to be balanced in order to optimize performance: batch mixing (the same cell types should be similar across independent studies) and cell type or cluster purity (where different cell types or clusters should be distinct).
 While these can be visually assessed by looking at marker gene expression across the 2D t-SNE or UMAP projection, it is more rigorous and scalable to quantify this process.
-We define batch as being each biological sample.
-We assume each study is at least one unique sample.
-We also studied the metadata and methods of each study to determine if the particular study contained multiple samples.
-In the end we identified `r meta_filter$batch %>% table() %>% length()` batches to be integrated.
 
-# scPOP wraps several different methods at measuring integration performance.
+We define batch as being each biological sample and assume each study is at least one unique sample.
+We studied the metadata and methods of each study to determine if the particular study contained multiple unique biological samples.
+In the end we identified `r umapRef %>% pull(batch) %>% unique() %>% length()` batches across `r umapRef %>% pull(study_accession) %>% unique() %>% length()` deposited datasets in `r umapRef %>% left_join(gse) %>% pull(PMID) %>% unique() %>% length()` published papers.
 
-Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, are used as well as the Adjusted Rand Index (ARI), Normalized Mutual Information (NMI), and PCR metrics as implemented by scIB.
+## scPOP wraps several different methods at measuring integration performance.
+
+scPOP brings together the Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, and the Adjusted Rand Index (ARI), Normalized Mutual Information (NMI), and PCR metrics from scIB.
 The LISI and ASW were used to measure batch mixing (where lower is better), cell type mixing (higher is better), and cluster mixing (higher is better).
@@
 The PCR metric (higher is better) runs PCA before and after integration to assess how much variance is explained by the batch corrected dimensional space.
-To visualize the interplay between batch mixing and cell type distinction we plot in FIG XXXXX the batch mixing LISI score (which has been multiplied by -1) was plotted on the y-axis (higher is better) against the cell type LISI on the x-axis (higher is better).
-The best performer on both metrics will be in the top right corner.
 
-```{r figBenchmark, fig.width=6, fig.height=4, fig.cap=figBenchmark_cap, echo=FALSE, message = FALSE}
+To visualize the interplay between batch mixing and cell type distinction we plot the batch mixing LISI score (which has been multiplied by -1) was plotted on the y-axis (higher is better) against the cluster LISI on the x-axis (higher is better).
+The best performer on both metrics will be in the top right corner  (`r supFig_cap(name = 'supFig_benchmark', display = 'cite')`a).
+In the same manner we plot the silhouette metric (`r supFig_cap(name = 'supFig_benchmark', display = 'cite')` b).
+
+## scPOP reveals different priorities across the different batch integration methods
+
+On one extreme we have combat, which merges together different batches very well, but also mixes together the distinct cell types (`r fig_cap(name = 'figBenchmark', display = 'cite')`a).
+The other extreme is running the pipeline without any batch integration method, where you see very distinct groups of cells, but each study also is a distinct group in the 2D UMAP space (`r fig_cap(name = 'figBenchmark', display = 'cite')`b).
+In this dataset we see that like combat, Harmony is weighed more towards batch mixing then cluster and cell type purity.
+Scanorama, fastMNN, and bbknn are more weighed to cleanly separate the clusters.
+With out scEiaD insct, scArches, and magic do not perform particularly well in batch mixing or cluster purity.
+
+## Different normalization methods alter integration performance
+
+https://www.frontiersin.org/articles/10.3389/fgene.2020.00041/full
+Another substantial area of uncertainty is expression normalization.
+A popular method (which we call "standard") that Seurat and anndata use by default is log normalization where the raw counts and divided by the sum counts for the cell, multiplied by a scaling factor, then log transformed.
+This helps make the data more normal in the distribution pattern, which is an assumption that some algorithms require.
+The scran normalization method groups cells into pools and normalizes across the pool summed counts instead of the individual cell counts.
+We also use the relatively simple square root (sqrt) normalization which replaces the log transformation with a square root.
+Library size (libSize) normalization omits the sqrt or log transformation.
+Finally some methods like scVI directly use the raw counts data for modelling the data.
+
+We see in the across the sumZScale metric, which we define as sumZScale = zscale(PCR) + zscale(LISI batch) + zscale(LISI cluster) + zscale(LISI cell type) + zscale(silhouette batch) + zscale(silhouette cluster) + zscale(silhouette cell type) + zscale(NMI) + zscale(ARI) that scVI overall performs best.
+As expected the libSize normalization which omits the log or square root scaling generally performs the worst (`r fig_cap(name = 'figBenchmark', display = 'cite')`c).
+We see that the remaining normalization approaches alters the batch correction performance, though the exact outcome differs across the different methods.
+For users who have custom needs, the weights of each metrics can be customized in scPOP to prioritize batch mixing or cluster/cell type separation.
+
+```{r figBenchmark, fig.width=13, fig.height=8, fig.cap=figBenchmark_cap, echo=FALSE, message = FALSE}
 figBenchmark_cap <- fig_cap(name='figBenchmark', caption =  'a.
-Example of a method (Combat) which has a high level of batch blending, but poor separation of cell types (colored by cell type).
+Example of a method (combat) which has a high level of batch blending, but poor separation of cell types (colored by cell type).
 b.
-example of an approach (no batch correction) that cleanly separates cell types but does not mix batches (colored by study).
-d.
-Optimal parameters for scVI method.
-sumZScale not directly to c')
+no batch correction cleanly separates cell types but does not mix batches (colored by study).
+c.
+sumZScale (higher is better) for each method across a variety of data normalizations.
+All methods shown here use 2000 HVG, louvain clustering, and 8 latent dimensions d.
+Optimal parameters for scVI methods across 1000 to 10000 HVGs and 4 to 100 latent dimensions.')
 
@@
 ```
-
 ## scVI has highest integration performance
 
-To rank the integration algorithms across the various benchmarks, scPOP Z scale normalizes all the metrics so that each score has an equal weight.
-The scores are summed which allows us to rank the performance of each integration tool across the diverse benchmarks.
-For users who have custom needs, the weights of each metrics can be customized to prioritize batch mixing or cluster/cell type separation.
 scPOP finds that for our dataset scVI has the strongest performance, with a sum score of SOMETHING.
 The next highest performers are fastMNN, harmony, and scanorama.
-We do not claim that these results are gegnral - rather we emphasize that for each (meta) atlas creation that scPOP or some similar approach be used to identify the optimal method for each dataset.
+We do not claim that these results are general - rather we emphasize that for each (meta) atlas creation that scPOP or some similar unbiased approach where a range of several key parameters (e.g.
+HVG number) are tested in a quantitative framework.
 
@@
 To find the best set of parameters for integration and clustering we did a grid search across key parameters: hyper variable genes (HVG), latent dimensions, and k nearest neighbors.
-Using scPOP we found the optimal parameters to be 2000 HVG, 8 latent dimensions, and a resolution of 0.6 with the leiden algorithm.
+Using scPOP we found the optimal parameters to be 2000 HVG, 8 latent dimensions, and a resolution of 0.6 with the leiden algorithm implemented by XXXXX (`r fig_cap(name = 'figBenchmark', display = 'cite')`d).
 We also varied the UMAP projection values of nearest neighbors and minimum distance to qualitatively pick a 2D projection, selecting a minimum distance of 0.001 and 500 nearest neighbors.
+
+## move this to the methods
 While we removed droplet-based cells with more than 3,000 transcripts as these are highly likely to be doublets (more than one cell per droplet) there are likely many more doublets.
@@
 Traditionally this is done by clustering the cells, then using cell type specific markers to label the clusters.
-However, as we had hundreds of thousands of pre-labelled cells we leveraged the expert labels by building a xgboost-based machine learning model that used the half of the labeled cells as a training set (see methods for more details) to train a cell type predictor for scEiaD input.
+However, as we had hundreds of thousands of pre-labelled cells across `r umapRef %>% filter(!is.na(CellType)) %>% left_join(gse) %>% pull(PMID) %>% unique() %>% length()` studies (`r fig_cap(name = 'fig1', display = 'cite')`a) we leveraged the expert labels by building a xgboost-based machine learning model that used the half of the labeled cells as a training set (see methods for more details) to train a cell type predictor for scEiaD input.
 The trained model was used to predict the cell type assignments for all cells in scEiaD.
@@
 
-~~ALLUVIAL PLOT FIGURE~~
+~~ALLUVIAL PLOT AND REF UMAP FIGURE~~
 
@@
 ## diff testing reveals....
-pseudo and haystack
+pseudo and (haystack <- save for next paper?)
 
@@
 # Supplemental
-```{r supFig1, fig.width=2, fig.height=3, out.width=300, fig.cap=supFig1_cap, echo=FALSE, message = FALSE}
+```{r supFig1, fig.width=4, fig.height=6, out.width=300, fig.cap=supFig1_cap, echo=FALSE, message = FALSE}
 supFig1_cap <- supFig_cap(name='supFig1_cap', caption =  'The left bar delineates the number of cells for each organism - technology combination.
@@
 
-```{r supFig1, fig.width=5, fig.height=5, out.width=300, fig.cap=supFig2_cap, echo=FALSE, message = FALSE}
-supFig2_cap <- supFig_cap(name='supFig2_cap', caption =  'Performance of the various batch correction tools across various benchmarking metrics.
+```{r supFig_benchmark, fig.width=8, fig.height=8, out.width=300, fig.cap=supFig_benchmark_cap, echo=FALSE, message = FALSE}
+supFig_benchmark_cap <- supFig_cap(name='supFig_benchmark', caption =  'Performance of the various batch correction tools across various benchmarking metrics.
 For the LISI and Silhouette plots in A, B higher means better batch mixing and further to the right means better cluster purity.
@@
 
-```{r supTab1, fig.width=4, fig.height=7, out.width=300, fig.cap=supTab1_cap, echo=FALSE, message = FALSE}
+```{r supTab1, fig.width=8, fig.height=14, out.width=300, fig.cap=supTab1_cap, echo=FALSE, message = FALSE}
 supTab1_cap <- supTab_cap(name='supTab1_cap', caption =  'Counts for cell type labels.
@@
 source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
-ctTable
+fontsize(ctTable, size = 8, part = 'all')
 ```
+
+```{r supTab_benchmark, fig.width=8, fig.height=14, out.width=300, fig.cap=supTab1_cap, echo=FALSE, message = FALSE}
+supTab_benchmark_cap <- supTab_cap(name='supTab_benchmark', caption =  'Counts for cell type labels.
+Published are the author created labels from the published datasets.
+Transferred area the cell labels that were transferred by a machine learning model onto the entire scEiaD dataset.')
+
+source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
+fontsize(ctTable, size = 8, part = 'all')
+```

### 2020-11-10 figs_and_tables/fig_umap_benchmarking.R

@@ -12,6 +12,8 @@ umap_plot_maker <- function(umap, color_against = 'CellType', red = 'UMAP', ptsi
   umapFig <- umap %>% 
     #rename(Stage = integration_group) %>% 
     mutate(CellType = gsub('Rod Bipolar Cells', 'Bipolar Cells', !!as.symbol(celltype_col))) %>% 
+    
+    filter(!CellType %in% c('RPE', 'Vein', 'Smooth Muscle Cell', 'Schwann','Macrophage','Mast')) %>% 
     filter(!is.na(CellType), 
            !is.na(study_accession), 
            !CellType %in% c('Doublet', 'Doublets'),
@@ -22,7 +24,7 @@ umap_plot_maker <- function(umap, color_against = 'CellType', red = 'UMAP', ptsi
   # attach colors to cell types
   cell_types <- umapFig %>% 
     pull(CellType) %>% unique() %>% sort()
-  type_val <- setNames(c(pals::alphabet(), pals::alphabet2())[1:length(cell_types)], cell_types)
+  type_val <- setNames(c(pals::brewer.set2(8),pals::alphabet())[1:length(cell_types)], cell_types)
   type_col <- scale_colour_manual(values = type_val)
   type_fill <- scale_fill_manual(values = type_val)
   # cell type known

### 2020-11-10 manuscript_00.Rmd
text updates

@@
 ---
-title: 'An Million Transcriptome Ocular Meta-Atlas for Discovery and Development'
+title: 'Building the Million Transcriptome Ocular Meta-Atlas'
 author:
@@
 
-## move this to the methods
-While we removed droplet-based cells with more than 3,000 transcripts as these are highly likely to be doublets (more than one cell per droplet) there are likely many more doublets.
-We ran DoubletDetect and scrublet and calculated the distribution of DoubletDetect and scrublet scores across all clusters and removed clusters with a DoubletDetect and scrublet score more than 4 standard deviations above the mean.
-This removed another `r format((umapRef %>% nrow() %>% as.integer()) - (meta_filter %>% nrow() %>% as.integer()), big.mark=",", scientific=FALSE)` cells, leaving `r format((meta_filter %>% nrow()) %>% as.integer(), big.mark=",", scientific=FALSE)` in total.
-
 ## xgboost model built to label unknown cell types
@@
 Traditionally this is done by clustering the cells, then using cell type specific markers to label the clusters.
-However, as we had hundreds of thousands of pre-labelled cells across `r umapRef %>% filter(!is.na(CellType)) %>% left_join(gse) %>% pull(PMID) %>% unique() %>% length()` studies (`r fig_cap(name = 'fig1', display = 'cite')`a) we leveraged the expert labels by building a xgboost-based machine learning model that used the half of the labeled cells as a training set (see methods for more details) to train a cell type predictor for scEiaD input.
+However, as we had hundreds of thousands of pre-labelled cells across `r umapRef %>% filter(!is.na(CellType)) %>% left_join(gse) %>% pull(PMID) %>% unique() %>% length()` studies (`r fig_cap(name = 'fig1', display = 'cite')`a) we leveraged the expert labels by building a xgboost-based machine learning model that used half of the labeled cells as a training set (see methods for more details) to train a cell type predictor for scEiaD input.
 The trained model was used to predict the cell type assignments for all cells in scEiaD.
-In this manner we both label most cells (cells which cannot be assigned a cell type assignment with a confidence above 0.5 were left unlabelled) and correct a small number of probable mislabels in the truth set.
+In this manner we both label most cells (cells which cannot be assigned a cell type assignment with a confidence above 0.5 were left unlabelled) and correct a small number of probable mislabels in the truth set (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a).
 
-~~ALLUVIAL PLOT AND REF UMAP FIGURE~~
+## ML model reveals certain retina cell types have more stable community consensus IDs
+
+Certain cell types (rod/cones/MG?) can be very reliably predicted
+
+The RPC/neurogenic pops arise from one group and a lower number of studies and aren't predicted as reliably
+
+```{r figXGboost_umap, fig.width=13, fig.height=20, fig.cap=figXGboost_umap_cap, echo=FALSE, message = FALSE}
+figXGboost_umap_cap <- fig_cap(name='figXGboost_umap', caption =  '')
+
+#source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
+plot_grid(ct_alluvial, labels = c('a'))
+
+```
 
@@
 
+Leiden by ? in anndata
+
+## Multi-step doublet removal
+While we removed droplet-based cells with more than 3,000 transcripts as these are highly likely to be doublets (more than one cell per droplet) there are likely many more doublets.
+We ran DoubletDetect and scrublet and calculated the distribution of DoubletDetect and scrublet scores across all clusters and removed clusters with a DoubletDetect and scrublet score more than 4 standard deviations above the mean.
+This removed another `r format((umapRef %>% nrow() %>% as.integer()) - (meta_filter %>% nrow() %>% as.integer()), big.mark=",", scientific=FALSE)` cells, leaving `r format((meta_filter %>% nrow()) %>% as.integer(), big.mark=",", scientific=FALSE)` in total.
+
 ## pseudobulk differential expression testing

### 2020-11-11 figs_and_tables/fig2__integrationPerf.R
keep cca

@@ -10,8 +10,7 @@ perf_well <- perf %>% unique() %>% filter(set == 'onlyWELL') %>%
 
 perf_tabula <- perf %>% unique() %>% filter(set == 'TabulaDroplet') %>% 
   select(Score, Group, Value, set, dims:normalization) %>% 
-  filter(Score %in% c('LISI','Silhouette', 'ARI', 'PCR', 'NMI')) %>% 
-  filter(method != 'CCA')
+  filter(Score %in% c('LISI','Silhouette', 'ARI', 'PCR', 'NMI')) 
 
 lisi <- perf_tabula %>% 
   filter(Score == 'LISI', nf == 2000, dims %in% c(30), knn == 7) %>%
@@ -44,7 +43,7 @@ legend <- get_legend(
 )
 
 zscore_tabula <- perf_tabula %>% 
-  filter(nf == 2000, dims %in% c(30), knn == 7) %>% 
+  filter(nf == 2000, dims %in% c(8), knn == 7) %>% 
   pivot_wider(names_from = c('Score','Group'), values_from = Value) %>% 
   mutate(sumZScale = 
            -scale(LISI_CellType)[,1] +

### 2020-11-11 figs_and_tables/sup_fig__xgboost_PR.R
xgboost PR plot

@@ -1,0 +1,48 @@
+library(tidyverse)
+library(matrixStats)
+library(multiROC)
+load('~/PREDICTIONSMus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features2000__counts__TabulaDroplet__batch__scVI__dims8__preFilter__mindist0.001__nneighbors500.Rdata')
+load('Mus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features2000__counts__TabulaDroplet__batch__scVI__dims8__preFilter__mindist0.001__nneighbors500.umap.Rdata')
+umapRef <- umap
+test_predictions <- predictions %>%  
+  mutate(max_pred_prob = rowMaxs(.[,-(31:33)] %>% as.matrix )) %>% 
+  select(-cell_type_id) %>% 
+  rename(PredCellType = CellType) %>% 
+  inner_join(umapRef %>% select(Barcode, TrueCellType = CellType)) %>% 
+  filter(!is.na(TrueCellType)) %>% 
+  mutate(pred_correct  = ifelse(PredCellType == TrueCellType, 'Correct', 'incorrect')) 
+sum(test_predictions$pred_correct == 'Correct') / nrow(test_predictions) 
+
+cell_type2id <- predictions %>% select(cell_type_id, CellType) %>% distinct %>% filter(CellType!='None')
+target_cell_types <-unique(cell_type2id$CellType)
+pr_df <- test_predictions %>% 
+  select(all_of(target_cell_types), PredCellType, TrueCellType) 
+pr_df <- filter(pr_df, TrueCellType  %in% colnames(pr_df)) 
+
+one_hot_labled <- model.matrix(~0+as.factor(pr_df$TrueCellType)) %>% as.data.frame
+colnames(one_hot_labled) <- strsplit(colnames(one_hot_labled), ')', fixed = T ) %>% 
+  sapply(function(x)x[length(x)]) %>% paste0( '_true')
+pr_df <- pr_df %>% select(all_of(target_cell_types))
+colnames(pr_df) <- paste0(colnames(pr_df), '_pred_xgb') 
+
+pr_df <- bind_cols( one_hot_labled, pr_df)
+colnames(pr_df) <- str_replace_all(colnames(pr_df), ' ','-')
[…]

### 2020-11-12 manuscript_00.Rmd
more figs

@@
 
-```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
+```{r Setup, message=FALSE, warning=FALSE, include=FALSE}
 #knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
@@
 library(formattable)
+library(ggalluvial)
+library(scattermore)
 # setup caption-ing
@@
 meta_filter <- fst::read_fst('~/git/plaeApp/inst/app/www/meta_filter.fst') %>% as_tibble()
-
+meta_filter <- meta_filter %>% mutate(CellType_predict = case_when(CellType_predict == 'Photoreceptor Precursors' ~ 'PR Precursors',
+                                                                   CellType_predict == 'AC/HC_Precurs' ~ 'AC/HC Precursors',
+                                                                   CellType_predict == 'RPC' ~ 'RPCs',
+                                                                   CellType_predict == 'Mesenchymal/RPE/Endothelial' ~ 'Endothelial',
+                                                                   TRUE ~ CellType_predict)) 
 # cells before QC removal (mito / likely doublets)
@@
 As this information is crucial to assess dataset integration and provide an accurate reference for user querying, we extracted individual cell labels with a combination of hand-searching the Gene Expression Omnibus, supplemental information from the publication, web resources (if an web app was made available), and personal correspondence.
-After normalizing cell type names, we obtained labels for `r meta_filter$CellType %>% table() %>% sum()` cells across `r meta_filter$CellType %>% table() %>% length()` cell types (`r supTab_cap(name = 'supTab1', display = 'cite')`).
+After normalizing cell type names, we obtained labels for `r meta_filter$CellType %>% table() %>% sum()` cells across `r meta_filter$CellType %>% table() %>% length()` cell types (`r supTab_cap(name = 'supTab1_CTcounts', display = 'cite')`).
 
@@
 The other extreme is running the pipeline without any batch integration method, where you see very distinct groups of cells, but each study also is a distinct group in the 2D UMAP space (`r fig_cap(name = 'figBenchmark', display = 'cite')`b).
-In this dataset we see that like combat, Harmony is weighed more towards batch mixing then cluster and cell type purity.
+In this dataset we see that like combat, Harmony  and CCA is weighed more towards batch mixing then cluster and cell type purity.
 Scanorama, fastMNN, and bbknn are more weighed to cleanly separate the clusters.
-With out scEiaD insct, scArches, and magic do not perform particularly well in batch mixing or cluster purity.
+With our scEiaD meta-atlas insct, scArches, and magic do not perform particularly well in batch mixing or cluster purity.
 
@@
 Another substantial area of uncertainty is expression normalization.
-A popular method (which we call "standard") that Seurat and anndata use by default is log normalization where the raw counts and divided by the sum counts for the cell, multiplied by a scaling factor, then log transformed.
-This helps make the data more normal in the distribution pattern, which is an assumption that some algorithms require.
-The scran normalization method groups cells into pools and normalizes across the pool summed counts instead of the individual cell counts.
+A popular approach (which we call "standard") that Seurat and anndata use by default is to, per cell, divid the counts by the sum counts for the cell, multiply by a scaling factor, then log transform.
+This helps make the data more normal, which is an assumption that some algorithms require.
+In contrast, the scran normalization method groups cells into pools and normalizes across the pool summed counts instead of the individual cell counts.
 We also use the relatively simple square root (sqrt) normalization which replaces the log transformation with a square root.
 Library size (libSize) normalization omits the sqrt or log transformation.
-Finally some methods like scVI directly use the raw counts data for modelling the data.
+Finally some methods like scVI directly use the raw counts data for modeling the data.
 
@@
 
-## xgboost model built to label unknown cell types
+## xgboost ML model built to label unknown cell types
 To further study cell type specific expression patterning we needed to label the `r format(umapRef %>% nrow() - meta_filter %>% filter(!is.na(CellType)) %>% nrow(), big.mark=",", scientific=FALSE)` unlabelled cells.
 Traditionally this is done by clustering the cells, then using cell type specific markers to label the clusters.
-However, as we had hundreds of thousands of pre-labelled cells across `r umapRef %>% filter(!is.na(CellType)) %>% left_join(gse) %>% pull(PMID) %>% unique() %>% length()` studies (`r fig_cap(name = 'fig1', display = 'cite')`a) we leveraged the expert labels by building a xgboost-based machine learning model that used half of the labeled cells as a training set (see methods for more details) to train a cell type predictor for scEiaD input.
+However, as we had hundreds of thousands of expert labelled cells across `r umapRef %>% filter(!is.na(CellType)) %>% left_join(gse) %>% pull(PMID) %>% unique() %>% length()` studies (`r fig_cap(name = 'fig1', display = 'cite')`a) we built a xgboost-based machine learning model that used half of the labeled cells as a training set (see methods for more details) to train a cell type predictor for scEiaD input.
 The trained model was used to predict the cell type assignments for all cells in scEiaD.
@@
 
-## ML model reveals certain retina cell types have more stable community consensus IDs
+## ML cell type predictor has high accuracy across most cell types
+The precision recall (PR) curve visualizes the ability of the model to precisely label known cells at a given confidence.
+The area under the PR curve (AUC) summarizes the effectiveness of the model across different cell types, with 1 being the highest performance.
+The xgboost model can predict rods, bipolar cells, and Muller glia with near perfect performance (`r fig_cap(name = 'supFig_PR', display = 'cite')`).
+Most of the remaining cell types can be predicted with an AUC of well over 0.9.
+The cell types with the that are the most challenging for the ML to predict are the artery, choriocapillaris, and vein, which are next most often labelled as endothelial cells.
+This is a unsurprising results as artery, choriocapillaris, and vein are constructed from endothelial cells.
+The ML model also performs under 0.9 for predicting the neurogenic cell type label; however the most next common label that is predicts are the highly related retinal progenitor populations (RPCs) (`r supTab_cap(name = 'supTab_CTmislabels', display = 'cite')`).
+Overall, our xgboost based ML model shows strong accuracy across the major cell types of the retina.
 
-Certain cell types (rod/cones/MG?) can be very reliably predicted
+## Certain retina cell types have more stable community consensus IDs
 
-The RPC/neurogenic pops arise from one group and a lower number of studies and aren't predicted as reliably
+With the wide diversity of the scEiaD we can get a sense on how consistent the retina community is at labelling the major cell types of the retina.
+By calculating the AUC against each individual cell type and study, we can see which cell types are consistently labeled correctly across most (`r fig_cap(name = 'supFig_PR_cap', display = 'cite')`).
+Most studies show very similar performance with our ML model.
+The study with the consistently worse correspondence between published and predicted label is Macoksko et al; this study is notable as it was the first demonstration of the droplet based scRNA technique.
 
-```{r figXGboost_umap, fig.width=13, fig.height=20, fig.cap=figXGboost_umap_cap, echo=FALSE, message = FALSE}
+```{r figXGboost_umap, fig.width=8, fig.height=20, out.width=300, fig.cap=figXGboost_umap_cap, echo=FALSE, message = FALSE}
 figXGboost_umap_cap <- fig_cap(name='figXGboost_umap', caption =  '')
 
-#source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
-plot_grid(ct_alluvial, labels = c('a'))
+source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
+source('figs_and_tables/make_meta_scatter_umap_plot.R')
+input <- list()
+input[['meta_column']] <- 'CellType_predict'
+input[['pt_size_meta']] <- 1
+input[['gene_and_meta_scatter_tech']] <- 'Droplet'
+input[['meta_column_transform']] <- 'None'
+input[['label_toggle']] <- 2
+#input[['meta_filter_cat']] <- 'CellType_predict'
+#input[['meta_filter_on']] <- 'Bipolar Cells'
+ctp <- make_meta_scatter_umap_plot(input, mf, meta_filter,
+                                   celltype_predict_labels,
+                                   celltype_labels,
+                                   tabulamuris_predict_labels,
+                                   cluster_labels,
+                                   cat_to_color_df
+)
+
+input <- list()
+input[['meta_column']] <- 'organism'
+input[['pt_size_meta']] <- 1
+input[['gene_and_meta_scatter_tech']] <- 'Droplet'
+input[['meta_column_transform']] <- 'None'
+#input[['meta_filter_cat']] <- 'CellType_predict'
+#input[['meta_filter_on']] <- 'Bipolar Cells'
+org <- make_meta_scatter_umap_plot(input, mf, meta_filter,
+                                   celltype_predict_labels,
+                                   celltype_labels,
+                                   tabulamuris_predict_labels,
+                                   cluster_labels,
+                                   cat_to_color_df
+)
+plot_grid(ct_alluvial, 
+          plot_grid(ctp$plot, 
+                    ctp$plot + theme(legend.position = "none") + facet_wrap(~CellType_predict),
+                    org$plot + facet_wrap(~organism) + theme(legend.position = "none"), 
+                    ncol = 1,
+                    labels = c('b', 'c', 'd'),
+                    rel_heights = c(1,1,0.5)), 
+          rel_widths = c(0.2,1), 
+          labels = c('a'))
 
 ```
+
+## ML cell type labels allow for unprecedented biological in cell types
 
@@
 
-```{r supFig_benchmark, fig.width=8, fig.height=8, out.width=300, fig.cap=supFig_benchmark_cap, echo=FALSE, message = FALSE}
+```{r supFig_benchmark, fig.width=12, fig.height=8, out.width=300, fig.cap=supFig_benchmark_cap, echo=FALSE, message = FALSE}
 supFig_benchmark_cap <- supFig_cap(name='supFig_benchmark', caption =  'Performance of the various batch correction tools across various benchmarking metrics.
@@
 cowplot::plot_grid(
-  cowplot::plot_grid(lisi + theme(legend.position="none"), silhouette, NULL, rel_widths = c(1,1,0.2), nrow = 1, labels = c('A' ,'B')), 
-  cowplot::plot_grid(ari, nmi, legend, rel_widths = c(1,1, 0.2), ncol = 3, labels = c('C', 'D')), 
+  cowplot::plot_grid(lisi + theme(legend.position="none"), silhouette, NULL, rel_widths = c(1,1,0.2), nrow = 1, labels = c('a' ,'b')), 
+  cowplot::plot_grid(ari, nmi, pcr, legend, rel_widths = c(1,1, 1, 0.2), ncol = 4, labels = c('c', 'd', 'e')), 
   nrow = 2)
@@
 
-```{r supTab1, fig.width=8, fig.height=14, out.width=300, fig.cap=supTab1_cap, echo=FALSE, message = FALSE}
-supTab1_cap <- supTab_cap(name='supTab1_cap', caption =  'Counts for cell type labels.
+```{r supTab1_CTcounts, fig.width=8, fig.height=14, out.width=300, fig.cap=supTab1_CTcounts_cap, echo=FALSE, message = FALSE}
+supTab1_CTcounts_cap <- supTab_cap(name='supTab1_CTcounts_cap', caption =  'Counts for cell type labels.
 Published are the author created labels from the published datasets.
@@
 
-```{r supTab_benchmark, fig.width=8, fig.height=14, out.width=300, fig.cap=supTab1_cap, echo=FALSE, message = FALSE}
-supTab_benchmark_cap <- supTab_cap(name='supTab_benchmark', caption =  'Counts for cell type labels.
-Published are the author created labels from the published datasets.
-Transferred area the cell labels that were transferred by a machine learning model onto the entire scEiaD dataset.')
+```{r supFig_PR, fig.width=12, fig.height=10,  fig.cap=supFig_PR_cap, echo=FALSE, message = FALSE}
+supFig_PR_cap <- supFig_cap(name='supFig_PR', caption =  '')
 
-source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
-fontsize(ctTable, size = 8, part = 'all')
+source('figs_and_tables/sup_fig__xgboost_PR.R')
+pr_split_curves
 ```
+
+```{r supTab_AUCxgboost, fig.width=14,  fig.cap=supTab_AUCxgboost_cap, echo=FALSE, message = FALSE}
+supTab_AUCxgboost_cap <- supTab_cap(name='supTab_AUCxgboost', caption =  'TEXT')
+
+#source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
+fontsize(xgboost_pr_table_full %>% arrange(`Cell Type`, Study) %>% flextable() %>% width(width = 6), size = 8, part = 'all')
+```
+
+```{r supTab_CTmislabels, fig.width=8, fig.height=14, out.width=300, fig.cap=supTab_CTmislabels_cap, echo=FALSE, message = FALSE}
+
+supTab_CTmislabels_cap <- supTab_cap(name='supTab_CTmislabels', caption =  'TEXT2')
+
+#source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
+fontsize(predictions %>%  
+           mutate(max_pred_prob = rowMaxs(.[,-(31:ncol(predictions))] %>% as.matrix )) %>% 
+           select(-cell_type_id) %>% 
+           rename(PredCellType = CellType) %>% 
+           inner_join(umapRef %>% select(Barcode, TrueCellType = CellType)) %>% 
+           filter(!is.na(TrueCellType)) %>% 
+           mutate(pred_correct  = ifelse(PredCellType == TrueCellType, 'Correct', 'incorrect')) %>% 
+           group_by(PredCellType, TrueCellType) %>% 
+           summarise(Count = n()) %>% 
+           mutate(Ratio  = round(Count / sum(Count),2 )) %>% 
+           filter(Ratio > 0.05) %>% 
+           filter(!grepl('Doubl|Periocu', TrueCellType)) %>% 
+           flextable() %>% 
+           width(width = 6), 
+         size = 8, 
+         part = 'all')
+```

### 2020-11-12 figs_and_tables/fig2__integrationPerf.R

@@ -21,15 +21,21 @@ lisi <- perf_tabula %>%
 
 ari <- perf_tabula %>% 
   filter(Score == 'ARI', Group == 'CellType-Cluster', nf == 2000, dims %in% c(30), knn == 7) %>%
-  ggplot(aes(y=Value, x=normalization)) + geom_point(aes(color=method, shape = normalization), size = 5) +
-  cowplot::theme_cowplot() + scale_color_manual(values = pals::alphabet() %>% unname()) + ylab('ARI (higher is better)') +
+  ggplot(aes(y=Value, x=normalization)) + geom_jitter(aes(color=method, shape = normalization), size = 5, width = 0.2) +
+  cowplot::theme_cowplot() + scale_color_manual(values = pals::alphabet() %>% unname()) + ylab('ARI') +
   ggtitle('ARI') + theme(legend.position="none")
 
 nmi <- perf_tabula %>% 
   filter(Score == 'NMI', Group == 'CellType-Cluster', nf == 2000, dims %in% c(30), knn == 7) %>%
-  ggplot(aes(y=Value, x=normalization)) + geom_point(aes(color=method, shape = normalization), size = 5) +
+  ggplot(aes(y=Value, x=normalization)) + geom_jitter(aes(color=method, shape = normalization), size = 5, width = 0.2) +
   cowplot::theme_cowplot() + scale_color_manual(values = pals::alphabet() %>% unname()) + ylab('NMI') +
   ggtitle('NMI') + theme(legend.position="none")
+
+pcr <- perf_tabula %>% 
+  filter(Score == 'PCR', nf == 2000, dims %in% c(30), knn == 7) %>%
+  ggplot(aes(y=Value, x=normalization)) + geom_jitter(aes(color=method, shape = normalization), size = 5, width = 0.2) +
+  cowplot::theme_cowplot() + scale_color_manual(values = pals::alphabet() %>% unname()) + ylab('PCR') +
+  ggtitle('PCR') + theme(legend.position="none")
 
 silhouette <- perf_tabula %>% filter(Score == 'Silhouette', nf == 2000, dims %in% c(30), knn == 7) %>% 
   pivot_wider(values_from = Value, names_from = c('Group')) %>% 

### 2020-11-12 figs_and_tables/fig_umap_benchmarking.R

@@ -8,6 +8,9 @@ umapCombat <- umap
 
 umap_plot_maker <- function(umap, color_against = 'CellType', red = 'UMAP', ptsize = 2){
   celltype_col <- 'CellType' 
+  if (color_against == 'CellType_predict' ){
+    umap$CellType = umap$CellType_predict
+  }
   # filter
   umapFig <- umap %>% 
     #rename(Stage = integration_group) %>% 
@@ -28,7 +31,7 @@ umap_plot_maker <- function(umap, color_against = 'CellType', red = 'UMAP', ptsi
   type_col <- scale_colour_manual(values = type_val)
   type_fill <- scale_fill_manual(values = type_val)
   # cell type known
-  if (color_against == 'CellType'){
+  if (color_against == 'CellType' | color_against == 'CellType_predict'){
     umapFig %>% 
       ggplot() + 
       geom_scattermore(aes(x=umapFig[,paste0(red,'_1')] %>% pull(1), 

### 2020-11-12 figs_and_tables/make_meta_scatter_umap_plot.R

@@ -1,0 +1,192 @@
+
+library(ggplot2)
+library(Cairo)
+library(scattermore)
+library(pals)
+library(ggrepel)
+library(patchwork)
+library(purrr)
+library(pool)
+library(RSQLite)
+library(dplyr)
+library(magick)
+library(stringr)
+library(fst)
+
+tabulamuris_predict_labels <-scEiaD_2020_v01 %>% tbl('tabulamuris_predict_labels') %>% collect
+celltype_predict_labels <-scEiaD_2020_v01 %>% tbl('celltype_predict_labels') %>% 
+  mutate(CellType_predict = case_when(CellType_predict == 'RPC' ~ 'RPCs',
+                                      CellType_predict == 'Mesenchymal/RPE/Endothelial' ~ 'Endothelial',
+                                      TRUE ~ CellType_predict)) %>% collect
+celltype_labels <-scEiaD_2020_v01 %>% tbl('celltype_labels') %>% collect
+cluster_labels <-scEiaD_2020_v01 %>% tbl('cluster_labels')
+mf <- meta_filter 
+
+# generate color_mappings
+categorical_columns <- c("Phase","batch","study_accession","library_layout","organism","Platform",
+                         "Covariate","CellType","CellType_predict","TabulaMurisCellType","TabulaMurisCellType_predict",
+                         "GSE","Summary","Design","Citation","PMID","Stage","cluster",
+                         "Doublet","TechType", "SubCellType", 'subcluster' )
+#"SubCellType" and subcluster are problems
+meta_filter <- meta_filter %>% mutate(SubCellType = tidyr::replace_na(SubCellType, 'None'),
+                                      subcluster = as.character(subcluster)) 
+
+  
+map_color <- function(column, meta_filter){
+  master_colorlist <- c(pals::alphabet(), pals::alphabet2())
+  values <- meta_filter %>% pull(!!column) %>% unique %>% sort
+  if(length(values) > length(master_colorlist) ){
+    r= round(length(values) / length(master_colorlist)) +1
+    master_colorlist <- rep(master_colorlist, r)
+  }
+  
+  colors <- master_colorlist[1:length(values)]
+  return(tibble(meta_category = column,value = values, color=colors))
+  
+}
+
+
+cat_to_color_df <- lapply(categorical_columns, function(col) map_color(col, meta_filter)) %>% bind_rows()
+# %>%
+#   bind_rows(sub_cluster_map)
+
+make_meta_scatter_umap_plot <- function(input, mf, meta_filter,
[…]

### 2020-11-12 figs_and_tables/sup_fig__xgboost_PR.R

@@ -1,48 +1,121 @@
 library(tidyverse)
 library(matrixStats)
 library(multiROC)
-load('~/PREDICTIONSMus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features2000__counts__TabulaDroplet__batch__scVI__dims8__preFilter__mindist0.001__nneighbors500.Rdata')
-load('Mus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features2000__counts__TabulaDroplet__batch__scVI__dims8__preFilter__mindist0.001__nneighbors500.umap.Rdata')
-umapRef <- umap
-test_predictions <- predictions %>%  
-  mutate(max_pred_prob = rowMaxs(.[,-(31:33)] %>% as.matrix )) %>% 
-  select(-cell_type_id) %>% 
-  rename(PredCellType = CellType) %>% 
-  inner_join(umapRef %>% select(Barcode, TrueCellType = CellType)) %>% 
-  filter(!is.na(TrueCellType)) %>% 
-  mutate(pred_correct  = ifelse(PredCellType == TrueCellType, 'Correct', 'incorrect')) 
-sum(test_predictions$pred_correct == 'Correct') / nrow(test_predictions) 
 
-cell_type2id <- predictions %>% select(cell_type_id, CellType) %>% distinct %>% filter(CellType!='None')
-target_cell_types <-unique(cell_type2id$CellType)
-pr_df <- test_predictions %>% 
-  select(all_of(target_cell_types), PredCellType, TrueCellType) 
-pr_df <- filter(pr_df, TrueCellType  %in% colnames(pr_df)) 
+load('~/data/massive_integrated_eye_scRNA/PREDICTIONSMus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features2000__counts__TabulaDroplet__batch__scVI__dims8__preFilter__mindist0.001__nneighbors500.Rdata')
+# load('Mus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features2000__counts__TabulaDroplet__batch__scVI__dims8__preFilter__mindist0.001__nneighbors500.umap.Rdata')
+# umapRef <- umap
+pr_calculator <- function(predictions, partition){
+  if (partition != 'All'){
[…]

### 2020-11-12 manuscript_00.Rmd
tweak plots

@@
 Most of the remaining cell types can be predicted with an AUC of well over 0.9.
-The cell types with the that are the most challenging for the ML to predict are the artery, choriocapillaris, and vein, which are next most often labelled as endothelial cells.
-This is a unsurprising results as artery, choriocapillaris, and vein are constructed from endothelial cells.
+The cell types  that were the most challenging for the model to predict were the artery, choriocapillaris, and vein.
+Artery, choriocapillaris, and vein are constructed from endothelial cells.
+Indeed for all three of these, endothelial was the second most common label.
 The ML model also performs under 0.9 for predicting the neurogenic cell type label; however the most next common label that is predicts are the highly related retinal progenitor populations (RPCs) (`r supTab_cap(name = 'supTab_CTmislabels', display = 'cite')`).
@@
 
-```{r figXGboost_umap, fig.width=8, fig.height=20, out.width=300, fig.cap=figXGboost_umap_cap, echo=FALSE, message = FALSE}
+```{r figXGboost_umap, fig.width=8, fig.height=13, out.width=600, fig.cap=figXGboost_umap_cap, echo=FALSE, message = FALSE}
 figXGboost_umap_cap <- fig_cap(name='figXGboost_umap', caption =  '')
@@
 
-## scEiaD UMAP projection blends batches and separates cell types
-The 2D UMAP projection of the 8 scVI-calculated latent dimensional blends the `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` studies together while also maintaining distinct space for the `r meta_filter$CellType_predict %>% table() %>% length()` unique cell types.
-We see the neurogenic population from which the retinal cell types are derived near the center of the UMAP visualization.
-Cones and rods, which are the photoreceptors of the retina which are responsible for color and low-light vision, respectively, are near each other in the UMAP.
+## The scVI-based scEiaD UMAP projection blends batches and separates cell types
+The 2D UMAP projection of the scVI-calculated batch corrected 8 latent dimensional space blends the `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` studies together while also maintaining distinct space for the `r meta_filter$CellType_predict %>% table() %>% length()` unique cell types.
+We see the neurogenic populations from which the retinal cell types are derived near the center of the UMAP visualization.
+The photoreceptors, cones and rods, of the retina which are responsible for color and low-light vision, respectively, are near each other in the UMAP space.
 The amacrine cells ....say more about cell type in space
 
-## diff testing reveals....
-pseudo and (haystack <- save for next paper?)
+## Massive pseudbulk differential testing leverages high biological diversity
 
@@
 #source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
-fontsize(xgboost_pr_table_full %>% arrange(`Cell Type`, Study) %>% flextable() %>% width(width = 6), size = 8, part = 'all')
+fontsize(xgboost_pr_table_full %>% arrange(`Cell Type`, Study) %>% flextable() %>% width(width = 3), size = 8, part = 'all')
 ```
@@
            flextable() %>% 
-           width(width = 6), 
+           width(width = 3), 
          size = 8, 

### 2020-11-12 figs_and_tables/make_meta_scatter_umap_plot.R

@@ -140,7 +140,7 @@ make_meta_scatter_umap_plot <- function(input, mf, meta_filter,
                                         pixels=c(1750,1750),
                                         interpolate=FALSE) +
                        #geom_point(data=data.frame(x=double(0)), aes(x,x,color=x))  +
-                       geom_point(data=color_data, aes(x,x,color=value)) +
+                       geom_point(data=color_data, aes(x,x,color=value), alpha = 0) +
                        scale_colour_manual(name= meta_column %>% gsub('_',' (', .) %>% gsub('idct$',')',.),
                                            values = color_list) +
                        theme_cowplot() +

### 2020-11-16 manuscript_00.Rmd
fixes

@@
 library(scattermore)
+library(grid)
+library(ComplexHeatmap)
+library(viridis)
 # setup caption-ing
@@
 
-```{r figBenchmark, fig.width=13, fig.height=8, fig.cap=figBenchmark_cap, echo=FALSE, message = FALSE}
+```{r figBenchmark, fig.width=8, fig.height=8, fig.cap=figBenchmark_cap, echo=FALSE, message = FALSE}
 figBenchmark_cap <- fig_cap(name='figBenchmark', caption =  'a.
@@
 ```
+
 ## scVI has highest integration performance
@@
 
-```{r figXGboost_umap, fig.width=8, fig.height=13, out.width=600, fig.cap=figXGboost_umap_cap, echo=FALSE, message = FALSE}
+```{r figXGboost_umap, fig.width=18, fig.height=13, fig.cap=figXGboost_umap_cap, echo=FALSE, message = FALSE}
 figXGboost_umap_cap <- fig_cap(name='figXGboost_umap', caption =  '')
@@
 input <- list()
+input[['meta_column']] <- 'CellType_predict'
+input[['pt_size_meta']] <- 3
+input[['gene_and_meta_scatter_tech']] <- 'Droplet'
+input[['meta_column_transform']] <- 'None'
+#input[['meta_filter_cat']] <- 'CellType_predict'
+#input[['meta_filter_on']] <- 'Bipolar Cells'
+ctp2 <- make_meta_scatter_umap_plot(input, mf, meta_filter,
+                                   celltype_predict_labels,
+                                   celltype_labels,
+                                   tabulamuris_predict_labels,
+                                   cluster_labels,
+                                   cat_to_color_df
+)
+
+input <- list()
 input[['meta_column']] <- 'organism'
@@
 )
-plot_grid(ct_alluvial, 
-          plot_grid(ctp$plot, 
-                    ctp$plot + theme(legend.position = "none") + facet_wrap(~CellType_predict),
+plot_grid(grid.grabExpr(draw(ct_confusion)), 
+          ctp$plot + theme(legend.position = "none"), 
+          ctp2$plot + theme(legend.position = "none") + facet_wrap(~CellType_predict),
           org$plot + facet_wrap(~organism) + theme(legend.position = "none"), 
-                    ncol = 1,
-                    labels = c('b', 'c', 'd'),
-                    rel_heights = c(1,1,0.5)), 
-          rel_widths = c(0.2,1), 
-          labels = c('a'))
+          ncol = 2,
+          labels = c('a', 'b', 'c', 'd'),
+          rel_heights = c(1,0.8))
 
@@
 
-## ML cell type labels allow for unprecedented biological in cell types
+## ML cell type labels allow for high diversity in studies for each cell type
 
@@
 We see the neurogenic populations from which the retinal cell types are derived near the center of the UMAP visualization.
-The photoreceptors, cones and rods, of the retina which are responsible for color and low-light vision, respectively, are near each other in the UMAP space.
+The photoreceptors (cones and rods) of the retina which are responsible for color and low-light vision, respectively, are near each other in the UMAP space.
 The amacrine cells ....say more about cell type in space
 
-## Massive pseudbulk differential testing leverages high biological diversity
+## Pseudobulk differential testing leverages high biological diversity
+
+Discuss
 
 ## Pan mouse as non-ocular reference
+
+V. briefly discuss
 
@@
 - ARI
+- NMI
 - Silhouette
@@
 ## Clustering
-Louvain-jaccard as implemented by Seurat
-
-Leiden by ? in anndata
+Louvain-jaccard as implemented by Seurat and leiden (implemented by PARC)
 
@@
 
-```{r supTab_AUCxgboost, fig.width=14,  fig.cap=supTab_AUCxgboost_cap, echo=FALSE, message = FALSE}
+```{r supTab_AUCxgboost, fig.cap=supTab_AUCxgboost_cap, echo=FALSE, message = FALSE}
 supTab_AUCxgboost_cap <- supTab_cap(name='supTab_AUCxgboost', caption =  'TEXT')
@@
 
-```{r supTab_CTmislabels, fig.width=8, fig.height=14, out.width=300, fig.cap=supTab_CTmislabels_cap, echo=FALSE, message = FALSE}
+```{r supTab_CTmislabels, fig.cap=supTab_CTmislabels_cap, echo=FALSE, message = FALSE}
 
@@
            flextable() %>% 
-           width(width = 3), 
+           width(width = 2), 
          size = 8, 

### 2020-11-16 figs_and_tables/sup_fig2__celltype_xgboost_labelling.R

@@ -25,21 +25,45 @@ joinedCT <- left_join(prelabelledCT, predictedCT)
 joinedCT[is.na(joinedCT)] <- 0
 ctTable <- joinedCT %>% flextable()
 ##########
+# 
+# ct_alluvial <- ct_processing %>%
+#   select(CellType, CellType_predict) %>%
+#   group_by(CellType, CellType_predict) %>%
+#   summarise(Count = n()) %>%
+#   ggplot(aes(y = sqrt(Count), axis1 = CellType, axis2 = CellType_predict)) +
+#   geom_alluvium(aes(fill = `CellType`), alpha = 0.8) +
+#   geom_stratum(alpha = 0) +
+#   ggrepel::geom_label_repel(stat = "stratum",
+#                             direction = 'x',
+#                             fill = alpha(c("white"),0.5),
+#                             size = 2,
+#                             aes(label = after_stat(stratum))) +
+#   coord_cartesian(xlim = c(0.5,2.5)) +
+#   theme_void() +
+#   scale_fill_manual(values = c(pals::polychrome() %>% unname(),
+#                                pals::alphabet() %>% unname())) +
+#   theme(legend.position = "none")
 
-ct_alluvial <- ct_processing %>% 
-  select(CellType, CellType_predict) %>% 
-  group_by(CellType, CellType_predict) %>% 
+#########
+#  confusion matrix
+########
+
+ct_mat <- ct_processing %>%
+  select(CellType, CellType_predict) %>%
+  group_by(CellType, CellType_predict) %>%
   summarise(Count = n()) %>% 
-  ggplot(aes(y = sqrt(Count), axis1 = CellType, axis2 = CellType_predict)) +
-  geom_alluvium(aes(fill = `CellType`), alpha = 0.8) +
-  geom_stratum(alpha = 0) +
-  ggrepel::geom_label_repel(stat = "stratum", 
-                            direction = 'x', 
-                            fill = alpha(c("white"),0.5),
-                            size = 2, 
-                            aes(label = after_stat(stratum))) +
-  coord_cartesian(xlim = c(0.5,2.5)) +
-  theme_void() +
-  scale_fill_manual(values = c(pals::polychrome() %>% unname(),
-                               pals::alphabet() %>% unname())) +
[…]

### 2020-11-16 figs_and_tables/sup_fig__xgboost_PR.R

@@ -3,63 +3,49 @@ library(matrixStats)
 library(multiROC)
 
 load('~/data/massive_integrated_eye_scRNA/PREDICTIONSMus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features2000__counts__TabulaDroplet__batch__scVI__dims8__preFilter__mindist0.001__nneighbors500.Rdata')
-# load('Mus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features2000__counts__TabulaDroplet__batch__scVI__dims8__preFilter__mindist0.001__nneighbors500.umap.Rdata')
-# umapRef <- umap
-pr_calculator <- function(predictions, partition){
+# # load('Mus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features2000__counts__TabulaDroplet__batch__scVI__dims8__preFilter__mindist0.001__nneighbors500.umap.Rdata')
+# # umapRef <- umap
+test_predictions <- data.table::fread('~/Desktop/856474_CTpredictortest_data_probabilities.csv.gz') %>% as_tibble %>% 
+  mutate(max_pred_prob = rowMaxs(.[,-(31:34)] %>% as.matrix ), 
+         pred_correct = ifelse(cell_type_id == true_cell_id, 'Correct', 'incorrect'))
+
+pr_calculator <- function(test_predictions, partition){
   if (partition != 'All'){
-    predictionsC <- predictions %>% 
+    test_predictions <- test_predictions %>% 
       left_join(umapRef %>% select(Barcode, study_accession), by = 'Barcode') %>% 
       filter(study_accession == partition)
-  } else {
-    predictionsC <- predictions
-  }
-  test_predictions <- predictionsC %>%  
-    mutate(max_pred_prob = rowMaxs(.[,-(31:ncol(predictionsC))] %>% as.matrix )) %>% 
-    select(-cell_type_id) %>% 
-    rename(PredCellType = CellType) %>% 
-    inner_join(umapRef %>% select(Barcode, TrueCellType = CellType)) %>% 
-    filter(!is.na(TrueCellType)) %>% 
-    mutate(pred_correct  = ifelse(PredCellType == TrueCellType, 'Correct', 'incorrect')) 
[…]

### 2020-11-18 manuscript_00.Rmd
text updates. fig tweaks

@@
 
-```{r figBenchmark, fig.width=8, fig.height=8, fig.cap=figBenchmark_cap, echo=FALSE, message = FALSE}
+```{r figBenchmark, fig.width=14, fig.height=8, fig.cap=figBenchmark_cap, echo=FALSE, message = FALSE}
 figBenchmark_cap <- fig_cap(name='figBenchmark', caption =  'a.
@@
 Most studies show very similar performance with our ML model.
-The study with the consistently worse correspondence between published and predicted label is Macoksko et al; this study is notable as it was the first demonstration of the droplet based scRNA technique.
+The study with the lowest correspondence between published and predicted label is Macoksko et al; this study is notable as it was the first demonstration of the droplet based scRNA technique.
 
@@
 input[['meta_column']] <- 'CellType_predict'
-input[['pt_size_meta']] <- 3
+input[['pt_size_meta']] <- 2
 input[['gene_and_meta_scatter_tech']] <- 'Droplet'
@@
 
-## ML cell type labels allow for high diversity in studies for each cell type
+## ML cell type labels result in study diversity for each cell type
+
+After ML projection of cell type labels from the original `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` labels onto a total of `r meta_filter %>% filter(!is.na(CellType_predict) | !is.na(TabulaMurisCellType_predict)) %>% nrow()` cells we have substantially improved the number of studies per cell type.
+For example, we went from `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(2)` human studies with labelled Muller Glia to `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(5)` after labelling (`r supTab_cap(name = 'supTab_SAcounts', display = 'cite')`).
+Overall we go from an average of `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(2) %>% mean()` studies per human cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(5) %>% mean()` and `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(4) %>% mean()` studies per mouse cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(7) %>% mean()` after transferring the cell type labels.
 
@@
 
-## Pseudobulk differential testing leverages high biological diversity
+## Pseudobulk differential testing leverages high biological diversity to provide robust results
 
-Discuss
+Differential gene expression testing between clusters or cell types in the single cell field usually involve using running the statistical test of choice with all individual cells across the group.
+In bulk RNA-seq the tests are done in relation to replicates, which are generally required.
+As we have a large number of replicates, we can sum our gene expression counts by group (e.g.
+celltype or cluster, split by organism and study).
+After this summing, we now have created a "pseudo-bulk" matrix.
+The statistical properties (CITE SOMETHING) are now similar to the traditional bulk RNA-seq experiment, allowing us to leverage the robust tooling.
+We use the edgeR test to run our differential expression tests.
 
-## Pan mouse as non-ocular reference
+## Pan mouse as non-ocular reference to improve retina-specific gene search
 
-V. briefly discuss
+Do some sloppy/easy thing where we identify retina specific genes by searching across the TM data?
 
@@
 
-## Quantification
+## Reproducblity
 
-kallisto/bustools
+The full Snakemake pipeline that takes in the raw fastq sequence and outputs the scEiaD is at https://github.com/davemcg/scEiad.
+The publication commit is BLOOP.
+Furthermore, the repository has been deposited at zenodo under BLOOP.
+We will briefly discuss the pipeline choices, programs and algorithms, and versions below.
+For the R packages, we provide the versions as the supplementary file "R_session_info.txt"
 
-## Batch normalization
-Tested:
-- scVI
-- CCA
-- insct
-- combat
-- nothing
-- magic
-- scanorama
-- harmony
-- fastMNN
-- scArches
+## Quantification of gene counts
 
-## Grid search to pick optimal parameters
+Kallisto (0.46.2) was the pseudoaligner.
+To reference transcriptomes used to build the kallisto index were the Gencode "gencode.vM25.annotation.gtf.gz" and "gencode.v35.annotation.gtf.gz" for mouse and human, respectively.
+The Ensembl release 99 "Macaca_mulatta.Mmul_10.99.gtf.gz" was used for the macaque transcriptome, For the relatively few single ended samples, the params "--single -l 200 -s 30" were used for kallisto quant.
+Otherwise the "--bias" flag was added.
+For the droplet-based samples, bustools (0.39.4) was used to extract the single cell barcodes, sort the output, and fix small barcoding errors.
 
-Varied:
-- number of Hyper Variable Genes (HVG): 1000, 2000,5000,1000
-- dims: 4,6,8,10,20,30,50,100
-- normalization: scran, seurat standard, libSize, sqrt, counts (scVI only)
-- nneighbors (UMAP viz only)
-- knn: 5,7,10 (clustering)
+## Remove empty droplets and further QC.
 
-benchmarked with:
-- LISI
-- ARI
-- NMI
-- Silhouette
-- PCA variance explained before/after correction
+After bustools count, we used the R environment (3.6.2) to remove empty droplets.
+The BUSpaRse package was used to input the bustools counts outputted mtx file.
+The DropletUtils package with the "barcodeRanks" function was used to automatically detect the inflection point in the barcode count ranks that distinguishes the likely empty droplets.
+After merging the individual count matrices into one matrix, we created a Seurat version 3 object and removed cells with a percent mitochondrial reads of >10%, fewer than 200 detected unique genes, and for the droplet data, more than 3000 detected genes.
 
-## Clustering
-Louvain-jaccard as implemented by Seurat and leiden (implemented by PARC)
+## Normalization and batch effect correction
+
+We tested several gene count normalization approaches as we were not certain which would produce an optimal outcome: standard (default Seurat, library size normalization, then log transform), sqrt (same, but with sqrt normalization), libSize (omit the log or sqrt normalization), scran, SCT from Seurat, and for scVI, no normalization (counts).
+Our R implementation of the normalization approaches as well as how we constructed the Seurat v3 object can be found in the supplementary file "make_seurat_obj_functions.R"
+
+## Batch normalization under a grid search procedure
+
+We tested `r methods` against 2000 HVGs, the different gene count normalization procedures discussed above, and both 8 and 30 outputted batch corrected latent dimensions.
+The latent dimensions are the input for clustering, the 2D UMAP visualization, and the xgboost machine learning to transfer cell type labels to unlabelled cells.
+We were unable to run every method successfully with every normalization method.
+Magic could not complete with the standard or libSize normalization.
+CCA could not complete with the libSize normalization.
+We also tried the DESC, liber, and Conos batch corrections methods but were unable to get them to work reliably so they were not included.
+The batch correction step implementation can be found the supplementary file "merge_methods.R".
+
+## Clustering and UMAP
+Louvain-Jaccard clustering against the batch corrected latend dimensions used the Seurat implementation.
+We altered the k-nearest neighbors (knn) parameters to try 5, 7, and 10 (where 5 gives more clusters than 10).
+We also used the leiden algorith as implemented by PARC with a resolution of 0.6 and 0.8.
+These two resolutions were chosen as they roughly gave the same number of clusters at the Seurat Louvain-Jaccard with a knn 7.
+
+The UMAP visualization was also calculated with the Seurat implementation with the uwot R package.
+We altered the min.dist parameter with 0.001 and 0.1 and tried n.neighbors with 15, 30, 100, and 500.
+A smaller min.dist value gives "tighter" groupings while a higher number of n.neighbors uses a number of near cells the calculate the global positioning.
+
+## Benchmarking
+
+We wrote the scPOP R package to unify the LISI and Silhouette metrics from Harmony and kBet, respectively, along with NMI, ARI, and PCA variance explained before/after correction (PCR) which are implemented by scIB.
+As scIB is a python package, we brought over the tools with the basilisk package, which places a conda versioned python package which allows for easy installation without having to manage the python dependencies.
+
+To merge these metrics into a single score, we Z scale each and sum them.
+scPOP produces both tables and visualizations allowing the user to quickly see both the interplay of batch mixing and cluster/cell type separation and the overall performance.
+If a user wishes to prioritize batch mixing or cluster/cell type separation we let the user provide a custom batch/cluster-cell type scaling value (1 is the default).
 
@@
 ## pseudobulk differential expression testing
-- celltype (published) against remaining
-- celltype (published) pairwise against celltype (published)
-- species specific within celltype
-- same for celltype (projected)
-- same for cluster
+
+Our pseudo-bulk differential testing was inspired by the OSCA guide.
+We computed differential tests for: cell type (published/transferred) against remaining cell types, celltype (published/transferred) pairwise against other individual celltype (published/transferred), specific cluster against remaining clusters, specific cluster against different individual cluster (pair-wise).
+Briefly, we grouped cells by the categories (e.g.
+each unique celltype (transferred) - study combination) and summed all gene counts.
+After summing the scores, we built a model matrix as 0 ~ categories + organism to remove any species specific effects.
+When then used the edgeR "estimateDisp" and tested the fit with "glmQLFit" and the parameter "robust = TRUE".
+Specific contrasts (e.g.
+Rods vs Cones) were differentially tested with "glmQLFTest".
 
 # Conclusion
+
+## Self built meta-atlases are easy as 1,2, benchmark a bunch, 3
+
+## Projection of cell type labels from a smaller number of studies onto the remainining cells is a powerful way to increase diversity in a meta-atlas
+
+## The scEiaD dataset provides a highly diverse, large N resource for methods development
+
+## The scEiaD is a unique ocular community resource
+
+## Fast algorithms are important
 
@@
 ```
+
+Words
+
+Spacing
 
@@
 
+Words
+
+Spacing
+
 ```{r supTab1_CTcounts, fig.width=8, fig.height=14, out.width=300, fig.cap=supTab1_CTcounts_cap, echo=FALSE, message = FALSE}
@@
 
+Words
+
+Spacing
+
+```{r supTab_SAcounts, fig.width=8, fig.height=14, out.width=300, fig.cap=supTab_SAcounts_cap, echo=FALSE, message = FALSE}
+supTab_SAcounts_cap <- supTab_cap(name='supTab_SAcounts_cap', caption =  'Counts for number of studies with cell types labels before and after cell type label transfer')
+
+source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
+fontsize(saTable, size = 8, part = 'all')
+```
+
+Words
+
+Spacing
+
 ```{r supFig_PR, fig.width=12, fig.height=10,  fig.cap=supFig_PR_cap, echo=FALSE, message = FALSE}
@@
 
-```{r supTab_CTmislabels, fig.cap=supTab_CTmislabels_cap, echo=FALSE, message = FALSE}
+Words
+
+Spacing
+
+```{r supTab_CTmislabels, fig.width=2, fig.cap=supTab_CTmislabels_cap, echo=FALSE, message = FALSE}
 
@@
            flextable() %>% 
-           width(width = 2), 
+           autofit(), 
          size = 8, 
@@
 ```
+
+Words
+
+Spacing

### 2020-11-18 figs_and_tables/fig2__integrationPerf.R

@@ -93,13 +93,13 @@ zscore_droplet_scVI_optimize <- perf_tabula %>%
            scale(LISI_Batch)[,1] + 
            -scale(LISI_Cluster)[,1] +
            -scale(Silhouette_Batch)[,1] + 
-           scale(Silhouette_Cluster)[,1] +
+           scale(Silhouette_Cluster)[,1] #+
            
-           scale(`PCR_After-Before`)[,1]
+           #scale(`PCR_After-Before`)[,1]
   ) %>% 
   left_join(cluster_stats) %>% 
   filter(!is.na(clusterN)) %>% 
-  filter(knn == 0.6, method == 'scVI') %>% 
+  #filter(knn == 0.6, method == 'scVI') %>% 
   filter(!is.na(clusterN)) %>% 
   mutate(nf = as.factor(nf),
          `scVI latent dims` = as.factor(dims)) %>% 

### 2020-11-18 figs_and_tables/sup_fig2__celltype_xgboost_labelling.R

@@ -13,6 +13,7 @@ ct_processing <-  meta_filter %>%
                                       TRUE ~ CellType_predict)) 
 
 ##########
+# Cell Counts
 predictedCT <- ct_processing %>% 
   group_by(organism, CellType_predict) %>% 
   summarise(`Published Count` = n()) %>% pivot_wider(values_from = `Published Count`, names_from = c(organism))
@@ -24,6 +25,25 @@ colnames(prelabelledCT) <- c('CellType','HS Published','MF Published','MM Publis
 joinedCT <- left_join(prelabelledCT, predictedCT)
 joinedCT[is.na(joinedCT)] <- 0
 ctTable <- joinedCT %>% flextable()
+##########
+
+##########
+# Study Counts
+predictedSA <- ct_processing %>% 
+  filter(!is.na(CellType_predict)) %>% 
+  select(CellType_predict, organism, study_accession) %>% unique() %>% 
+  group_by(organism, CellType_predict) %>% 
+  summarise(`Study Count` = n()) %>% pivot_wider(values_from = `Study Count`, names_from = c(organism))
+colnames(predictedSA) <- c('CellType','HS Studies (transferred)','MF Studies (transferred)','MM Studies (transferred)')
+prelabelledSA <- ct_processing %>% 
+  filter(!is.na(CellType)) %>% 
+  select(CellType, organism, study_accession) %>% unique() %>% 
+  group_by(organism, CellType) %>% 
+  summarise(`Study Count` = n()) %>% pivot_wider(values_from = `Study Count`, names_from = c(organism))
+colnames(prelabelledSA) <- c('CellType','HS Studies (published)','MF Studies (published)','MM Studies (published)')
+joinedSA <- left_join(prelabelledSA, predictedSA)
+joinedSA[is.na(joinedSA)] <- 0
+saTable <- joinedSA %>% flextable()
 ##########
 # 
 # ct_alluvial <- ct_processing %>%

### 2020-12-09 manuscript_00.Rmd
text updates

@@
 library(viridis)
+library(Matrix)
+library(matrixStats)
 # setup caption-ing
@@
 
-scEiaD_2020_v01 <- dbPool(drv = SQLite(), dbname = "~/data/massive_integrated_eye_scRNA/MOARTABLES__anthology_limmaFALSE___Mus_musculus_Macaca_fascicularis_Homo_sapiens-0-2000-counts-TabulaDroplet-batch-scVI-8-0.001-500-0.6.sqlite", idleTimeout = 3600000)
+scEiaD_2020_v01 <- dbPool(drv = SQLite(), dbname = "~/data/massive_integrated_eye_scRNA/MOARTABLES__anthology_limmaFALSE___Mus_musculus_Macaca_fascicularis_Homo_sapiens-0-5000-counts-universe-batch-scVIprojectionSO-10-0.1-500-0.6.sqlite", idleTimeout = 3600000)
 
@@
                                                                    TRUE ~ CellType_predict)) 
-# cells before QC removal (mito / likely doublets)
-load('~/data/massive_integrated_eye_scRNA/cell_info_labelled.Rdata')
+# cell num  before QC removal (mito / likely doublets)
+#qc <- read_tsv('~/data/massive_integrated_eye_scRNA/mito_counts.tsv')
 
 # meta data before doublet removal
-load('~/data/massive_integrated_eye_scRNA/Mus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features2000__counts__TabulaDroplet__batch__scVI__dims8__preFilter__mindist0.001__nneighbors500.umap.Rdata')
+load('~/data/massive_integrated_eye_scRNA/Mus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features5000__counts__universe__batch__scVIprojectionSO__dims10__preFilter.scEiaDprojected__mindist0.1__nneighbors500.umap.Rdata')
 umapRef <- umap
@@
 qc <- readr::read_tsv('~/data/massive_integrated_eye_scRNA/QC.tsv.gz')
+
+# cell type prediction
+load('~/data/massive_integrated_eye_scRNA/PREDICTIONSMus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features5000__counts__universe__batch__scVIprojectionSO__dims10__preFilter.scEiaDprojected__mindist0.1__nneighbors500.Rdata')
+test_predictions <- data.table::fread('~/data/massive_integrated_eye_scRNA/554065_CTpredictortest_data_probabilities.csv.gz') %>% as_tibble %>% 
+  mutate(max_pred_prob = rowMaxs(.[,-(31:34)] %>% as.matrix ), 
+         pred_correct = ifelse(cell_type_id == true_cell_id, 'Correct', 'incorrect'))
 ```
@@
 ## Explosion of single cell transcriptomic atlases
-The  recent introduction of relatively low cost and high throughput single cell sequencing technlogy has led to an explosion of research across many fields.
-As of 2020, over 30 million cells have been sequenced across over 1,000 studies (Svennson Pachter http://www.nxn.se/single-cell-studies).
-As of November 2020, the average size of each individual dataset is near 100,000 cells.
-Some of the seminal work in the high throughput single cell transcriptomics field (Macosko et al) have used the retina.
-As of late 2020, over a dozen published studies containing over a million cells have used single cell technology to profile cell type specific gene expression patterns, cell fate trajectory, tissue and cell differentiation, and disease perturbation.
+The  recent introduction of lower cost and high throughput single cell sequencing technology has led to an explosion of research across many fields.
+As of 2020, over 30 million cells have been sequenced across over 1,000 studies (Svennson Pachter http://www.nxn.se/single-cell-studies) and the average size of each  study is near 100,000 cells.
+One of the earliest works in the high throughput single cell transcriptomics field (Macosko et al) used the retina and as of late 2020, over a dozen published studies containing over a million cells have used single cell technology to profile cell type specific gene expression patterns, cell fate trajectory, tissue and cell differentiation, and disease perturbation.
 
@@
 
-We propose that by re-processing publicly available raw single cell transcriptome data in a consistent bioinformatic framework and optimally using batch correction tools we can create a meta-atlas, which combines individual datasets into one unified (or meta-) atlas.
-As there are thousands of possible permutations of single cell tools, references, and parameter choices, we create our meta-atlas (which we refer to as the single cell eye in a disk or scEiaD) by benchmarking integration outcomes across multiple important single cell RNA-seq processing parameters (integration tool, number of hyper-variable genes, clustering resolution, etc).
-The benchmarking system we developed uses a wide variety of metrics and we make available to all via an R package scPOP (single-cell Pick Optimal Parameters).
-The scEiaD will be of utility to two communities: the ocular community who can search scEiaD for gene expression across many dimensions (e.g.
-cluster, cell type, study) and the computational community who can use this very large, well-curated dataset to test algorithms for compute efficiency and performance in a diverse environment.
-As we believe data re-use is a powerful and accessible means to drive forward research, we provide our meta-atlas code-base and propose general guidelines to optimally create custom meta-atlases.
+We propose that by re-processing publicly available raw single cell transcriptome data in a consistent bioinformatic framework and optimally using batch correction tools we can create a meta-atlas.
+As there are thousands of possible permutations of single cell tools, references, and parameter choices, we create our meta-atlas (which we refer to as the single cell Eye in a Disk or scEiaD) by benchmarking integration outcomes across multiple important single cell RNA-seq processing parameters (batch removal method, number of hyper-variable genes (HVGs), clustering resolution, etc).
+The benchmarking system we developed uses a wide variety of metrics and we make available via the R package scPOP (single-cell Pick Optimal Parameters).
+The scEiaD will be of utility to two communities: the ocular community who can both search scEiaD for gene expression across many dimensions (e.g.
+cluster, cell type, study) and project their own single cell data onto scEiaD for comparison and automatic cell labelling.
+Also the computational community  can use this very large, well-curated dataset to test algorithms for compute efficiency and performance in a diverse environment.
+As we believe data re-use is a powerful and efficient approach to facilitate discovery, we provide our meta-atlas code-base and propose general guidelines to optimally create custom meta-atlases.
 
@@
 As of September 2020 we identified `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` deposited datasets that have been published in `r meta_filter %>% pull(Citation) %>% unique() %>% length()` publications (`r fig_cap(name = 'fig1', display = 'cite')`).
-To provide a non-ocular reference we also downloaded the raw sequence data for the Tabula Muris project.
+To provide a non-ocular reference we also downloaded the raw sequence data for the Tabula Muris project for re-processing.
+
 In cases where the fastq file from the SRA was not processed properly, we acquired the original bam files (personal correspondence or the SRA) and re-extracted the fastq.
@@
 
-## `r cell_info_labels %>% nrow() %>% as.integer() %>% format(., big.mark=",", scientific=FALSE)` cells before quality control
+## `r qc %>% nrow() %>% as.integer() %>% format(., big.mark=",", scientific=FALSE)` cells before quality control
 
 Gene-level counts were quantified with the kallisto bustool pseudo-aligner for both the droplet and well based samples.
-For the droplet-based technology data, the inflection point between empty and non-empty droplets was automatically determined using the DropletUtils barcodeRank function in R.
-After empty droplet removal, we had `r format(cell_info_labels %>% nrow() %>% as.integer() , big.mark=",", scientific=FALSE)` cells.
+After empty droplet removal, we had `r format(qc %>% nrow() %>% as.integer() , big.mark=",", scientific=FALSE)` cells.
 We then created a Seurat v3 object, calculated the percent mitochondrial genes counts, and removed cells which had more than 10% mitochondrial reads across all gene counts.
@@
 After these standard quality control steps we were left with `r format(umapRef %>% nrow, big.mark = ",", scientific=FALSE)` cells.
-
-## Extraction and curation of cell type labels
 
@@
 Removal of the technical effects from the data is a crucial step in assessing the biological patterns out of the a meta-dataset.
-A wide variety of tools have been designed for this function.
-As we were uncertain which would perform the best and it is difficult to manually assess performance we ran `r methods %>% length()` tools with a commonly used set of key parameters that affect integration: 2000 hyper variable genes, 8 and 30 latent dimensions returned, and the louvain clustering algorithm with 7 k-nearest neighbors
+A wide variety of methods have been written for this function.
+As we were uncertain which would perform the best and it is difficult to manually assess performance we ran `r methods %>% length()` tools with a commonly used set of key parameters that affect integration: 2000 hyper variable genes, 8 and 30 latent dimensions returned, and the louvain clustering algorithm with 7 k-nearest neighbors.
+The Snakemake system was used to automate the running of the wide variety of patterns.
+In total, BLOOP jobs were run to quantify expression and assess integration performance.
 
-There are two key metrics which have to be balanced in order to optimize performance: batch mixing (the same cell types should be similar across independent studies) and cell type or cluster purity (where different cell types or clusters should be distinct).
-While these can be visually assessed by looking at marker gene expression across the 2D t-SNE or UMAP projection, it is more rigorous and scalable to quantify this process.
+The two key metrics which have to be balanced in order to optimize performance are batch mixing (the same cell types should be similar across independent studies) and cell type or cluster purity (where different cell types or clusters should be distinct).
+While these can be visually assessed by looking at marker gene expression across the 2D t-SNE or UMAP projection, it is more rigorous and scalable to quantify these diametrically opposed characteristics.
 
@@
 
-scPOP brings together the Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, and the Adjusted Rand Index (ARI), Normalized Mutual Information (NMI), and PCR metrics from scIB.
+scPOP brings together the Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, and re-implmenets the Adjusted Rand Index (ARI) and Normalized Mutual Information (NMI) algorithms.
 The LISI and ASW were used to measure batch mixing (where lower is better), cell type mixing (higher is better), and cluster mixing (higher is better).
 NMI and ARI were used to assess the consistency of cell type to cluster assignment (where 1 is perfect correspondence between cluster and cell type).
-The PCR metric (higher is better) runs PCA before and after integration to assess how much variance is explained by the batch corrected dimensional space.
 
@@
 On one extreme we have combat, which merges together different batches very well, but also mixes together the distinct cell types (`r fig_cap(name = 'figBenchmark', display = 'cite')`a).
-The other extreme is running the pipeline without any batch integration method, where you see very distinct groups of cells, but each study also is a distinct group in the 2D UMAP space (`r fig_cap(name = 'figBenchmark', display = 'cite')`b).
-In this dataset we see that like combat, Harmony  and CCA is weighed more towards batch mixing then cluster and cell type purity.
-Scanorama, fastMNN, and bbknn are more weighed to cleanly separate the clusters.
+The other extreme is not using any batch integration method, where you see very distinct groups of cells, but also each study is a distinct in the 2D UMAP space (`r fig_cap(name = 'figBenchmark', display = 'cite')`b).
+In our scEiaD dataset we see that like combat, Harmony and CCA are weighed more towards batch mixing then cluster and cell type purity.
+Scanorama, fastMNN, and bbknn more prioritize cleanly separating the clusters.
 With our scEiaD meta-atlas insct, scArches, and magic do not perform particularly well in batch mixing or cluster purity.
@@
 Another substantial area of uncertainty is expression normalization.
-A popular approach (which we call "standard") that Seurat and anndata use by default is to, per cell, divid the counts by the sum counts for the cell, multiply by a scaling factor, then log transform.
-This helps make the data more normal, which is an assumption that some algorithms require.
+The approach (which we call "standard") that Seurat and anndata use by default is to, per cell, divide the counts by the sum counts for the cell, multiply by a scaling factor, then log transform.
+This helps make the count distribution more normal, which is an assumption that many algorithms require.
 In contrast, the scran normalization method groups cells into pools and normalizes across the pool summed counts instead of the individual cell counts.
-We also use the relatively simple square root (sqrt) normalization which replaces the log transformation with a square root.
+We also use the square root (sqrt) normalization which replaces the log transformation with a square root.
 Library size (libSize) normalization omits the sqrt or log transformation.
@@
 
-We see in the across the sumZScale metric, which we define as sumZScale = zscale(PCR) + zscale(LISI batch) + zscale(LISI cluster) + zscale(LISI cell type) + zscale(silhouette batch) + zscale(silhouette cluster) + zscale(silhouette cell type) + zscale(NMI) + zscale(ARI) that scVI overall performs best.
+We see in the across the sumZScale metric, which we define as sumZScale = zscale(LISI batch) + zscale(LISI cluster) + zscale(LISI cell type) + zscale(silhouette batch) + zscale(silhouette cluster) + zscale(silhouette cell type) + zscale(NMI) + zscale(ARI) that scVI overall performs best.
 As expected the libSize normalization which omits the log or square root scaling generally performs the worst (`r fig_cap(name = 'figBenchmark', display = 'cite')`c).
 We see that the remaining normalization approaches alters the batch correction performance, though the exact outcome differs across the different methods.
-For users who have custom needs, the weights of each metrics can be customized in scPOP to prioritize batch mixing or cluster/cell type separation.
+We also see that changing the number of latent dimensions (8 or 30) can occassionally dramatically change performance.
+
+<!-- For users who have custom needs, the weights of each metrics can be customized in scPOP to prioritize batch mixing or cluster/cell type separation. -->
 
@@
 
-## scVI has highest integration performance
+<!-- ## scVI has highest integration performance -->
 
-scPOP finds that for our dataset scVI has the strongest performance, with a sum score of SOMETHING.
+<!-- scPOP finds that for our dataset scVI has the strongest performance, with a sum score of SOMETHING.
 The next highest performers are fastMNN, harmony, and scanorama.
@@
 HVG number) are tested in a quantitative framework.
+-->
 
-## Further optimization of scVI with grid search
+## Further optimization of scVI with grid search and projection
 
-To find the best set of parameters for integration and clustering we did a grid search across key parameters: hyper variable genes (HVG), latent dimensions, and k nearest neighbors.
-Using scPOP we found the optimal parameters to be 2000 HVG, 8 latent dimensions, and a resolution of 0.6 with the leiden algorithm implemented by XXXXX (`r fig_cap(name = 'figBenchmark', display = 'cite')`d).
-We also varied the UMAP projection values of nearest neighbors and minimum distance to qualitatively pick a 2D projection, selecting a minimum distance of 0.001 and 500 nearest neighbors.
+To find the best set of parameters for the creation of scEiaD we did a grid search across key parameters: hyper variable genes (HVG), latent dimensions, and k nearest neighbors.
+Furthermore we used a recent advance in scVI capability that allows one to build a reference atlas with a subset of cells and project the remaining cells onto it - we qualitatively saw that the human only integration had the most reliable performance across the key parameters.
+We thus built a human scVI model and projected the mouse and macaque data onto it.
+We then compared the "scVIprojected" against the against the straight scVI across all the previously mentioned parameters.
+Using scPOP we first saw that the scVIprojection approach generally performed better than running scVI with all of the data.
+We found the optimal parameters to be 5000 HVG, 10 latent dimensions, and a resolution of 0.6 with the leiden algorithm implemented by PARC (`r fig_cap(name = 'figBenchmark', display = 'cite')`d).
+We also varied the UMAP projection values of nearest neighbors and minimum distance to qualitatively pick a 2D projection, selecting a minimum distance of 0.1 and 500 nearest neighbors.
 
 ## xgboost ML model built to label unknown cell types
+
 To further study cell type specific expression patterning we needed to label the `r format(umapRef %>% nrow() - meta_filter %>% filter(!is.na(CellType)) %>% nrow(), big.mark=",", scientific=FALSE)` unlabelled cells.
@@
 ## ML cell type predictor has high accuracy across most cell types
+
 The precision recall (PR) curve visualizes the ability of the model to precisely label known cells at a given confidence.
@@
 The cell types that were the most challenging for the model to predict were the artery, choriocapillaris, and vein.
-Artery, choriocapillaris, and vein are constructed from endothelial cells.
-Indeed for all three of these, endothelial was the second most common label.
+Artery, choriocapillaris, and vein are constructed from endothelial cells and we find that for all three of these, endothelial was the second most common label.
 The ML model also performs under 0.9 for predicting the neurogenic cell type label; however the most next common label that is predicts are the highly related retinal progenitor populations (RPCs) (`r supTab_cap(name = 'supTab_CTmislabels', display = 'cite')`).
@@
 
-## ML cell type labels result in study diversity for each cell type
+## ML cell type labels result in high study diversity for each cell type
 
@@
 
-## The scVI-based scEiaD UMAP projection blends batches and separates cell types
-The 2D UMAP projection of the scVI-calculated batch corrected 8 latent dimensional space blends the `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` studies together while also maintaining distinct space for the `r meta_filter$CellType_predict %>% table() %>% length()` unique cell types.
+## The scVI-based scEiaD UMAP projection blends batches while separating cell types
+The 2D UMAP projection of the scVI-calculated batch corrected 10 latent dimensional space blends the `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` studies together while also maintaining distinct space for the `r meta_filter$CellType_predict %>% table() %>% length()` unique cell types.
 We see the neurogenic populations from which the retinal cell types are derived near the center of the UMAP visualization.
 The photoreceptors (cones and rods) of the retina which are responsible for color and low-light vision, respectively, are near each other in the UMAP space.
-The amacrine cells ....say more about cell type in space
+The amacrine, muller glia, horizontal, and retinal ganglia are in distinct spaces.
 
-## Pseudobulk differential testing leverages high biological diversity to provide robust results
+## Pseudobulk testing leverages high biological diversity to provide robust differential expression results
 
@@
 We use the edgeR test to run our differential expression tests.
+
+## Projection of new data onto scEiaD
+
+The hard work of creating this resource can be
 

### 2020-12-09 figs_and_tables/fig1.R

@@ -47,9 +47,9 @@ b <- as_tbl_graph(dagified) %>% mutate(bop = name) %>% ggraph(layout = 'graphopt
   
 
 # CellType
-ct_order <- cell_info_labels %>% mutate(CellType = case_when(is.na(CellType) ~ 'Unlabelled', grepl('RPC', CellType) ~ 'RPCs', TRUE ~ CellType)) %>% group_by(CellType) %>% count() %>% arrange(-n) %>% filter(n > 10000 | CellType %in% c('Astrocytes','RPE', 'Horizontal Cells'), CellType != 'Unlabelled') %>% pull(CellType)
+ct_order <- qc %>% mutate(CellType = case_when(is.na(CellType) ~ 'Unlabelled', grepl('RPC', CellType) ~ 'RPCs', TRUE ~ CellType)) %>% group_by(CellType) %>% count() %>% arrange(-n) %>% filter(n > 10000 | CellType %in% c('Astrocytes','RPE', 'Horizontal Cells'), CellType != 'Unlabelled') %>% pull(CellType)
 
-c <- cell_info_labels %>% 
+c <- qc %>% 
   mutate(CellType = case_when(is.na(CellType) ~ 'Unlabelled', grepl('RPC', CellType) ~ 'RPCs', TRUE ~ CellType)) %>% 
   filter(CellType %in% ct_order) %>% 
   filter(!is.na(CellType), 

### 2020-12-09 figs_and_tables/fig2__integrationPerf.R

@@ -1,4 +1,4 @@
-load('~/data/massive_integrated_eye_scRNA/metrics_2020_10_07.Rdata')
+load('~/data/massive_integrated_eye_scRNA/metrics_2020_11_28.Rdata')
 
 perf <- perf %>% 
   mutate(knn = case_when(method == 'scArches' ~ 7, TRUE ~ knn))
@@ -8,7 +8,7 @@ perf_well <- perf %>% unique() %>% filter(set == 'onlyWELL') %>%
   select(Score, Group, Value, set, dims:normalization) %>% 
   filter(Score %in% c('LISI','Silhouette')) 
 
-perf_tabula <- perf %>% unique() %>% filter(set == 'TabulaDroplet') %>% 
+perf_tabula <- perf %>% unique() %>% filter(set %in% c('TabulaDroplet', 'universe')) %>% 
   select(Score, Group, Value, set, dims:normalization) %>% 
   filter(Score %in% c('LISI','Silhouette', 'ARI', 'PCR', 'NMI')) 
 
@@ -49,7 +49,7 @@ legend <- get_legend(
 )
 
 zscore_tabula <- perf_tabula %>% 
-  filter(nf == 2000, dims %in% c(8), knn == 7) %>% 
+  filter(nf == 2000, dims %in% c(8, 30), knn == 7) %>% 
   pivot_wider(names_from = c('Score','Group'), values_from = Value) %>% 
   mutate(sumZScale = 
            -scale(LISI_CellType)[,1] +
@@ -64,25 +64,27 @@ zscore_tabula <- perf_tabula %>%
            scale(LISI_Batch)[,1] + # Z score
            -scale(LISI_Cluster)[,1] +
            -scale(Silhouette_Batch)[,1] + # Z score
-           scale(Silhouette_Cluster)[,1] +
-           scale(`PCR_After-Before`)[,1]) %>% 
+           scale(Silhouette_Cluster)[,1]) %>% 
+           #scale(`PCR_After-Before`)[,1]) %>% 
   left_join(cluster_stats) %>% 
   filter(!is.na(clusterN))
 
 zscore_sum_all_methods <- zscore_tabula  %>% 
+  filter(!grepl('proj', method)) %>% 
   mutate(normalization = case_when(method == 'scArches' ~ 'standard',
                                    TRUE ~ normalization)) %>% 
   arrange(-sumZScale) %>% 
[…]

### 2020-12-09 figs_and_tables/sup_fig1__qc_failure.R

@@ -1,5 +1,7 @@
 library(ggalluvial)
 counts <- c(qc %>% 
+              mutate(organism = case_when(is.na(organism) ~ 'Homo sapiens',
+                                          TRUE ~ organism)) %>% 
               mutate(QC = gsub(' QC','', QC),
                      QC = gsub('Uni', '\nUni', QC), 
                      QC = gsub('Doubl', '\nDoubl', QC), 
@@ -22,6 +24,8 @@ counts <- c(qc %>%
 ) 
 
 supFig1_plot <- qc %>% 
+  mutate(organism = case_when(is.na(organism) ~ 'Homo sapiens',
+                                                   TRUE ~ organism)) %>% 
   mutate(QC = gsub(' QC','', QC),
          QC = gsub('Uni', '\nUni', QC), 
          QC = gsub('Doubl', '\nDoubl', QC), 
@@ -37,7 +41,7 @@ supFig1_plot <- qc %>%
   geom_stratum(aes(fill = `OrgTech`), alpha = 0.6) +
   geom_stratum(aes(fill = `QC`), alpha = 0.6) + 
   ggrepel::geom_text_repel(stat = "stratum", segment.color = NA,
-                           nudge_x = c(rep(c(-20),5), rep(c(20), 4)),
+                           nudge_x = c(rep(c(-20),5), rep(c(20), 5)),
                            direction = "y", size = 4,
                            aes(label = after_stat(stratum))) + theme_void() +
   shadowtext::geom_shadowtext(stat = "stratum", label=counts, bg.color = 'white', color = 'black') +

### 2020-12-09 figs_and_tables/sup_fig2__celltype_xgboost_labelling.R

@@ -67,7 +67,6 @@ saTable <- joinedSA %>% flextable()
 #########
 #  confusion matrix
 ########
-
 ct_mat <- ct_processing %>%
   select(CellType, CellType_predict) %>%
   group_by(CellType, CellType_predict) %>%

### 2020-12-09 figs_and_tables/sup_fig__xgboost_PR.R

@@ -1,13 +1,11 @@
 library(tidyverse)
-library(matrixStats)
+
 library(multiROC)
 
-load('~/data/massive_integrated_eye_scRNA/PREDICTIONSMus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features2000__counts__TabulaDroplet__batch__scVI__dims8__preFilter__mindist0.001__nneighbors500.Rdata')
+
 # # load('Mus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features2000__counts__TabulaDroplet__batch__scVI__dims8__preFilter__mindist0.001__nneighbors500.umap.Rdata')
 # # umapRef <- umap
-test_predictions <- data.table::fread('~/Desktop/856474_CTpredictortest_data_probabilities.csv.gz') %>% as_tibble %>% 
-  mutate(max_pred_prob = rowMaxs(.[,-(31:34)] %>% as.matrix ), 
-         pred_correct = ifelse(cell_type_id == true_cell_id, 'Correct', 'incorrect'))
+
 
 pr_calculator <- function(test_predictions, partition){
   if (partition != 'All'){

### 2020-12-09 manuscript_00.Rmd
text updates

@@
 
-## The meta-atlas
+## The projectable meta-atlas
 
@@
 As there are thousands of possible permutations of single cell tools, references, and parameter choices, we create our meta-atlas (which we refer to as the single cell Eye in a Disk or scEiaD) by benchmarking integration outcomes across multiple important single cell RNA-seq processing parameters (batch removal method, number of hyper-variable genes (HVGs), clustering resolution, etc).
-The benchmarking system we developed uses a wide variety of metrics and we make available via the R package scPOP (single-cell Pick Optimal Parameters).
+The benchmarking system we developed uses a wide variety of metrics that we make easily available via the R package scPOP (single-cell Pick Optimal Parameters).
 The scEiaD will be of utility to two communities: the ocular community who can both search scEiaD for gene expression across many dimensions (e.g.
@@
 
-scPOP brings together the Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, and re-implmenets the Adjusted Rand Index (ARI) and Normalized Mutual Information (NMI) algorithms.
+scPOP brings together the Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, and re-implemnets the Adjusted Rand Index (ARI) and Normalized Mutual Information (NMI) algorithms.
 The LISI and ASW were used to measure batch mixing (where lower is better), cell type mixing (higher is better), and cluster mixing (higher is better).
@@
 
-## scPOP reveals different priorities across the different batch integration methods
+<!-- ## scPOP reveals different priorities across the different batch integration methods -->
 
 On one extreme we have combat, which merges together different batches very well, but also mixes together the distinct cell types (`r fig_cap(name = 'figBenchmark', display = 'cite')`a).
-The other extreme is not using any batch integration method, where you see very distinct groups of cells, but also each study is a distinct in the 2D UMAP space (`r fig_cap(name = 'figBenchmark', display = 'cite')`b).
+The other extreme is not using any batch integration method, where you see very distinct groups of cells, but also each nearly study is has a distinct space in the 2D (`r fig_cap(name = 'figBenchmark', display = 'cite')`b).
 In our scEiaD dataset we see that like combat, Harmony and CCA are weighed more towards batch mixing then cluster and cell type purity.
@@
 
-https://www.frontiersin.org/articles/10.3389/fgene.2020.00041/full
-Another substantial area of uncertainty is expression normalization.
-The approach (which we call "standard") that Seurat and anndata use by default is to, per cell, divide the counts by the sum counts for the cell, multiply by a scaling factor, then log transform.
+<!-- https://www.frontiersin.org/articles/10.3389/fgene.2020.00041/full -->
+There are several normalization approaches that have been used or published.
+The "standard" approach that the popular analysis packages Seurat and anndata use by default is to, per cell, divide the counts by the sum counts for the cell, multiply by a scaling factor, then log transform.
 This helps make the count distribution more normal, which is an assumption that many algorithms require.
@@
 We see that the remaining normalization approaches alters the batch correction performance, though the exact outcome differs across the different methods.
-We also see that changing the number of latent dimensions (8 or 30) can occassionally dramatically change performance.
+We also see that changing the number of latent dimensions (8 or 30) can occasionally dramatically change performance.
+These results demonstrate the importance of assessing performance in a rigorous manner aceoss many parameters.
 
@@
 To find the best set of parameters for the creation of scEiaD we did a grid search across key parameters: hyper variable genes (HVG), latent dimensions, and k nearest neighbors.
-Furthermore we used a recent advance in scVI capability that allows one to build a reference atlas with a subset of cells and project the remaining cells onto it - we qualitatively saw that the human only integration had the most reliable performance across the key parameters.
-We thus built a human scVI model and projected the mouse and macaque data onto it.
+Furthermore we used a recent advance in scVI capability that allows one to build a reference model and project the new cells onto it.
+We thus built a human-only scVI model and then projected the mouse and macaque data onto it.
 We then compared the "scVIprojected" against the against the straight scVI across all the previously mentioned parameters.
@@
 
-## xgboost ML model built to label unknown cell types
+## High accuracy xgboost ML model built to label unknown cell types
 
@@
 
-## ML cell type predictor has high accuracy across most cell types
+<!-- ## ML cell type predictor has high accuracy across most cell types -->
 
@@
 
-## Certain retina cell types have more stable community consensus IDs
+<!-- ## Certain retina cell types have more stable community consensus IDs -->
 
-With the wide diversity of the scEiaD we can get a sense on how consistent the retina community is at labelling the major cell types of the retina.
+<!-- With the wide diversity of the scEiaD we can get a sense on how consistent the retina community is at labelling the major cell types of the retina.
 By calculating the AUC against each individual cell type and study, we can see which cell types are consistently labeled correctly across most (`r fig_cap(name = 'supFig_PR_cap', display = 'cite')`).
@@
 The study with the lowest correspondence between published and predicted label is Macoksko et al; this study is notable as it was the first demonstration of the droplet based scRNA technique.
+-->
 
@@
 ## The scVI-based scEiaD UMAP projection blends batches while separating cell types
+
 The 2D UMAP projection of the scVI-calculated batch corrected 10 latent dimensional space blends the `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` studies together while also maintaining distinct space for the `r meta_filter$CellType_predict %>% table() %>% length()` unique cell types.

### 2020-12-11 manuscript_00.Rmd
text updates

@@
 The first step in building a meta-atlas is identifying studies to draw the data from.
-We identified ocular single cell RNA seq (scRNA) studies by querying PubMed, the Sequence Read Archive (SRA), and the European Nucleotide Archive (ENA) for the inclusive terms "retina", "single cell", "scRNA", "ocular", "eye", "transcriptome" and then hand filtering the results to only keep normal (non-perturbed or mutagenized) data from single cell RNA-seq technology.
+We identified ocular single cell RNA sequencing (scRNA) studies by querying PubMed, the Sequence Read Archive (SRA), and the European Nucleotide Archive (ENA) for the inclusive terms "retina", "single cell", "scRNA", "ocular", "eye", "transcriptome." We then hand filtered the results to only keep ocular and normal (non-perturbed or mutagenized) data from single cell RNA-seq technology.
 As of September 2020 we identified `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` deposited datasets that have been published in `r meta_filter %>% pull(Citation) %>% unique() %>% length()` publications (`r fig_cap(name = 'fig1', display = 'cite')`).
-To provide a non-ocular reference we also downloaded the raw sequence data for the Tabula Muris project for re-processing.
+To provide a non-ocular reference we also downloaded the raw sequence data from the Tabula Muris project for re-processing.
 
-In cases where the fastq file from the SRA was not processed properly, we acquired the original bam files (personal correspondence or the SRA) and re-extracted the fastq.
+In cases where the fastq file from the SRA was not processed properly, we acquired the original bam files (SRA or personal correspondence) and re-extracted the fastq.
 After downloading all the data we had 5.7 TB across `r scEiaD_2020_v01 %>% tbl('metadata_filter') %>% pull(sample_accession) %>% unique() %>% length()` fastq file sets.
@@
 
-A core objective of many scRNA based studies is labeling distinct cell types.
-As this information is crucial to assess dataset integration and provide an accurate reference for user querying, we extracted individual cell labels with a combination of hand-searching the Gene Expression Omnibus, supplemental information from the publication, web resources (if an web app was made available), and personal correspondence.
-After normalizing cell type names, we obtained labels for `r meta_filter$CellType %>% table() %>% sum()` cells across `r meta_filter$CellType %>% table() %>% length()` cell types (`r supTab_cap(name = 'supTab1_CTcounts', display = 'cite')`).
+A core objective of many scRNA based studies is labeling the cell types.
+As this information is crucial to assess dataset integration and provide an accurate reference for user querying, we extracted individual cell labels with a combination of the GEO, supplemental information from the publication, web resources (e.g.
+a web app was created for the paper), and personal correspondence.
+After normalizing cell type name nomenclature, we obtained labels for `r meta_filter$CellType %>% table() %>% sum()` cells across `r meta_filter$CellType %>% table() %>% length()` cell types (`r supTab_cap(name = 'supTab1_CTcounts', display = 'cite')`).
 
@@
 
-Removal of the technical effects from the data is a crucial step in assessing the biological patterns out of the a meta-dataset.
+Disentangling the technical and biological effects from the datasets is crucial.
 A wide variety of methods have been written for this function.
-As we were uncertain which would perform the best and it is difficult to manually assess performance we ran `r methods %>% length()` tools with a commonly used set of key parameters that affect integration: 2000 hyper variable genes, 8 and 30 latent dimensions returned, and the louvain clustering algorithm with 7 k-nearest neighbors.
-The Snakemake system was used to automate the running of the wide variety of patterns.
-In total, BLOOP jobs were run to quantify expression and assess integration performance.
+As we were uncertain which would perform the best and it is difficult to manually assess performance we ran `r methods %>% length()` tools with a commonly used set of key parameters: 2000 hyper variable genes, 8 and 30 latent dimensions returned, and the louvain clustering algorithm with 7 k-nearest neighbors.
+The Snakemake system was used to automate the running of the wide variety of tools.
+In total 15,226 jobs were run to quantify expression and assess integration performance.
 
 The two key metrics which have to be balanced in order to optimize performance are batch mixing (the same cell types should be similar across independent studies) and cell type or cluster purity (where different cell types or clusters should be distinct).
-While these can be visually assessed by looking at marker gene expression across the 2D t-SNE or UMAP projection, it is more rigorous and scalable to quantify these diametrically opposed characteristics.
+While these can be visually assessed by looking at marker gene expression across the 2D UMAP projection, it is more rigorous and scalable to quantify these diametrically opposed characteristics.
 
@@
 
-scPOP brings together the Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, and re-implemnets the Adjusted Rand Index (ARI) and Normalized Mutual Information (NMI) algorithms.
+scPOP is a lightweight, low dependency R package which brings together the Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, and re-implements the Adjusted Rand Index (ARI) and Normalized Mutual Information (NMI) algorithms.
 The LISI and ASW were used to measure batch mixing (where lower is better), cell type mixing (higher is better), and cluster mixing (higher is better).
@@
 
-To visualize the interplay between batch mixing and cell type distinction we plot the batch mixing LISI score (which has been multiplied by -1) was plotted on the y-axis (higher is better) against the cluster LISI on the x-axis (higher is better).
+To visualize the interplay between batch mixing and cell type distinction we plot the batch mixing LISI score (which has been multiplied by -1) on plotted on the y-axis (higher is better) against the cluster LISI on the x-axis (higher is better).
 The best performer on both metrics will be in the top right corner  (`r supFig_cap(name = 'supFig_benchmark', display = 'cite')`a).
@@
 input[['meta_column']] <- 'CellType_predict'
-input[['pt_size_meta']] <- 1
+input[['pt_size_meta']] <- 0
 input[['gene_and_meta_scatter_tech']] <- 'Droplet'
@@
 input[['meta_column']] <- 'CellType_predict'
-input[['pt_size_meta']] <- 2
+input[['pt_size_meta']] <- 1
 input[['gene_and_meta_scatter_tech']] <- 'Droplet'
@@
 input[['meta_column']] <- 'organism'
-input[['pt_size_meta']] <- 1
+input[['pt_size_meta']] <- 0
 input[['gene_and_meta_scatter_tech']] <- 'Droplet'
@@
 
+## Pseudotime recapitulates and improves resolution of developmental processes
+
 ## Projection of new data onto scEiaD
 
-The hard work of creating this resource can be
+The hard work of creating this resource can be leveraged by the wider community with a few relatively simple steps.
+Very briefly, if outside groups quantify their mouse or human scRNA with kallisto and the same references (see methods), they can overlay their data on top of scEiaD by 1.
+installing scVI >=0.8.0, 2.
+downloading our scVI model, and 3.
+following the jupyter notebook we provide as supplemental file BLOOP.
+We demonstrate by adding new retina data to scEiaD that was published after our last data pull.
 
-## Pan mouse as non-ocular reference to improve retina-specific gene search
+##
+<!-- ## Pan mouse as non-ocular reference to improve retina-specific gene search -->
 
-Do some sloppy/easy thing where we identify retina specific genes by searching across the TM data?
+<!-- Do some sloppy/easy thing where we identify retina specific genes by searching across the TM data? -->
 
@@
 
-## Reproducblity
+## Reproduciblity
 
@@
 We will briefly discuss the pipeline choices, programs and algorithms, and versions below.
-For the R packages, we provide the versions as the supplementary file "R_session_info.txt"
+For the R packages, we provide versions as the supplementary file "R_session_info.txt"
 
@@
 
+## Intersection of gene names between mouse, macaque, and human
+
+VINAY DISCUSS
+
+## Custom macaque reference quant
+
+As we noticed that several retina marker genes (WHICH ONES!!!!) had unusually low expression we quantified the macaque data twice: once with the Ensembl reference and again with the same Gencode human reference used for the human data.
+We compared the gene-level counts for each cell and replaced the macaque gene count with the human counts if the human counts were higher.
+
 ## Remove empty droplets and further QC.
@@
 The DropletUtils package with the "barcodeRanks" function was used to automatically detect the inflection point in the barcode count ranks that distinguishes the likely empty droplets.
-After merging the individual count matrices into one matrix, we created a Seurat version 3 object and removed cells with a percent mitochondrial reads of >10%, fewer than 200 detected unique genes, and for the droplet data, more than 3000 detected genes.
+We then removed cells with percent mitochondrial reads of >10%.
+After merging the individual count matrices into one sparse matrix, we created a Seurat version 3 object and removed cells with fewer than 200 detected unique genes, and for the droplet data, more than 3000 detected genes (these are likely to be doublets).
 
@@
 CCA could not complete with the libSize normalization.
-We also tried the DESC, liber, and Conos batch corrections methods but were unable to get them to work reliably so they were not included.
+We also tried the DESC, liger, and Conos batch corrections methods but were unable to get them to work reliably so they were not included.
 The batch correction step implementation can be found the supplementary file "merge_methods.R".
@@
 Louvain-Jaccard clustering against the batch corrected latend dimensions used the Seurat implementation.
-We altered the k-nearest neighbors (knn) parameters to try 5, 7, and 10 (where 5 gives more clusters than 10).
-We also used the leiden algorith as implemented by PARC with a resolution of 0.6 and 0.8.
-These two resolutions were chosen as they roughly gave the same number of clusters at the Seurat Louvain-Jaccard with a knn 7.
+We tried the k-nearest neighbors (knn) parameters 5, 7, and 10 (where 5 gives more clusters than 10).
+We also used the leiden algorithm as implemented by PARC with a resolution of 0.6 and 0.8.
+These two resolutions were chosen as they roughly gave the same number of clusters at the Seurat Louvain-Jaccard aprroach with a knn 7.
 
-The UMAP visualization was also calculated with the Seurat implementation with the uwot R package.
-We altered the min.dist parameter with 0.001 and 0.1 and tried n.neighbors with 15, 30, 100, and 500.
-A smaller min.dist value gives "tighter" groupings while a higher number of n.neighbors uses a number of near cells the calculate the global positioning.
+The UMAP visualization was calculated with the Seurat "RunUMAP" using the uwot R package.
+We altered the min.dist parameter with 0.001 and 0.1 and tried n.neighbors across 15, 30, 100, and 500.
+A smaller min.dist value gives "tighter" groupings while a higher number of n.neighbors uses a larger number of near cells to calculate the global positioning.
 
@@
 
-We wrote the scPOP R package to unify the LISI and Silhouette metrics from Harmony and kBet, respectively, along with NMI, ARI, and PCA variance explained before/after correction (PCR) which are implemented by scIB.
-As scIB is a python package, we brought over the tools with the basilisk package, which places a conda versioned python package which allows for easy installation without having to manage the python dependencies.
+We wrote the scPOP R package to unify the LISI and Silhouette metrics from Harmony and kBet, respectively, along with NMI and ARI.
 
@@
 ## Multi-step doublet removal
-While we removed droplet-based cells with more than 3,000 transcripts as these are highly likely to be doublets (more than one cell per droplet) there are likely many more doublets.
+While we removed droplet-based cells with more than 3,000 transcripts as these are highly likely to be doublets (more than one cell per droplet) there were likely many more doublets.
 We ran DoubletDetect and scrublet and calculated the distribution of DoubletDetect and scrublet scores across all clusters and removed clusters with a DoubletDetect and scrublet score more than 4 standard deviations above the mean.
 This removed another `r format((umapRef %>% nrow() %>% as.integer()) - (meta_filter %>% nrow() %>% as.integer()), big.mark=",", scientific=FALSE)` cells, leaving `r format((meta_filter %>% nrow()) %>% as.integer(), big.mark=",", scientific=FALSE)` in total.
+
+## xgboost based cell type model
+
+VINAY WRITE
 
@@
 
+## trajectory
+
+DAVID
+
+## velocity
+
+VINAY
+
+## Projection of new data onto scEiaD
+
+DAVID
+
 # Conclusion
@@
 
-## Projection of cell type labels from a smaller number of studies onto the remainining cells is a powerful way to increase diversity in a meta-atlas
+## Transfer of cell type labels from a smaller number of studies onto the remainining cells is a powerful way to increase diversity in a meta-atlas
 
@@
 
-## Fast algorithms are important
+## Projection allows community knowledge to be leveraged by all
 

### 2020-12-11 manuscript_00.Rmd
intro update

@@
 The  recent introduction of lower cost and high throughput single cell sequencing technology has led to an explosion of research across many fields.
-As of 2020, over 30 million cells have been sequenced across over 1,000 studies (Svennson Pachter http://www.nxn.se/single-cell-studies) and the average size of each  study is near 100,000 cells.
-One of the earliest works in the high throughput single cell transcriptomics field (Macosko et al) used the retina and as of late 2020, over a dozen published studies containing over a million cells have used single cell technology to profile cell type specific gene expression patterns, cell fate trajectory, tissue and cell differentiation, and disease perturbation.
+As of 2020, over 30 million cells have been sequenced across over 1,000 studies (Svennson Pachter http://www.nxn.se/single-cell-studies) and the average size of each study is nearing 100,000 cells.
+The retina was used as the source rissue doe one of the earliest works in the high throughput single cell transcriptomics field (Macosko et al) and as of late 2020, over a dozen published studies containing over a million cells have used single cell technology to profile cell type specific gene expression patterns, cell fate trajectory, tissue and cell differentiation, and disease perturbation.
 
-## Many researchers use processed count tables for methods development and hypothesis testing
+<!-- ## Many researchers use processed count tables for methods development and hypothesis testing -->
 
@@
 
-Unfortunately, there still remain study specific batch effects due to the variety in single cell technologies used and variation in tissue handling and processing across different scientific groups.
-The single cell community has recognized that removal of these technical (also referred to as batch) effects is a critical issue and have independently developed many tools.
-It remains unclear which tool(s) is optimal for a particular dataset.
+Even after re-quantication under identical conditions there remain study specific batch effects due to the variety in single cell technologies used and variation in tissue handling and processing across eajh scientific groups.
+The single cell community has recognized that removal of these technical (also referred to as batch) effects is a critical issue and have independently developed many tools, though it remains unclear which tools and parameters are optimal for a particular dataset.
 
@@
 The scEiaD will be of utility to two communities: the ocular community who can both search scEiaD for gene expression across many dimensions (e.g.
-cluster, cell type, study) and project their own single cell data onto scEiaD for comparison and automatic cell labelling.
+cluster, cell type, study) and project their own single cell data onto scEiaD for comparison and rich automatic cell labelling.
 Also the computational community  can use this very large, well-curated dataset to test algorithms for compute efficiency and performance in a diverse environment.
@@
 We identified ocular single cell RNA sequencing (scRNA) studies by querying PubMed, the Sequence Read Archive (SRA), and the European Nucleotide Archive (ENA) for the inclusive terms "retina", "single cell", "scRNA", "ocular", "eye", "transcriptome." We then hand filtered the results to only keep ocular and normal (non-perturbed or mutagenized) data from single cell RNA-seq technology.
-As of September 2020 we identified `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` deposited datasets that have been published in `r meta_filter %>% pull(Citation) %>% unique() %>% length()` publications (`r fig_cap(name = 'fig1', display = 'cite')`).
+On September 2020 we identified `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` deposited datasets that have been published in `r meta_filter %>% pull(Citation) %>% unique() %>% length()` publications (`r fig_cap(name = 'fig1', display = 'cite')`).
+In December 2020 we ran our search again, find N studies to project onto scEiaD.
 To provide a non-ocular reference we also downloaded the raw sequence data from the Tabula Muris project for re-processing.
-
 In cases where the fastq file from the SRA was not processed properly, we acquired the original bam files (SRA or personal correspondence) and re-extracted the fastq.

### 2020-12-12 figs_and_tables/fig2__integrationPerf.R
fig tweaks, text updates

@@ -106,7 +106,7 @@ zscore_droplet_scVI_optimize <- perf_tabula %>%
   mutate(nf = as.factor(nf),
          `scVI latent dims` = as.factor(dims)) %>% 
   ggplot(aes(x=sumZScale, y = nf, color = `scVI latent dims`, shape = method)) + 
-  ggbeeswarm::geom_quasirandom(size = 4, groupOnX=FALSE) +
+  ggbeeswarm::geom_quasirandom(size = 3, groupOnX=FALSE) +
   scale_color_manual(values = pals::brewer.set1(n = 10) %>% unname()) +
   cowplot::theme_cowplot() +
   ylab('Number of\nHVGs')

### 2020-12-12 figs_and_tables/make_meta_scatter_umap_plot.R

@@ -33,7 +33,7 @@ meta_filter <- meta_filter %>% mutate(SubCellType = tidyr::replace_na(SubCellTyp
 
   
 map_color <- function(column, meta_filter){
-  master_colorlist <- c(pals::alphabet(), pals::alphabet2())
+  master_colorlist <- c(pals::polychrome()[3:length(pals::polychrome())], pals::alphabet2())
   values <- meta_filter %>% pull(!!column) %>% unique %>% sort
   if(length(values) > length(master_colorlist) ){
     r= round(length(values) / length(master_colorlist)) +1
@@ -64,13 +64,13 @@ make_meta_scatter_umap_plot <- function(input, mf, meta_filter,
   pt_size <- input$pt_size_meta %>% as.numeric()
   filter_column <- input$meta_column
   # cut down to match tech selected
-  tech <- input$gene_and_meta_scatter_tech
-  mf <- mf %>% filter(TechType == tech)
-  meta_filter <- meta_filter %>% filter(TechType == tech)
-  celltype_predict_labels <- celltype_predict_labels %>% filter(TechType == tech)
-  celltype_labels <- celltype_labels %>% filter(TechType == tech)
-  tabulamuris_predict_labels <- tabulamuris_predict_labels %>% filter(TechType == tech)
-  cluster_labels <- cluster_labels %>% filter(TechType == tech)
+  #tech <- input$gene_and_meta_scatter_tech
+  #mf <- mf %>% filter(TechType == tech)
+  #meta_filter <- meta_filter %>% filter(TechType == tech)
+  #celltype_predict_labels <- celltype_predict_labels %>% filter(TechType == tech)
+  #celltype_labels <- celltype_labels %>% filter(TechType == tech)
+  #tabulamuris_predict_labels <- tabulamuris_predict_labels %>% filter(TechType == tech)
+  #cluster_labels <- cluster_labels %>% filter(TechType == tech)
   
   if (transform == 'log2' && is.numeric(meta_filter[,meta_column] %>% pull(1))){
     cat('log2 time')
@@ -141,7 +141,7 @@ make_meta_scatter_umap_plot <- function(input, mf, meta_filter,
                                         interpolate=FALSE) +
                        #geom_point(data=data.frame(x=double(0)), aes(x,x,color=x))  +
                        geom_point(data=color_data, aes(x,x,color=value), alpha = 0) +
[…]

### 2020-12-12 figs_and_tables/heatmap.R
heatmap prototyping

@@ -1,0 +1,13 @@
+dar <- scEiaD_2020_v01 %>% tbl('cpm') %>%
+  filter(Gene %in% c('CRX (ENSG00000105392)', 'TFAP2A (ENSG00000137203)')) %>% collect()
+
+dar_w <- dar %>% pivot_wider(values_from = cpm, names_from = Barcode)
+
+dar_m <- meta_filter$Barcode[!meta_filter$Barcode %in%  dar$Barcode ]
+
+mat_empty <- matrix(0L, nrow = nrow(dar_w), ncol = length(dar_m))
+colnames(mat_empty) <- dar_m
+
+
+sample_down <- dar_all[,sample(ncol(dar_all), 1e4)]
+sample_down[is.na(sample_down)] <- 0

### 2020-12-12 manuscript_00.Rmd

@@
                                                                    CellType_predict == 'Mesenchymal/RPE/Endothelial' ~ 'Endothelial',
+                                                                   TRUE ~ CellType_predict)) %>%
+  mutate(CellType_predict = case_when(!is.na(TabulaMurisCellType_predict) ~ 'Tabula Muris',
                    TRUE ~ CellType_predict))
@@
 input[['pt_size_meta']] <- 0
-input[['gene_and_meta_scatter_tech']] <- 'Droplet'
+#input[['gene_and_meta_scatter_tech']] <- 'Droplet'
 input[['meta_column_transform']] <- 'None'
@@
 input <- list()
-input[['meta_column']] <- 'organism'
+input[['meta_column']] <- 'CellType_predict'
 input[['pt_size_meta']] <- 0
-input[['gene_and_meta_scatter_tech']] <- 'Droplet'
+#input[['gene_and_meta_scatter_tech']] <- 'Droplet'
 input[['meta_column_transform']] <- 'None'
@@
 The 2D UMAP projection of the scVI-calculated batch corrected 10 latent dimensional space blends the `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` studies together while also maintaining distinct space for the `r meta_filter$CellType_predict %>% table() %>% length()` unique cell types.
-We see the neurogenic populations from which the retinal cell types are derived near the center of the UMAP visualization.
+We also see good mixing across all the droplet and well based single cell technologies (`r supFig_cap(name = 'supFig_platformInt', display = 'cite')`).
+We see the neurogenic and progenitor populations from which the retinal cell types are derived near the center of the UMAP visualization.
+The photoreceptor precursors are adjacent to the neurogenic population and flow into the rods and cones.
+The amacrine and horizontal precursors (AC/HC) likewise flow from the neurogenic center into the mature amacrine and horizontal cells.
+
 The photoreceptors (cones and rods) of the retina which are responsible for color and low-light vision, respectively, are near each other in the UMAP space.
-The amacrine, muller glia, horizontal, and retinal ganglia are in distinct spaces.
+The major remaining retina cell types, by proportion in the mammalian eye are the Muller Glia, which are a glial cell type which help support the neurons of the retina.
+Next we have the neural cell types which transmit and help interpret the signals from the photoreceptors before they leave the retina via the optic stalk: the amacrine cells, retinal ganglia, horizontal, and bipolar cells.
+All of these cells are in well defined spaces in the UMAP.
+Finally we see even across species that the major cell types still group near each other (`r supFig_cap(name = 'figXGboost_umap', display = 'cite')`d).
+The macaque retina cells are not present in the precursor/neurogenic center of the UMAP as expected because only fully developed tissues were sampled in these data.
 
-## Pseudobulk testing leverages high biological diversity to provide robust differential expression results
+<!-- ## Pseudobulk testing leverages high biological diversity to provide robust differential expression results -->
 
-Differential gene expression testing between clusters or cell types in the single cell field usually involve using running the statistical test of choice with all individual cells across the group.
+<!-- Differential gene expression testing between clusters or cell types in the single cell field usually involve using running the statistical test of choice with all individual cells across the group.
 In bulk RNA-seq the tests are done in relation to replicates, which are generally required.
@@
 We use the edgeR test to run our differential expression tests.
+-->
 
@@
 
-As we noticed that several retina marker genes (WHICH ONES!!!!) had unusually low expression we quantified the macaque data twice: once with the Ensembl reference and again with the same Gencode human reference used for the human data.
+As we noticed that several retina marker genes (WHICH ONES!!!!) had unusually low expression in the macaque data we quantified the scRNA data twice: once with the Ensembl reference and again with the same Gencode human reference used for the human data.
 We compared the gene-level counts for each cell and replaced the macaque gene count with the human counts if the human counts were higher.
+VINAY AM I MISSING DETAILS?
 
@@
 
-After bustools count, we used the R environment (3.6.2) to remove empty droplets.
-The BUSpaRse package was used to input the bustools counts outputted mtx file.
+After bustools count, we used R (3.6.2) to remove empty droplets.
+The BUSpaRse package was used to input the bustools counts mtx file.
 The DropletUtils package with the "barcodeRanks" function was used to automatically detect the inflection point in the barcode count ranks that distinguishes the likely empty droplets.
@@
 ## Clustering and UMAP
-Louvain-Jaccard clustering against the batch corrected latend dimensions used the Seurat implementation.
+Louvain-Jaccard clustering against the batch corrected latent dimensions used the Seurat implementation.
 We tried the k-nearest neighbors (knn) parameters 5, 7, and 10 (where 5 gives more clusters than 10).
@@
 We wrote the scPOP R package to unify the LISI and Silhouette metrics from Harmony and kBet, respectively, along with NMI and ARI.
+LISI and Silhouette require a dense matrix, which is a problem for our data as a 1e6 x 20,000 matrix cannot fit even in a 1.5 TB memory node.
+We down-sampled the dataset to ~100,000 cells, taking care to keep rarer cell types for the LISI and Silhouette benchmarking.
 
@@
 
-## pseudobulk differential expression testing
+<!-- ## pseudobulk differential expression testing -->
 
-Our pseudo-bulk differential testing was inspired by the OSCA guide.
+<!-- Our pseudo-bulk differential testing was inspired by the OSCA guide.
 We computed differential tests for: cell type (published/transferred) against remaining cell types, celltype (published/transferred) pairwise against other individual celltype (published/transferred), specific cluster against remaining clusters, specific cluster against different individual cluster (pair-wise).
@@
 Rods vs Cones) were differentially tested with "glmQLFTest".
+-->
 
@@
 Spacing
+
+```{r supFig_platformInt, fig.width=12, fig.height=10,  fig.cap=supFig_platformInt_cap, echo=FALSE, message = FALSE}
+supFig_platformInt_cap <- supFig_cap(name='supFig_platformInt', caption =  '')
+input <- list()
+input[['meta_column']] <- 'CellType_predict'
+input[['pt_size_meta']] <- 5
+input[['gene_and_meta_scatter_tech']] <- 'Droplet'
+input[['meta_column_transform']] <- 'None'
+#input[['meta_filter_cat']] <- 'CellType_predict'
+#input[['meta_filter_on']] <- 'Bipolar Cells'
+ctp3 <- make_meta_scatter_umap_plot(input, mf, meta_filter,
+                                   celltype_predict_labels,
+                                   celltype_labels,
+                                   tabulamuris_predict_labels,
+                                   cluster_labels,
+                                   cat_to_color_df
+)
+
+ctp3$plot + facet_wrap(~Platform)
+```

### 2020-12-14 figs_and_tables/dotplot.R
still prototyping

@@ -1,0 +1,270 @@
+library(ggplot2)
+library(Cairo)
+library(scattermore)
+library(pals)
+library(ggrepel)
+library(patchwork)
+library(purrr)
+library(pool)
+library(RSQLite)
+library(dplyr)
+library(magick)
+library(stringr)
+library(ggtext)
+load('~/data/massive_integrated_eye_scRNA/Mus_musculus_Macaca_fascicularis_Homo_sapiens__0__5000__counts__universe__batch__scVIprojectionSO__dims10__0.6__500__0.1__CellType_predict.sceWilcox_summary.Rdata')
+#genes <- c('ENSG00000144834','ENSG00000139970','ENSG00000109846','ENSG00000026025','ENSG00000067715','ENSG00000123307','ENSG00000139053','ENSG00000143774','ENSG00000105372','ENSG00000130066','ENSG00000104435','ENSG00000139970','ENSG00000050165','ENSG00000135821','ENSG00000104435','ENSG00000131711','ENSG00000016082','ENSG00000198668','ENSG00000149489','ENSG00000129535','ENSG00000105372','ENSG00000140988')
+
+#markers <- scEiaD_2020_v01 %>% tbl('genes') %>% collect() %>% mutate(ENS = str_extract(Gene, 'ENSG\\d+')) %>% filter(ENS %in% genes) %>% pull(Gene)
+
+# marker_info_DT <- marker_list %>% 
+#   bind_rows(.id = 'cluster') %>% 
+#   group_by(cluster) %>% 
+#   #top_n(5, L4) %>% 
+#   filter(grepl('Amacrine|Rod|Cone|Retinal|Muller|Horizon|Bipol|Astro', cluster)) %>% 
+#   filter(!Gene %in%
+#            (marker_list %>% 
+#            bind_rows(.id = 'cluster') %>% 
+#            group_by(cluster) %>% 
+#            top_n(5, L4) %>% 
+#            filter(grepl('Amacrine|Rod|Cone|Retinal|Muller|Horizon|Bipol|Astro', cluster)) %>% 
+#            group_by(Gene) %>% 
+#            summarise(Count = n()) %>% 
+#            arrange(-Count) %>% 
+#            filter(Count > 1) %>% 
+#            pull(Gene))) %>% 
+#   ungroup() %>% 
+#   group_by(cluster) %>% 
+#   top_n(3, L4)  
+
+marker_info <-  markers_summary %>% 
[…]

### 2020-12-14 figs_and_tables/heatmap.R

@@ -1,13 +1,95 @@
-dar <- scEiaD_2020_v01 %>% tbl('cpm') %>%
-  filter(Gene %in% c('CRX (ENSG00000105392)', 'TFAP2A (ENSG00000137203)')) %>% collect()
+genes <- c('ENSG00000175130','ENSG00000104435','ENSG00000067715','ENSG00000144834','ENSG00000139970','ENSG00000109846','ENSG00000010278','ENSG00000172216','ENSG00000026025','ENSG00000177606','ENSG00000067715','ENSG00000123307','ENSG00000154975','ENSG00000188133','ENSG00000129244','ENSG00000139053','ENSG00000143774','ENSG00000021300','ENSG00000090097','ENSG00000112041','ENSG00000087086','ENSG00000105372','ENSG00000010278','ENSG00000026025','ENSG00000130066','ENSG00000104435','ENSG00000137285','ENSG00000119547','ENSG00000144834','ENSG00000139970','ENSG00000115310','ENSG00000050165','ENSG00000140522','ENSG00000162998','ENSG00000135821','ENSG00000104435','ENSG00000137285','ENSG00000131711','ENSG00000172020','ENSG00000139970','ENSG00000016082','ENSG00000253368','ENSG00000198668','ENSG00000183036','ENSG00000084234','ENSG00000149489','ENSG00000129535','ENSG00000185527','ENSG00000112619','ENSG00000130561','ENSG00000105372','ENSG00000231500','ENSG00000177600','ENSG00000083845','ENSG00000140988')
 
-dar_w <- dar %>% pivot_wider(values_from = cpm, names_from = Barcode)
+markers <- scEiaD_2020_v01 %>% tbl('genes') %>% collect() %>% mutate(ENS = str_extract(Gene, 'ENSG\\d+')) %>% filter(ENS %in% genes) %>% pull(Gene)
+dat <- scEiaD_2020_v01 %>% tbl('cpm') %>%
+  filter(Gene %in% markers) %>% collect()
 
-dar_m <- meta_filter$Barcode[!meta_filter$Barcode %in%  dar$Barcode ]
+dat_w <- dat %>% pivot_wider(values_from = cpm, names_from = Barcode)
 
-mat_empty <- matrix(0L, nrow = nrow(dar_w), ncol = length(dar_m))
-colnames(mat_empty) <- dar_m
+# only keep main retina cell types
+dat_w <- dat_w[,colnames(dat_w) %in% 
[…]

### 2020-12-14 figs_and_tables/dotplot.R
chunk added

@@ -36,6 +36,30 @@ load('~/data/massive_integrated_eye_scRNA/Mus_musculus_Macaca_fascicularis_Homo_
 #   group_by(cluster) %>% 
 #   top_n(3, L4)  
 
+marker_list <- list()
+for (i in (meta_filter$CellType_predict %>% unique())){
+  if (is.na(i)){
+    next()
+  } else {
+    var <- i
+    print(var)
+    test <- glue("%-1*{var}%")
+    ct <- glue("%{var}%")
+    markers <- scEiaD_2020_v01 %>% 
+      tbl('PB_results') %>% 
+      filter(PB_Test == 'Pairwise CellType (Predict) against CellType (Predict)', 
+             test %like% (!!ct), 
+             logCPM > 5)  %>% 
+      mutate(logFC = case_when(comparison %like% (!!test) ~ abs(logFC), 
+                               TRUE ~ logFC)) %>%  group_by(Gene) %>% 
+      summarise(L1 = sum(logFC > 1), L4 = sum(logFC > 4), 
+                Diff = mean(logFC)) %>% 
+      arrange(-Diff, -L4) %>% 
+      collect()
+    marker_list[[var]] <- markers
+  }
+}
+
 marker_info <-  markers_summary %>% 
   #filter(pval < 1) %>% 
   group_by(cluster) %>% 

### 2020-12-14 figs_and_tables/dotplot.R
query

@@ -97,6 +97,7 @@ exp_stats <- scEiaD_2020_v01 %>% tbl('grouped_stats') %>%
 
 marker_info %>% left_join(exp_stats %>% dplyr::rename(cluster = CellType_predict))
 
+# marker_info %>% left_join(exp_stats %>% dplyr::rename(cluster = CellType_predict)) %>% filter((FDR < 1 |  med_auc > 0.4 | D_KL > 0.5), D_KL > 0.1,`%` > 30) %>% filter(cluster == 'Horizontal Cells')
 
 make_dotplot <- function(input, db, meta_filter, cat_to_color_df){
   ### this makes it a little easier to test

### 2020-12-30 manuscript_00.Rmd
text updates

@@
       - '--lua-filter=author-info-blocks.lua'
-bibliography: massive_integrated_eye_scRNA.bib
+bibliography: references2.bib
 csl: investigative-ophthalmology-and-visual-science.csl
@@
 
-scEiaD_2020_v01 <- dbPool(drv = SQLite(), dbname = "~/data/massive_integrated_eye_scRNA/MOARTABLES__anthology_limmaFALSE___Mus_musculus_Macaca_fascicularis_Homo_sapiens-0-5000-counts-universe-batch-scVIprojectionSO-10-0.1-500-0.6.sqlite", idleTimeout = 3600000)
+scEiaD_2020_v01 <- dbPool(drv = SQLite(), dbname = "~/data/massive_integrated_eye_scRNA/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-500-0.6.sqlite", idleTimeout = 3600000)
 
@@
                                                                    CellType_predict == 'Mesenchymal/RPE/Endothelial' ~ 'Endothelial',
-                                                                   TRUE ~ CellType_predict)) %>%
+                                                                   TRUE ~ CellType_predict),
+                                      UMAP_1 = UMAP_1 * -1,
+                                      UMAP_2 = UMAP_2 * -1) %>%
   mutate(CellType_predict = case_when(!is.na(TabulaMurisCellType_predict) ~ 'Tabula Muris',
@@
 # meta data before doublet removal
-load('~/data/massive_integrated_eye_scRNA/Mus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features5000__counts__universe__batch__scVIprojectionSO__dims10__preFilter.scEiaDprojected__mindist0.1__nneighbors500.umap.Rdata')
+load('~/data/massive_integrated_eye_scRNA/n_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__preFilter.scEiaDprojected__dist-0.1__neighbors-500.umap.Rdata')
 umapRef <- umap
@@
 # cell type prediction
-load('~/data/massive_integrated_eye_scRNA/PREDICTIONSMus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features5000__counts__universe__batch__scVIprojectionSO__dims10__preFilter.scEiaDprojected__mindist0.1__nneighbors500.Rdata')
-test_predictions <- data.table::fread('~/data/massive_integrated_eye_scRNA/554065_CTpredictortest_data_probabilities.csv.gz') %>% as_tibble %>% 
-  mutate(max_pred_prob = rowMaxs(.[,-(31:34)] %>% as.matrix ), 
+load('~/data/massive_integrated_eye_scRNA/PREDICTIONSn_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__preFilter.scEiaDprojected__dist-0.1__neighbors-500.Rdata')
+test_predictions <- data.table::fread('~/data/massive_integrated_eye_scRNA/277874_CTpredictortest_data_probabilities.csv.gz') %>% as_tibble %>% 
+  mutate(max_pred_prob = rowMaxs(.[,-(33:36)] %>% as.matrix ), 
          pred_correct = ifelse(cell_type_id == true_cell_id, 'Correct', 'incorrect'))
+# created by running 'source' on src/make_dotplot_markers.R
+load('~/data/massive_integrated_eye_scRNA/top_markers.Rdata')
+
+# pmid search
+load('~/data/massive_integrated_eye_scRNA/top_marker_pmid.Rdata')
 ```
@@
 The  recent introduction of lower cost and high throughput single cell sequencing technology has led to an explosion of research across many fields.
-As of 2020, over 30 million cells have been sequenced across over 1,000 studies (Svennson Pachter http://www.nxn.se/single-cell-studies) and the average size of each study is nearing 100,000 cells.
-The retina was used as the source rissue doe one of the earliest works in the high throughput single cell transcriptomics field (Macosko et al) and as of late 2020, over a dozen published studies containing over a million cells have used single cell technology to profile cell type specific gene expression patterns, cell fate trajectory, tissue and cell differentiation, and disease perturbation.
+As of November 2020, over 30 million cells have been sequenced across over 1,000 studies and the average size of each study is nearing 100,000 cells [@svenssonCuratedDatabaseReveals2020].
+The retina was used as the source tissue in one of the earliest works in the high throughput single cell transcriptomics field and as of late 2020, over a dozen published studies containing over a million cells have used single cell technology to profile cell type specific gene expression patterns, cell fate trajectory, tissue and cell differentiation, and disease perturbation across multiple mammalian organisms [@buenaventuraIdentificationGenesEnriched2019; @clarkComprehensiveAnalysisRetinal2018; @collinCRXExpressionPluripotent2019; @collinDeconstructingRetinalOrganoids2019; @cowanCellTypesHuman2020; @diacouSix3Six6Are2018; @eldredThyroidHormoneSignaling2018; @hengHypoxiaToleranceNorrindeficient2019; @huDissectingTranscriptomeLandscape2019; @logiudiceSinglecellTranscriptionalLogic2019; @lukowskiSinglecellTranscriptomeAtlas2019; @luSingleCellAnalysisHuman2020; @macoskoHighlyParallelGenomewide2015; @menonSinglecellTranscriptomicAtlas2019; @okorenMicroglialFunctionDistinct2019; @pengMolecularClassificationComparative2019; @phillipsNovelApproachSingle2018; @rheaumeSingleCellTranscriptome2018; @ruedaHippoPathwayBlocks2019; @shekharComprehensiveClassificationRetinal2016; @sridharSingleCellTranscriptomicComparison2020; @tranSingleCellProfilesRetinal2019; @voigtMolecularCharacterizationFoveal2019; @voigtSinglecellTranscriptomicsHuman2019; @yanCellAtlasHuman2020].
 
@@
 This issue can be addressed only by re-quantifying the data in a uniform environment.
-Fortunately, due to the continued development of computationally light-weight gene quantification tools into the single-cell space (kbtools, alevin-something), re-quantification does not require massive compute resources.
+Fortunately, due to the continued development of computationally light-weight gene quantification tools into the single-cell space (e.g kallisto bustools, alevin), re-quantification does not require massive compute resources [@melstedModularEfficientPreprocessing2019; @srivastavaAlevinEfficientlyEstimates2019].
 
-Even after re-quantication under identical conditions there remain study specific batch effects due to the variety in single cell technologies used and variation in tissue handling and processing across eajh scientific groups.
-The single cell community has recognized that removal of these technical (also referred to as batch) effects is a critical issue and have independently developed many tools, though it remains unclear which tools and parameters are optimal for a particular dataset.
+Even after re-quantification under identical conditions there remain study specific batch effects due to the variety in single cell technologies used and variation in tissue handling and processing across each scientific group.
+The single cell community has recognized that removal of these technical (also referred to as batch) effects is a critical issue and have independently developed many tools, though it remains unclear which tools and parameters are optimal for a particular dataset [@butlerIntegratingSinglecellTranscriptomic2018; @haghverdiBatchEffectsSinglecell2018; @hieEfficientIntegrationHeterogeneous2019; @johnsonAdjustingBatchEffects2007; @korsunskyFastSensitiveAccurate2019; @liuJointlyDefiningCell2020; @lopezDeepGenerativeModeling2018; @polanskiBBKNNFastBatch2020; @QueryReferenceSinglecell; @simonINSCTIntegratingMillions2020; @stuartComprehensiveIntegrationSingleCell2019; @vandijkRecoveringGeneInteractions2018].
 
@@
 To find the best set of parameters for the creation of scEiaD we did a grid search across key parameters: hyper variable genes (HVG), latent dimensions, and k nearest neighbors.
-Furthermore we used a recent advance in scVI capability that allows one to build a reference model and project the new cells onto it.
-We thus built a human-only scVI model and then projected the mouse and macaque data onto it.
+Furthermore we used a recent advance in scVI capability adapted from scArches that allows one to build a reference model and project the new cells onto it.
+We built a human-only scVI model and then projected the mouse and macaque data onto it.
 We then compared the "scVIprojected" against the against the straight scVI across all the previously mentioned parameters.
@@
 
-To further study cell type specific expression patterning we needed to label the `r format(umapRef %>% nrow() - meta_filter %>% filter(!is.na(CellType)) %>% nrow(), big.mark=",", scientific=FALSE)` unlabelled cells.
+To further study cell type specific expression patterning we needed to label the `r format(umapRef %>% nrow() - meta_filter %>% filter(!is.na(CellType)) %>% nrow(), big.mark=",", scientific=FALSE)` unlabeled cells.
 Traditionally this is done by clustering the cells, then using cell type specific markers to label the clusters.
@@
 The trained model was used to predict the cell type assignments for all cells in scEiaD.
-In this manner we both label most cells (cells which cannot be assigned a cell type assignment with a confidence above 0.5 were left unlabelled) and correct a small number of probable mislabels in the truth set (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a).
+In this manner we both label most cells (cells which cannot be assigned a cell type assignment with a confidence above 0.5 were left unlabeled) and correct a small number of probable mislabels in the truth set (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a).
 
@@
 ```{r figXGboost_umap, fig.width=18, fig.height=13, fig.cap=figXGboost_umap_cap, echo=FALSE, message = FALSE}
-figXGboost_umap_cap <- fig_cap(name='figXGboost_umap', caption =  '')
+figXGboost_umap_cap <- fig_cap(name='figXGboost_umap', caption =  'filler text hello world')
 
@@
 source('figs_and_tables/make_meta_scatter_umap_plot.R')
+source('figs_and_tables/dotplot.R')
+
+# markers for dotplot
+input <- list()
+input[['dotplot_Gene']] <- top_markers %>%  
+  filter(grepl('Amacrine|Rod|Cone|Retinal|Muller|Horizon|Bipol|Astro|RPE|Micro', cluster), D_KL > 0.2) %>%  
+  group_by(cluster)  %>% 
+  slice_max(med_auc, n =5) %>% 
+  pull(Gene)
+input[['dotplot_groups']] <- c('CellType_predict')
+input[['dotplot_filter_cat']] <- 'CellType_predict'
+input[['dotplot_filter_on']] <- (meta_filter %>%
+                                   filter(grepl('Amacrine|Rod|Cone|Retinal|Muller|Horizon|Bipol|Astro|RPE|Micro', CellType_predict)) %>% 
+                                   pull(CellType_predict) %>% unique())
+d <- make_dotplot(input, scEiaD_2020_v01, meta_filter, cat_to_color_df)
+dplot <- d  + facet_grid(vars(cluster), scale = 'free_y', space = 'free') + theme(
+  strip.background = element_blank(),
+  strip.text.y = element_blank()
+)
+
 input <- list()
@@
 input[['meta_column_transform']] <- 'None'
-#input[['meta_filter_cat']] <- 'CellType_predict'
-#input[['meta_filter_on']] <- 'Bipolar Cells'
+input[['meta_filter_cat']] <- 'CellType_predict'
+input[['meta_filter_on']] <- c('AC/HC Precursors','Amacrine Cells','Astrocytes','Bipolar Cells','Cones','Early RPCs','Endothelial','Horizontal Cells','Late RPCs','Microglia','Muller Glia','Neurogenic Cells','Pericytes','PR Precursors','Red Blood Cells','Retinal Ganglion Cells','Rod Bipolar Cells','Rods','RPCs','RPE','Schwann','Tabula Muris')
 ctp2 <- make_meta_scatter_umap_plot(input, mf, meta_filter,
@@
 )
-plot_grid(grid.grabExpr(draw(ct_confusion)), 
-          ctp$plot + theme(legend.position = "none"), 
+
+plot_grid(dplot + theme(legend.title = element_text(angle = 90),
+                        axis.text = element_text(size = 8)),
+          plot_grid(ctp$plot + theme(legend.position = "none"), 
                     ctp2$plot + theme(legend.position = "none") + facet_wrap(~CellType_predict),
+                    grid.grabExpr(draw(ct_confusion)),
                     org$plot + facet_wrap(~organism) + theme(legend.position = "none"), 
                     ncol = 2,
-          labels = c('a', 'b', 'c', 'd'),
-          rel_heights = c(1,0.8))
+                    labels = c('b', 'c', 'd', 'e'),
+                    rel_heights = c(1,0.8)), 
+          ncol = 2, 
+          rel_widths = c(1,3), 
+          labels = c('a',NULL))
 
@@
 Overall we go from an average of `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(2) %>% mean()` studies per human cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(5) %>% mean()` and `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(4) %>% mean()` studies per mouse cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(7) %>% mean()` after transferring the cell type labels.
+
+<!-- # REWRITE? (UNFORTUNATELY) FOR NEW DATA WHEN(?) IT COMES IN -->
+As another check on the quality of the cell type assignments, we ran pairwise differential expression tests between the predicted cell types (see methods for further details).
+We show the five most consistently differentially expressed genes for each of the major retina cell types (`r supFig_cap(name = 'figXGboost_umap', display = 'cite')`a).
+As a simple metric to identify known and unknown genes relating to the cell type specific expression we search PubMed for the number of publications with two searches per gene.
+The first search is the more precise "gene AND cell type" (e.g.
+"PDE6H AND Cones") and the second search is the more inclusive "gene AND retina" (e.g.
+"PDE6G AND Retina").
+Of the 55 genes in (`r supFig_cap(name = 'figXGboost_umap', display = 'cite')`a), 37 had one or more citations in the gene/celltype search (table X) and 47 had one or more citation in the inclusive search.
+The 55 genes had a mean of `r top_markers %>% group_by(cluster)  %>% slice_max(med_auc, n =5) %>% pull(c2) %>% mean() %>% round()` studies (with the inclusive gene by "retina" search).
+In contrast, 100 randomly chosen genes had a mean of `r pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% pull(value) %>% mean() %>% round`.
+
+The four genes with zero hits were IGSF21 (Bipolar Cells), BTBD8 (Bipolar Cells), FRZB (Muller Glia), and TAGLN3 (Retinal Ganglion Cells).
+In a more exhaustive hand literature search we find that IGSF21 is associated with Diabetic Retinopathy in a Han Chinese population (https://pubmed.ncbi.nlm.nih.gov/27607899/).
+BTBD8 contains BTB/POZ domains associated with transcription factor protein-protein interactions (https://www.spandidos-publications.com/10.3892/ijmm.13.1.193).
+FRZB was previously referred to as SFRP3 and this Wnt pathway member has been previously shown to be active in Muller Glia (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2330018/).
+TAGLN3 has been previously reported in chick to be expressed in developing Retinal Ganglion Cells (https://onlinelibrary.wiley.com/doi/abs/10.1002/cne.24241).
+<!-- # REWRITE? (UNFORTUNATELY) FOR NEW DATA WHEN(?) IT COMES IN -->
 
@@
 downloading our scVI model, and 3.
-following the jupyter notebook we provide as supplemental file BLOOP.
-We demonstrate by adding new retina data to scEiaD that was published after our last data pull.
+following the R notebook we provide as supplemental file BLOOP and at http://www.ZENODOSOMETHING.
+We demonstrate by adding new retina data to scEiaD that was published after our last data pull.....
 
-##
 <!-- ## Pan mouse as non-ocular reference to improve retina-specific gene search -->
@@
 
-The full Snakemake pipeline that takes in the raw fastq sequence and outputs the scEiaD is at https://github.com/davemcg/scEiad.
+The full Snakemake pipeline that takes in the raw fastq sequence and outputs the scEiaD is at https://github.com/davemcg/scEiad [@kosterSnakemakeScalableBioinformatics2012].
 The publication commit is BLOOP.
-Furthermore, the repository has been deposited at zenodo under BLOOP.
+Furthermore, the repository has been deposited at Zenodo under BLOOP.
 We will briefly discuss the pipeline choices, programs and algorithms, and versions below.
-For the R packages, we provide versions as the supplementary file "R_session_info.txt"
+For the R packages, we provide package versions as the supplementary file "R_session_info.txt"
 
@@
 
-Kallisto (0.46.2) was the pseudoaligner.
-To reference transcriptomes used to build the kallisto index were the Gencode "gencode.vM25.annotation.gtf.gz" and "gencode.v35.annotation.gtf.gz" for mouse and human, respectively.
-The Ensembl release 99 "Macaca_mulatta.Mmul_10.99.gtf.gz" was used for the macaque transcriptome, For the relatively few single ended samples, the params "--single -l 200 -s 30" were used for kallisto quant.
+Kallisto (0.46.2) was the pseudoaligner [@brayNearoptimalProbabilisticRNAseq2016].
+The reference transcriptomes used to build the kallisto index were the Gencode "gencode.vM25.annotation.gtf.gz" and "gencode.v35.annotation.gtf.gz" for mouse and human, respectively [@harrowGENCODEReferenceHuman2012; @jGENCODEReferenceHuman2012].
+The Ensembl release 101 "Macaca_mulatta.Mmul_101.gtf.gz" was used for the macaque transcriptome [@yatesEnsembl20202020].
+For the relatively few single ended samples, the params "--single -l 200 -s 30" were used for kallisto quant.
 Otherwise the "--bias" flag was added.
-For the droplet-based samples, bustools (0.39.4) was used to extract the single cell barcodes, sort the output, and fix small barcoding errors.
+For the droplet-based samples, bustools (0.39.4) was used to extract the single cell barcodes, sort the output, and fix barcoding errors [@melstedBarcodeUMISet2019].
 
@@
 The BUSpaRse package was used to input the bustools counts mtx file.
-The DropletUtils package with the "barcodeRanks" function was used to automatically detect the inflection point in the barcode count ranks that distinguishes the likely empty droplets.
+The DropletUtils package with the "barcodeRanks" function was used to automatically detect the inflection point in the barcode count ranks that delineates the likely empty droplets [@lunEmptyDropsDistinguishingCells2019].
 We then removed cells with percent mitochondrial reads of >10%.
-After merging the individual count matrices into one sparse matrix, we created a Seurat version 3 object and removed cells with fewer than 200 detected unique genes, and for the droplet data, more than 3000 detected genes (these are likely to be doublets).
+After merging the individual count matrices into one sparse matrix, we created a Seurat version 3 object and removed cells with fewer than 200 detected unique genes, and for the droplet data, more than 3000 detected genes (these are likely to be doublets) [@butlerIntegratingSinglecellTranscriptomic2018].
 
@@
 
-We tested several gene count normalization approaches as we were not certain which would produce an optimal outcome: standard (default Seurat, library size normalization, then log transform), sqrt (same, but with sqrt normalization), libSize (omit the log or sqrt normalization), scran, SCT from Seurat, and for scVI, no normalization (counts).
+We tested several gene count normalization approaches as we were not certain which would produce an optimal outcome: standard (default Seurat, library size normalization, then log transform), sqrt (same, but with sqrt normalization), libSize (omit the log or sqrt normalization), scran, SCT from Seurat, and for scVI, no normalization (counts) [@lunStepbystepWorkflowLowlevel2016].
 Our R implementation of the normalization approaches as well as how we constructed the Seurat v3 object can be found in the supplementary file "make_seurat_obj_functions.R"
@@
 CCA could not complete with the libSize normalization.
-We also tried the DESC, liger, and Conos batch corrections methods but were unable to get them to work reliably so they were not included.
+We also tried the DESC, liger, and Conos batch corrections methods but were unable to get them to work reliably so they were dropped.
 The batch correction step implementation can be found the supplementary file "merge_methods.R".
@@
 ## Clustering and UMAP
-Louvain-Jaccard clustering against the batch corrected latent dimensions used the Seurat implementation.
-We tried the k-nearest neighbors (knn) parameters 5, 7, and 10 (where 5 gives more clusters than 10).
-We also used the leiden algorithm as implemented by PARC with a resolution of 0.6 and 0.8.
+Louvain-Jaccard clustering against the batch corrected latent dimensions used the Seurat implementation [@blondelFastUnfoldingCommunities2008].
+We tried the k-nearest neighbors (knn) parameters 5 and 7 (where 5 gives more clusters than 7).
+We also used the leiden algorithm as implemented by PARC with a resolution of 0.6 and 0.8 (higher results in more clusters) [@traagLouvainLeidenGuaranteeing2019; @stassenPARCUltrafastAccurate2020].
 These two resolutions were chosen as they roughly gave the same number of clusters at the Seurat Louvain-Jaccard aprroach with a knn 7.
 
-The UMAP visualization was calculated with the Seurat "RunUMAP" using the uwot R package.
+The UMAP visualization was calculated with the Seurat "RunUMAP" using the uwot R package [@mcinnesUMAPUniformManifold2020].
 We altered the min.dist parameter with 0.001 and 0.1 and tried n.neighbors across 15, 30, 100, and 500.
@@
 
-We wrote the scPOP R package to unify the LISI and Silhouette metrics from Harmony and kBet, respectively, along with NMI and ARI.
-LISI and Silhouette require a dense matrix, which is a problem for our data as a 1e6 x 20,000 matrix cannot fit even in a 1.5 TB memory node.
-We down-sampled the dataset to ~100,000 cells, taking care to keep rarer cell types for the LISI and Silhouette benchmarking.
+We wrote the scPOP R package to unify the LISI and Silhouette metrics from Harmony and kBet, respectively, along with NMI and ARI [@buttnerTestMetricAssessing2019; @korsunskyFastSensitiveAccurate2019].
+LISI and Silhouette require a dense matrix, which is a problem for our data as a 7e5 x 20,000 matrix cannot fit in a 1.5 TB memory node.
+We down-sampled the dataset to ~100,000 cells, taking care to keep all rarer cell types for the LISI and Silhouette benchmarking.
 
-To merge these metrics into a single score, we Z scale each and sum them.
+To merge these metrics into a balanced single score, we Z scale each and sum them.
 scPOP produces both tables and visualizations allowing the user to quickly see both the interplay of batch mixing and cluster/cell type separation and the overall performance.
@@
 While we removed droplet-based cells with more than 3,000 transcripts as these are highly likely to be doublets (more than one cell per droplet) there were likely many more doublets.
-We ran DoubletDetect and scrublet and calculated the distribution of DoubletDetect and scrublet scores across all clusters and removed clusters with a DoubletDetect and scrublet score more than 4 standard deviations above the mean.
+We ran DoubletDetect and scrublet and calculated the distribution of DoubletDetect and scrublet scores across all clusters and removed clusters with a score in both metrics greather than 4 standard deviations above the mean [@adamgayosoJonathanShorDoubletDetectionDoubletdetection2020; @wolockScrubletComputationalIdentification2019].
 This removed another `r format((umapRef %>% nrow() %>% as.integer()) - (meta_filter %>% nrow() %>% as.integer()), big.mark=",", scientific=FALSE)` cells, leaving `r format((meta_filter %>% nrow()) %>% as.integer(), big.mark=",", scientific=FALSE)` in total.
@@
 
-## trajectory
+## Marker gene identification
 
-DAVID
+scran findmarkers (wilcox test) along with singleCellHaystack. Filtering to keep genes that FDR < 1 or
+
+## Trajectory
+
+To calculate continuous trajectories for individual terminally differentiated retina cell types, we used the Slingshot R package [@streetSlingshotCellLineage2018].
+Slingshot first uses the clusters to build a minimum spanning tree to identify the global lineage structure.
+We specified the starting cluster by identifying the cluster for human and mouse with a high proportion of the early RPC cells and specified the terminal clusters that were predominantly a terminal retina cell type (Amacrine, Bipolar, Cones, Horizontal, Muller Glia, Retinal Ganglion, Rods).
+XXXX trajectories were built and we hand-pruned out redundant or non-sensical (e.g.
+Photoreceptor precursors to Cone to Rod Bipolar) trajectories, leaving XXXX trajectories for downstream analysis.
+To identify major points of gene expression changes during differentiation we built linear models across XXXXXXX bins.
 
@@
 ## Projection of new data onto scEiaD
-
-DAVID
 
@@
   cowplot::plot_grid(lisi + theme(legend.position="none"), silhouette, NULL, rel_widths = c(1,1,0.2), nrow = 1, labels = c('a' ,'b')), 
-  cowplot::plot_grid(ari, nmi, pcr, legend, rel_widths = c(1,1, 1, 0.2), ncol = 4, labels = c('c', 'd', 'e')), 
+  cowplot::plot_grid(ari, nmi, legend, rel_widths = c(1,1, 1, 0.2), ncol = 4, labels = c('c', 'd')), 
   nrow = 2)
@@
 ```{r supFig_PR, fig.width=12, fig.height=10,  fig.cap=supFig_PR_cap, echo=FALSE, message = FALSE}
-supFig_PR_cap <- supFig_cap(name='supFig_PR', caption =  '')
+supFig_PR_cap <- supFig_cap(name='supFig_PR', caption =  'Precision recall curves for our xgboost cell type predictor model across each cell type predicted')
 
@@
 ```{r supTab_AUCxgboost, fig.cap=supTab_AUCxgboost_cap, echo=FALSE, message = FALSE}
-supTab_AUCxgboost_cap <- supTab_cap(name='supTab_AUCxgboost', caption =  'TEXT')
+supTab_AUCxgboost_cap <- supTab_cap(name='supTab_AUCxgboost', caption =  'Area under the precision recall curve (AUC) for each cell type, split by study')
 

### 2020-12-30 figs_and_tables/dotplot.R

@@ -1,117 +1,21 @@
-library(ggplot2)
-library(Cairo)
-library(scattermore)
-library(pals)
-library(ggrepel)
-library(patchwork)
-library(purrr)
-library(pool)
-library(RSQLite)
-library(dplyr)
-library(magick)
-library(stringr)
 library(ggtext)
-load('~/data/massive_integrated_eye_scRNA/Mus_musculus_Macaca_fascicularis_Homo_sapiens__0__5000__counts__universe__batch__scVIprojectionSO__dims10__0.6__500__0.1__CellType_predict.sceWilcox_summary.Rdata')
-#genes <- c('ENSG00000144834','ENSG00000139970','ENSG00000109846','ENSG00000026025','ENSG00000067715','ENSG00000123307','ENSG00000139053','ENSG00000143774','ENSG00000105372','ENSG00000130066','ENSG00000104435','ENSG00000139970','ENSG00000050165','ENSG00000135821','ENSG00000104435','ENSG00000131711','ENSG00000016082','ENSG00000198668','ENSG00000149489','ENSG00000129535','ENSG00000105372','ENSG00000140988')
-
-#markers <- scEiaD_2020_v01 %>% tbl('genes') %>% collect() %>% mutate(ENS = str_extract(Gene, 'ENSG\\d+')) %>% filter(ENS %in% genes) %>% pull(Gene)
-
-# marker_info_DT <- marker_list %>% 
-#   bind_rows(.id = 'cluster') %>% 
-#   group_by(cluster) %>% 
-#   #top_n(5, L4) %>% 
-#   filter(grepl('Amacrine|Rod|Cone|Retinal|Muller|Horizon|Bipol|Astro', cluster)) %>% 
-#   filter(!Gene %in%
-#            (marker_list %>% 
-#            bind_rows(.id = 'cluster') %>% 
-#            group_by(cluster) %>% 
-#            top_n(5, L4) %>% 
-#            filter(grepl('Amacrine|Rod|Cone|Retinal|Muller|Horizon|Bipol|Astro', cluster)) %>% 
-#            group_by(Gene) %>% 
-#            summarise(Count = n()) %>% 
-#            arrange(-Count) %>% 
-#            filter(Count > 1) %>% 
-#            pull(Gene))) %>% 
-#   ungroup() %>% 
-#   group_by(cluster) %>% 
-#   top_n(3, L4)  
-
-marker_list <- list()
[…]

### 2020-12-30 figs_and_tables/fig1.R

@@ -31,8 +31,9 @@ dagified <- dagify('counts' ~ 'fastq',
                    'DiffTesting' ~ 'projectedCellType',
                    'DiffTesting' ~ 'CellType',
                    'DiffTesting' ~ 'cluster',
-                   'DiffTesting' ~ 'LatentDims',
                    'projectedCellType' ~ 'LatentDims',
+                   'Trajectory' ~ 'projectedCellType',
+                   'Trajectory' ~ 'LatentDims',
                    'projectedCellType' ~ 'CellType')
 
 tidy_dagitty(dagified)

### 2020-12-30 figs_and_tables/fig2__integrationPerf.R

@@ -1,47 +1,66 @@
 load('~/data/massive_integrated_eye_scRNA/metrics_2020_11_28.Rdata')
+perfO <- perf
+
+load('~/data/massive_integrated_eye_scRNA/merged_stats_2020_12_28.Rdata')
 
 perf <- perf %>% 
   mutate(knn = case_when(method == 'scArches' ~ 7, TRUE ~ knn))
 cluster_stats <- perf %>% select(clusterN, clusterMedian, dims, nf, knn, method, normalization, set) %>% unique() 
 
-perf_well <- perf %>% unique() %>% filter(set == 'onlyWELL') %>% 
-  select(Score, Group, Value, set, dims:normalization) %>% 
-  filter(Score %in% c('LISI','Silhouette')) 
+perfO <- perfO %>% 
+  mutate(knn = case_when(method == 'scArches' ~ 7, TRUE ~ knn),
+         normalization = case_when(method == 'scArches' ~ 'standard',
+                                   TRUE ~ normalization))
 
-perf_tabula <- perf %>% unique() %>% filter(set %in% c('TabulaDroplet', 'universe')) %>% 
+#cluster_stats <- perf %>% select(clusterN, clusterMedian, dims, nf, knn, method, normalization, set) %>% unique() 
+# 
+# perf_well <- perf %>% unique() %>% filter(set == 'onlyWELL') %>% 
+#   select(Score, Group, Value, set, dims:normalization) %>% 
+#   filter(Score %in% c('LISI','Silhouette')) 
+
+perf_tabula <- perfO %>% unique() %>% filter(set %in% c('TabulaDroplet')) %>% 
+  select(Score, Group, Value, set, dims:normalization) %>% 
+  filter(Score %in% c('LISI','Silhouette', 'ARI', 'PCR', 'NMI')) 
+
+perf_scVI <- perf %>% unique() %>% filter(set %in% c('TabulaDroplet', 'universe')) %>% 
   select(Score, Group, Value, set, dims:normalization) %>% 
   filter(Score %in% c('LISI','Silhouette', 'ARI', 'PCR', 'NMI')) 
 
 lisi <- perf_tabula %>% 
-  filter(Score == 'LISI', nf == 2000, dims %in% c(30), knn == 7) %>%
+  filter(Score == 'LISI', nf == 2000, dims %in% c(8,30), knn == 7) %>%
   pivot_wider(values_from = Value, names_from = c('Group')) %>% 
[…]

### 2020-12-30 figs_and_tables/fig_umap_benchmarking.R

@@ -27,7 +27,7 @@ umap_plot_maker <- function(umap, color_against = 'CellType', red = 'UMAP', ptsi
   # attach colors to cell types
   cell_types <- umapFig %>% 
     pull(CellType) %>% unique() %>% sort()
-  type_val <- setNames(c(pals::brewer.set2(8),pals::alphabet())[1:length(cell_types)], cell_types)
+  type_val <- setNames(c(pals::cols25(),pals::alphabet())[1:length(cell_types)], cell_types)
   type_col <- scale_colour_manual(values = type_val)
   type_fill <- scale_fill_manual(values = type_val)
   # cell type known

### 2020-12-30 figs_and_tables/make_meta_scatter_umap_plot.R

@@ -13,13 +13,13 @@ library(magick)
 library(stringr)
 library(fst)
 
-tabulamuris_predict_labels <-scEiaD_2020_v01 %>% tbl('tabulamuris_predict_labels') %>% collect
+tabulamuris_predict_labels <-scEiaD_2020_v01 %>% tbl('tabulamuris_predict_labels') %>%  collect
 celltype_predict_labels <-scEiaD_2020_v01 %>% tbl('celltype_predict_labels') %>% 
   mutate(CellType_predict = case_when(CellType_predict == 'RPC' ~ 'RPCs',
                                       CellType_predict == 'Mesenchymal/RPE/Endothelial' ~ 'Endothelial',
                                       TRUE ~ CellType_predict)) %>% collect
-celltype_labels <-scEiaD_2020_v01 %>% tbl('celltype_labels') %>% collect
-cluster_labels <-scEiaD_2020_v01 %>% tbl('cluster_labels')
+celltype_labels <-scEiaD_2020_v01 %>% tbl('celltype_labels') %>% collect() %>% mutate(UMAP_1 = UMAP_1 * -1, UMAP_2 = UMAP_2 * -1) 
+cluster_labels <-scEiaD_2020_v01 %>% tbl('cluster_labels') %>% collect() %>% mutate(UMAP_1 = UMAP_1 * -1, UMAP_2 = UMAP_2 * -1) 
 mf <- meta_filter 
 
 # generate color_mappings
@@ -33,7 +33,8 @@ meta_filter <- meta_filter %>% mutate(SubCellType = tidyr::replace_na(SubCellTyp
 
   
 map_color <- function(column, meta_filter){
-  master_colorlist <- c(pals::polychrome()[3:length(pals::polychrome())], pals::alphabet2())
+ # master_colorlist <- c(pals::polychrome()[3:length(pals::polychrome())], pals::alphabet2())
+  master_colorlist <- c(pals::cols25()[1:23],pals::alphabet())
   values <- meta_filter %>% pull(!!column) %>% unique %>% sort
   if(length(values) > length(master_colorlist) ){
     r= round(length(values) / length(master_colorlist)) +1
@@ -186,7 +187,7 @@ celltype_predict_labels <- celltype_predict_labels %>%
            case_when(CellType_predict %in% 
[…]

### 2020-12-30 src/make_dotplot_markers.R

@@ -1,0 +1,126 @@
+library(pool)
+library(RSQLite)
+library(tidyverse)
+library(glue)
+scEiaD_2020_v01 <- dbPool(drv = SQLite(), dbname = "~/data/massive_integrated_eye_scRNA/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-500-0.6.sqlite", idleTimeout = 3600000)
+
+meta_filter <- fst::read_fst('~/git/plaeApp/inst/app/www/meta_filter.fst') %>% as_tibble()
+meta_filter <- meta_filter %>% mutate(CellType_predict = case_when(CellType_predict == 'Photoreceptor Precursors' ~ 'PR Precursors',
+                                                                   CellType_predict == 'AC/HC_Precurs' ~ 'AC/HC Precursors',
+                                                                   CellType_predict == 'RPC' ~ 'RPCs',
+                                                                   CellType_predict == 'Mesenchymal/RPE/Endothelial' ~ 'Endothelial',
+                                                                   TRUE ~ CellType_predict)) %>%
+  mutate(CellType_predict = case_when(!is.na(TabulaMurisCellType_predict) ~ 'Tabula Muris',
+                                      TRUE ~ CellType_predict))
+load('~/data/massive_integrated_eye_scRNA/n_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__knn-0.6__neighbors-500__dist-0.1__CellType_predict.sceWilcox.Rdata')
+#genes <- c('ENSG00000144834','ENSG00000139970','ENSG00000109846','ENSG00000026025','ENSG00000067715','ENSG00000123307','ENSG00000139053','ENSG00000143774','ENSG00000105372','ENSG00000130066','ENSG00000104435','ENSG00000139970','ENSG00000050165','ENSG00000135821','ENSG00000104435','ENSG00000131711','ENSG00000016082','ENSG00000198668','ENSG00000149489','ENSG00000129535','ENSG00000105372','ENSG00000140988')
+
[…]

### 2020-12-30 src/pubmed_query.R

@@ -1,0 +1,75 @@
+# return PMID for query
+library(easyPubMed)
+library(tidyverse)
+#load('~/data/massive_integrated_eye_scRNA/top_markers.Rdata')
+library(pool)
+library(RSQLite)
+# scEiaD_2020_v01 <- dbPool(drv = SQLite(), dbname = "~/data/massive_integrated_eye_scRNA/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-500-0.6.sqlite", idleTimeout = 3600000)
+
+pubmed_counter <- function(query, sleep_time = 1){
+  Sys.sleep(sleep_time)
+  pub_query <- query
+  entrez_id <- get_pubmed_ids(pub_query)
+  if (entrez_id$Count == 0){
+    out <- NA
+  } else {
+    abstracts_txt <- fetch_pubmed_data(entrez_id, format = "abstract") 
+    out <- grep('PMID', abstracts_txt, value = TRUE) %>% str_extract(., '\\d\\d\\d+') %>% unique()}
+  out
+}
+
+# x <- marker_info %>% 
+#   left_join(exp_stats %>% dplyr::rename(cluster = CellType_predict)) %>% 
+#   #filter((FDR < 1 |  med_auc > 0.20 | L1 > 50), D_KL > 0.15,`%` > 20) %>%
+#   filter((FDR < 1 |  med_auc > 0.20 | L1 > 50), `%` > 20) %>% 
+#   #group_by(cluster) %>% 
+#   #top_n(100, med_auc) %>% 
+#   ungroup() %>% 
+#   group_by(Gene) %>% 
+#   slice_max(order_by = `%`) %>% 
+#   mutate(pm_query = paste0(gsub(' \\(.*','', Gene), ' AND ', cluster) %>% 
+#            gsub('Cells','',.)) %>% 
+#   mutate(pm_query2 = paste0(gsub(' \\(.*','', Gene), ' AND Retina'))
+
+x <- top_markers %>% 
+    mutate(pm_query = paste0(gsub(' \\(.*','', Gene), ' AND ', cluster) %>%
+             gsub('Cells','',.)) %>%
+    mutate(pm_query2 = paste0(gsub(' \\(.*','', Gene), ' AND Retina'))
+pmid <- list()
+for (i in x$pm_query){
+  print(i)
[…]

### 2021-01-15 manuscript_00.Rmd
updates

@@
       - '--lua-filter=author-info-blocks.lua'
-bibliography: references2.bib
+bibliography: references.bib
 csl: investigative-ophthalmology-and-visual-science.csl
@@
 We developed an R package (scPOP) that can scan the results of multiple methods and key parameter choices and create a ranked list of the best performing method / parameter combinations.
-We use this package along with a Snakefile based workflow system to demonstrate how to optimally merge over 1.4 million cells from 30 datsets and three species to create a massive ocular single cell transcriptome meta-atlas.
+We use this package along with a Snakefile based workflow system to demonstrate how to optimally merge over one million cells from 30 datsets and three species to create a massive ocular single cell transcriptome meta-atlas.
 This provides a model how to efficiently create meta-atlases for tissues and cells of interest."
@@
 
-scEiaD_2020_v01 <- dbPool(drv = SQLite(), dbname = "~/data/massive_integrated_eye_scRNA/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-500-0.6.sqlite", idleTimeout = 3600000)
+scEiaD_2020_v01 <- dbPool(drv = SQLite(), dbname = "~/data/scEiaD/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.2-500-0.6.sqlite", idleTimeout = 3600000)
 
@@
                                                                    TRUE ~ CellType_predict),
-                                      UMAP_1 = UMAP_1 * -1,
+                                      UMAP_1 = UMAP_1 * 1,
                                       UMAP_2 = UMAP_2 * -1) %>%
@@
                                       TRUE ~ CellType_predict))
-# cell num  before QC removal (mito / likely doublets)
-#qc <- read_tsv('~/data/massive_integrated_eye_scRNA/mito_counts.tsv')
 
 # meta data before doublet removal
-load('~/data/massive_integrated_eye_scRNA/n_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__preFilter.scEiaDprojected__dist-0.1__neighbors-500.umap.Rdata')
+load('~/data/scEiaD/n_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__preFilter.scEiaDprojected__dist-0.2__neighbors-500.umap.Rdata')
 umapRef <- umap
 
-methods <- factor(c('scArches', 'bbknn','insct','magic', 'scVI','CCA', 'scanorama', 'harmony', 'fastMNN', 'combat', 'none', 'liger'), levels = c('bbknn','CCA','combat','fastMNN','harmony','insct','liger','magic','scanorama','scArches','scVI', 'none'))
+methods <- factor(c('scArches', 'bbknn','insct','magic', 'scVI','CCA', 'scanorama', 'harmony', 'fastMNN', 'combat', 'none'), levels = c('bbknn','CCA','combat','fastMNN','harmony','insct','magic','scanorama','scArches','scVI', 'none'))
 
-gse <- read_tsv('~/git/massive_integrated_eye_scRNA/data/GEO_Study_Level_Metadata.tsv')
+gse <- read_tsv('~/git/scEiaD/data/GEO_Study_Level_Metadata.tsv')
 
-qc <- readr::read_tsv('~/data/massive_integrated_eye_scRNA/QC.tsv.gz')
+qc <- readr::read_tsv('~/data/scEiaD/QC.tsv.gz')
 
 # cell type prediction
-load('~/data/massive_integrated_eye_scRNA/PREDICTIONSn_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__preFilter.scEiaDprojected__dist-0.1__neighbors-500.Rdata')
-test_predictions <- data.table::fread('~/data/massive_integrated_eye_scRNA/277874_CTpredictortest_data_probabilities.csv.gz') %>% as_tibble %>% 
+load('~/data/scEiaD/PREDICTIONSn_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__preFilter.scEiaDprojected__dist-0.2__neighbors-500.Rdata')
+test_predictions <- data.table::fread('~/data/scEiaD/981791_CTpredictortest_data_probabilities.csv.gz') %>% as_tibble %>% 
   mutate(max_pred_prob = rowMaxs(.[,-(33:36)] %>% as.matrix ), 
@@
 # created by running 'source' on src/make_dotplot_markers.R
-load('~/data/massive_integrated_eye_scRNA/top_markers.Rdata')
+load('~/data/scEiaD/top_markers.Rdata')
 
 # pmid search
-load('~/data/massive_integrated_eye_scRNA/top_marker_pmid.Rdata')
+load('~/data/scEiaD/top_marker_pmid.Rdata')
 ```
@@
   group_by(cluster)  %>% 
-  slice_max(med_auc, n =5) %>% 
+  slice_max(mean_auc, n =5) %>% 
   pull(Gene)
@@
 plot_grid(dplot + theme(legend.title = element_text(angle = 90),
-                        axis.text = element_text(size = 8)),
+                        axis.text = element_text(size = 6)),
           plot_grid(ctp$plot + theme(legend.position = "none"), 
@@
 
-<!-- # REWRITE? (UNFORTUNATELY) FOR NEW DATA WHEN(?) IT COMES IN -->
 As another check on the quality of the cell type assignments, we ran pairwise differential expression tests between the predicted cell types (see methods for further details).
@@
 "PDE6G AND Retina").
-Of the 55 genes in (`r supFig_cap(name = 'figXGboost_umap', display = 'cite')`a), 37 had one or more citations in the gene/celltype search (table X) and 47 had one or more citation in the inclusive search.
-The 55 genes had a mean of `r top_markers %>% group_by(cluster)  %>% slice_max(med_auc, n =5) %>% pull(c2) %>% mean() %>% round()` studies (with the inclusive gene by "retina" search).
+Of the 55 genes in (`r supFig_cap(name = 'figXGboost_umap', display = 'cite')`a), 40 had one or more citations in the gene/celltype search (table X) and 51 had one or more citation in the inclusive search.
+The 55 genes had a mean of `r top_markers %>% group_by(cluster)  %>% slice_max(mean_auc, n =5) %>% pull(c2) %>% mean() %>% round()` studies (with the inclusive gene by "retina" search).
 In contrast, 100 randomly chosen genes had a mean of `r pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% pull(value) %>% mean() %>% round`.
 
-The four genes with zero hits were IGSF21 (Bipolar Cells), BTBD8 (Bipolar Cells), FRZB (Muller Glia), and TAGLN3 (Retinal Ganglion Cells).
-In a more exhaustive hand literature search we find that IGSF21 is associated with Diabetic Retinopathy in a Han Chinese population (https://pubmed.ncbi.nlm.nih.gov/27607899/).
-BTBD8 contains BTB/POZ domains associated with transcription factor protein-protein interactions (https://www.spandidos-publications.com/10.3892/ijmm.13.1.193).
-FRZB was previously referred to as SFRP3 and this Wnt pathway member has been previously shown to be active in Muller Glia (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2330018/).
-TAGLN3 has been previously reported in chick to be expressed in developing Retinal Ganglion Cells (https://onlinelibrary.wiley.com/doi/abs/10.1002/cne.24241).
-<!-- # REWRITE? (UNFORTUNATELY) FOR NEW DATA WHEN(?) IT COMES IN -->
+The four genes with zero hits were SCRG1 (Astrocytes), BTBD8 (Bipolar Cells), GATM (Microglia), and FRZB (Muller Glia).
+In a more exhaustive literature search we find that SCRG1 is a secreted protein which marks brain neurons, regulates human mesenchymal stem cells differentiation, and is overexpressed in brain astrocyte precursors [@aomatsuNovelSCRG1BST12014; @chaboubTemporalProfilingAstrocyte2016].
+BTBD8 contains BTB/POZ domains known to generally be associated with with transcription factor mediated protein-protein interactions [@xuMolecularCloningCharacterization2004; @zollmanBTBDomainFound1994].
+GATM is mitochondriale nzyme involved in creatine biosynthesis and was previously shown to upregulated in a brain microglia population [@hammondSingleCellRNASequencing2019].
+FRZB was previously referred to as SFRP3 and this Wnt pathway member has been previously shown to be active in Muller Glia [@zhuCharacterizationCanonicalWnt2018].
+Overall, we find very high concordance with known retina cell type markers.
 
@@
 
-The 2D UMAP projection of the scVI-calculated batch corrected 10 latent dimensional space blends the `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` studies together while also maintaining distinct space for the `r meta_filter$CellType_predict %>% table() %>% length()` unique cell types.
+The 2D UMAP projection of the scVI-calculated batch corrected 8 latent dimensional space blends the `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` studies together while also maintaining distinct space for the `r meta_filter$CellType_predict %>% table() %>% length()` unique cell types.
 We also see good mixing across all the droplet and well based single cell technologies (`r supFig_cap(name = 'supFig_platformInt', display = 'cite')`).
@@
 
-## Pseudotime recapitulates and improves resolution of developmental processes
-
 ## Projection of new data onto scEiaD
 
-The hard work of creating this resource can be leveraged by the wider community with a few relatively simple steps.
+The hard work of creating this resource can be leveraged and extended by the wider community with relatively few simple steps and modest compute requirements.
 Very briefly, if outside groups quantify their mouse or human scRNA with kallisto and the same references (see methods), they can overlay their data on top of scEiaD by 1.
 installing scVI >=0.8.0, 2.
-downloading our scVI model, and 3.
-following the R notebook we provide as supplemental file BLOOP and at http://www.ZENODOSOMETHING.
-We demonstrate by adding new retina data to scEiaD that was published after our last data pull.....
+downloading our ~13 megabyte scVI model, and 3.
+following the Jupyter notebook on Google colab that we provide as supplemental file BLOOP and at http://colab.goo.
+We demonstrate by adding self-generated iPSC-based RPE along with previously published retina organoid and cell line data to scEiaD.
+
+Projection of new data onto scEiaD potentially allows for semi-direct comparison of other species, *in vitro* systems, and pharamacologic / mutational perturbation of mammalian retina cells.
 
@@
 
-scran findmarkers (wilcox test) along with singleCellHaystack. Filtering to keep genes that FDR < 1 or
+scran findmarkers (wilcox test) along with singleCellHaystack. Filtering to keep genes that mean AUC > 0.2 or with a D KL > 0.2.
 
-## Trajectory
+<!-- ## Trajectory -->
 
-To calculate continuous trajectories for individual terminally differentiated retina cell types, we used the Slingshot R package [@streetSlingshotCellLineage2018].
+<!-- To calculate continuous trajectories for individual terminally differentiated retina cell types, we used the Slingshot R package [@streetSlingshotCellLineage2018] on the human subset.
 Slingshot first uses the clusters to build a minimum spanning tree to identify the global lineage structure.
-We specified the starting cluster by identifying the cluster for human and mouse with a high proportion of the early RPC cells and specified the terminal clusters that were predominantly a terminal retina cell type (Amacrine, Bipolar, Cones, Horizontal, Muller Glia, Retinal Ganglion, Rods).
+We specified the starting cluster by identifying the cluster for human and mouse with a high proportion of the early RPC cells and specified the end clusters that were predominantly a terminal retina cell type (Amacrine, (Rod) Bipolar, Cones, Horizontal, Muller Glia, Retinal Ganglion, Rods).
+As some clusters were combinations of different (but similar) cell types, we "subclustered" the clusters by splitting them into smaller groups if they contained two or more cell types with > 20% ratio of a cell type.
 XXXX trajectories were built and we hand-pruned out redundant or non-sensical (e.g.
 Photoreceptor precursors to Cone to Rod Bipolar) trajectories, leaving XXXX trajectories for downstream analysis.
-To identify major points of gene expression changes during differentiation we built linear models across XXXXXXX bins.
+-->
 
-## velocity
+<!-- To identify major points of gene expression changes during differentiation we built linear models across XXXXXXX bins.  -->
 
-VINAY
+<!-- ## velocity -->
+
+<!-- To id -->
 

### 2021-01-15 manuscript_00.Rmd
updates

@@
 
+We demonstrate with a provided iPSC-based RPE cell line that we sequenced that non-tissue can be projected onto the scEiaD
+
 <!-- ## Pan mouse as non-ocular reference to improve retina-specific gene search -->
@@
 
-The full Snakemake pipeline that takes in the raw fastq sequence and outputs the scEiaD is at https://github.com/davemcg/scEiad [@kosterSnakemakeScalableBioinformatics2012].
+The set of Snakemake pipelines that takes in the raw fastq sequence and outputs the scEiaD is at https://github.com/davemcg/scEiad [@kosterSnakemakeScalableBioinformatics2012].
 The publication commit is BLOOP.
@@
 
+Gene quantification is handled by the SnakQUANT snakefile.
 Kallisto (0.46.2) was the pseudoaligner [@brayNearoptimalProbabilisticRNAseq2016].
@@
 
-VINAY DISCUSS
+To facilitate comparison of gene expression across species, where possible we converted mouse and macaque gene ids and names to human ones.
+We downloaded a mapping of orthologous genes between human, mouse, and macaque using the Ensembl BioMart web browser in November 2020.
+We identified 15XXX human genes that could be directly mapped to mouse and macaque orthologs.
+Genes present in mouse or macaque that were not found in human were not used for scVI batch correction, but were retaomed and used for differential gene expression.
 
-## Custom macaque reference quant
+## Custom macaque reference quantification
 
-As we noticed that several retina marker genes (WHICH ONES!!!!) had unusually low expression in the macaque data we quantified the scRNA data twice: once with the Ensembl reference and again with the same Gencode human reference used for the human data.
-We compared the gene-level counts for each cell and replaced the macaque gene count with the human counts if the human counts were higher.
-VINAY AM I MISSING DETAILS?
+As we noticed that several retina marker genes (e.g.
+NRL and CRX) had very low expression in the macaque data we quantified the scRNA data twice: once with the Ensembl reference and again with the same Gencode human reference used for the human data.
+We compared the gene-level counts for each cell and replaced the macaque gene count with the human counts if the human counts were greater than the macaque counts and, to prevent genes with very few total counts from being used, we required the counts greater than the first quartile of non-zero macaque gene expression.
 
@@
 
-VINAY WRITE
+In order to identify cell types for the `r meta_filter %>% filter(is.na(CellType), is.na(TabulaMurisCellType))` unlabelledcells we designed a custom xgboost based cell type classifier.
+We took labeled data and split it into training and test sets, stratified by cell type.
+The input features used to train the model are the scVI latent dimensions, the total number of reads in each cell, the number genes detected in each cell, and the percent mitochondrial gene expression of each cell.
+We additionally generated features using the age of each sample by group sample into three developmental categories (Early Development, Late Development, and Adult) and then generated a one-hot encoded feature for each category.
+In order to speed up training times, we used the gpu implementation of the xgboost algorithm from the the xgboost python library.
+The model was trained using default parameters.
+The trained model had an overall F1 score of XXX an average precision and recall of XXX and XXX for each major retinal cell type.
+This model was then used to identify labels for each unlabelled cell.
+Unlabelled data was preprocessed identically to training data and fed into model to generate a vector of label probabilities for each cell.
+We selected the highest label probability for each cell, and required a minimum probability of .5 to assign a label to a cell.
 

### 2021-01-15 figs_and_tables/fig1.R
code update

@@ -21,16 +21,21 @@ a <- meta_filter %>%
 dagified <- dagify('counts' ~ 'fastq',
                    'SnakePOP' ~ 'CellType',
                    'SnakePOP' ~ 'counts',
-                   'LatentDims' ~ 'SnakePOP',
+                   'LatentDims' ~ 'scVIrefquery',
                    'cluster' ~ 'LatentDims',
                    'umap' ~ 'LatentDims' ,
                    'ML' ~ 'CellType',
                    'projectedCellType' ~ 'ML',
                    'ML' ~ 'LatentDims',
+                   'scVIrefquery' ~ 'SnakePOP',
                    'DiffTesting' ~ 'counts',
                    'DiffTesting' ~ 'projectedCellType',
                    'DiffTesting' ~ 'CellType',
                    'DiffTesting' ~ 'cluster',
+                   'SnakePOP' ~ 'BatchInfo',
+                   'scVIrefquery' ~ 'BatchInfo',
+                   'Trajectory' ~ 'cluster',
+                   'Velocity' ~ 'LatentDims',
                    'projectedCellType' ~ 'LatentDims',
                    'Trajectory' ~ 'projectedCellType',
                    'Trajectory' ~ 'LatentDims',
@@ -38,14 +43,25 @@ dagified <- dagify('counts' ~ 'fastq',
 
 tidy_dagitty(dagified)
 set.seed(2534)
-b <- as_tbl_graph(dagified) %>% mutate(bop = name) %>% ggraph(layout = 'graphopt') + 
+b <- tidy_dagitty(dagified) %>% as_tbl_graph() %>% 
+  mutate(type = case_when(name %in% c('fastq', 'CellType', 'BatchInfo') ~ 'Input',
+                          name %in% c('DiffTesting', 'Velocity', 'Trajectory','umap', 'projectedCellTypes','cluster') ~ 'Outputs'),
+         name = case_when(name == 'CellType' ~ 'Published\nCell Types',
+                          name == 'projectedCellType' ~ 'Learned\nCell Types',
+                          name == 'DiffTesting' ~ 'Diff\nTesting',
+                          name == 'scVIrefquery' ~ 'scVI Ref +\nQuery',
+                          name == 'LatentDims' ~ 'Latent Dims',
+                          name == 'BatchInfo' ~ 'Batch Info',
[…]

### 2021-01-15 figs_and_tables/fig2__integrationPerf.R

@@ -1,7 +1,7 @@
-load('~/data/massive_integrated_eye_scRNA/metrics_2020_11_28.Rdata')
+load('~/data/scEiaD//metrics_2020_11_28.Rdata')
 perfO <- perf
 
-load('~/data/massive_integrated_eye_scRNA/merged_stats_2020_12_28.Rdata')
+load('~/data/scEiaD//merged_stats_2021_01_01.Rdata')
 
 perf <- perf %>% 
   mutate(knn = case_when(method == 'scArches' ~ 7, TRUE ~ knn))

### 2021-01-15 figs_and_tables/fig_umap_benchmarking.R

@@ -1,9 +1,9 @@
 library(scattermore)
-source('figs_and_tables/fig2__integrationPerf.R')
-load('~/data/massive_integrated_eye_scRNA/Mus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features2000__libSize__TabulaDroplet__batch__none__dims30__preFilter__mindist0.3__nneighbors30.umap.Rdata')
+source('~/git/scEiaD_manuscript/figs_and_tables/fig2__integrationPerf.R')
+load('~/data/scEiaD/n_features-2000__transform-standard__partition-TabulaDroplet__covariate-batch__method-none__dims-30__dist-0.3__neighbors-15__knn-7__umap.Rdata')
 umapNone <- umap
 
-load('~/data/massive_integrated_eye_scRNA/Mus_musculus_Macaca_fascicularis_Homo_sapiens__n_spec_genes-0__n_features2000__libSize__TabulaDroplet__batch__combat__dims30__preFilter__mindist0.3__nneighbors30.umap.Rdata')
+load('~/data/scEiaD/n_features-2000__transform-standard__partition-TabulaDroplet__covariate-batch__method-combat__dims-30__dist-0.3__neighbors-15__knn-7__umap.Rdata')
 umapCombat <- umap
 
 umap_plot_maker <- function(umap, color_against = 'CellType', red = 'UMAP', ptsize = 2){

### 2021-01-15 figs_and_tables/make_meta_scatter_umap_plot.R

@@ -17,9 +17,10 @@ tabulamuris_predict_labels <-scEiaD_2020_v01 %>% tbl('tabulamuris_predict_labels
 celltype_predict_labels <-scEiaD_2020_v01 %>% tbl('celltype_predict_labels') %>% 
   mutate(CellType_predict = case_when(CellType_predict == 'RPC' ~ 'RPCs',
                                       CellType_predict == 'Mesenchymal/RPE/Endothelial' ~ 'Endothelial',
-                                      TRUE ~ CellType_predict)) %>% collect
-celltype_labels <-scEiaD_2020_v01 %>% tbl('celltype_labels') %>% collect() %>% mutate(UMAP_1 = UMAP_1 * -1, UMAP_2 = UMAP_2 * -1) 
-cluster_labels <-scEiaD_2020_v01 %>% tbl('cluster_labels') %>% collect() %>% mutate(UMAP_1 = UMAP_1 * -1, UMAP_2 = UMAP_2 * -1) 
+                                      TRUE ~ CellType_predict)) %>% collect() %>% 
+  mutate(UMAP_1 = UMAP_1 * 1, UMAP_2 = UMAP_2 * -1) 
+celltype_labels <-scEiaD_2020_v01 %>% tbl('celltype_labels') %>% collect() %>% mutate(UMAP_1 = UMAP_1 * 1, UMAP_2 = UMAP_2 * -1) 
+cluster_labels <-scEiaD_2020_v01 %>% tbl('cluster_labels') %>% collect() %>% mutate(UMAP_1 = UMAP_1 * 1, UMAP_2 = UMAP_2 * -1) 
 mf <- meta_filter 
 
 # generate color_mappings

### 2021-01-16 figs_and_tables/make_meta_scatter_umap_plot.R
plotting code fix

@@ -1,4 +1,3 @@
-
 library(ggplot2)
 library(Cairo)
 library(scattermore)
@@ -79,17 +78,15 @@ make_meta_scatter_umap_plot <- function(input, mf, meta_filter,
     meta_filter[,meta_column] <- log2(meta_filter[,meta_column] + 1)
   }
   p_data <- meta_filter %>%
-    filter(!grepl('Doub|\\/Margin\\/Periocular', CellType)) %>%
-    filter(!is.na(!!as.symbol(meta_column)))
+    #filter(!grepl('Doub|\\/Margin\\/Periocular', CellType)) %>%
+    filter_at(vars(all_of(input$meta_filter_cat)), all_vars(. %in% input$meta_filter_on))
   
   # metadata NUMERIC plot --------------
   if (is.numeric(meta_filter[,meta_column] %>% pull(1)) ){
     color_range <- range(p_data[,meta_column] %>% pull(1))
     suppressWarnings(plot <- ggplot() +
-                       geom_scattermost(cbind(mf %>%
-                                                filter(is.na(!!as.symbol(meta_column))) %>% pull(UMAP_1),
-                                              mf %>%
-                                                filter(is.na(!!as.symbol(meta_column))) %>% pull(UMAP_2)),
+                       geom_scattermost(cbind(mf %>% pull(UMAP_1),
+                                              mf %>% pull(UMAP_2)),
                                         pointsize = pt_size - 1, color = '#D3D3D333',
                                         pixels = c(1750,750)) +
                        geom_scattermost(cbind(p_data$UMAP_1, p_data$UMAP_2),
@@ -129,17 +126,15 @@ make_meta_scatter_umap_plot <- function(input, mf, meta_filter,
     names(color_list) <- color_data$value
     
     suppressWarnings(plot <- ggplot() +
-                       geom_scattermost(cbind(mf %>%
-                                                filter(is.na(!!as.symbol(meta_column))) %>% pull(UMAP_1),
-                                              mf %>%
-                                                filter(is.na(!!as.symbol(meta_column))) %>% pull(UMAP_2)),
-                                        pointsize = pt_size, color = '#D3D3D333',
+                       geom_scattermost(cbind(mf  %>% pull(UMAP_1),
+                                              mf %>% pull(UMAP_2)),
+                                        pointsize = input$pt_size_back, color = '#D3D3D333',
                                         pixels = c(1750,1750)) +
-                       geom_scattermore(data = p_data,
+                       geom_scattermore(data = p_data, alpha = 0.2,
[…]

### 2021-01-16 manuscript_00.Rmd
text updates

@@
 
-The hard work of creating this resource can be leveraged and extended by the wider community with relatively few simple steps and modest compute requirements.
+The hard work of creating this resource can be leveraged and extended by the wider community with a few relatively simple steps and modest compute requirements.
 Very briefly, if outside groups quantify their mouse or human scRNA with kallisto and the same references (see methods), they can overlay their data on top of scEiaD by 1.
@@
 downloading our ~13 megabyte scVI model, and 3.
-following the Jupyter notebook on Google colab that we provide as supplemental file BLOOP and at http://colab.goo.
-We demonstrate by adding self-generated iPSC-based RPE along with previously published retina organoid and cell line data to scEiaD.
+following the Jupyter notebook on Google colab that we provide as supplemental file BLOOP and at http://colab.google.
+We demonstrate the power of this approach in two ways.
 
-Projection of new data onto scEiaD potentially allows for semi-direct comparison of other species, *in vitro* systems, and pharamacologic / mutational perturbation of mammalian retina cells.
+First, we project another retina dataset onto scEiaD from Fadl et al..
+As this is a pan-retina tissue sample, we would expect most of the major cell types of the retina to be present.
+In the Google colab notebook we provide a short set of the reads from SRR12130660 can be processed from the raw reads to a UMAP overlapping scEiaD in under 15 minutes.
+Indeed, as expected we see many amacrine, rods, muller glia, bipolar, a few cones and retinal ganglion, and a small number of non-retina cells (`r supFig_cap(name = 'projection_fig', display = 'cite')`b).
 
-We demonstrate with a provided iPSC-based RPE cell line that we sequenced that non-tissue can be projected onto the scEiaD
-
-<!-- ## Pan mouse as non-ocular reference to improve retina-specific gene search -->
-
-<!-- Do some sloppy/easy thing where we identify retina specific genes by searching across the TM data? -->
+Second, we demonstrate how the RPE derived from the Bharti iPSC differentiation method closely match tissue RPE from other studies (the largest being the Cowan et al data; `r supFig_cap(name = 'projection_fig', display = 'cite')`a) This provides another layer of evidence that the Bharti iPSC-RPE differentation protocol generates RPE cells that are functionally equivalent to human adult RPE.
 

### 2021-01-16 manuscript_00.Rmd
conclusion first draft

@@
 BTBD8 contains BTB/POZ domains known to generally be associated with with transcription factor mediated protein-protein interactions [@xuMolecularCloningCharacterization2004; @zollmanBTBDomainFound1994].
-GATM is mitochondriale nzyme involved in creatine biosynthesis and was previously shown to upregulated in a brain microglia population [@hammondSingleCellRNASequencing2019].
+GATM is mitochondrial enzyme involved in creatine biosynthesis and was previously shown to up-regulated in a brain microglia population [@hammondSingleCellRNASequencing2019].
 FRZB was previously referred to as SFRP3 and this Wnt pathway member has been previously shown to be active in Muller Glia [@zhuCharacterizationCanonicalWnt2018].
@@
 
-## Projection of new data onto scEiaD
+## Projection of new data onto scEiaD demonstrates how iPSC-based cells are similar to tissue-based cells
 
@@
 Very briefly, if outside groups quantify their mouse or human scRNA with kallisto and the same references (see methods), they can overlay their data on top of scEiaD by 1.
-installing scVI >=0.8.0, 2.
-downloading our ~13 megabyte scVI model, and 3.
-following the Jupyter notebook on Google colab that we provide as supplemental file BLOOP and at http://colab.google.
+installing scVI >=0.8.1, 2.
+downloading our 13 megabyte scVI model, and 3.
+following the Jupyter notebook on Google colab that we provide as supplemental file BLOOP and as a live demo at http://colab.google.
 We demonstrate the power of this approach in two ways.
 
-First, we project another retina dataset onto scEiaD from Fadl et al..
+First, we project another retina dataset onto scEiaD from Fadl et al .
 As this is a pan-retina tissue sample, we would expect most of the major cell types of the retina to be present.
-In the Google colab notebook we provide a short set of the reads from SRR12130660 can be processed from the raw reads to a UMAP overlapping scEiaD in under 15 minutes.
+In the Google colab notebook we show how a set of the reads from SRR12130660 can be processed from the raw reads to a UMAP visualization overlapping our scEiaD in under 15 minutes.
 Indeed, as expected we see many amacrine, rods, muller glia, bipolar, a few cones and retinal ganglion, and a small number of non-retina cells (`r supFig_cap(name = 'projection_fig', display = 'cite')`b).
 
-Second, we demonstrate how the RPE derived from the Bharti iPSC differentiation method closely match tissue RPE from other studies (the largest being the Cowan et al data; `r supFig_cap(name = 'projection_fig', display = 'cite')`a) This provides another layer of evidence that the Bharti iPSC-RPE differentation protocol generates RPE cells that are functionally equivalent to human adult RPE.
+Second, we demonstrate how the RPE derived from the Bharti iPSC differentiation method closely match tissue RPE from other studies (the largest being the Cowan et al data; `r supFig_cap(name = 'projection_fig', display = 'cite')`a).
+Cells generated with this process have been extensively validated with functional RPE tests.
+Our scRNA approach provides another layer of evidence that the Bharti iPSC-RPE differentation protocol generates RPE cells that are functionally equivalent to human adult RPE in an unbiased manner.
 
@@
 
-scran findmarkers (wilcox test) along with singleCellHaystack. Filtering to keep genes that mean AUC > 0.2 or with a D KL > 0.2.
+To identify marker genes across the CellType (predict) and cluster groups, we used the scran findmarkers (wilcox test) along with the singleCellHaystack algorithm.
+The scran findmarkers test runs a wilcox test in a pairwise manner (e.g.
+Rods vs all other cell types).
+It returns an overall p-value (and FDR) that assesses how well the gene is at separating the group of interest from all other cells.
+It also returns for each pair-wise comparison an area under the curve (AUC) score, where 1 is a perfect power to distinguish and 0 is no power.
+The singleCellHaystack uses a Kullback-Leibler divergence measurement of the scVI lower dimensional space to identify genes with non-random distribution.
+A higher "D KL" score represents a gene with "specific" expression in teh UMAP/lower dimensional space.
+We filtered to keep genes with FDR < 0.01 and a mean AUC > 0.2 or  a D KL > 0.2.
 
@@
 
-## Projection of new data onto scEiaD
-
 # Conclusion
 
-## Self built meta-atlases are easy as 1,2, benchmark a bunch, 3
+# Limitations
+
+The scVI model is first built the on human data.
+The mouse and macaque data are then projected (or queried) onto it with the scVI/scArches approach.
+While this system works very well to integrate information between these three species, this approach may not scale to more distantly related species.
+Another limitations is that discrepancies between cell type labels between different labs makes certain transitioning cell type labels a bit imprecise.
+One example is how the rods and photoreceptor precursor labels overlap somewhat.
+Though we attempted to ameliorate the issue by removing cell type labels in large disagreement with the consensus our cell type labels are effectively "crowd-sourced" and some disagreements can propagate into our labeled cells.
+
+## The scEiaD is a unique ocular resource that provides a highly diverse, large N dataset with a relatively small amount of compute power
+
+We have assembled the largest ocular single cell transcriptome database to date.
+The rapid of advancement of algorithms to batch correct and process data continue to reduce the computational requirements to handle huge numbers of cells.
+The scVI batch correction step on around one million cells runs within a few hours on a GPU and 150GB of memory.
+This places this crucial step within the capabilities of a moderately large physical server or a cloud compute node.
+Further downstream proessing can largely be done a computers with 64+ Gb of memory and a few hundred GB of disk space.
+We believe that our efforts can be replicated in any other tissue / system with a large number of independent studies by an individual lab or small number of scientists.
+We provide the completed analysis as both Scanpy (h5ad) and Seurat objects at https://github.com/davemcg/scEiaD.
+
+## Benchmarking and quantitation of integration performance is crucial for meta-atlas studies
+
+We originally intended to just use the Seurat CCA method to integrate the datasets.
+However, the long run-times of CCA and poor integration of our known cell types led us to benchmark more methods and parameters.
+We also originally attempted to "hand-assess" the integration results by using the UMAP 2D projection view.
+This proved to scale poorly and this led use to curate some of the most useful benchmarking algorithms (NMI, ARI, silhouette, and LISI) that roughly matched our "hand-assessed" results into the scPOP R package.
+While we chose the scVI algorithm, we do not believe that it will work best for all datasets.
+We strongly suggest any other groups attempting a similar meta-atlas construction chose a quantifiable set of criteria so optimal methods and parameters can be picked.
 
@@
 
-## The scEiaD dataset provides a highly diverse, large N resource for methods development
-
-## The scEiaD is a unique ocular community resource
+We hand curate over 300,000 cell type labels.
+With a standard xgboost algorithm using the latent dimensions, cell age classification (developing or matured), and the UMAP coordinates, we can very accurately label the remaining cells.
+Many other cell type labelling algorithms and systems exist for those groups less willing or able to tune a machine learning algorithm.
+For example, the developers of scVI also have a cell type label projection algorithm called scANVI.
+Whichever approach you use, taking a smaller number of high quality labels and projecting them onto the remaining cells is a powerful way to leverage community knowledge across a huge diverse dataset.
 
 ## Projection allows community knowledge to be leveraged by all
+
+Many retina "atlases" have been published to date.
+We argue that we have created the first atlas that is generally useful because 1.
+our dataset/atlas is several times larger than any other published set, 2.
+our data is available via download in several forms at https://github.com/davemcg/scEiaD, and crucially 3.
+we provide a Google colab/Jupyter notebook which exactly lays out how to use scVI's to project (or query) outside data onto our scEiaD with minimum compute resources.
+Finally we also provide our xgboost model so outside groups can label their own datasets with our robust retina cell type labelling system.
 

### 2021-01-16 manuscript_00.Rmd
text update

--- 2021-01-15 manuscript_00.Rmd
@@
 
-scPOP is a lightweight, low dependency R package which brings together the Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, and re-implements the Adjusted Rand Index (ARI) and Normalized Mutual Information (NMI) algorithms.
+scPOP is a lightweight, low dependency R package which brings together the Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, and brings in the Adjusted Rand Index (ARI) and Normalized Mutual Information (NMI) algorithms from aricode.
 The LISI and ASW were used to measure batch mixing (where lower is better), cell type mixing (higher is better), and cluster mixing (higher is better).
@@
 
-We see in the across the sumZScale metric, which we define as sumZScale = zscale(LISI batch) + zscale(LISI cluster) + zscale(LISI cell type) + zscale(silhouette batch) + zscale(silhouette cluster) + zscale(silhouette cell type) + zscale(NMI) + zscale(ARI) that scVI overall performs best.
+To merge the different scores we define $sumZScale = \sum scale(m)$ where m is a metric (LISI by batch, LISI by cluster, LISI by cell type, silhouette by batch, silhouette by cluster, silhouette by celltype, NMI, and ARI).
+In this scaled metric that weights the different benchmarking algorithms evently, scVI performs best.
 As expected the libSize normalization which omits the log or square root scaling generally performs the worst (`r fig_cap(name = 'figBenchmark', display = 'cite')`c).
@@
 
-## Projection of new data onto scEiaD
+## Projection of new data onto scEiaD shows that outside datasets for both tissue and iPSC-based samples overlap the scEiaD/scVI model
 
@@
 
-Projection of new data onto scEiaD potentially allows for semi-direct comparison of other species, *in vitro* systems, and pharamacologic / mutational perturbation of mammalian retina cells.
+Projection of new data onto scEiaD allows for semi-direct comparison of other species, *in vitro* systems, and pharamacologic / mutational perturbation of mammalian retina cells.
+In the previous example we demonstrated how a outside mouse pan-retina dataset maps onto
 
-We demonstrate with a provided iPSC-based RPE cell line that we sequenced that non-tissue can be projected onto the scEiaD
+We demonstrate with a iPSC-based RPE cell line that we ran with the 10x v2 system that non-primary cells can be projected onto the scEiaD.
+The differentiated iPSC-based RPE from this highly optimized differentiation protocol (CITE BHARTI) overlap post-mortem adult human RPE cells from the Cowan et al.
+work.
+(MAKE FIGURE)
 

### 2021-01-17 manuscript_00.Rmd
updates

@@
 input[['meta_column']] <- 'CellType_predict'
+input[['pt_size_back']] <- 0
 input[['pt_size_meta']] <- 0
@@
 input[['label_toggle']] <- 2
-#input[['meta_filter_cat']] <- 'CellType_predict'
-#input[['meta_filter_on']] <- 'Bipolar Cells'
+input[['meta_filter_cat']] <- 'CellType_predict'
+input[['meta_filter_on']] <- NULL
 ctp <- make_meta_scatter_umap_plot(input, mf, meta_filter,
@@
 input[['meta_column']] <- 'CellType_predict'
+input[['pt_size_back']] <- 1
 input[['pt_size_meta']] <- 1
@@
 input[['meta_column_transform']] <- 'None'
-#input[['meta_filter_cat']] <- 'CellType_predict'
-#input[['meta_filter_on']] <- 'Bipolar Cells'
+input[['meta_filter_cat']] <- 'CellType_predict'
+input[['meta_filter_on']] <- NULL
 org <- make_meta_scatter_umap_plot(input, mf, meta_filter,
@@
 
-First, we project another retina dataset onto scEiaD from Fadl et al .
+First, we project another retina dataset onto scEiaD from Fadl et al.
 As this is a pan-retina tissue sample, we would expect most of the major cell types of the retina to be present.
@@
 
-In order to identify cell types for the `r meta_filter %>% filter(is.na(CellType), is.na(TabulaMurisCellType))` unlabelledcells we designed a custom xgboost based cell type classifier.
+In order to identify cell types for the `r format(meta_filter %>% filter(is.na(CellType), is.na(TabulaMurisCellType)) %>% nrow(), big.mark=",", scientific=FALSE)` unlabelledcells we designed a custom xgboost based cell type classifier.
 We took labeled data and split it into training and test sets, stratified by cell type.
@@
 input[['meta_column']] <- 'CellType_predict'
+input[['pt_size_back']] <- 5
 input[['pt_size_meta']] <- 5
@@
 ```
+
+```{r projection_fig, fig.width=12, fig.height=10,  fig.cap=projection_fig_cap, echo=FALSE, message = FALSE}
+projection_fig_cap <- supFig_cap(name='projection_fig', caption =  'bloop')
+1 + 1
+```

### 2021-01-17 figs_and_tables/make_meta_scatter_umap_plot.R

@@ -77,9 +77,13 @@ make_meta_scatter_umap_plot <- function(input, mf, meta_filter,
     cat('log2 time')
     meta_filter[,meta_column] <- log2(meta_filter[,meta_column] + 1)
   }
+  if (!is.null(input$meta_filter_on)){
   p_data <- meta_filter %>%
     #filter(!grepl('Doub|\\/Margin\\/Periocular', CellType)) %>%
     filter_at(vars(all_of(input$meta_filter_cat)), all_vars(. %in% input$meta_filter_on))
+  } else {
+    p_data <- meta_filter %>% filter(!is.na(!!as.symbol(input$meta_column)))
+  }
   
   # metadata NUMERIC plot --------------
   if (is.numeric(meta_filter[,meta_column] %>% pull(1)) ){

### 2021-01-17 manuscript_00.Rmd
sup fig

@@
 ```{r projection_fig, fig.width=12, fig.height=10,  fig.cap=projection_fig_cap, echo=FALSE, message = FALSE}
-projection_fig_cap <- supFig_cap(name='projection_fig', caption =  'bloop')
-1 + 1
+input[['pt_size_back']] <- 0
+input[['pt_size_meta']] <- 4
+#input[['gene_and_meta_scatter_tech']] <- 'Droplet'
+input[['meta_column_transform']] <- 'None'
+input[['meta_filter_cat']] <- 'CellType_predict'
+input[['meta_filter_on']] <- c('RPE')
+rpe <- make_meta_scatter_umap_plot(input, mf, meta_filter,
+                                   celltype_predict_labels,
+                                   celltype_labels,
+                                   tabulamuris_predict_labels,
+                                   cluster_labels,
+                                   cat_to_color_df
+)
+#
+
+colab_p <- ggdraw() + draw_image("figs_and_tables/colab_screenshot.png", scale = 0.7)
+
+plot_grid(rpe$plot , colab_p)
+
 ```

### 2021-01-17 manuscript_00.Rmd
text update

--- 2021-01-17 manuscript_00.Rmd
@@
 We propose that by re-processing publicly available raw single cell transcriptome data in a consistent bioinformatic framework and optimally using batch correction tools we can create a meta-atlas.
-As there are thousands of possible permutations of single cell tools, references, and parameter choices, we create our meta-atlas (which we refer to as the single cell Eye in a Disk or scEiaD) by benchmarking integration outcomes across multiple important single cell RNA-seq processing parameters (batch removal method, number of hyper-variable genes (HVGs), clustering resolution, etc).
+As there are thousands of possible permutations of single cell tools, references, and parameter choices, we create our meta-atlas (which we refer to as the single cell Eye in a Disk or scEiaD) by benchmarking integration outcomes across multiple important single cell RNA-seq processing parameters (batch removal method, number of hyper-variable genes (HVGs), clustering resolution, etc.).
 The benchmarking system we developed uses a wide variety of metrics that we make easily available via the R package scPOP (single-cell Pick Optimal Parameters).
@@
 Also the computational community can use this very large, well-curated dataset to test algorithms for compute efficiency and performance in a diverse environment.
-As we believe data re-use is a powerful and efficient approach to facilitate discovery, we provide our meta-atlas code-base and propose general guidelines to optimally create custom meta-atlases.
+As we believe data re-use is a powerful and efficient approach to facilitate discovery, we provide our meta-atlas code-base, the meta-atlas in several data formats, and propose general guidelines to optimally create custom meta-atlases.
 
@@
 ```{r projection_fig, fig.width=12, fig.height=10,  fig.cap=projection_fig_cap, echo=FALSE, message = FALSE}
-projection_fig_cap <- supFig_cap(name='projection_fig', caption =  'bloop')
-1 + 1
+projection_fig_cap <- supFig_cap(name='projection_fig', caption =  '')
+input[['pt_size_back']] <- 0
+input[['pt_size_meta']] <- 6
+input[['meta_column']] <- 'study_accession'
+#input[['gene_and_meta_scatter_tech']] <- 'Droplet'
+input[['meta_column_transform']] <- 'None'
+input[['meta_filter_cat']] <- 'CellType_predict'
+input[['meta_filter_on']] <- c('RPE')
+rpe <- make_meta_scatter_umap_plot(input, mf, meta_filter,
+                                   celltype_predict_labels,
+                                   celltype_labels,
+                                   tabulamuris_predict_labels,
+                                   cluster_labels,
+                                   cat_to_color_df
+)
+#
+
+colab_p <- ggdraw() + draw_image("figs_and_tables/colab_screenshot.png", scale = 0.7)
+
+plot_grid(rpe$plot , colab_p)
+
 ```

### 2021-01-17 figs_and_tables/fig1.R

@@ -34,12 +34,7 @@ dagified <- dagify('counts' ~ 'fastq',
                    'DiffTesting' ~ 'cluster',
                    'SnakePOP' ~ 'BatchInfo',
                    'scVIrefquery' ~ 'BatchInfo',
-                   'Trajectory' ~ 'cluster',
-                   'Velocity' ~ 'LatentDims',
-                   'projectedCellType' ~ 'LatentDims',
-                   'Trajectory' ~ 'projectedCellType',
-                   'Trajectory' ~ 'LatentDims',
-                   'projectedCellType' ~ 'CellType')
+                   'projectedCellType' ~ 'LatentDims')
 
 tidy_dagitty(dagified)
 set.seed(2534)

### 2021-01-18 manuscript_00.Rmd
text updates

@@
 The  recent introduction of lower cost and high throughput single cell sequencing technology has led to an explosion of research across many fields.
-As of November 2020, over 30 million cells have been sequenced across over 1,000 studies and the average size of each study is nearing 100,000 cells [@svenssonCuratedDatabaseReveals2020].
-The retina was used as the source tissue in one of the earliest works in the high throughput single cell transcriptomics field and as of late 2020, over a dozen published studies containing over a million cells have used single cell technology to profile cell type specific gene expression patterns, cell fate trajectory, tissue and cell differentiation, and disease perturbation across multiple mammalian organisms [@buenaventuraIdentificationGenesEnriched2019; @clarkComprehensiveAnalysisRetinal2018; @collinCRXExpressionPluripotent2019; @collinDeconstructingRetinalOrganoids2019; @cowanCellTypesHuman2020; @diacouSix3Six6Are2018; @eldredThyroidHormoneSignaling2018; @hengHypoxiaToleranceNorrindeficient2019; @huDissectingTranscriptomeLandscape2019; @logiudiceSinglecellTranscriptionalLogic2019; @lukowskiSinglecellTranscriptomeAtlas2019; @luSingleCellAnalysisHuman2020; @macoskoHighlyParallelGenomewide2015; @menonSinglecellTranscriptomicAtlas2019; @okorenMicroglialFunctionDistinct2019; @pengMolecularClassificationComparative2019; @phillipsNovelApproachSingle2018; @rheaumeSingleCellTranscriptome2018; @ruedaHippoPathwayBlocks2019; @shekharComprehensiveClassificationRetinal2016; @sridharSingleCellTranscriptomicComparison2020; @tranSingleCellProfilesRetinal2019; @voigtMolecularCharacterizationFoveal2019; @voigtSinglecellTranscriptomicsHuman2019; @yanCellAtlasHuman2020].
+As of January 2021, over 40 million cells have been sequenced across over 1,200 studies and the average size of each study past 2020 is over 100,000 cells [@svenssonCuratedDatabaseReveals2020].
+The retina was used as the source tissue in one of the earliest works in the high throughput single cell transcriptomics field [@macoskoHighlyParallelGenomewide2015] and as of late 2020, over a dozen published studies containing over a million cells have used single cell technology to profile cell type specific gene expression patterns, cell fate trajectory, tissue and cell differentiation, and disease perturbation across multiple mammalian organisms [@buenaventuraIdentificationGenesEnriched2019; @clarkComprehensiveAnalysisRetinal2018; @collinCRXExpressionPluripotent2019; @collinDeconstructingRetinalOrganoids2019; @cowanCellTypesHuman2020; @diacouSix3Six6Are2018; @eldredThyroidHormoneSignaling2018; @hengHypoxiaToleranceNorrindeficient2019; @huDissectingTranscriptomeLandscape2019; @logiudiceSinglecellTranscriptionalLogic2019; @lukowskiSinglecellTranscriptomeAtlas2019; @luSingleCellAnalysisHuman2020;  @menonSinglecellTranscriptomicAtlas2019; @okorenMicroglialFunctionDistinct2019; @pengMolecularClassificationComparative2019; @phillipsNovelApproachSingle2018; @rheaumeSingleCellTranscriptome2018; @ruedaHippoPathwayBlocks2019; @shekharComprehensiveClassificationRetinal2016; @sridharSingleCellTranscriptomicComparison2020; @tranSingleCellProfilesRetinal2019; @voigtMolecularCharacterizationFoveal2019; @voigtSinglecellTranscriptomicsHuman2019; @yanCellAtlasHuman2020].
 
@@
 
-Even after re-quantification under identical conditions there remain study specific batch effects due to the variety in single cell technologies used and variation in tissue handling and processing across each scientific group.
+Still, even after re-quantification under identical conditions there remain study specific batch effects due to the variety in single cell technologies used and variation in tissue handling and processing across each scientific group.
 The single cell community has recognized that removal of these technical (also referred to as batch) effects is a critical issue and have independently developed many tools, though it remains unclear which tools and parameters are optimal for a particular dataset [@butlerIntegratingSinglecellTranscriptomic2018; @haghverdiBatchEffectsSinglecell2018; @hieEfficientIntegrationHeterogeneous2019; @johnsonAdjustingBatchEffects2007; @korsunskyFastSensitiveAccurate2019; @liuJointlyDefiningCell2020; @lopezDeepGenerativeModeling2018; @polanskiBBKNNFastBatch2020; @QueryReferenceSinglecell; @simonINSCTIntegratingMillions2020; @stuartComprehensiveIntegrationSingleCell2019; @vandijkRecoveringGeneInteractions2018].
@@
 To provide a non-ocular reference we also downloaded the raw sequence data from the Tabula Muris project for re-processing.
-In cases where the fastq file from the SRA was not processed properly, we acquired the original bam files (SRA or personal correspondence) and re-extracted the fastq.
+In cases where the fastq file from the SRA was not processed properly, we acquired the bam files (SRA or personal correspondence) and re-extracted the fastq.
 After downloading all the data we had 5.7 TB across `r scEiaD_2020_v01 %>% tbl('metadata_filter') %>% pull(sample_accession) %>% unique() %>% length()` fastq file sets.
@@
 A core objective of many scRNA based studies is labeling the cell types.
-As this information is crucial to assess dataset integration and provide an accurate reference for user querying, we extracted individual cell labels with a combination of the GEO, supplemental information from the publication, web resources (e.g.
+As this information is crucial to assess dataset integration and provide an accurate reference for user querying, we extracted individual cell labels with a combination of inspecting the GEO web site, supplemental information from the publication, custom web resources (e.g.
 a web app was created for the paper), and personal correspondence.
@@
 A wide variety of methods have been written for this function.
-As we were uncertain which would perform the best and it is difficult to manually assess performance we ran `r methods %>% length()` tools with a commonly used set of key parameters: 2000 hyper variable genes, 8 and 30 latent dimensions returned, and the louvain clustering algorithm with 7 k-nearest neighbors.
+As we were uncertain which would perform the best and it is difficult to manually assess performance we ran `r methods %>% length()` tools with a commonly used set of key parameters: 2000 hyper variable genes (HVG), 8 and 30 latent dimensions returned, and the louvain clustering algorithm with 7 k-nearest neighbors.
 The Snakemake system was used to automate the running of the wide variety of tools.
@@
 Scanorama, fastMNN, and bbknn more prioritize cleanly separating the clusters.
-With our scEiaD meta-atlas insct, scArches, and magic do not perform particularly well in batch mixing or cluster purity.
+With our scEiaD meta-atlas insct, trVAE, and magic do not perform particularly well in batch mixing or cluster purity.
 
@@
 
-To find the best set of parameters for the creation of scEiaD we did a grid search across key parameters: hyper variable genes (HVG), latent dimensions, and k nearest neighbors.
-Furthermore we used a recent advance in scVI capability adapted from scArches that allows one to build a reference model and project the new cells onto it.
+To find the best set of parameters for the creation of scEiaD we did a grid search across key parameters: HVG, latent dimensions, and k nearest neighbors.
+Furthermore we used a recent advance in scVI capability adapted from scArches that allows one to build a reference model and project the new cells onto it [@QueryReferenceSinglecell].
 We built a human-only scVI model and then projected the mouse and macaque data onto it.
@@
 Using scPOP we first saw that the scVIprojection approach generally performed better than running scVI with all of the data.
-We found the optimal parameters to be 5000 HVG, 10 latent dimensions, and a resolution of 0.6 with the leiden algorithm implemented by PARC (`r fig_cap(name = 'figBenchmark', display = 'cite')`d).
-We also varied the UMAP projection values of nearest neighbors and minimum distance to qualitatively pick a 2D projection, selecting a minimum distance of 0.1 and 500 nearest neighbors.
+We found the optimal overall parameters to be 5000 HVG, 8 latent dimensions, and a resolution of 0.6 with the leiden algorithm implemented by PARC (`r fig_cap(name = 'figBenchmark', display = 'cite')`d).
+We also varied the UMAP projection values of nearest neighbors and minimum distance to qualitatively pick a 2D projection, selecting a minimum distance of 0.2 and 500 nearest neighbors.
 
@@
 Traditionally this is done by clustering the cells, then using cell type specific markers to label the clusters.
-However, as we had hundreds of thousands of expert labelled cells across `r umapRef %>% filter(!is.na(CellType)) %>% left_join(gse) %>% pull(PMID) %>% unique() %>% length()` studies (`r fig_cap(name = 'fig1', display = 'cite')`a) we built a xgboost-based machine learning model that used half of the labeled cells as a training set (see methods for more details) to train a cell type predictor for scEiaD input.
+However, as we had hundreds of thousands of expert labeled cells across `r umapRef %>% filter(!is.na(CellType)) %>% left_join(gse) %>% pull(PMID) %>% unique() %>% length()` studies (`r fig_cap(name = 'fig1', display = 'cite')`a) we built a xgboost-based machine learning model that used half of the labeled cells as a training set (see methods for more details) to train a cell type predictor for scEiaD input.
 The trained model was used to predict the cell type assignments for all cells in scEiaD.
-In this manner we both label most cells (cells which cannot be assigned a cell type assignment with a confidence above 0.5 were left unlabeled) and correct a small number of probable mislabels in the truth set (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a).
+In this manner we both label most cells (cells which cannot assign a cell type with a probability above 0.5 were left unlabeled) and correct a small number of probable mislabels in the truth set (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a).
 
@@
 As a simple metric to identify known and unknown genes relating to the cell type specific expression we search PubMed for the number of publications with two searches per gene.
+We expect most of the genes identified to be known in the literature.
 The first search is the more precise "gene AND cell type" (e.g.
@@
 "PDE6G AND Retina").
-Of the 55 genes in (`r supFig_cap(name = 'figXGboost_umap', display = 'cite')`a), 40 had one or more citations in the gene/celltype search (table X) and 51 had one or more citation in the inclusive search.
+Of the 55 genes in (`r supFig_cap(name = 'figXGboost_umap', display = 'cite')`a), 40 had one or more citations in the gene - cell type search (table X) and 51 had one or more citation in the inclusive search.
 The 55 genes had a mean of `r top_markers %>% group_by(cluster)  %>% slice_max(mean_auc, n =5) %>% pull(c2) %>% mean() %>% round()` studies (with the inclusive gene by "retina" search).
@@
 In a more exhaustive literature search we find that SCRG1 is a secreted protein which marks brain neurons, regulates human mesenchymal stem cells differentiation, and is overexpressed in brain astrocyte precursors [@aomatsuNovelSCRG1BST12014; @chaboubTemporalProfilingAstrocyte2016].
-BTBD8 contains BTB/POZ domains known to generally be associated with with transcription factor mediated protein-protein interactions [@xuMolecularCloningCharacterization2004; @zollmanBTBDomainFound1994].
+BTBD8 contains BTB/POZ domains known to generally be associated with with transcription factor mediated protein-protein interactions and has been seen to be overexpresed in enteric neurons [@xuMolecularCloningCharacterization2004; @zollmanBTBDomainFound1994; @drokhlyansky_human_2020].
 GATM is mitochondrial enzyme involved in creatine biosynthesis and was previously shown to up-regulated in a brain microglia population [@hammondSingleCellRNASequencing2019].
 FRZB was previously referred to as SFRP3 and this Wnt pathway member has been previously shown to be active in Muller Glia [@zhuCharacterizationCanonicalWnt2018].
-Overall, we find very high concordance with known retina cell type markers.
+Overall, as expected, we find very high concordance with known retina cell type markers.
 
@@
 The photoreceptors (cones and rods) of the retina which are responsible for color and low-light vision, respectively, are near each other in the UMAP space.
-The major remaining retina cell types, by proportion in the mammalian eye are the Muller Glia, which are a glial cell type which help support the neurons of the retina.
+The major remaining retina cell types, by proportion in the mammalian eye are the muller glia, which are a glial cell type which help support the neurons of the retina.
 Next we have the neural cell types which transmit and help interpret the signals from the photoreceptors before they leave the retina via the optic stalk: the amacrine cells, retinal ganglia, horizontal, and bipolar cells.

### 2021-01-20 manuscript_00.Rmd
updates

@@
 supFig_cap <- captioner("Supplemental Figure")
-tab_cap <- captioner("Table")
+#tab_cap <- captioner("Table")
 supTab_cap <- captioner("Supplemental Table")
@@
 # Introduction
-## Explosion of single cell transcriptomic atlases
+
+## The retina, single cell gene analysis, and a recent explosion
+
+The retina contains a multitude of cell types that, in total, are responsible for turning light information into signal for the brain to interpret as vision.
+Very briefly, the photoreceptors (rods and cones) are responsible for capturing the photons.
+The retinal pigmented epithelium (RPE) behind the photoreceptors physically support the rods and cones by processing byproducts of the visual cycle.
+Müller glia serve as support cells for the neurons.
+The retinal bipolar cells transmit the electrical signal from the photoreceptors to the retinal ganglion cells.
+Horizontal and amacrine regulate and help interpret signals from the photoreceptors.
+The signal is relayed via the retinal ganglion projections through the optic stalk to the brain (for review see [@maslandNeuronalOrganizationRetina2012]).
+Since 2000 many groups have investigated gene expression in small numbers of individual cells of the retina [@anniesIsoformPatternAChR2002a; @hagstromConePigmentGene2000; @trimarchiMolecularHeterogeneityDeveloping2007a; @wahlinMethodAnalysisGene2004].
+
 The  recent introduction of lower cost and high throughput single cell sequencing technology has led to an explosion of research across many fields.
 As of January 2021, over 40 million cells have been sequenced across over 1,200 studies and the average size of each study past 2020 is over 100,000 cells [@svenssonCuratedDatabaseReveals2020].
-The retina was used as the source tissue in one of the earliest works in the high throughput single cell transcriptomics field [@macoskoHighlyParallelGenomewide2015] and as of late 2020, over a dozen published studies containing over a million cells have used single cell technology to profile cell type specific gene expression patterns, cell fate trajectory, tissue and cell differentiation, and disease perturbation across multiple mammalian organisms [@buenaventuraIdentificationGenesEnriched2019; @clarkComprehensiveAnalysisRetinal2018; @collinCRXExpressionPluripotent2019; @collinDeconstructingRetinalOrganoids2019; @cowanCellTypesHuman2020; @diacouSix3Six6Are2018; @eldredThyroidHormoneSignaling2018; @hengHypoxiaToleranceNorrindeficient2019; @huDissectingTranscriptomeLandscape2019; @logiudiceSinglecellTranscriptionalLogic2019; @lukowskiSinglecellTranscriptomeAtlas2019; @luSingleCellAnalysisHuman2020;  @menonSinglecellTranscriptomicAtlas2019; @okorenMicroglialFunctionDistinct2019; @pengMolecularClassificationComparative2019; @phillipsNovelApproachSingle2018; @rheaumeSingleCellTranscriptome2018; @ruedaHippoPathwayBlocks2019; @shekharComprehensiveClassificationRetinal2016; @sridharSingleCellTranscriptomicComparison2020; @tranSingleCellProfilesRetinal2019; @voigtMolecularCharacterizationFoveal2019; @voigtSinglecellTranscriptomicsHuman2019; @yanCellAtlasHuman2020].
+The retina was used as the source tissue in one of the earliest works in the high throughput single cell transcriptomics field [@macoskoHighlyParallelGenomewide2015] and as of late 2020, over a dozen published studies containing over a million cells have used single cell technology to profile cell type specific gene expression patterns, cell fate trajectory, tissue and cell differentiation, and disease perturbation across multiple mammalian species [@buenaventuraIdentificationGenesEnriched2019; @clarkComprehensiveAnalysisRetinal2018; @collinCRXExpressionPluripotent2019; @collinDeconstructingRetinalOrganoids2019; @cowanCellTypesHuman2020; @diacouSix3Six6Are2018; @eldredThyroidHormoneSignaling2018; @hengHypoxiaToleranceNorrindeficient2019; @huDissectingTranscriptomeLandscape2019; @logiudiceSinglecellTranscriptionalLogic2019; @lukowskiSinglecellTranscriptomeAtlas2019; @luSingleCellAnalysisHuman2020;  @menonSinglecellTranscriptomicAtlas2019; @okorenMicroglialFunctionDistinct2019; @pengMolecularClassificationComparative2019; @phillipsNovelApproachSingle2018; @rheaumeSingleCellTranscriptome2018; @ruedaHippoPathwayBlocks2019; @shekharComprehensiveClassificationRetinal2016; @sridharSingleCellTranscriptomicComparison2020; @tranSingleCellProfilesRetinal2019; @voigtMolecularCharacterizationFoveal2019; @voigtSinglecellTranscriptomicsHuman2019; @yanCellAtlasHuman2020].
 
@@
 
-```{r fig1, fig.width=13, fig.height=6, fig.cap=fig1_cap, echo=FALSE, message = FALSE}
+```{r fig1, fig.width=12, fig.height=6, fig.cap=fig1_cap, echo=FALSE, message = FALSE, warning = FALSE}
 fig1_cap <- fig_cap(name='fig1', caption =  'a.
-Counts of (study) accession, published papers, and batches for each scRNA technlogy, split by organism.
-b.
+Cartoon schematic of retina with major cell types delinearted b.
 Simplified directed workflow of major steps in scEiaD creation from raw counts to gene counts, benchmarking optimal integration methods (SnakePOP) to produce batch corrected latent dimensions (LatentDims), then downstream analysis outputs like clustering, differential gene testing (DiffTesting), and 2D UMAP visualization.
 c.
+Counts of (study) accession, published papers, and batches for each scRNA technlogy, split by organism.
+d.
 Cell type counts extracted from published studies for the more common retina cell types, split by species.
@@
 source('figs_and_tables/fig1.R')
-plot_grid(plot_grid(a, c, align = 'v',  axis = 'lr', ncol = 1, labels = c('','c')), b,  nrow = 1, rel_widths = c(1.1,1), labels = c('a','b'))
+retina_cartoon <- ggdraw() + draw_image("figs_and_tables/retina_cells_flat2.png", scale = 1)
+plot_grid(
+  plot_grid(retina_cartoon, b + coord_cartesian(clip = "off"), nrow = 1, labels = c('','b'), rel_widths = c(0.5,1)), 
+  plot_grid(a, c, align = 'v',  axis = 'lr', nrow = 1, labels = c('','d')),
+  nrow = 2, 
+  rel_heights= c(1,1), 
+  labels = c('a','c'))
 ```
@@
 
-## Running `r methods %>% length()` tools in a Snakefile-based system
+## Running `r methods %>% length()` tools in a Snakemake-based system
 
@@
 
-```{r figBenchmark, fig.width=14, fig.height=8, fig.cap=figBenchmark_cap, echo=FALSE, message = FALSE}
+```{r figBenchmark, fig.width=14, fig.height=8, fig.cap=figBenchmark_cap, echo=FALSE, message = FALSE, warning = FALSE}
 figBenchmark_cap <- fig_cap(name='figBenchmark', caption =  'a.
@@
 
-```{r figXGboost_umap, fig.width=18, fig.height=13, fig.cap=figXGboost_umap_cap, echo=FALSE, message = FALSE}
-figXGboost_umap_cap <- fig_cap(name='figXGboost_umap', caption =  'filler text hello world')
+```{r figXGboost_umap, fig.width=18, fig.height=13, fig.cap=figXGboost_umap_cap, echo=FALSE, message = FALSE, warning = FALSE}
+figXGboost_umap_cap <- fig_cap(name='figXGboost_umap', caption =  'a.
+Top genes that are differentially expressed across the major cell types of the retina.
+Genes are colored by which cell type they are differentially expressed in.
+The dot size is proportional to the percentage of those cells that have detectable levels of the gene.
+The color of the dot is the log2 scaled CPM expression.
+b.
+2D UMAP projection of scEiaD, colored by cell type.
+The Tabula Muris data is colored all in gray.
+c.
+Facet plot that demonstrates how each major cell type of the retina is contained within a distinct space.
+d.
+Confusion matrix of cell type prediction performance of our xgboost labeller between predicted (y axis) and known (x axis) with data withheld from the machine learner.
+Most of the cell types are indeed labelled as their true type.
+e.
+Faceting of 2D UMAP by species and colored by cell type demonstrate how the major cell types of the retina share space with like cell types, despite being from mouse, human, and macaque.')
 
@@
 input[['meta_column']] <- 'CellType_predict'
-input[['pt_size_back']] <- 0
-input[['pt_size_meta']] <- 0
+input[['pt_size_back']] <- 1
+input[['pt_size_meta']] <- 1
 #input[['gene_and_meta_scatter_tech']] <- 'Droplet'
@@
 input[['meta_column']] <- 'CellType_predict'
-input[['pt_size_back']] <- 1
-input[['pt_size_meta']] <- 1
+input[['pt_size_back']] <- 2
+input[['pt_size_meta']] <- 2
 input[['gene_and_meta_scatter_tech']] <- 'Droplet'
@@
 After ML projection of cell type labels from the original `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` labels onto a total of `r meta_filter %>% filter(!is.na(CellType_predict) | !is.na(TabulaMurisCellType_predict)) %>% nrow()` cells we have substantially improved the number of studies per cell type.
-For example, we went from `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(2)` human studies with labelled Muller Glia to `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(5)` after labelling (`r supTab_cap(name = 'supTab_SAcounts', display = 'cite')`).
+For example, we went from `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(2)` human studies with labelled Muller Glia to `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(5)` after labeling (`r supTab_cap(name = 'supTab_SAcounts', display = 'cite')`).
 Overall we go from an average of `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(2) %>% mean()` studies per human cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(5) %>% mean()` and `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(4) %>% mean()` studies per mouse cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(7) %>% mean()` after transferring the cell type labels.
@@
 As another check on the quality of the cell type assignments, we ran pairwise differential expression tests between the predicted cell types (see methods for further details).
-We show the five most consistently differentially expressed genes for each of the major retina cell types (`r supFig_cap(name = 'figXGboost_umap', display = 'cite')`a).
+We show the five most consistently differentially expressed genes for each of the major retina cell types (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a).
 As a simple metric to identify known and unknown genes relating to the cell type specific expression we search PubMed for the number of publications with two searches per gene.
@@
 "PDE6G AND Retina").
-Of the 55 genes in (`r supFig_cap(name = 'figXGboost_umap', display = 'cite')`a), 40 had one or more citations in the gene - cell type search (table X) and 51 had one or more citation in the inclusive search.
+Of the 55 genes in (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a), 40 had one or more citations in the gene - cell type search (table X) and 51 had one or more citation in the inclusive search.
 The 55 genes had a mean of `r top_markers %>% group_by(cluster)  %>% slice_max(mean_auc, n =5) %>% pull(c2) %>% mean() %>% round()` studies (with the inclusive gene by "retina" search).
@@
 All of these cells are in well defined spaces in the UMAP.
-Finally we see even across species that the major cell types still group near each other (`r supFig_cap(name = 'figXGboost_umap', display = 'cite')`d).
+Finally we see even across species that the major cell types still group near each other (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`d).
 The macaque retina cells are not present in the precursor/neurogenic center of the UMAP as expected because only fully developed tissues were sampled in these data.
@@
 we provide a Google colab/Jupyter notebook which exactly lays out how to use scVI's to project (or query) outside data onto our scEiaD with minimum compute resources.
-Finally we also provide our xgboost model so outside groups can label their own datasets with our robust retina cell type labelling system.
+Finally we also provide our xgboost model so outside groups can label their own datasets with our robust retina cell type labeling system.
 
-# Supplemental
-```{r supFig1, fig.width=4, fig.height=6, out.width=300, fig.cap=supFig1_cap, echo=FALSE, message = FALSE}
-supFig1_cap <- supFig_cap(name='supFig1_cap', caption =  'The left bar delineates the number of cells for each organism - technology combination.
-The right bar specifies the number of each cells in each post QC category.
-In silico doublets were identified with scrublet and DoubletDetector.')
-
-source('figs_and_tables/sup_fig1__qc_failure.R')
-supFig1_plot
+# Supplemental Information
+## Tables
+```{r supTab1_CTcounts, fig.width=8, fig.height=14, out.width=300, echo = F, results='asis', warning = F, fig.cap = supTab1_CTcounts_cap, fig.pos = 'H'}
+supTab1_CTcounts_cap <- supTab_cap(name='supTab1_CTcounts_cap', caption =  'Counts for cell type labels.
+Published are the author created labels from the published datasets.
+Transferred are the cell labels that were transferred by our xgboost-based machine learning model onto the entire scEiaD dataset.')
+cat("<div id=\"tbl:supTab1_CTcounts\">")
+source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
+fontsize(ctTable, size = 8, part = 'all')
+#Output Caption (* * for italics)
+cat(paste0("\n*",supTab1_CTcounts_cap,"*\n"))
+# Output </div>
+cat("</div>")
 ```
 
-Words
+```{r supTab_CTmislabels, fig.width=2, fig.cap=supTab_CTmislabels_cap, echo=FALSE, results='asis', message = FALSE, message = FALSE, warning = FALSE}
 
-Spacing
+supTab_CTmislabels_cap <- supTab_cap(name='supTab_CTmislabels', caption =  'Counts of cell type labels with our xgboost machine learning system (PredCellType) and the published cell type labels (TrueCellType).')
 
-```{r supFig_benchmark, fig.width=12, fig.height=8, out.width=300, fig.cap=supFig_benchmark_cap, echo=FALSE, message = FALSE}
-supFig_benchmark_cap <- supFig_cap(name='supFig_benchmark', caption =  'Performance of the various batch correction tools across various benchmarking metrics.
-For the LISI and Silhouette plots in A, B higher means better batch mixing and further to the right means better cluster purity.
-For the ARI and NMI metrics (which reflects how well cluster matches with cell type) in C, D, higher means a better score.')
-
-source('figs_and_tables/fig2__integrationPerf.R')
-cowplot::plot_grid(
-  cowplot::plot_grid(lisi + theme(legend.position="none"), silhouette, NULL, rel_widths = c(1,1,0.2), nrow = 1, labels = c('a' ,'b')), 
-  cowplot::plot_grid(ari, nmi, legend, rel_widths = c(1,1, 1, 0.2), ncol = 4, labels = c('c', 'd')), 
-  nrow = 2)
-```
-
-Words
-
-Spacing
-
-```{r supTab1_CTcounts, fig.width=8, fig.height=14, out.width=300, fig.cap=supTab1_CTcounts_cap, echo=FALSE, message = FALSE}
-supTab1_CTcounts_cap <- supTab_cap(name='supTab1_CTcounts_cap', caption =  'Counts for cell type labels.
-Published are the author created labels from the published datasets.
-Transferred area the cell labels that were transferred by a machine learning model onto the entire scEiaD dataset.')
-
-source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
-fontsize(ctTable, size = 8, part = 'all')
-```
-
-Words
-
-Spacing
-
-```{r supTab_SAcounts, fig.width=8, fig.height=14, out.width=300, fig.cap=supTab_SAcounts_cap, echo=FALSE, message = FALSE}
-supTab_SAcounts_cap <- supTab_cap(name='supTab_SAcounts_cap', caption =  'Counts for number of studies with cell types labels before and after cell type label transfer')
-
-source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
-fontsize(saTable, size = 8, part = 'all')
-```
-
-Words
-
-Spacing
-
-```{r supFig_PR, fig.width=12, fig.height=10,  fig.cap=supFig_PR_cap, echo=FALSE, message = FALSE}
-supFig_PR_cap <- supFig_cap(name='supFig_PR', caption =  'Precision recall curves for our xgboost cell type predictor model across each cell type predicted')
-
-source('figs_and_tables/sup_fig__xgboost_PR.R')
-pr_split_curves
-```
-
-Words
-
-Spacing
-
-```{r supTab_AUCxgboost, fig.cap=supTab_AUCxgboost_cap, echo=FALSE, message = FALSE}
-supTab_AUCxgboost_cap <- supTab_cap(name='supTab_AUCxgboost', caption =  'Area under the precision recall curve (AUC) for each cell type, split by study')
-
-#source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
-fontsize(xgboost_pr_table_full %>% arrange(`Cell Type`, Study) %>% flextable() %>% width(width = 3), size = 8, part = 'all')
-```
-
-Words
-
-Spacing
-
-```{r supTab_CTmislabels, fig.width=2, fig.cap=supTab_CTmislabels_cap, echo=FALSE, message = FALSE}
-
-supTab_CTmislabels_cap <- supTab_cap(name='supTab_CTmislabels', caption =  'TEXT2')
-
+cat("<div id=\"tbl:supTab_CTmislabels\">")
 #source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
@@
          part = 'all')
+#Output Caption (* * for italics)
+cat(paste0("\n*",supTab_CTmislabels_cap,"*\n"))
+# Output </div>
+cat("</div>")
 ```
 
-Words
+```{r supTab_SAcounts, fig.width=8, fig.height=14, out.width=300, fig.cap=supTab_SAcounts_cap, echo=FALSE, results='asis', message = FALSE, warning = FALSE}
+cat("<div id=\"tbl:supTab_SAcounts\">")
+supTab_SAcounts_cap <- supTab_cap(name='supTab_SAcounts_cap', caption =  'Counts for number of studies with cell types labels before and after cell type label transfer')
 
-Spacing
+source('figs_and_tables/sup_fig__xgboost_PR.R')
+fontsize(saTable, size = 8, part = 'all')
+#Output Caption (* * for italics)
+cat(paste0("\n*",supTab_SAcounts_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
 
-```{r supFig_platformInt, fig.width=12, fig.height=10,  fig.cap=supFig_platformInt_cap, echo=FALSE, message = FALSE}
-supFig_platformInt_cap <- supFig_cap(name='supFig_platformInt', caption =  '')
+```{r supTab_AUCxgboost, fig.cap=supTab_AUCxgboost_cap, echo=FALSE, results='asis', message = FALSE, warning = FALSE}
+supTab_AUCxgboost_cap <- supTab_cap(name='supTab_AUCxgboost', caption =  'Area under the precision recall curve (AUC) for each cell type, split by study')
+
+cat("<div id=\"tbl:supTab_AUCxgboost\">")
+#source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
+fontsize(xgboost_pr_table_full %>% arrange(`Cell Type`, Study) %>% flextable() %>% width(width = 3), size = 8, part = 'all')
+#Output Caption (* * for italics)
+cat(paste0("\n*",supTab_AUCxgboost_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
+
+## Figures
+
+```{r supFig_benchmark, fig.width=12, fig.height=8, out.width=300, fig.cap=supFig_benchmark_cap, echo=FALSE, message = FALSE, warning = FALSE}
+supFig_benchmark_cap <- supFig_cap(name='supFig_benchmark', caption =  'Performance of the various batch correction tools across various benchmarking metrics.
+For the LISI and Silhouette plots in A, B higher (y-axis) means better batch mixing and further to the right (x-axis) means better cluster purity.
+For the ARI and NMI metrics (which reflects how well cluster matches with cell type) in C, D, higher means a better score.')
+
+source('figs_and_tables/fig2__integrationPerf.R')
+cowplot::plot_grid(
+  cowplot::plot_grid(lisi + theme(legend.position="none"), silhouette, NULL, rel_widths = c(1,1,0.2), nrow = 1, labels = c('a' ,'b')), 
+  cowplot::plot_grid(ari, nmi, legend, rel_widths = c(1,1, 1, 0.2), ncol = 4, labels = c('c', 'd')), 
+  nrow = 2)
+```
+
+```{r supFig_platformInt, fig.width=12, fig.height=10,  fig.cap=supFig_platformInt_cap, echo=FALSE, message = FALSE, warning = FALSE}
+supFig_platformInt_cap <- supFig_cap(name='supFig_platformInt', caption =  'Distribution of cell types across the 2D UMAP confirms that the cell types are being properly placed despite the wide variety of single sequencing platforms present.')
 input <- list()
@@
 
-```{r projection_fig, fig.width=12, fig.height=10,  fig.cap=projection_fig_cap, echo=FALSE, message = FALSE}
-projection_fig_cap <- supFig_cap(name='projection_fig', caption =  '')
+```{r projection_fig, fig.width=12, fig.height=10,  fig.cap=projection_fig_cap, echo=FALSE, message = FALSE, warning = FALSE}
+projection_fig_cap <- supFig_cap(name='projection_fig', caption =  'a.
+RPE distribution colored by study demonstrates how the RPE are grouped in a similar location.
+iPSC-based RPE we provided are located within the tissue based RPE cells.
+b.
+Screen shot of Google colab notebook that demonstrates how to integrate external data into the scEiaD resource.
+We see in the screenshort how the dataset contains many of the retinal cell types.')
 input[['pt_size_back']] <- 0
@@
 ```
+
+```{r supFig1, fig.width=4, fig.height=6, out.width=300, fig.cap=supFig1_cap, echo=FALSE, message = FALSE, warning = FALSE}
+supFig1_cap <- supFig_cap(name='supFig1_cap', caption =  'The left bar delineates the number of cells for each organism - technology combination.
+The right bar specifies the number of each cells in each post QC category.
+In silico doublets were identified with scrublet and DoubletDetector.')
+
+source('figs_and_tables/sup_fig1__qc_failure.R')
+supFig1_plot
+```
+
+```{r supFig_PR, fig.width=12, fig.height=10,  fig.cap=supFig_PR_cap, echo=FALSE, message = FALSE, warning = FALSE}
+supFig_PR_cap <- supFig_cap(name='supFig_PR', caption =  'Precision recall curves for our xgboost cell type predictor model across each cell type predicted')
+
+pr_split_curves
+```
+
+# Works Cited

### 2021-01-20 figs_and_tables/fig1.R

@@ -37,7 +37,7 @@ dagified <- dagify('counts' ~ 'fastq',
                    'projectedCellType' ~ 'LatentDims')
 
 tidy_dagitty(dagified)
-set.seed(2534)
+set.seed(51345)
 b <- tidy_dagitty(dagified) %>% as_tbl_graph() %>% 
   mutate(type = case_when(name %in% c('fastq', 'CellType', 'BatchInfo') ~ 'Input',
                           name %in% c('DiffTesting', 'Velocity', 'Trajectory','umap', 'projectedCellTypes','cluster') ~ 'Outputs'),
@@ -46,9 +46,9 @@ b <- tidy_dagitty(dagified) %>% as_tbl_graph() %>%
                           name == 'DiffTesting' ~ 'Diff\nTesting',
                           name == 'scVIrefquery' ~ 'scVI Ref +\nQuery',
                           name == 'LatentDims' ~ 'Latent Dims',
-                          name == 'BatchInfo' ~ 'Batch Info',
+                          name == 'BatchInfo' ~ 'Batch\nInfo',
                           TRUE ~ name)) %>% 
-  ggraph(layout = 'kk') + 
+  ggraph(layout = 'gem') + 
   geom_edge_link(arrow = arrow(length = unit(3, 'mm')), 
                  end_cap = circle(9, 'mm'), start_cap = circle(2, 'mm')) + 
   #geom_node_point(size = 10) +

### 2021-01-20 src/make_dotplot_markers.R
function updates

@@ -2,7 +2,7 @@ library(pool)
 library(RSQLite)
 library(tidyverse)
 library(glue)
-scEiaD_2020_v01 <- dbPool(drv = SQLite(), dbname = "~/data/massive_integrated_eye_scRNA/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-500-0.6.sqlite", idleTimeout = 3600000)
+scEiaD_2020_v01 <- dbPool(drv = SQLite(), dbname = "~/data/scEiaD/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.2-500-0.6.sqlite", idleTimeout = 3600000)
 
 meta_filter <- fst::read_fst('~/git/plaeApp/inst/app/www/meta_filter.fst') %>% as_tibble()
 meta_filter <- meta_filter %>% mutate(CellType_predict = case_when(CellType_predict == 'Photoreceptor Precursors' ~ 'PR Precursors',
@@ -12,71 +12,15 @@ meta_filter <- meta_filter %>% mutate(CellType_predict = case_when(CellType_pred
                                                                    TRUE ~ CellType_predict)) %>%
   mutate(CellType_predict = case_when(!is.na(TabulaMurisCellType_predict) ~ 'Tabula Muris',
                                       TRUE ~ CellType_predict))
-load('~/data/massive_integrated_eye_scRNA/n_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__knn-0.6__neighbors-500__dist-0.1__CellType_predict.sceWilcox.Rdata')
-#genes <- c('ENSG00000144834','ENSG00000139970','ENSG00000109846','ENSG00000026025','ENSG00000067715','ENSG00000123307','ENSG00000139053','ENSG00000143774','ENSG00000105372','ENSG00000130066','ENSG00000104435','ENSG00000139970','ENSG00000050165','ENSG00000135821','ENSG00000104435','ENSG00000131711','ENSG00000016082','ENSG00000198668','ENSG00000149489','ENSG00000129535','ENSG00000105372','ENSG00000140988')
 
[…]

### 2021-01-20 src/pubmed_query.R

@@ -67,9 +67,9 @@ for (i in x2$pm_query2){
 #   pmid2[[i]] <- pubmed_counter(i)
 # }
 
-save(pmid, pmid2, pmid3, file = '~/data/massive_integrated_eye_scRNA/top_marker_pmid.Rdata')
+save(pmid, pmid2, pmid3, file = '~/data/scEiaD/top_marker_pmid.Rdata')
 top_markers$c1 <- pmid %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% pull(value)
 top_markers$c2 <- pmid2 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% pull(value)
-save(top_markers, marker_info, exp_stats, file = '~/data/massive_integrated_eye_scRNA/top_markers.Rdata')
+save(top_markers, marker_info, exp_stats, file = '~/data/scEiaD/top_markers.Rdata')
 #pmid %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe()
 #pmid %>% map(function(x) paste %>% unlist() %>% enframe()

### 2021-02-07 manuscript_00.Rmd
updates

@@
 library(Matrix)
+library(MatrixGenerics)
 library(matrixStats)
@@
 
-scEiaD_2020_v01 <- dbPool(drv = SQLite(), dbname = "~/data/scEiaD/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.2-500-0.6.sqlite", idleTimeout = 3600000)
+scEiaD_2020_v01 <- dbPool(drv = SQLite(), dbname = "~/data/scEiaD/2021_02_05_MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-0.6.sqlite", idleTimeout = 3600000)
 
@@
 # meta data before doublet removal
-load('~/data/scEiaD/n_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__preFilter.scEiaDprojected__dist-0.2__neighbors-500.umap.Rdata')
+load('~/data/scEiaD/n_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__preFilter.scEiaDprojected__dist-0.1__neighbors-50.umap.Rdata')
 umapRef <- umap
@@
 # cell type prediction
-load('~/data/scEiaD/PREDICTIONSn_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__preFilter.scEiaDprojected__dist-0.2__neighbors-500.Rdata')
-test_predictions <- data.table::fread('~/data/scEiaD/981791_CTpredictortest_data_probabilities.csv.gz') %>% as_tibble %>% 
+load('~/data/scEiaD/PREDICTIONSn_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__preFilter.scEiaD__dist-0.1__neighbors-50.Rdata')
+test_predictions <- data.table::fread('~/data/scEiaD/390498_CTpredictortest_data_probabilities.csv.gz') %>% as_tibble %>% 
   mutate(max_pred_prob = rowMaxs(.[,-(33:36)] %>% as.matrix ), 
@@
 The area under the PR curve (AUC) summarizes the effectiveness of the model across different cell types, with 1 being the highest performance.
-The xgboost model can predict rods, bipolar cells, and Muller glia with near perfect performance (`r fig_cap(name = 'supFig_PR', display = 'cite')`).
+The xgboost model can predict rods, bipolar cells, and Müller glia with near perfect performance (`r fig_cap(name = 'supFig_PR', display = 'cite')`).
 Most of the remaining cell types can be predicted with an AUC of well over 0.9.
@@
 input[['dotplot_Gene']] <- top_markers %>%  
-  filter(grepl('Amacrine|Rod|Cone|Retinal|Muller|Horizon|Bipol|Astro|RPE|Micro', cluster), D_KL > 0.2) %>%  
-  group_by(cluster)  %>% 
-  slice_max(mean_auc, n =5) %>% 
+  filter(grepl('Amacrine|Rod|Cone|Retinal|Muller|Horizon|Bipol|Astro|RPE', cluster), !grepl('C1orf|MT-', Gene), `log.p.adj` < -10000) %>%  
+  group_by(cluster) %>% slice_max(mean_auc, n =5) %>% 
   pull(Gene)
@@
 input[['meta_filter_cat']] <- 'CellType_predict'
-input[['meta_filter_on']] <- c('AC/HC Precursors','Amacrine Cells','Astrocytes','Bipolar Cells','Cones','Early RPCs','Endothelial','Horizontal Cells','Late RPCs','Microglia','Muller Glia','Neurogenic Cells','Pericytes','PR Precursors','Red Blood Cells','Retinal Ganglion Cells','Rod Bipolar Cells','Rods','RPCs','RPE','Schwann','Tabula Muris')
+input[['meta_filter_on']] <- c('AC/HC Precursors','Amacrine Cells','Astrocytes','Bipolar Cells','Cones','Early RPCs','Endothelial','Horizontal Cells','Late RPCs','Microglia','Muller Glia','Neurogenic Cells','Pericytes','PR Precursors','Retinal Ganglion Cells','Rod Bipolar Cells','Rods','RPCs','RPE','Tabula Muris')
 ctp2 <- make_meta_scatter_umap_plot(input, mf, meta_filter,
@@
 "PDE6G AND Retina").
-Of the 55 genes in (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a), 40 had one or more citations in the gene - cell type search (table X) and 51 had one or more citation in the inclusive search.
+Of the 55 genes in (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a), 39 had one or more citations in the gene - cell type search (table X) and 48 had one or more citation in the inclusive search.
 The 55 genes had a mean of `r top_markers %>% group_by(cluster)  %>% slice_max(mean_auc, n =5) %>% pull(c2) %>% mean() %>% round()` studies (with the inclusive gene by "retina" search).
@@
 The scVI model is first built the on human data.
-The mouse and macaque data are then projected (or queried) onto it with the scVI/scArches approach.
+The mouse and macaque data are then projected (or queried) onto it with the scVI - scArches approach.
 While this system works very well to integrate information between these three species, this approach may not scale to more distantly related species.
-Another limitations is that discrepancies between cell type labels between different labs makes certain transitioning cell type labels a bit imprecise.
+Another limitation is that discrepancies between cell type labels between different labs makes certain transitioning cell type labels a bit imprecise.
 One example is how the rods and photoreceptor precursor labels overlap somewhat.
 Though we attempted to ameliorate the issue by removing cell type labels in large disagreement with the consensus our cell type labels are effectively "crowd-sourced" and some disagreements can propagate into our labeled cells.
+As much of the data we used was 3' biased (10X, DropSeq), we decided to collapse all transcript specific information into the gene level.
 
@@
 fontsize(predictions %>%  
-           mutate(max_pred_prob = rowMaxs(.[,-(31:ncol(predictions))] %>% as.matrix )) %>% 
-           select(-cell_type_id) %>% 
+           mutate(max_pred_prob = rowMaxs(.[,-(30:ncol(predictions))] %>% as.matrix )) %>% 
+           select(-CellTypeID) %>% 
            rename(PredCellType = CellType) %>% 

### 2021-02-07 figs_and_tables/dotplot.R

@@ -14,8 +14,8 @@ make_dotplot <- function(input, db, meta_filter, cat_to_color_df){
   grouping_features <- input$dotplot_groups
   
   # ct order
-  ct_order <- data.frame(CT = c('Cones','Rods','RPE', 'Retinal Ganglion Cells', 'Astrocytes','Amacrine Cells','Bipolar Cells','Rod Bipolar Cells','Horizontal Cells','Muller Glia', 'Microglia'),
-                         Order = c(1,2,3,4,5,6,7,8,9,10, 11))
+  ct_order <- data.frame(CT = c('Cones','Rods','RPE', 'Retinal Ganglion Cells', 'Astrocytes','Amacrine Cells','Bipolar Cells','Rod Bipolar Cells','Horizontal Cells','Muller Glia'),
+                         Order = c(1,2,3,4,5,6,7,8,9,10))
   if (input$dotplot_filter_cat != ''){
     dotplot_data <- db %>% tbl('grouped_stats') %>%
       filter(Gene %in% gene) %>%

### 2021-02-07 figs_and_tables/fig2__integrationPerf.R

@@ -1,7 +1,7 @@
 load('~/data/scEiaD//metrics_2020_11_28.Rdata')
 perfO <- perf
 
-load('~/data/scEiaD//merged_stats_2021_01_01.Rdata')
+load('~/data/scEiaD//merged_stats_2021-02-05.Rdata')
 
 perf <- perf %>% 
   mutate(knn = case_when(method == 'scArches' ~ 7, TRUE ~ knn))
@@ -100,7 +100,7 @@ zscore_sum_all_methods <- zscore_tabula  %>%
 
 zscore_droplet_scVI_optimize <- perf_scVI%>% 
   filter(knn > 0.4, knn < 10) %>% 
-  filter(grepl('scVI', method)) %>% 
+  filter(grepl('scVI', method), epochs == 5) %>% 
   pivot_wider(names_from = c('Score','Group'), values_from = Value) %>% 
   mutate(sumZScale = 
            -scale(LISI_CellType)[,1] +

### 2021-02-07 figs_and_tables/sup_fig2__celltype_xgboost_labelling.R

@@ -1,6 +1,6 @@
 ct_processing <-  meta_filter %>% 
   #filter(CellType %in% common) %>% 
-  filter(is.na(CellType) | CellType != 'Mast') %>% 
+  filter(!is.na(CellType), CellType_predict != 'Tabula Muris') %>% 
   filter(!(!Platform %in% c('10xv2','10xv3','DropSeq') & CellType_predict == 'Unlabelled'  )) %>% 
   filter(!(!Platform %in% c('10xv2','10xv3','DropSeq') & CellType_predict == 'Muller Glia Progenitor'  )) %>% 
   mutate(CellType = case_when(is.na(CellType) ~ 'Unlabelled', 
@@ -73,7 +73,9 @@ ct_mat <- ct_processing %>%
   summarise(Count = n()) %>% 
   mutate(Ratio = Count / sum(Count)) %>% 
   select(-Count) %>% 
-  pivot_wider(values_from = Ratio, names_from = CellType_predict)
+  #filter(!CellType_predict %in% c('Astrocytes', 'Artery',  'Choriocapillaris', 'Tabula Muris','B-Cell','T-Cell', 'Smooth Muscle Cell', 'Unlabelled'), !CellType %in% c('Astrocytes','Artery','Choriocapillaris','Tabula Muris','B-Cell','T-Cell', 'Smooth Muscle Cell')) %>% 
+  #filter(!CellType_predict %in% c('Astrocytes', 'Artery',  'Choriocapillaris', 'Tabula Muris','Smooth Muscle Cell', 'Unlabelled'), !CellType %in% c('Astrocytes','Artery','Choriocapillaris', 'Smooth Muscle Cell')) %>% 
+  pivot_wider(values_from = Ratio, names_from = CellType_predict) 
 ct_mat[is.na(ct_mat)] <- 0
 #ct_mat <- data.frame(ct_mat)
 

### 2021-02-07 src/make_dotplot_markers.R

@@ -2,7 +2,7 @@ library(pool)
 library(RSQLite)
 library(tidyverse)
 library(glue)
-scEiaD_2020_v01 <- dbPool(drv = SQLite(), dbname = "~/data/scEiaD/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.2-500-0.6.sqlite", idleTimeout = 3600000)
+scEiaD_2020_v01 <- dbPool(drv = SQLite(), dbname = "~/data/scEiaD/2021_02_05_MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-0.6.sqlite", idleTimeout = 3600000)
 
 meta_filter <- fst::read_fst('~/git/plaeApp/inst/app/www/meta_filter.fst') %>% as_tibble()
 meta_filter <- meta_filter %>% mutate(CellType_predict = case_when(CellType_predict == 'Photoreceptor Precursors' ~ 'PR Precursors',
@@ -16,27 +16,28 @@ meta_filter <- meta_filter %>% mutate(CellType_predict = case_when(CellType_pred
 markers_summary <- scEiaD_2020_v01 %>% 
   tbl('wilcox_diff_testing') %>% 
   group_by(Base, Gene) %>% 
-  summarise(count = sum(AUC > 0.7), 
+  summarise(auc_count = sum(AUC > 0.5), 
             pval = min(`p.value`),
             FDR = min(FDR),
             mean_auc = mean(AUC)) %>% 
-  mutate(cluster = Base)
+  mutate(cluster = Base) %>% 
+  as_tibble()
 
 marker_info <-  markers_summary %>% 
   #filter(pval < 1) %>% 
-  group_by(cluster) %>% 
+  #group_by(cluster) %>% 
   #top_n(4, meauc) %>% 
   #filter(grepl('Amacrine|Rod|Cone|Retinal|Muller|Horizon|Bipol|RPE|Astro', cluster)) %>% 
-  data.frame() %>% 
+  as_tibble() %>% 
   left_join(scEiaD_2020_v01 %>% 
               tbl('genes') %>% 
               collect() %>% 
               mutate(gene = str_extract(Gene, 'ENSG\\d+'))) %>% 
-  left_join(marker_list %>% 
-              bind_rows(.id = 'cluster') %>% 
[…]

### 2021-02-10 manuscript_00.Rmd
updates

@@
 load('~/data/scEiaD/top_markers.Rdata')
-
-# pmid search
 load('~/data/scEiaD/top_marker_pmid.Rdata')
@@
 We identified ocular single cell RNA sequencing (scRNA) studies by querying PubMed, the Sequence Read Archive (SRA), and the European Nucleotide Archive (ENA) for the inclusive terms "retina", "single cell", "scRNA", "ocular", "eye", "transcriptome." We then hand filtered the results to only keep ocular and normal (non-perturbed or mutagenized) data from single cell RNA-seq technology.
-On September 2020 we identified `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` deposited datasets that have been published in `r meta_filter %>% pull(Citation) %>% unique() %>% length()` publications (`r fig_cap(name = 'fig1', display = 'cite')`).
-In December 2020 we ran our search again, find N studies to project onto scEiaD.
+On December 2020 we identified `r meta_filter %>% pull(study_accession) %>% unique() %>% length() - 1` deposited datasets that have been published in `r meta_filter %>% pull(Citation) %>% unique() %>% length()` publications (`r fig_cap(name = 'fig1', display = 'cite')`).
 To provide a non-ocular reference we also downloaded the raw sequence data from the Tabula Muris project for re-processing.
@@
 A core objective of many scRNA based studies is labeling the cell types.
-As this information is crucial to assess dataset integration and provide an accurate reference for user querying, we extracted individual cell labels with a combination of inspecting the GEO web site, supplemental information from the publication, custom web resources (e.g.
+As this information is crucial to assess dataset integration and provide an accurate reference for user querying, we extracted individual cell labels with a combination of inspecting the GEO web site, supplemental information from the publication, and web resources (e.g.
 a web app was created for the paper), and personal correspondence.
@@
 input[['dotplot_Gene']] <- top_markers %>%  
-  filter(grepl('Amacrine|Rod|Cone|Retinal|Muller|Horizon|Bipol|Astro|RPE', cluster), !grepl('C1orf|MT-', Gene), `log.p.adj` < -10000) %>%  
+  filter(grepl('Amacrine|Rod|Cone|Retinal|Muller|Horizon|Bipol|Astro|RPE', cluster)) %>%  
   group_by(cluster) %>% slice_max(mean_auc, n =5) %>% 
   pull(Gene)
+# bipolar have none with auc_count > 0, so run again
+# input[['dotplot_Gene']] <- c(input[['dotplot_Gene']], 
+#                              top_markers %>%  
+#   filter(cluster %in% c('Amacrine Cells', 'Bipolar Cells'), !grepl('C1orf|MT-', Gene), `log.p.adj` < -10000) %>%  
+#   group_by(cluster) %>% slice_min(`log.p.adj`, n =5) %>% 
+#   pull(Gene))
+dplot_markers <- input[['dotplot_Gene']]
 input[['dotplot_groups']] <- c('CellType_predict')
@@
 input[['dotplot_filter_on']] <- (meta_filter %>%
-                                   filter(grepl('Amacrine|Rod|Cone|Retinal|Muller|Horizon|Bipol|Astro|RPE|Micro', CellType_predict)) %>% 
+                                   filter(grepl('Amacrine|Rod|Cone|Retinal|Muller|Horizon|Bipol|Astro|RPE', CellType_predict)) %>% 
                                    pull(CellType_predict) %>% unique())
@@
 
-As another check on the quality of the cell type assignments, we ran pairwise differential expression tests between the predicted cell types (see methods for further details).
+As another check on the quality of the cell type assignments, we ran the cell and cluster independent haystack gene search and pairwise differential expression tests between the predicted cell types (see methods for further details).
 We show the five most consistently differentially expressed genes for each of the major retina cell types (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a).
@@
 "PDE6H AND Cones") and the second search is the more inclusive "gene AND retina" (e.g.
-"PDE6G AND Retina").
-Of the 55 genes in (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a), 39 had one or more citations in the gene - cell type search (table X) and 48 had one or more citation in the inclusive search.
-The 55 genes had a mean of `r top_markers %>% group_by(cluster)  %>% slice_max(mean_auc, n =5) %>% pull(c2) %>% mean() %>% round()` studies (with the inclusive gene by "retina" search).
-In contrast, 100 randomly chosen genes had a mean of `r pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% pull(value) %>% mean() %>% round`.
+"PDE6H AND Retina").
+Of the 50 genes in (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a), `r 50 - top_markers %>% filter(Gene %in% dplot_markers) %>% filter(c1 == 0) %>% nrow` had one or more citations in the gene - cell type search (table X) and `r 50 - top_markers %>% filter(Gene %in% dplot_markers) %>% filter(c1 == 0, c2 == 0) %>% nrow` had one or more citation in the inclusive search.
+The 50 genes had a mean of `r top_markers %>% filter(Gene %in% dplot_markers) %>% pull(c2) %>% mean() %>% round()` studies (with the inclusive gene by "retina" search).
+In contrast, 100 randomly chosen genes had a mean of `r pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% pull(value) %>% mean() %>% round` (wilcox test p < 2.5 x 10**-12).
 
-The four genes with zero hits were SCRG1 (Astrocytes), BTBD8 (Bipolar Cells), GATM (Microglia), and FRZB (Muller Glia).
-In a more exhaustive literature search we find that SCRG1 is a secreted protein which marks brain neurons, regulates human mesenchymal stem cells differentiation, and is overexpressed in brain astrocyte precursors [@aomatsuNovelSCRG1BST12014; @chaboubTemporalProfilingAstrocyte2016].
-BTBD8 contains BTB/POZ domains known to generally be associated with with transcription factor mediated protein-protein interactions and has been seen to be overexpresed in enteric neurons [@xuMolecularCloningCharacterization2004; @zollmanBTBDomainFound1994; @drokhlyansky_human_2020].
-GATM is mitochondrial enzyme involved in creatine biosynthesis and was previously shown to up-regulated in a brain microglia population [@hammondSingleCellRNASequencing2019].
-FRZB was previously referred to as SFRP3 and this Wnt pathway member has been previously shown to be active in Muller Glia [@zhuCharacterizationCanonicalWnt2018].
-Overall, as expected, we find very high concordance with known retina cell type markers.
+<!-- (broom::tidy(wilcox.test(pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% pull(value), top_markers %>% filter(Gene %in% dplot_markers) %>% pull(c2))))$p.value -->
 
@@
 It also returns for each pair-wise comparison an area under the curve (AUC) score, where 1 is a perfect power to distinguish and 0 is no power.
-The singleCellHaystack uses a Kullback-Leibler divergence measurement of the scVI lower dimensional space to identify genes with non-random distribution.
-A higher "D KL" score represents a gene with "specific" expression in teh UMAP/lower dimensional space.
-We filtered to keep genes with FDR < 0.01 and a mean AUC > 0.2 or  a D KL > 0.2.
+The singleCellHaystack algorithm uses a Kullback-Leibler divergence (D KL) measurement of the scVI lower dimensional space to identify genes with non-random distribution.
+A higher D KL score represents a gene with "specific" expression in the lower dimensional space and is used to calculate a FDR corrected p value against the full distribution of D KL values.
+We filtered to keep genes with scran FDR < 1, a mean AUC > 0.2, and a log10(D KL FDR) < -10000.
+No more than 50 genes for each cell type were retained (sorted by mean AUC).
 
@@
 One example is how the rods and photoreceptor precursor labels overlap somewhat.
-Though we attempted to ameliorate the issue by removing cell type labels in large disagreement with the consensus our cell type labels are effectively "crowd-sourced" and some disagreements can propagate into our labeled cells.
+Though we attempted to ameliorate the issue by removing cell type labels in large disagreement with the consensus, some disagreements can propagate into our labeled cells.
 As much of the data we used was 3' biased (10X, DropSeq), we decided to collapse all transcript specific information into the gene level.
+While scEiaD distinguishes the cell types very well, for some of the cell types there exist many "sub types" - notably the Amacrine cells have a huge variety in morphology with a recent study identifying over 60 different types of amacrine cells.
+At this time our batch corrected pan retina cell space does not resolve individual sub cell types with high resolution.
 
@@
 
-We originally intended to just use the Seurat CCA method to integrate the datasets.
+We originally intended to use the Seurat CCA method to integrate the datasets.
 However, the long run-times of CCA and poor integration of our known cell types led us to benchmark more methods and parameters.
-We also originally attempted to "hand-assess" the integration results by using the UMAP 2D projection view.
-This proved to scale poorly and this led use to curate some of the most useful benchmarking algorithms (NMI, ARI, silhouette, and LISI) that roughly matched our "hand-assessed" results into the scPOP R package.
-While we chose the scVI algorithm, we do not believe that it will work best for all datasets.
-We strongly suggest any other groups attempting a similar meta-atlas construction chose a quantifiable set of criteria so optimal methods and parameters can be picked.
+After adding more integration methods we first attempted "hand-assess" the integration results by using the UMAP 2D projection view.
+This proved to scale poorly and this led use to curate some of the more useful benchmarking algorithms (NMI, ARI, silhouette, and LISI) that roughly matched our "hand-assessed" results into the scPOP R package.
+While we chose the scVI algorithm, we strongly suggest any other groups attempting a similar meta-atlas construction chose a quantifiable set of criteria so optimal methods and parameters can be picked.
 
@@
 
-We hand curate over 300,000 cell type labels.
-With a standard xgboost algorithm using the latent dimensions, cell age classification (developing or matured), and the UMAP coordinates, we can very accurately label the remaining cells.
-Many other cell type labelling algorithms and systems exist for those groups less willing or able to tune a machine learning algorithm.
+We first hand curated over 300,000 cell type labels.
+With a xgboost algorithm using the latent dimensions, cell age classification (developing or matured), and the UMAP coordinates, we can very accurately label the remaining cells.
+Many other cell type labeling algorithms and systems exist for those groups less willing or able to tune a machine learning algorithm.
 For example, the developers of scVI also have a cell type label projection algorithm called scANVI.
@@
            summarise(Count = n()) %>% 
-           mutate(Ratio  = round(Count / sum(Count),2 )) %>% 
+           mutate(Ratio  = round(Count / sum(Count),3 )) %>% 
            filter(Ratio > 0.05) %>% 
@@
          size = 8, 
-         part = 'all')
+         part = 'all') %>% 
+  colformat_double(digits = 2)
 #Output Caption (* * for italics)
@@
 #source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
-fontsize(xgboost_pr_table_full %>% arrange(`Cell Type`, Study) %>% flextable() %>% width(width = 3), size = 8, part = 'all')
+fontsize(xgboost_pr_table_full %>% arrange(`Cell Type`, Study) %>% flextable() %>% width(width = 3), size = 8, part = 'all') %>% colformat_double(digits = 2)
 #Output Caption (* * for italics)
@@
 input[['pt_size_back']] <- 5
-input[['pt_size_meta']] <- 5
+input[['pt_size_meta']] <- 7
 input[['gene_and_meta_scatter_tech']] <- 'Droplet'

### 2021-02-10 figs_and_tables/fig1.R

@@ -40,7 +40,7 @@ tidy_dagitty(dagified)
 set.seed(51345)
 b <- tidy_dagitty(dagified) %>% as_tbl_graph() %>% 
   mutate(type = case_when(name %in% c('fastq', 'CellType', 'BatchInfo') ~ 'Input',
-                          name %in% c('DiffTesting', 'Velocity', 'Trajectory','umap', 'projectedCellTypes','cluster') ~ 'Outputs'),
+                          name %in% c('DiffTesting', 'Velocity', 'Trajectory','umap', 'projectedCellType','cluster') ~ 'Outputs'),
          name = case_when(name == 'CellType' ~ 'Published\nCell Types',
                           name == 'projectedCellType' ~ 'Learned\nCell Types',
                           name == 'DiffTesting' ~ 'Diff\nTesting',

### 2021-02-10 src/make_dotplot_markers.R

@@ -16,10 +16,12 @@ meta_filter <- meta_filter %>% mutate(CellType_predict = case_when(CellType_pred
 markers_summary <- scEiaD_2020_v01 %>% 
   tbl('wilcox_diff_testing') %>% 
   group_by(Base, Gene) %>% 
-  summarise(auc_count = sum(AUC > 0.5), 
+  summarise(auc_count_05 = sum(AUC > 0.5),
+            auc_count_02 = sum(AUC > 0.2), 
             pval = min(`p.value`),
             FDR = min(FDR),
-            mean_auc = mean(AUC)) %>% 
+            mean_auc = mean(AUC),
+            median_auc = median(AUC)) %>% 
   mutate(cluster = Base) %>% 
   as_tibble()
 
@@ -58,14 +60,17 @@ exp_stats <- scEiaD_2020_v01 %>% tbl('grouped_stats') %>%
          Expression = round(cpm * (`%` / 100), 2)) %>%
   select_at(vars(one_of(c('Gene', grouping_features, 'cell_exp_ct', 'Count', '%', 'Expression')))) %>%
   arrange(-Expression) 
-# D_KL > 0.15
+
 top_markers <- marker_info %>% 
   left_join(exp_stats %>% dplyr::rename(cluster = CellType_predict)) %>% 
   mutate(pval = as.numeric(pval), FDR = as.numeric(FDR)) %>% 
-  filter((FDR < 1 |  mean_auc > 0.10), D_KL > 0.15, `%` > 10) %>% 
+  filter((FDR < 1 |  mean_auc > 0.10), `log.p.adj` < -1000, `%` > 10, auc_count_02 > 0) %>% 
   group_by(Gene) %>% 
   slice_max(order_by = `%`) %>% 
-  filter(grepl('Amacrine|Rod|Cone|Retinal|Muller|Horizon|Mast|Bipol|Astro|RPE|Micro', cluster), !grepl('^NA', Gene))
+  filter(grepl('Amacrine|Rod|Cone|Retinal|Muller|Horizon|Mast|Bipol|Astro|RPE', cluster), !grepl('^NA', Gene)) %>% 
+  ungroup() %>% 
+  group_by(Base) %>% 
+  slice_max(order_by = `mean_auc`, n = 50)
 
 #save(top_markers, marker_info, exp_stats, file = '~/data/massive_integrated_eye_scRNA/top_markers.Rdata')
 source('src/pubmed_query.R')

### 2021-03-22 manuscript_00.Rmd
updates

@@
 We developed an R package (scPOP) that can scan the results of multiple methods and key parameter choices and create a ranked list of the best performing method / parameter combinations.
-We use this package along with a Snakefile based workflow system to demonstrate how to optimally merge over one million cells from 30 datsets and three species to create a massive ocular single cell transcriptome meta-atlas.
+We use this package along with a Snakefile based workflow system to demonstrate how to optimally merge 750,000 cells from 32 datsets and three species to create a massive ocular single cell transcriptome meta-atlas.
 This provides a model how to efficiently create meta-atlases for tissues and cells of interest."
@@
 
-scEiaD_2020_v01 <- dbPool(drv = SQLite(), dbname = "~/data/scEiaD/2021_02_05_MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-0.6.sqlite", idleTimeout = 3600000)
+scEiaD_2020_v01 <- dbPool(drv = SQLite(), dbname = "~/data/scEiaD/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite", idleTimeout = 3600000)
+x_dir <- -1
+y_dir <- 1
 
-meta_filter <- fst::read_fst('~/git/plaeApp/inst/app/www/meta_filter.fst') %>% as_tibble()
+meta_filter <- fst::read_fst('~/data/scEiaD/2021_03_17_meta_filter.fst') %>% as_tibble()
 meta_filter <- meta_filter %>% mutate(CellType_predict = case_when(CellType_predict == 'Photoreceptor Precursors' ~ 'PR Precursors',
@@
                                                                    TRUE ~ CellType_predict),
-                                      UMAP_1 = UMAP_1 * 1,
-                                      UMAP_2 = UMAP_2 * -1) %>%
+                                      CellType = case_when(CellType == 'Photoreceptor Precursors' ~ 'PR Precursors',
+                                                           CellType == 'AC/HC_Precurs' ~ 'AC/HC Precursors',
+                                                           CellType == 'RPC' ~ 'RPCs',
+                                                           CellType == 'Mesenchymal/RPE/Endothelial' ~ 'Endothelial',
+                                                           TRUE ~ CellType)) %>%
+  mutate(UMAP_a = UMAP_2 * x_dir,
+         UMAP_b = UMAP_1 * y_dir) %>%
+  mutate(UMAP_1 = UMAP_a, UMAP_2 = UMAP_b) %>%
   mutate(CellType_predict = case_when(!is.na(TabulaMurisCellType_predict) ~ 'Tabula Muris',
@@
 # meta data before doublet removal
-load('~/data/scEiaD/n_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__preFilter.scEiaDprojected__dist-0.1__neighbors-50.umap.Rdata')
+load('~/data/scEiaD/n_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__preFilter.scEiaD__dist-0.1__neighbors-50.umap.Rdata')
 umapRef <- umap
@@
 
-gse <- read_tsv('~/git/scEiaD/data/GEO_Study_Level_Metadata.tsv')
+gse <- read_tsv('~/git/scEiaD/data/GEO_Study_Level_Metadata.tsv') %>% filter(study_accession %in% meta_filter$study_accession)
 
@@
 # cell type prediction
-load('~/data/scEiaD/PREDICTIONSn_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__preFilter.scEiaD__dist-0.1__neighbors-50.Rdata')
-test_predictions <- data.table::fread('~/data/scEiaD/390498_CTpredictortest_data_probabilities.csv.gz') %>% as_tibble %>% 
-  mutate(max_pred_prob = rowMaxs(.[,-(33:36)] %>% as.matrix ), 
-         pred_correct = ifelse(cell_type_id == true_cell_id, 'Correct', 'incorrect'))
+load('~/data/scEiaD/predictions_n_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__preFilter.scEiaD__dist-0.1__neighbors-50.Rdata')
+# test_predictions <- data.table::fread('~/data/scEiaD/390498_CTpredictortest_data_probabilities.csv.gz') %>% as_tibble %>% 
+#   mutate(max_pred_prob = rowMaxs(.[,-(33:36)] %>% as.matrix ), 
+#          pred_correct = ifelse(cell_type_id == true_cell_id, 'Correct', 'incorrect'))
+test_predictions <- model_out$test_dat %>% as_tibble %>% 
+  mutate(max_pred_prob = rowMaxs(.[,-(31:34)] %>% as.matrix), 
+         pred_correct = ifelse(CellTypeID == true_label_id, 'Correct', 'incorrect'))
 # created by running 'source' on src/make_dotplot_markers.R
@@
 load('~/data/scEiaD/top_marker_pmid.Rdata')
+
+# replace cone bipolar with bipolar
+meta_filter$CellType <- gsub('Cone Bipolar','Bipolar', meta_filter$CellType)
+
+# kallisto stats
+run_info <- read_tsv('~/data/scEiaD/aggregated_run_info.tsv.gz')
+write_tsv(run_info %>% 
+            filter(sample_accession %in% meta_filter$sample_accession) %>% 
+            select(-file) %>% 
+            left_join(meta_filter %>% 
+                        group_by(sample_accession) %>% 
+                        summarise(nCount_RNA = mean(nCount_RNA), nFeature_RNA = mean(nFeature_RNA), percent_mt = mean(percent_mt)) %>% 
+                        select(sample_accession, nCount_RNA, nFeature_RNA, percent_mt)), 
+          file = 'supplemental_file__kallisto_stats.tsv')
 ```
@@
 This issue can be addressed only by re-quantifying the data in a uniform environment.
-Fortunately, due to the continued development of computationally light-weight gene quantification tools into the single-cell space (e.g kallisto bustools, alevin), re-quantification does not require massive compute resources [@melstedModularEfficientPreprocessing2019; @srivastavaAlevinEfficientlyEstimates2019].
+Fortunately, due to the continued development of computationally light-weight gene quantification tools into the single-cell space (e.g kallisto bustools, alevin-fry), re-quantification does not require massive compute resources [@melstedModularEfficientPreprocessing2019; @srivastavaAlevinEfficientlyEstimates2019].
 
-Still, even after re-quantification under identical conditions there remain study specific batch effects due to the variety in single cell technologies used and variation in tissue handling and processing across each scientific group.
+Still, even after re-quantification under identical computational conditions there remain study specific batch effects due to the diversity in single cell technologies used and variation in tissue handling and processing across each scientific group.
 The single cell community has recognized that removal of these technical (also referred to as batch) effects is a critical issue and have independently developed many tools, though it remains unclear which tools and parameters are optimal for a particular dataset [@butlerIntegratingSinglecellTranscriptomic2018; @haghverdiBatchEffectsSinglecell2018; @hieEfficientIntegrationHeterogeneous2019; @johnsonAdjustingBatchEffects2007; @korsunskyFastSensitiveAccurate2019; @liuJointlyDefiningCell2020; @lopezDeepGenerativeModeling2018; @polanskiBBKNNFastBatch2020; @QueryReferenceSinglecell; @simonINSCTIntegratingMillions2020; @stuartComprehensiveIntegrationSingleCell2019; @vandijkRecoveringGeneInteractions2018].
@@
 As there are thousands of possible permutations of single cell tools, references, and parameter choices, we create our meta-atlas (which we refer to as the single cell Eye in a Disk or scEiaD) by benchmarking integration outcomes across multiple important single cell RNA-seq processing parameters (batch removal method, number of hyper-variable genes (HVGs), clustering resolution, etc.).
-The benchmarking system we developed uses a wide variety of metrics that we make easily available via the R package scPOP (single-cell Pick Optimal Parameters).
-The scEiaD will be of utility to two communities: the ocular community who can both search scEiaD for gene expression across many dimensions (e.g.
+The benchmarking system we developed uses a wide variety of metrics that combine in the R package scPOP (single-cell Pick Optimal Parameters).
+The scEiaD will be of utility to two communities.
+First, the ocular community who can both search scEiaD for gene expression across many dimensions (e.g.
 cluster, cell type, study) and project their own single cell data onto scEiaD for comparison and rich automatic cell labelling.
-Also the computational community can use this very large, well-curated dataset to test algorithms for compute efficiency and performance in a diverse environment.
+Second, the computational community can use this very large, well-curated dataset to test algorithms for compute efficiency and performance in a diverse environment.
 As we believe data re-use is a powerful and efficient approach to facilitate discovery, we provide our meta-atlas code-base, the meta-atlas in several data formats, and propose general guidelines to optimally create custom meta-atlases.
@@
 In cases where the fastq file from the SRA was not processed properly, we acquired the bam files (SRA or personal correspondence) and re-extracted the fastq.
-After downloading all the data we had 5.7 TB across `r scEiaD_2020_v01 %>% tbl('metadata_filter') %>% pull(sample_accession) %>% unique() %>% length()` fastq file sets.
+After downloading all the data we had 11 TB across `r scEiaD_2020_v01 %>% tbl('metadata_filter') %>% pull(sample_accession) %>% unique() %>% length()` fastq file sets.
 
@@
 fig1_cap <- fig_cap(name='fig1', caption =  'a.
-Cartoon schematic of retina with major cell types delinearted b.
-Simplified directed workflow of major steps in scEiaD creation from raw counts to gene counts, benchmarking optimal integration methods (SnakePOP) to produce batch corrected latent dimensions (LatentDims), then downstream analysis outputs like clustering, differential gene testing (DiffTesting), and 2D UMAP visualization.
+Schematic of the retina with major cell types delineated b.
+Simplified directed workflow of major steps in scEiaD creation from raw counts to gene counts, benchmarking optimal integration methods (SnakePOP) to produce batch corrected latent dimensions (Latent Dims), then downstream analysis outputs like clustering, differential gene testing (Diff Testing), and 2D UMAP visualization.
 c.
-Counts of (study) accession, published papers, and batches for each scRNA technlogy, split by organism.
-d.
+Counts of published papers and batches (unique biological samples) for each scRNA technlogy, split by organism d.
 Cell type counts extracted from published studies for the more common retina cell types, split by species.
@@
 
+## Transcriptome quantification across multiple technologies
+Droplet and well based scRNA-seq technologies require different quantification approaches as the former have UMI and multiple cells are quantified within a single file.
+We wrote a Snakemake based pipeline (SnakeQUANT, see methods) to quantify and merge both droplet and well based technologies into a single matrix for downstream processing.
+For well based data we perform both gene and isoform level quantification; for droplet base technologies we quantify both exonic and intronic gene-level expression to facilitate calculation of RNA velocity.
+In total we quantify `r run_info %>% filter(sample_accession %in% meta_filter$sample_accession) %>% pull(n_processed) %>% sum() %>% formatC(., digits = 2)` molecules, finding `r run_info %>% filter(sample_accession %in% meta_filter$sample_accession) %>% pull(n_unique) %>% sum() %>% formatC(., digits = 2)` unique molecules with a mean pseudoalignment rate of `r ((run_info$n_pseudoaligned %>% sum() ) / (run_info$n_processed %>% sum()) * 100)  %>% formatC(., digits = 3)`%.
+Across the `r meta_filter %>% nrow() %>% format(., big.mark=",", scientific=FALSE)` cells (post QC) we have an average of `r meta_filter$nCount_RNA %>% mean() %>% formatC(., digits = 4, big.mark=",")` RNA counts across `r meta_filter$nFeature_RNA %>% mean() %>% formatC(., digits = 3, big.mark=",")` unique genes (see Supplemental File kallisto_stats.tsv and splicing_stats.tsv for more details).
+
 ## `r qc %>% nrow() %>% as.integer() %>% format(., big.mark=",", scientific=FALSE)` cells before quality control
@@
 After empty droplet removal, we had `r format(qc %>% nrow() %>% as.integer() , big.mark=",", scientific=FALSE)` cells.
-We then created a Seurat v3 object, calculated the percent mitochondrial genes counts, and removed cells which had more than 10% mitochondrial reads across all gene counts.
-For the droplet-based data, we also removed cells which had more than 3000 unique genes detected as these are likely to be doublets.
+We then created a Seurat v3 object and removed cells which had more than 10% mitochondrial reads across all gene counts or fewer than 250 unique genes quantified.
 After these standard quality control steps we were left with `r format(umapRef %>% nrow, big.mark = ",", scientific=FALSE)` cells.
@@
 A core objective of many scRNA based studies is labeling the cell types.
-As this information is crucial to assess dataset integration and provide an accurate reference for user querying, we extracted individual cell labels with a combination of inspecting the GEO web site, supplemental information from the publication, and web resources (e.g.
+As this information is crucial to assess dataset integration and provide an accurate reference for user querying, we extracted individual cell labels with a combination of inspecting the GEO web site, supplemental information from the publication, web resources (e.g.
 a web app was created for the paper), and personal correspondence.
-After normalizing cell type name nomenclature, we obtained labels for `r meta_filter$CellType %>% table() %>% sum()` cells across `r meta_filter$CellType %>% table() %>% length()` cell types (`r supTab_cap(name = 'supTab1_CTcounts', display = 'cite')`).
+After normalizing cell type name nomenclature, we obtained labels for `r meta_filter$CellType %>% table() %>% sum() %>% format(., big.mark=",", scientific=FALSE)` cells across `r meta_filter$CellType %>% table() %>% length() ` cell types (`r supTab_cap(name = 'supTab1_CTcounts', display = 'cite')`).
 
@@
 
-Disentangling the technical and biological effects from the datasets is crucial.
-A wide variety of methods have been written for this function.
-As we were uncertain which would perform the best and it is difficult to manually assess performance we ran `r methods %>% length()` tools with a commonly used set of key parameters: 2000 hyper variable genes (HVG), 8 and 30 latent dimensions returned, and the louvain clustering algorithm with 7 k-nearest neighbors.
+Disentangling the technical and biological effects when integrating multiple datasets is crucial.
+We define batch as each unique biological sample and assume each study is at least one unique sample.
+We studied the metadata and methods of each study to identify the unique biological samples.
+Within the current scEiaD data set we identified `r umapRef %>% pull(batch) %>% unique() %>% length()` batches across `r umapRef %>% pull(study_accession) %>% unique() %>% length()` deposited datasets in `r umapRef %>% left_join(gse) %>% pull(PMID) %>% unique() %>% length()` published papers.
+
+A wide variety of methods have been written for scRNA-seq integration.
+As we were uncertain which would perform the best, we ran `r methods %>% length()` tools with a commonly used set of key parameters like number of hyper variable genes (HVG), size of latent dimensional space, and the number of nearest neighbors for the louvain clustering algorithm.
 The Snakemake system was used to automate the running of the wide variety of tools.
-In total 15,226 jobs were run to quantify expression and assess integration performance.
+In total 5,591 jobs were run to quantify gene expression and build the unified Seurat objects (SnakeQUANT) and 2,446 jobs were run to assess integration performance (SnakePOP).
 
-The two key metrics which have to be balanced in order to optimize performance are batch mixing (the same cell types should be similar across independent studies) and cell type or cluster purity (where different cell types or clusters should be distinct).
+The two key metrics which have to be balanced in order to optimize integration performance are cell type or cluster purity (where different cell types or clusters should be homogenous) and batch mixing (the same cell types should be similar across independent studies).
 While these can be visually assessed by looking at marker gene expression across the 2D UMAP projection, it is more rigorous and scalable to quantify these diametrically opposed characteristics.
 
-We define batch as being each biological sample and assume each study is at least one unique sample.
-We studied the metadata and methods of each study to determine if the particular study contained multiple unique biological samples.
-In the end we identified `r umapRef %>% pull(batch) %>% unique() %>% length()` batches across `r umapRef %>% pull(study_accession) %>% unique() %>% length()` deposited datasets in `r umapRef %>% left_join(gse) %>% pull(PMID) %>% unique() %>% length()` published papers.
+## scPOP wraps several different methods for measuring integration performance.
 
-## scPOP wraps several different methods at measuring integration performance.
-
-scPOP is a lightweight, low dependency R package which brings together the Local Simpson Index (LISI) and Average Silhouette Width (ASW) metrics from Harmony and kBET, respectively, and brings in the Adjusted Rand Index (ARI) and Normalized Mutual Information (NMI) algorithms from aricode.
+Multiple methods have been proposed to quantitatively evaluate batch correction.
+Some of these metrics evaluate the concordance between sets of labels, while other compute distances between the individual data points of a given set of labels.
+While any one of these methods can be useful, we propose that calculating and evaluating them in tandem provides greater accuracy for dataset integration.
+We developed the R package scPOP, a lightweight, low dependency R package which brings together the Local Simpson Index (LISI), Average Silhouette Width (ASW), Adjusted Rand Index (ARI) and Normalized Mutual Information (NMI) metrics from the R packages Harmony, kBET and aricode, respectively.
 The LISI and ASW were used to measure batch mixing (where lower is better), cell type mixing (higher is better), and cluster mixing (higher is better).
@@
 Library size (libSize) normalization omits the sqrt or log transformation.
-Finally some methods like scVI directly use the raw counts data for modeling the data.
+Finally some methods, like scVI, directly use the raw counts data for modeling the data.
 
 To merge the different scores we define $sumZScale = \sum scale(m)$ where m is a metric (LISI by batch, LISI by cluster, LISI by cell type, silhouette by batch, silhouette by cluster, silhouette by celltype, NMI, and ARI).
-In this scaled metric that weights the different benchmarking algorithms evently, scVI performs best.
+In this scaled metric that weighs the different benchmarking algorithms evenly, scVI performs best.
 As expected the libSize normalization which omits the log or square root scaling generally performs the worst (`r fig_cap(name = 'figBenchmark', display = 'cite')`c).
@@
 We also see that changing the number of latent dimensions (8 or 30) can occasionally dramatically change performance.
-These results demonstrate the importance of assessing performance in a rigorous manner aceoss many parameters.
+These results demonstrate the importance of assessing performance in a rigorous manner across many parameters.
 
@@
 
-```{r figBenchmark, fig.width=14, fig.height=8, fig.cap=figBenchmark_cap, echo=FALSE, message = FALSE, warning = FALSE}
+```{r figBenchmark, fig.width=14, fig.height=9, fig.cap=figBenchmark_cap, echo=FALSE, message = FALSE, warning = FALSE}
 figBenchmark_cap <- fig_cap(name='figBenchmark', caption =  'a.
@@
 sumZScale (higher is better) for each method across a variety of data normalizations.
-All methods shown here use 2000 HVG, louvain clustering, and 8 latent dimensions d.
-Optimal parameters for scVI methods across 1000 to 10000 HVGs and 4 to 100 latent dimensions.')
+All methods shown here use 2000 HVG, louvain clustering, and 8 latent dimensions.
+Each color is a different method.
+d.
+Optimal parameters for scVI methods across 1000 to 10000 HVGs and 4 to 50 latent dimensions.')
 
@@
   nrow = 2,
-  rel_heights = c(1, 0.4)
+  rel_heights = c(1, 0.9)
 )
@@
 To find the best set of parameters for the creation of scEiaD we did a grid search across key parameters: HVG, latent dimensions, and k nearest neighbors.
-Furthermore we used a recent advance in scVI capability adapted from scArches that allows one to build a reference model and project the new cells onto it [@QueryReferenceSinglecell].
+Furthermore we used a recent advance in scVI capability (>= version 0.8.0) adapted from scArches that allows one to build a reference model and project the new cells onto it [@QueryReferenceSinglecell].
 We built a human-only scVI model and then projected the mouse and macaque data onto it.
 We then compared the "scVIprojected" against the against the straight scVI across all the previously mentioned parameters.
-Using scPOP we first saw that the scVIprojection approach generally performed better than running scVI with all of the data.
-We found the optimal overall parameters to be 5000 HVG, 8 latent dimensions, and a resolution of 0.6 with the leiden algorithm implemented by PARC (`r fig_cap(name = 'figBenchmark', display = 'cite')`d).
-We also varied the UMAP projection values of nearest neighbors and minimum distance to qualitatively pick a 2D projection, selecting a minimum distance of 0.2 and 500 nearest neighbors.
+Using scPOP we first saw that the scVIprojection approach generally performed better than running scVI with all of the data (`r fig_cap(name = 'figBenchmark', display = 'cite')`d).
+We found the optimal overall parameters to be 5000 HVG, 8 latent dimensions, and with 5 k-nearest neighbors for the cluster finding.
+We also varied the UMAP projection values of nearest neighbors and minimum distance to qualitatively pick a 2D projection, selecting a minimum distance of 0.1 and 50 nearest neighbors.
 
-## High accuracy xgboost ML model built to label unknown cell types
+## High accuracy xgboost ML model built to label unknown cell types across all technologies
 
@@
 Traditionally this is done by clustering the cells, then using cell type specific markers to label the clusters.
-However, as we had hundreds of thousands of expert labeled cells across `r umapRef %>% filter(!is.na(CellType)) %>% left_join(gse) %>% pull(PMID) %>% unique() %>% length()` studies (`r fig_cap(name = 'fig1', display = 'cite')`a) we built a xgboost-based machine learning model that used half of the labeled cells as a training set (see methods for more details) to train a cell type predictor for scEiaD input.
+However, as we had hundreds of thousands of expert labeled cells across `r umapRef %>% filter(!is.na(CellType)) %>% left_join(gse) %>% pull(PMID) %>% unique() %>% length()` publications (`r fig_cap(name = 'fig1', display = 'cite')`a) we built a xgboost-based machine learning model that used 2/3 of the labeled cells as a training set (see methods for more details) to train a cell type predictor for scEiaD input.
 The trained model was used to predict the cell type assignments for all cells in scEiaD.
@@
 
-<!-- ## ML cell type predictor has high accuracy across most cell types -->
-
-The precision recall (PR) curve visualizes the ability of the model to precisely label known cells at a given confidence.
+As a brief case study of our ML performance, we look at the cell type labels assigned to a Shekhar et al.
+study where they used SMART-seq on a retinal bipolar cell enriched population of cells [@shekharComprehensiveClassificationRetinal2016] .
+Even though our ML algorithm was trained only on droplet-based data, our algorithm labels most of this dataset as retinal bipolar cells, with the next most common cell types being Amacrine,  like Shekhkar et al.
+labeled them (`r fig_cap(name = 'supFigShekharSMART', display = 'cite')`).
+To more generally evaluate performance we use the precision recall (PR) curve, which visualizes the ability of the model to precisely label known cells at a given confidence.
 The area under the PR curve (AUC) summarizes the effectiveness of the model across different cell types, with 1 being the highest performance.
 The xgboost model can predict rods, bipolar cells, and Müller glia with near perfect performance (`r fig_cap(name = 'supFig_PR', display = 'cite')`).
-Most of the remaining cell types can be predicted with an AUC of well over 0.9.
+Most of the remaining cell types can be predicted with an AUC over 0.95 `r supTab_cap(name = 'supTab_AUCxgboost', display = 'cite')`.
 The cell types that were the most challenging for the model to predict were the artery, choriocapillaris, and vein.
@@
 The ML model also performs under 0.9 for predicting the neurogenic cell type label; however the most next common label that is predicts are the highly related retinal progenitor populations (RPCs) (`r supTab_cap(name = 'supTab_CTmislabels', display = 'cite')`).
-Overall, our xgboost based ML model shows strong accuracy across the major cell types of the retina.
+Overall, our xgboost based ML model shows strong accuracy across the major cell types of the retina (overall  AUC of 0.98).
 
@@
 d.
-Confusion matrix of cell type prediction performance of our xgboost labeller between predicted (y axis) and known (x axis) with data withheld from the machine learner.
+Confusion matrix of cell type prediction performance of our xgboost labeller between predicted (x axis) and known (y axis) using data withheld from the machine learner.
 Most of the cell types are indeed labelled as their true type.
@@
 input[['dotplot_Gene']] <- top_markers %>%  
-  filter(grepl('Amacrine|Rod|Cone|Retinal|Muller|Horizon|Bipol|Astro|RPE', cluster)) %>%  
+  filter(grepl('Amacrine|Rod|Cone|Retinal|Muller|Horizon|Bipol|Astro|RPE', cluster),
+         !grepl('C1orf|MT-', Gene)) %>%  
   group_by(cluster) %>% slice_max(mean_auc, n =5) %>% 
@@
 input[['meta_column']] <- 'CellType_predict'
-input[['pt_size_back']] <- 2
+input[['pt_size_back']] <- 4
 input[['pt_size_meta']] <- 2
@@
 input[['meta_filter_cat']] <- 'CellType_predict'
-input[['meta_filter_on']] <- c('AC/HC Precursors','Amacrine Cells','Astrocytes','Bipolar Cells','Cones','Early RPCs','Endothelial','Horizontal Cells','Late RPCs','Microglia','Muller Glia','Neurogenic Cells','Pericytes','PR Precursors','Retinal Ganglion Cells','Rod Bipolar Cells','Rods','RPCs','RPE','Tabula Muris')
+input[['meta_filter_on']] <- c('AC/HC Precursors','Amacrine Cells','Bipolar Cells','Cones','Early RPCs','Endothelial','Horizontal Cells','Late RPCs','Microglia','Muller Glia','Neurogenic Cells','Pericytes','PR Precursors','Retinal Ganglion Cells','Rod Bipolar Cells','Rods','RPCs','Tabula Muris')
 ctp2 <- make_meta_scatter_umap_plot(input, mf, meta_filter,
@@
           plot_grid(ctp$plot + theme(legend.position = "none"), 
-                    ctp2$plot + theme(legend.position = "none") + facet_wrap(~CellType_predict),
+                    ctp2$plot + theme(legend.position = "none") + facet_wrap(~CellType_predict, ncol = 4),
                     grid.grabExpr(draw(ct_confusion)),
@@
 After ML projection of cell type labels from the original `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` labels onto a total of `r meta_filter %>% filter(!is.na(CellType_predict) | !is.na(TabulaMurisCellType_predict)) %>% nrow()` cells we have substantially improved the number of studies per cell type.
-For example, we went from `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(2)` human studies with labelled Muller Glia to `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(5)` after labeling (`r supTab_cap(name = 'supTab_SAcounts', display = 'cite')`).
-Overall we go from an average of `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(2) %>% mean()` studies per human cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(5) %>% mean()` and `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(4) %>% mean()` studies per mouse cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(7) %>% mean()` after transferring the cell type labels.
+For example, we went from `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(2)` human studies with labeled Muller Glia to `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(5)` after labeling (`r supTab_cap(name = 'supTab_SAcounts', display = 'cite')`).
+Overall we go from an average of `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(2) %>% mean() %>% round()` studies per human cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(5) %>% mean() %>% round()` and `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(4) %>% mean() %>% round()` studies per mouse cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(7) %>% mean() %>% round()` after transferring the cell type labels.
 
@@
 "PDE6H AND Retina").
-Of the 50 genes in (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a), `r 50 - top_markers %>% filter(Gene %in% dplot_markers) %>% filter(c1 == 0) %>% nrow` had one or more citations in the gene - cell type search (table X) and `r 50 - top_markers %>% filter(Gene %in% dplot_markers) %>% filter(c1 == 0, c2 == 0) %>% nrow` had one or more citation in the inclusive search.
+Of the 50 genes in (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a), `r 50 - top_markers %>% filter(Gene %in% dplot_markers) %>% filter(c1 == 0) %>% nrow` had one or more citations in the gene - cell type search (`r supTab_cap(name = 'supTab_topMarkers', display = 'cite')`) and `r 50 - top_markers %>% filter(Gene %in% dplot_markers) %>% filter(c1 == 0, c2 == 0) %>% nrow` had one or more citation in the inclusive search.
 The 50 genes had a mean of `r top_markers %>% filter(Gene %in% dplot_markers) %>% pull(c2) %>% mean() %>% round()` studies (with the inclusive gene by "retina" search).
-In contrast, 100 randomly chosen genes had a mean of `r pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% pull(value) %>% mean() %>% round` (wilcox test p < 2.5 x 10**-12).
+In contrast, 100 randomly chosen genes had a mean of `r pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% pull(value) %>% mean() %>% round` (wilcox test p < 2.5 x 10^-12).
 
@@
 
-## Projection of new data onto scEiaD demonstrates how iPSC-based cells are similar to tissue-based cells
+## Projection of outside data onto scEiaD demonstrates similarities and differences of iPSC and organoids to primary cells
 
@@
 Very briefly, if outside groups quantify their mouse or human scRNA with kallisto and the same references (see methods), they can overlay their data on top of scEiaD by 1.
-installing scVI >=0.8.1, 2.
+installing scVI (version 0.9.0 or higher), 2.
 downloading our 13 megabyte scVI model, and 3.
-following the Jupyter notebook on Google colab that we provide as supplemental file BLOOP and as a live demo at http://colab.google.
+following the Jupyter notebook on Google colab that we provide as a live demo at http://colab.google.
 We demonstrate the power of this approach in two ways.
 
-First, we project another retina dataset onto scEiaD from Fadl et al.
-As this is a pan-retina tissue sample, we would expect most of the major cell types of the retina to be present.
-In the Google colab notebook we show how a set of the reads from SRR12130660 can be processed from the raw reads to a UMAP visualization overlapping our scEiaD in under 15 minutes.
-Indeed, as expected we see many amacrine, rods, muller glia, bipolar, a few cones and retinal ganglion, and a small number of non-retina cells (`r supFig_cap(name = 'projection_fig', display = 'cite')`b).
+First, a wild-type retinal organoid dataset from Kallman et al was projected onto scEiaD [@kallmanInvestigatingConePhotoreceptor2020].
+In the Google colab notebook we show how a subset of the reads from Kallman et al.'s SRR12130660 can be processed from the raw reads to a UMAP visualization in under 10 minutes.
+Indeed we see how organoid-based retinal cell types can be detected overlapping primary cells for RPCs, photoreceptors, amacrine cells, and retinal ganglion cells (`r supFig_cap(name = 'projection_fig', display = 'cite')`d).
 
-Second, we demonstrate how the RPE derived from the Bharti iPSC differentiation method closely match tissue RPE from other studies (the largest being the Cowan et al data; `r supFig_cap(name = 'projection_fig', display = 'cite')`a).
-Cells generated with this process have been extensively validated with functional RPE tests.
-Our scRNA approach provides another layer of evidence that the Bharti iPSC-RPE differentation protocol generates RPE cells that are functionally equivalent to human adult RPE in an unbiased manner.
+Finally, we demonstrate how the RPE derived from the Bharti iPSC differentiation process express canonical RPE markers of TTR and RPE65 but end up in a slightly different position in the UMAP projection`r fig_cap(name = 'RPE_fig', display = 'cite')`a).
+Differential expression analysis identified that vimentin is nearly exclusively expressed in the iPSC-based RPE `r fig_cap(name = 'RPE_fig', display = 'cite')`b).
+This result is not surprising as Hunt et al.
+demonstrated that cultured and proliferating RPE express high levels of vimentin [@huntAlteredExpressionKeratin1990] and this same trend is seen in bulk RNA-seq from stem cell RPE and primary RPE `r supFig_cap(name = 'eyeIntegration_screenshot', display = 'cite')`) [@swamyEyeDiskEyeIntegration2019].
+```{r RPE_fig, fig.width=14, fig.height=5,  fig.cap=RPE_fig_cap, echo=FALSE, message = FALSE, warning = FALSE}
+RPE_fig_cap <- fig_cap(name='projection_fig', caption =  'a.
+RPE distribution colored by study demonstrates how the most RPE are in two locations.
+iPSC-based RPE we provided are located more enriched in cluster 14.
+Tissue RPE more enriched in cluster 46.
+b.
+Violin plot of 3 functional RPE markers (TTR, RPE65, TYRP1) and vimentin')
+input <- list()
+input[['pt_size_back']] <- 5
+input[['pt_size_meta']] <- 10
+input[['meta_column']] <- 'study_accession'
+#input[['gene_and_meta_scatter_tech']] <- 'Droplet'
+input[['meta_column_transform']] <- 'None'
+input[['meta_filter_cat']] <- 'CellType_predict'
+input[['meta_filter_on']] <- c('RPE')
+rpe <- make_meta_scatter_umap_plot(input, mf, meta_filter,
+                                   celltype_predict_labels,
+                                   celltype_labels,
+                                   tabulamuris_predict_labels,
+                                   cluster_labels,
+                                   cat_to_color_df
+)
+
+genes4 <- c('VIM (ENSG00000026025)', 'RPE65 (ENSG00000116745)', 'TTR (ENSG00000118271)')
+gene_list_dat <- list()
+for (i in genes4){
+  gene_list_dat[[i]] <- 
+    scEiaD_2020_v01 %>% 
+    tbl('metadata_filter') %>% 
+    filter(cluster %in% c('34','47'), CellType_predict == 'RPE') %>% 
+    left_join(., scEiaD_2020_v01 %>% tbl('counts') %>% 
+                filter(Gene %in% c(i))) %>% 
+    collect() %>%
+    mutate(Gene = case_when(is.na(counts) ~ i, TRUE ~ Gene), 
+           counts = case_when(is.na(counts) ~ 0, TRUE ~ counts))
+}
+library(gghalves) 
+cluster_coords <- meta_filter %>% filter(cluster %in% c('34','47'), CellType_predict == 'RPE') %>% group_by(cluster) %>% 
+  mutate(cluster = case_when(cluster == '47' ~ '47 (iPSC RPE)', TRUE ~ '34 (Tissue RPE)')) %>% summarise(UMAP_1 = mean(UMAP_1), UMAP_2 = mean(UMAP_2))
+
+rpe_plot <- rpe$plot + coord_cartesian(xlim = c(-11,-2), ylim = c(-15,-5)) +
+  geom_label_repel(data = cluster_coords, aes(x=UMAP_1, y= UMAP_2, label = cluster), force = 300)
+gene_list_dat %>% bind_rows() %>% group_by(cluster) %>% summarise(UMAP_1 = mean(UMAP_1), UMAP_2 = mean(UMAP_2))
+
+vln_plt <- gene_list_dat %>% bind_rows() %>%  mutate(cluster = case_when(cluster == '34' ~ '34 (Tissue RPE)', TRUE ~ '47 (iPSC RPE)')) %>% ggplot(aes(x=Gene, y=counts, color = cluster))   + geom_half_boxplot() + geom_half_violin(side = 'r', aes(fill=cluster), scale = 'width') + cowplot::theme_cowplot() #+ facet_wrap(~organism)
+
+plot_grid(rpe_plot ,
+          vln_plt + coord_flip() + ggsci::scale_fill_nejm() + ggsci::scale_color_nejm(),
+          labels = c('a','b'))
+```
 
@@
 
-Gene quantification is handled by the SnakQUANT snakefile.
-Kallisto (0.46.2) was the pseudoaligner [@brayNearoptimalProbabilisticRNAseq2016].
-The reference transcriptomes used to build the kallisto index were the Gencode "gencode.vM25.annotation.gtf.gz" and "gencode.v35.annotation.gtf.gz" for mouse and human, respectively [@harrowGENCODEReferenceHuman2012; @jGENCODEReferenceHuman2012].
-The Ensembl release 101 "Macaca_mulatta.Mmul_101.gtf.gz" was used for the macaque transcriptome [@yatesEnsembl20202020].
-For the relatively few single ended samples, the params "--single -l 200 -s 30" were used for kallisto quant.
-Otherwise the "--bias" flag was added.
-For the droplet-based samples, bustools (0.39.4) was used to extract the single cell barcodes, sort the output, and fix barcoding errors [@melstedBarcodeUMISet2019].
+Gene quantification is handled by the SnakeQUANT snakefile.
+First, we generated multiple quantification indices to facilitate calculation of RNA velocity.
+For each droplet technology, we generated a separate set of transcript sequences that contain both exonic and intronic sequences using the “get_velocity_files” function from the R package BUSpaRse.
+The reference transcriptome annotation used to build sets of transcript sequences were the Gencode “gencode.vM25.annotation.gtf.gz” and “gencode.v35.annotation.gtf.gz” for mouse and human, respectively [@harrowGENCODEReferenceHuman2012; @jGENCODEReferenceHuman2012].
+Because the Macaca Fascicularis genome is less well annotated we used the Ensembl release 101  genome and transcriptome annotation “Macaca_mulatta.Mmul_101.gtf.gz”  [@yatesEnsembl20202020].
+A single set of exonic transcript sequences was created for each species for all well techonologies.
+A kallisto quantification index was generated for each of these fastas using  Kallisto index (0.46.2) [@brayNearoptimalProbabilisticRNAseq2016].
+Well based samples were quantified using kallisto quant.
+For the relatively few single ended samples, the params “–single -l 200 -s 30” were used for kallisto quant.
+Otherwise the “–bias” flag was added.
+For the droplet-based samples, we adapted the bustools workflow for generating spliced and unspliced count matricies.
+(0.39.4) [@melstedBarcodeUMISet2019].
 
@@
 We downloaded a mapping of orthologous genes between human, mouse, and macaque using the Ensembl BioMart web browser in November 2020.
-We identified 15XXX human genes that could be directly mapped to mouse and macaque orthologs.
-Genes present in mouse or macaque that were not found in human were not used for scVI batch correction, but were retaomed and used for differential gene expression.
+We identified 15,759 human genes that could be directly mapped to mouse and macaque orthologs.
+Genes present in mouse or macaque that were not found in human were not used for HVG gene selection, but were retained and used for differential gene expression.
 
@@
 We wrote the scPOP R package to unify the LISI and Silhouette metrics from Harmony and kBet, respectively, along with NMI and ARI [@buttnerTestMetricAssessing2019; @korsunskyFastSensitiveAccurate2019].
-LISI and Silhouette require a dense matrix, which is a problem for our data as a 7e5 x 20,000 matrix cannot fit in a 1.5 TB memory node.
+LISI and Silhouette require a dense matrix, which is a problem for our data as a `r meta_filter %>% nrow()` cell by 8 latent dimension dense matrix cannot fit in a 1.5 TB memory node.
 We down-sampled the dataset to ~100,000 cells, taking care to keep all rarer cell types for the LISI and Silhouette benchmarking.
@@
 While we removed droplet-based cells with more than 3,000 transcripts as these are highly likely to be doublets (more than one cell per droplet) there were likely many more doublets.
-We ran DoubletDetect and scrublet and calculated the distribution of DoubletDetect and scrublet scores across all clusters and removed clusters with a score in both metrics greather than 4 standard deviations above the mean [@adamgayosoJonathanShorDoubletDetectionDoubletdetection2020; @wolockScrubletComputationalIdentification2019].
+We ran DoubletDetect and scrublet and calculated the distribution of DoubletDetect and scrublet scores across all clusters and removed clusters with a score in both metrics greater than 4 standard deviations above the mean [@adamgayosoJonathanShorDoubletDetectionDoubletdetection2020; @wolockScrubletComputationalIdentification2019].
 This removed another `r format((umapRef %>% nrow() %>% as.integer()) - (meta_filter %>% nrow() %>% as.integer()), big.mark=",", scientific=FALSE)` cells, leaving `r format((meta_filter %>% nrow()) %>% as.integer(), big.mark=",", scientific=FALSE)` in total.
@@
 The model was trained using default parameters.
-The trained model had an overall F1 score of XXX an average precision and recall of XXX and XXX for each major retinal cell type.
-This model was then used to identify labels for each unlabelled cell.
-Unlabelled data was preprocessed identically to training data and fed into model to generate a vector of label probabilities for each cell.
-We selected the highest label probability for each cell, and required a minimum probability of .5 to assign a label to a cell.
+The trained model had an overall macro and micro AUC score of 0.98 and 0.99, respectively.
+The AUC scores ranged from.
+This model was then used to identify labels for each all cell.
+Unlabelled data was pre-processed identically to training data and fed into model to generate a vector of label probabilities for each cell.
+We selected the highest label probability for each cell, and required a minimum probability of 0.5 to assign a label to a cell.
 
@@
 
-<!-- ## velocity -->
+## Calculation of RNA velocity
+All RNA velocity calculations were done through the the velocriaptor implementation of the scVelo python library [@bergenGeneralizingRNAVelocity2020].
+From the anndata objects generated by our Snakemake pipeline we calculated velocity across all genes.
+Genes without detectable velocity were dropped.
+The scVI generated latent space (instead of PCA) was used to calculate first and second order moments.
+The calculated moments were used to estimate RNA velocity.
+Differential velocity was tested between celltypes using pairwise wilcox rank sum tests.
 
@@
 The scVI model is first built the on human data.
-The mouse and macaque data are then projected (or queried) onto it with the scVI - scArches approach.
+The mouse and macaque data are then projected (or queried) onto it with the scVI implementation of the scArches method.
 While this system works very well to integrate information between these three species, this approach may not scale to more distantly related species.
@@
 One example is how the rods and photoreceptor precursor labels overlap somewhat.
-Though we attempted to ameliorate the issue by removing cell type labels in large disagreement with the consensus, some disagreements can propagate into our labeled cells.
-As much of the data we used was 3' biased (10X, DropSeq), we decided to collapse all transcript specific information into the gene level.
-While scEiaD distinguishes the cell types very well, for some of the cell types there exist many "sub types" - notably the Amacrine cells have a huge variety in morphology with a recent study identifying over 60 different types of amacrine cells.
-At this time our batch corrected pan retina cell space does not resolve individual sub cell types with high resolution.
+Though we attempted to ameliorate the issue by removing cell type labels in large disagreement with the consensus, some disagreements could propagate into our machine labeled cells type assignments.
+
+While scEiaD distinguishes the major cell types very well, some of the cell types contain many "sub types" - notably the Amacrine cells have a huge variety in morphology with a recent study identifying over 60 different types of Amacrine cells in mouse.
+At this time our batch corrected pan retina cell space does not precisely resolve these sub cell types with high resolution.
+We are actively working to "sub cluster" the cell types so we can robustly and reliably identify the high diversity of retinal cell types across the entire retina.
 
@@
 The scVI batch correction step on around one million cells runs within a few hours on a GPU and 150GB of memory.
-This places this crucial step within the capabilities of a moderately large physical server or a cloud compute node.
-Further downstream proessing can largely be done a computers with 64+ Gb of memory and a few hundred GB of disk space.
+This places this crucial step within the capabilities of a moderately powerful computer or a cloud compute node.
+Further downstream processing can largely be done a computers with 64+ Gb of memory and a few hundred GB of disk space.
 We believe that our efforts can be replicated in any other tissue / system with a large number of independent studies by an individual lab or small number of scientists.
@@
 
-We first hand curated over 300,000 cell type labels.
+We first hand curated over 400,000 published cell type labels.
 With a xgboost algorithm using the latent dimensions, cell age classification (developing or matured), and the UMAP coordinates, we can very accurately label the remaining cells.
@@
 
-Many retina "atlases" have been published to date.
+Many retina atlases have been published to date.
 We argue that we have created the first atlas that is generally useful because 1.
@@
 our data is available via download in several forms at https://github.com/davemcg/scEiaD, and crucially 3.
-we provide a Google colab/Jupyter notebook which exactly lays out how to use scVI's to project (or query) outside data onto our scEiaD with minimum compute resources.
+we provide a Google colab/Jupyter notebook which exactly lays out how to use scVI to project (or query) outside data onto our scEiaD with minimum compute resources.
 Finally we also provide our xgboost model so outside groups can label their own datasets with our robust retina cell type labeling system.
@@
 
-supTab_CTmislabels_cap <- supTab_cap(name='supTab_CTmislabels', caption =  'Counts of cell type labels with our xgboost machine learning system (PredCellType) and the published cell type labels (TrueCellType).')
+supTab_CTmislabels_cap <- supTab_cap(name='supTab_CTmislabels', caption =  'Counts of cell type labels with our xgboost machine learning system (PredCellType) and the published cell type labels (TrueCellType).
+Ratio is calculated as CellType that were labelled as CellType_predict.
+Ratio < 0.05 were filtered out from view in the table.
+')
 
@@
 #source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
-fontsize(predictions %>%  
-           mutate(max_pred_prob = rowMaxs(.[,-(30:ncol(predictions))] %>% as.matrix )) %>% 
-           select(-CellTypeID) %>% 
-           rename(PredCellType = CellType) %>% 
-           inner_join(umapRef %>% select(Barcode, TrueCellType = CellType)) %>% 
-           filter(!is.na(TrueCellType)) %>% 
-           mutate(pred_correct  = ifelse(PredCellType == TrueCellType, 'Correct', 'incorrect')) %>% 
-           group_by(PredCellType, TrueCellType) %>% 
+fontsize(meta_filter %>% filter(!is.na(CellType_predict), !is.na(CellType), is.na(TabulaMurisCellType_predict)) %>% 
+           mutate(pred_correct  = ifelse(CellType == CellType_predict, 'Correct', 'incorrect')) %>% 
+           group_by(CellType, CellType_predict) %>% 
            summarise(Count = n()) %>% 
@@
            filter(Ratio > 0.05) %>% 
-           filter(!grepl('Doubl|Periocu', TrueCellType)) %>% 
+           filter(!grepl('Doubl|Periocu', CellType)) %>% 
+           ungroup() %>% 
+           group_by(CellType, CellType_predict) %>% 
+           arrange(CellType, -Ratio) %>% 
            flextable() %>% 
@@
 ```{r supTab_AUCxgboost, fig.cap=supTab_AUCxgboost_cap, echo=FALSE, results='asis', message = FALSE, warning = FALSE}
-supTab_AUCxgboost_cap <- supTab_cap(name='supTab_AUCxgboost', caption =  'Area under the precision recall curve (AUC) for each cell type, split by study')
+supTab_AUCxgboost_cap <- supTab_cap(name='supTab_AUCxgboost', caption =  'Area under the precision recall curve (AUC) for each cell type, split by study.
+The "All" study is the AUC score across all cells within the cell type')
 
@@
 cat(paste0("\n*",supTab_AUCxgboost_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
+
+```{r supTab_topMarkers, fig.cap=supTab_topMarkers_cap, echo=FALSE, results='asis', message = FALSE, warning = FALSE}
+supTab_topMarkers_cap <- supTab_cap(name='supTab_topMarkers', caption =  'Top differentially expressed genes for each retinal cell type.
+C1 is the number of citations for the gene <-> cell type (e.g.
+Rho and Rods) and C2 is the number of citations for gene and retina (e.g.
+Rho and Retina).')
+
+cat("<div id=\"tbl:supTab_topMarkers\">")
+fontsize(top_markers %>% flextable() %>% width(width = 3), size = 8, part = 'all') %>% colformat_double(digits = 2)
+#Output Caption (* * for italics)
+cat(paste0("\n*",supTab_topMarkers_cap,"*\n"))
 # Output </div>
@@
 ```
+```{r eyeIntegration_screenshot, fig.width = 6, fig.height = 4}
+projection_fig_cap <- supFig_cap(name='projection_fig', caption =  'Screenshot of eyeIntegration bulk RNA-seq meta-analysis of vimentin expression in different RPE tissue sources')
+
+ggdraw() + draw_image("figs_and_tables/rpe_eyeintegration.png", scale = 1)
+```
 
 ```{r projection_fig, fig.width=12, fig.height=10,  fig.cap=projection_fig_cap, echo=FALSE, message = FALSE, warning = FALSE}
-projection_fig_cap <- supFig_cap(name='projection_fig', caption =  'a.
-RPE distribution colored by study demonstrates how the RPE are grouped in a similar location.
-iPSC-based RPE we provided are located within the tissue based RPE cells.
-b.
-Screen shot of Google colab notebook that demonstrates how to integrate external data into the scEiaD resource.
-We see in the screenshort how the dataset contains many of the retinal cell types.')
-input[['pt_size_back']] <- 0
-input[['pt_size_meta']] <- 6
-input[['meta_column']] <- 'study_accession'
-#input[['gene_and_meta_scatter_tech']] <- 'Droplet'
-input[['meta_column_transform']] <- 'None'
-input[['meta_filter_cat']] <- 'CellType_predict'
-input[['meta_filter_on']] <- c('RPE')
-rpe <- make_meta_scatter_umap_plot(input, mf, meta_filter,
-                                   celltype_predict_labels,
-                                   celltype_labels,
-                                   tabulamuris_predict_labels,
-                                   cluster_labels,
-                                   cat_to_color_df
-)
-#
+projection_fig_cap <- supFig_cap(name='projection_fig', caption =  'Screen shot of Google colab notebook that demonstrates how to integrate external data into the scEiaD resource.
+We see in the screenshot how the organoid dataset contains many of the retinal cell types.')
 
-colab_p <- ggdraw() + draw_image("figs_and_tables/colab_screenshot.png", scale = 0.7)
-
-plot_grid(rpe$plot , colab_p)
+ggdraw() + draw_image("figs_and_tables/colab_screenshot2.png", scale = 0.7)
 
@@
 
+```{r supFigShekharSMART, fig.width=3, fig.height=3,  fig.cap=supFigShekharSMART_cap, echo=FALSE, message = FALSE, warning = FALSE}
+supFigShekharSMART_cap <- supFig_cap(name='figShekharSMART', caption =  'Our xgboost ML properly labels this RBC enriched population as enriched in RBC')
+screenshot_paper <- ggdraw() + draw_image("figs_and_tables/shekhar_screenshot.png", scale = 1.2)
+
+shekhar_counts <- meta_filter %>% 
+  filter(study_accession == 'SRP073242', !is.na(CellType_predict)) %>% 
+  group_by(CellType_predict) %>% summarise(Count = n()) %>% 
+  ggplot(aes(x=CellType_predict, y = Count, label = Count)) + geom_bar(stat = 'identity') + 
+  geom_text(color = 'gray', size = 5) + 
+  cowplot::theme_cowplot(margin(t = 5, r = 5, b = 5, l = 5, unit = "pt")) + 
+  coord_flip()
+cowplot::plot_grid(screenshot_paper, shekhar_counts, labels = c('a','b'), ncol =1)
+```
+
 # Works Cited

### 2021-03-22 figs_and_tables/dotplot.R

@@ -31,7 +31,7 @@ make_dotplot <- function(input, db, meta_filter, cat_to_color_df){
   dotplot_data <- dotplot_data %>%
     filter(Gene %in% gene) %>%
     group_by_at(vars(one_of(c('Gene', grouping_features)))) %>%
-    summarise(cpm = sum(cpm * cell_exp_ct) / sum(cell_exp_ct),
+    summarise(counts = sum(counts * cell_exp_ct) / sum(cell_exp_ct),
               cell_exp_ct = sum(cell_exp_ct, na.rm = TRUE)) %>%
     collect() %>%
     tidyr::drop_na() %>%
@@ -40,7 +40,7 @@ make_dotplot <- function(input, db, meta_filter, cat_to_color_df){
                 summarise(Count = n())) %>%
     mutate(cell_exp_ct = ifelse(is.na(cell_exp_ct), 0, cell_exp_ct)) %>%
     mutate(`%` = round((cell_exp_ct / Count) * 100, 2),
-           Expression = cpm * (`%` / 100)) %>%
+           Expression = counts * (`%` / 100)) %>%
     select_at(vars(one_of(c('Gene', grouping_features, 'cell_exp_ct', 'Count', '%', 'Expression')))) %>%
     filter(!is.na(Gene))
   #meanE <- dotplot_data %>% group_by(Gene) %>% summarise(meanE = mean(Expression))

### 2021-03-22 figs_and_tables/fig1.R

@@ -6,16 +6,21 @@ library(cowplot)
 library(patchwork)
 a <- meta_filter %>% 
   mutate(Platform = gsub('C1', 'SMARTSeq_v4', Platform)) %>%
+  mutate(organism = case_when(grepl('Homo', organism) ~ 'HS',
+                              grepl('Mus', organism) ~ 'MM',
+                              TRUE ~ 'MF')) %>% 
   group_by(organism, Platform) %>% 
   summarise(accession = length(unique(study_accession)), paper = length(unique(PMID)), batch = length(unique(batch))) %>% 
   pivot_longer(cols = accession:batch) %>% 
+  filter(name %in% c('batch','paper')) %>% 
   mutate(Count = value,
          Type = name,
          organism = gsub(' ','\n',organism)) %>%  
-  ggplot(aes(x=Platform, y=Count, fill = Type, group = Count)) + 
-  facet_wrap(~organism) + 
+  ggplot(aes(x=Platform, y=Count, fill = organism, group = Count, label = Count)) + 
+  facet_wrap(~Type) +
   geom_bar(stat = 'identity', position = position_dodge2()) + 
   cowplot::theme_cowplot() + 
+  geom_text(position = position_dodge(width = 1)) + 
   coord_flip()
 
 dagified <- dagify('counts' ~ 'fastq',

### 2021-03-22 figs_and_tables/fig2__integrationPerf.R

@@ -1,7 +1,21 @@
+# celltype prediction accuracy
+load('~/data/scEiaD/merged_xgboost_2021-03-17.Rdata')
+
+accuracy <- accuracy %>% 
+  filter(score != 0) %>% 
+  mutate(
+    norm = str_extract(file, 'transform-[a-zA-Z]+') %>% gsub('transform-','',.),
+    nf = str_extract(file, 'n_features-\\d+') %>% gsub('n_features-', '', .) %>% as.numeric(),
+    dims = str_extract(file, 'dims-\\d+') %>% gsub('dims-', '', .) %>% as.numeric(),
+    method = str_extract(file, 'method-[a-zA-Z]+') %>% gsub('method-','',.),
+    knn = str_extract(file, 'knn-\\d+\\.\\d+|knn-\\d+') %>% gsub('knn-', '', .) %>% as.numeric(),
+    epochs = str_extract(file, 'epochs-\\d+') %>% gsub('epochs-','',.) %>% as.numeric())
+
+
 load('~/data/scEiaD//metrics_2020_11_28.Rdata')
 perfO <- perf
 
-load('~/data/scEiaD//merged_stats_2021-02-05.Rdata')
+load('~/data/scEiaD//merged_stats_2021-03-17.Rdata')
 
 perf <- perf %>% 
   mutate(knn = case_when(method == 'scArches' ~ 7, TRUE ~ knn))
@@ -22,7 +36,7 @@ perf_tabula <- perfO %>% unique() %>% filter(set %in% c('TabulaDroplet')) %>%
   select(Score, Group, Value, set, dims:normalization) %>% 
   filter(Score %in% c('LISI','Silhouette', 'ARI', 'PCR', 'NMI')) 
 
-perf_scVI <- perf %>% unique() %>% filter(set %in% c('TabulaDroplet', 'universe')) %>% 
+perf_scVI <- perf %>% unique() %>% filter(set %in% c('universe')) %>% 
   select(Score, Group, Value, set, dims:normalization) %>% 
   filter(Score %in% c('LISI','Silhouette', 'ARI', 'PCR', 'NMI')) 
 
@@ -98,22 +112,25 @@ zscore_sum_all_methods <- zscore_tabula  %>%
   scale_alpha_continuous(range = c(0.6, 1), breaks =c(8,30))
 
 
-zscore_droplet_scVI_optimize <- perf_scVI%>% 
[…]

### 2021-03-22 figs_and_tables/make_meta_scatter_umap_plot.R

@@ -17,9 +17,17 @@ celltype_predict_labels <-scEiaD_2020_v01 %>% tbl('celltype_predict_labels') %>%
   mutate(CellType_predict = case_when(CellType_predict == 'RPC' ~ 'RPCs',
                                       CellType_predict == 'Mesenchymal/RPE/Endothelial' ~ 'Endothelial',
                                       TRUE ~ CellType_predict)) %>% collect() %>% 
-  mutate(UMAP_1 = UMAP_1 * 1, UMAP_2 = UMAP_2 * -1) 
-celltype_labels <-scEiaD_2020_v01 %>% tbl('celltype_labels') %>% collect() %>% mutate(UMAP_1 = UMAP_1 * 1, UMAP_2 = UMAP_2 * -1) 
-cluster_labels <-scEiaD_2020_v01 %>% tbl('cluster_labels') %>% collect() %>% mutate(UMAP_1 = UMAP_1 * 1, UMAP_2 = UMAP_2 * -1) 
+  mutate( UMAP_a = UMAP_2 * x_dir,
+          UMAP_b = UMAP_1 * y_dir) %>%
+  mutate(UMAP_1 = UMAP_a, UMAP_2 = UMAP_b)
+celltype_labels <-scEiaD_2020_v01 %>% tbl('celltype_labels') %>% collect() %>% 
+  mutate(  UMAP_a = UMAP_2 * x_dir,
+           UMAP_b = UMAP_1 * y_dir) %>%
+  mutate(UMAP_1 = UMAP_a, UMAP_2 = UMAP_b)
+cluster_labels <-scEiaD_2020_v01 %>% tbl('cluster_labels') %>% collect() %>% 
+  mutate(  UMAP_a = UMAP_2 * x_dir,
+           UMAP_b = UMAP_1 * y_dir) %>%
+  mutate(UMAP_1 = UMAP_a, UMAP_2 = UMAP_b)
 mf <- meta_filter 
 
 # generate color_mappings
@@ -31,9 +39,9 @@ categorical_columns <- c("Phase","batch","study_accession","library_layout","org
 meta_filter <- meta_filter %>% mutate(SubCellType = tidyr::replace_na(SubCellType, 'None'),
                                       subcluster = as.character(subcluster)) 
 
-  
+
 map_color <- function(column, meta_filter){
- # master_colorlist <- c(pals::polychrome()[3:length(pals::polychrome())], pals::alphabet2())
[…]

### 2021-03-22 figs_and_tables/sup_fig1__qc_failure.R

@@ -41,7 +41,7 @@ supFig1_plot <- qc %>%
   geom_stratum(aes(fill = `OrgTech`), alpha = 0.6) +
   geom_stratum(aes(fill = `QC`), alpha = 0.6) + 
   ggrepel::geom_text_repel(stat = "stratum", segment.color = NA,
-                           nudge_x = c(rep(c(-20),5), rep(c(20), 5)),
+                           nudge_x = c(rep(c(-20),5), rep(c(20), 4)),
                            direction = "y", size = 4,
                            aes(label = after_stat(stratum))) + theme_void() +
   shadowtext::geom_shadowtext(stat = "stratum", label=counts, bg.color = 'white', color = 'black') +

### 2021-03-22 figs_and_tables/sup_fig2__celltype_xgboost_labelling.R

@@ -12,13 +12,24 @@ ct_processing <-  meta_filter %>%
                                       is.na(CellType_predict) ~ 'Unlabelled', 
                                       TRUE ~ CellType_predict)) 
 
+
+ct_processing_NF <-  meta_filter %>% 
+  mutate(CellType = case_when(is.na(CellType) ~ 'Unlabelled', 
+                              CellType == 'AC/HC_Precurs' ~ 'AC/HC Precursors',
+                              TRUE ~ CellType)) %>% 
+  mutate(CellType_predict = case_when(CellType_predict == 'RPC' ~ 'RPCs',
+                                      CellType_predict == 'AC/HC_Precurs' ~ 'AC/HC Precursors',
+                                      grepl('Mesenchymal', CellType_predict) ~ 'Endothelial',
+                                      is.na(CellType_predict) ~ 'Unlabelled', 
+                                      TRUE ~ CellType_predict)) 
+
 ##########
 # Cell Counts
-predictedCT <- ct_processing %>% 
+predictedCT <- ct_processing_NF %>% 
   group_by(organism, CellType_predict) %>% 
   summarise(`Published Count` = n()) %>% pivot_wider(values_from = `Published Count`, names_from = c(organism))
 colnames(predictedCT) <- c('CellType','HS Transferred','MF Transferred','MM Transferred')
-prelabelledCT <- ct_processing %>% 
+prelabelledCT <- ct_processing_NF %>% 
   group_by(organism, CellType) %>% 
   summarise(`Published Count` = n()) %>% pivot_wider(values_from = `Published Count`, names_from = c(organism))
 colnames(prelabelledCT) <- c('CellType','HS Published','MF Published','MM Published')
@@ -29,13 +40,13 @@ ctTable <- joinedCT %>% flextable()
 
 ##########
 # Study Counts
-predictedSA <- ct_processing %>% 
+predictedSA <- ct_processing_NF %>% 
   filter(!is.na(CellType_predict)) %>% 
   select(CellType_predict, organism, study_accession) %>% unique() %>% 
   group_by(organism, CellType_predict) %>% 
   summarise(`Study Count` = n()) %>% pivot_wider(values_from = `Study Count`, names_from = c(organism))
 colnames(predictedSA) <- c('CellType','HS Studies (transferred)','MF Studies (transferred)','MM Studies (transferred)')
-prelabelledSA <- ct_processing %>% 
[…]

### 2021-03-22 figs_and_tables/sup_fig__xgboost_PR.R

@@ -13,17 +13,17 @@ pr_calculator <- function(test_predictions, partition){
       left_join(umapRef %>% select(Barcode, study_accession), by = 'Barcode') %>% 
       filter(study_accession == partition)
   } 
-  cell_type2id <- test_predictions %>% select(cell_type_id, CellType) %>% distinct()
+  cell_type2id <- test_predictions %>% select(CellTypeID, CellType) %>% distinct()
   target_cell_types <-unique(cell_type2id$CellType)
   pr_df <- test_predictions %>% 
-    select(-cell_type_id, -CellType) %>% 
-    rename(cell_type_id = true_cell_id) %>% 
+    select(-CellTypeID, -CellType) %>% 
+    rename(CellTypeID = true_label_id) %>% 
     inner_join(cell_type2id) %>% 
     filter(CellType%in% target_cell_types) %>% 
-    select(all_of(target_cell_types), cell_type_id, CellType) 
+    select(all_of(target_cell_types), CellTypeID, CellType) 
   # test_predictions <- predictionsC %>%  
   #   mutate(max_pred_prob = rowMaxs(.[,-(31:ncol(predictionsC))] %>% as.matrix )) %>% 
-  #   select(-cell_type_id) %>% 
+  #   select(-CellTypeID) %>% 
   #   rename(PredCellType = CellType) %>% 
   #   inner_join(umapRef %>% select(Barcode, TrueCellType = CellType)) %>% 
   #   filter(!is.na(TrueCellType)) %>% 
@@ -32,7 +32,7 @@ pr_calculator <- function(test_predictions, partition){
   # 
   # sum(test_predictions$pred_correct == 'Correct') / nrow(test_predictions) 
   # 
-  # cell_type2id <- predictions %>% select(cell_type_id, CellType) %>% distinct %>% filter(CellType!='None')
+  # cell_type2id <- predictions %>% select(CellTypeID, CellType) %>% distinct %>% filter(CellType!='None')
   
   # pr_df <- test_predictions %>% 
   #   select(all_of(target_cell_types), PredCellType, TrueCellType) 
@@ -72,8 +72,7 @@ xgboost_pr_table_full <- pr_list %>% bind_rows() %>%
   select(AUC, `Cell Type`, Study) %>%
   unique() %>%
   arrange(-AUC) 
-  
-  
+ 
 # test <- pr_calculator(predictions, 'SRP157927')
[…]

### 2021-03-22 src/make_dotplot_markers.R

@@ -2,9 +2,9 @@ library(pool)
 library(RSQLite)
 library(tidyverse)
 library(glue)
-scEiaD_2020_v01 <- dbPool(drv = SQLite(), dbname = "~/data/scEiaD/2021_02_05_MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-0.6.sqlite", idleTimeout = 3600000)
+scEiaD_2020_v01 <- dbPool(drv = SQLite(), dbname = "~/data/scEiaD/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite", idleTimeout = 3600000)
 
-meta_filter <- fst::read_fst('~/git/plaeApp/inst/app/www/meta_filter.fst') %>% as_tibble()
+meta_filter <- fst::read_fst('~/data/scEiaD/2021_03_17_meta_filter.fst') %>% as_tibble()
 meta_filter <- meta_filter %>% mutate(CellType_predict = case_when(CellType_predict == 'Photoreceptor Precursors' ~ 'PR Precursors',
                                                                    CellType_predict == 'AC/HC_Precurs' ~ 'AC/HC Precursors',
                                                                    CellType_predict == 'RPC' ~ 'RPCs',
@@ -48,7 +48,7 @@ grouping_features <- 'CellType_predict'
 exp_stats <- scEiaD_2020_v01 %>% tbl('grouped_stats') %>%
   filter(Gene %in% genes) %>%
   group_by_at(vars(one_of(c('Gene', grouping_features)))) %>%
-  summarise(cpm = sum(cpm * cell_exp_ct) / sum(cell_exp_ct),
+  summarise(counts = sum(counts * cell_exp_ct) / sum(cell_exp_ct),
             cell_exp_ct = sum(cell_exp_ct, na.rm = TRUE)) %>%
   collect() %>%
   tidyr::drop_na() %>%
@@ -57,7 +57,7 @@ exp_stats <- scEiaD_2020_v01 %>% tbl('grouped_stats') %>%
               summarise(Count = n())) %>%
   mutate(cell_exp_ct = ifelse(is.na(cell_exp_ct), 0, cell_exp_ct)) %>%
   mutate(`%` = round((cell_exp_ct / Count) * 100, 2),
-         Expression = round(cpm * (`%` / 100), 2)) %>%
[…]

### 2021-03-22 manuscript_00.Rmd
vinay fixes for PR AUC

@@
 load('~/data/scEiaD/predictions_n_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__preFilter.scEiaD__dist-0.1__neighbors-50.Rdata')
-# test_predictions <- data.table::fread('~/data/scEiaD/390498_CTpredictortest_data_probabilities.csv.gz') %>% as_tibble %>% 
-#   mutate(max_pred_prob = rowMaxs(.[,-(33:36)] %>% as.matrix ), 
-#          pred_correct = ifelse(cell_type_id == true_cell_id, 'Correct', 'incorrect'))
+
 test_predictions <- model_out$test_dat %>% as_tibble %>% 

### 2021-03-22 figs_and_tables/sup_fig__xgboost_PR.R

@@ -49,6 +49,17 @@ pr_calculator <- function(test_predictions, partition){
   out
 }
 
+label2id <- predictions %>% select(CellType, CellTypeID) %>% distinct
+test_predictions <- predictions %>% 
+  inner_join(umapRef %>% select(Barcode, true_cell_label = CellType)) %>% 
+  filter(!is.na(true_cell_label), 
+         true_cell_label != 'None', 
+         CellType != 'None') %>% 
+  inner_join(label2id %>% select(true_cell_label = CellType, true_label_id = CellTypeID))
+
+
+
+
 pr_list <- list()
 for (i in c('All', umap$study_accession %>% unique())){
   #print(i)

### 2021-03-25 manuscript_00.Rmd
text updates

@@
 ---
-title: 'Building the Million Transcriptome Ocular Meta-Atlas'
+title: 'Building the Meta Single Cell Transcriptome Ocular Meta-Atlas'
 author:
@@
 csl: investigative-ophthalmology-and-visual-science.csl
-abstract: "The development of highly scalable single cell transcriptome technology has allowed for a high resolution view into huge numbers of single cell transcriptomes.
-Analyzing the transcriptomes between different projects is highly desirable as this would theoretically allow for better assessment of which effects are consistent across independent studies.
+abstract: "The development of highly scalable single cell transcriptome technology has resulted in the creation of thousands of datasets, over 30 in the retina alone.
+Analyzing the transcriptomes between different projects is highly desirable as this would allow for better assessment of which biological effects are consistent across independent studies.
 However it is difficult to compare and contrast data across different projects as there are substantial batch effects from computational processing, single cell technology utilized, and the natural biological variation.
-While many single cell transcriptome specific batch correction methods purport to remove the technical noise it is difficult to ascertain which method works best.
-We developed an R package (scPOP) that can scan the results of multiple methods and key parameter choices and create a ranked list of the best performing method / parameter combinations.
-We use this package along with a Snakefile based workflow system to demonstrate how to optimally merge 750,000 cells from 32 datsets and three species to create a massive ocular single cell transcriptome meta-atlas.
+While many single cell transcriptome specific batch correction methods purport to remove the technical noise it is difficult to ascertain which method functions works best.
+We developed a lightweight R package (scPOP) that brings in batch integration methods and uses a simple heuristic to balance batch merging and celltype/cluster purity.
+We use this package along with a Snakefile based workflow system to demonstrate how to optimally merge 766,615 cells from 34 retina datsets and three species to create a massive ocular single cell transcriptome meta-atlas.
 This provides a model how to efficiently create meta-atlases for tissues and cells of interest."
@@
 library(matrixStats)
+library(velociraptor)
+library(SingleCellExperiment)
 # setup caption-ing
@@
 
-qc <- readr::read_tsv('~/data/scEiaD/QC.tsv.gz')
-
+srt <- data.table::fread('~/git/scEiaD/data/sample_run_layout_organism_tech.tsv')
+qc <- data.table::fread('~/data/scEiaD/QC.tsv.gz')
+qc <- qc %>% left_join(srt %>% select(sample_accession, SX = Source) %>% unique(), by = 'sample_accession') %>% filter(SX %in% c('iPSC','Tissue'))
 # cell type prediction
@@
           file = 'supplemental_file__kallisto_stats.tsv')
+
+# write top markers to sup file
+write_tsv(top_markers, file = 'supplemental_file__celltype_markers.tsv')
+
+# load mini human velocity
+load('~/data/scEiaD/velo_scEiaD_2021-03-18_basemodel_homosapiens_tiny.Rdata')
+umap_mat <- reducedDim(velo_sceX, 'X_scviUMAP')
+grid.df <- gridVectors(umap_mat, reducedDim(velo_sceX, 'velocity_scviUMAP'))
+
+# load ML species transfer data
+# created with src/organism_celltype_ML.R
+load('~/data/scEiaD/species_transfer_celltype_model.Rdata')
 ```
@@
 
-## The retina, single cell gene analysis, and a recent explosion
+## A plethora of single-cell transcriptome studies in the retina
 
@@
 The  recent introduction of lower cost and high throughput single cell sequencing technology has led to an explosion of research across many fields.
-As of January 2021, over 40 million cells have been sequenced across over 1,200 studies and the average size of each study past 2020 is over 100,000 cells [@svenssonCuratedDatabaseReveals2020].
-The retina was used as the source tissue in one of the earliest works in the high throughput single cell transcriptomics field [@macoskoHighlyParallelGenomewide2015] and as of late 2020, over a dozen published studies containing over a million cells have used single cell technology to profile cell type specific gene expression patterns, cell fate trajectory, tissue and cell differentiation, and disease perturbation across multiple mammalian species [@buenaventuraIdentificationGenesEnriched2019; @clarkComprehensiveAnalysisRetinal2018; @collinCRXExpressionPluripotent2019; @collinDeconstructingRetinalOrganoids2019; @cowanCellTypesHuman2020; @diacouSix3Six6Are2018; @eldredThyroidHormoneSignaling2018; @hengHypoxiaToleranceNorrindeficient2019; @huDissectingTranscriptomeLandscape2019; @logiudiceSinglecellTranscriptionalLogic2019; @lukowskiSinglecellTranscriptomeAtlas2019; @luSingleCellAnalysisHuman2020;  @menonSinglecellTranscriptomicAtlas2019; @okorenMicroglialFunctionDistinct2019; @pengMolecularClassificationComparative2019; @phillipsNovelApproachSingle2018; @rheaumeSingleCellTranscriptome2018; @ruedaHippoPathwayBlocks2019; @shekharComprehensiveClassificationRetinal2016; @sridharSingleCellTranscriptomicComparison2020; @tranSingleCellProfilesRetinal2019; @voigtMolecularCharacterizationFoveal2019; @voigtSinglecellTranscriptomicsHuman2019; @yanCellAtlasHuman2020].
+As of early 2021, over 40 million cells have been sequenced across over 1,200 studies and the average size of each study starting in 2020 is over 100,000 cells [@svenssonCuratedDatabaseReveals2020].
+The retina was used as the source tissue in one of the earliest works in the high throughput single cell transcriptomics field [@macoskoHighlyParallelGenomewide2015].
+As of late 2020, over a twenty published studies, cumulatively containing over a million cells, have used single cell technology to profile cell type specific gene expression patterns, cell fate trajectory, tissue and cell differentiation, and disease perturbation across multiple mammalian species [@buenaventuraIdentificationGenesEnriched2019; @clarkComprehensiveAnalysisRetinal2018; @collinCRXExpressionPluripotent2019; @collinDeconstructingRetinalOrganoids2019; @cowanCellTypesHuman2020; @diacouSix3Six6Are2018; @eldredThyroidHormoneSignaling2018; @hengHypoxiaToleranceNorrindeficient2019; @huDissectingTranscriptomeLandscape2019; @logiudiceSinglecellTranscriptionalLogic2019; @lukowskiSinglecellTranscriptomeAtlas2019; @luSingleCellAnalysisHuman2020;  @menonSinglecellTranscriptomicAtlas2019; @okorenMicroglialFunctionDistinct2019; @pengMolecularClassificationComparative2019; @phillipsNovelApproachSingle2018; @rheaumeSingleCellTranscriptome2018; @ruedaHippoPathwayBlocks2019; @shekharComprehensiveClassificationRetinal2016; @sridharSingleCellTranscriptomicComparison2020; @tranSingleCellProfilesRetinal2019; @voigtMolecularCharacterizationFoveal2019; @voigtSinglecellTranscriptomicsHuman2019; @yanCellAtlasHuman2020].
 
@@
 
-While the gene - cell count tables are often made available in repositories like the Gene Expression Omnibus (GEO), there are no requirements to uniformly process the data.
+While the gene - cell count tables are generally made available in repositories like the Gene Expression Omnibus (GEO), there are no requirements to uniformly process the data.
 This means the count tables cannot be used in cross-study comparisons as even small differences in the computational pipeline (aligner, transcriptome reference, etc.) create study-specific effects.
-This issue can be addressed only by re-quantifying the data in a uniform environment.
-Fortunately, due to the continued development of computationally light-weight gene quantification tools into the single-cell space (e.g kallisto bustools, alevin-fry), re-quantification does not require massive compute resources [@melstedModularEfficientPreprocessing2019; @srivastavaAlevinEfficientlyEstimates2019].
+This issue can be addressed only by re-quantifying the data in a uniform computational environment.
+Fortunately, due to the continued development of computationally light-weight gene quantification tools in the single-cell space (e.g kallisto bustools, alevin-fry), re-quantification does not require massive compute and time resources [@melstedModularEfficientPreprocessing2019; @srivastavaAlevinEfficientlyEstimates2019].
 
@@
 
-We propose that by re-processing publicly available raw single cell transcriptome data in a consistent bioinformatic framework and optimally using batch correction tools we can create a meta-atlas.
+We propose that by re-processing publicly available raw single cell transcriptome data in a consistent bioinformatic framework and optimally using batch correction tools we can create a meta-atlas of retina single cell transcriptomes.
 As there are thousands of possible permutations of single cell tools, references, and parameter choices, we create our meta-atlas (which we refer to as the single cell Eye in a Disk or scEiaD) by benchmarking integration outcomes across multiple important single cell RNA-seq processing parameters (batch removal method, number of hyper-variable genes (HVGs), clustering resolution, etc.).
@@
 
-Gene-level counts were quantified with the kallisto bustool pseudo-aligner for both the droplet and well based samples.
+Gene-level counts were quantified with the kallisto bustools pseudo-aligner for both the droplet and well based samples.
 After empty droplet removal, we had `r format(qc %>% nrow() %>% as.integer() , big.mark=",", scientific=FALSE)` cells.
-We then created a Seurat v3 object and removed cells which had more than 10% mitochondrial reads across all gene counts or fewer than 250 unique genes quantified.
-After these standard quality control steps we were left with `r format(umapRef %>% nrow, big.mark = ",", scientific=FALSE)` cells.
+We then removed cells which had more than 10% mitochondrial reads across all gene counts or fewer than 200 unique genes quantified.
+After these standard quality control steps we were left with `r format(umapRef %>% nrow, big.mark = ",", scientific=FALSE)` cells (`r supFig_cap(name = 'supFig1', display = 'cite')`).
 
@@
 
-To visualize the interplay between batch mixing and cell type distinction we plot the batch mixing LISI score (which has been multiplied by -1) on plotted on the y-axis (higher is better) against the cluster LISI on the x-axis (higher is better).
+To visualize the interplay between batch mixing and cell type distinction we plot the batch mixing LISI score (which has been multiplied by -1) on the y-axis (higher is better) against the cluster LISI on the x-axis (higher is better).
 The best performer on both metrics will be in the top right corner  (`r supFig_cap(name = 'supFig_benchmark', display = 'cite')`a).
 In the same manner we plot the silhouette metric (`r supFig_cap(name = 'supFig_benchmark', display = 'cite')` b).
+To merge the different scores we define $sumZScale = \sum scale(m)$ where m is a metric (LISI by batch, LISI by cluster, LISI by cell type, silhouette by batch, silhouette by cluster, silhouette by celltype, NMI, and ARI).
 
-<!-- ## scPOP reveals different priorities across the different batch integration methods -->
-
-On one extreme we have combat, which merges together different batches very well, but also mixes together the distinct cell types (`r fig_cap(name = 'figBenchmark', display = 'cite')`a).
-The other extreme is not using any batch integration method, where you see very distinct groups of cells, but also each nearly study is has a distinct space in the 2D (`r fig_cap(name = 'figBenchmark', display = 'cite')`b).
-In our scEiaD dataset we see that like combat, Harmony and CCA are weighed more towards batch mixing then cluster and cell type purity.
-Scanorama, fastMNN, and bbknn more prioritize cleanly separating the clusters.
+On one extreme we have ComBat, which merges together different batches very well, but also mixes together the distinct cell types (`r fig_cap(name = 'figBenchmark', display = 'cite')`a).
+The other extreme is not using any batch integration method, where you see very distinct groups of cells, but also each nearly study is has a distinct region in the UMAP (`r fig_cap(name = 'figBenchmark', display = 'cite')`b).
+In our scEiaD dataset we see that like ComBat, Harmony and CCA are weighed more towards batch mixing then cluster and cell type purity.
+Scanorama, fastMNN, and bbknn prioritize cleanly separating the clusters.
 With our scEiaD meta-atlas insct, trVAE, and magic do not perform particularly well in batch mixing or cluster purity.
+Across the 11 integration methods we tested, scVI achieved the highest sumZScale for out scEiaD meta-atlas.
 
@@
 
-To merge the different scores we define $sumZScale = \sum scale(m)$ where m is a metric (LISI by batch, LISI by cluster, LISI by cell type, silhouette by batch, silhouette by cluster, silhouette by celltype, NMI, and ARI).
-In this scaled metric that weighs the different benchmarking algorithms evenly, scVI performs best.
 As expected the libSize normalization which omits the log or square root scaling generally performs the worst (`r fig_cap(name = 'figBenchmark', display = 'cite')`c).
-We see that the remaining normalization approaches alters the batch correction performance, though the exact outcome differs across the different methods.
+We see that the remaining normalization techniques alter the batch correction performance, though the outcome differs across the different methods.
 We also see that changing the number of latent dimensions (8 or 30) can occasionally dramatically change performance.
@@
 d.
-Optimal parameters for scVI methods across 1000 to 10000 HVGs and 4 to 50 latent dimensions.')
+Boxplot of 4 different clustering resolutions for across 1000 to 10000 HVG numbers and 4 to 50 scVI latent dimensions.
+Open boxes are using scVI-standard and gray boxes are scVI-projection (human reference with the remaining data projected) ')
 
@@
 
-To find the best set of parameters for the creation of scEiaD we did a grid search across key parameters: HVG, latent dimensions, and k nearest neighbors.
-Furthermore we used a recent advance in scVI capability (>= version 0.8.0) adapted from scArches that allows one to build a reference model and project the new cells onto it [@QueryReferenceSinglecell].
-We built a human-only scVI model and then projected the mouse and macaque data onto it.
-We then compared the "scVIprojected" against the against the straight scVI across all the previously mentioned parameters.
+To find the best set of parameters for scVI in out dataset we did a grid search across key parameters: HVG, latent dimensions, and k nearest neighbors.
+Furthermore we used a recent advance in scVI capability (>= version 0.8.0) adapted from scArches that allows one to build a reference model and query or project the new cells onto it [@QueryReferenceSinglecell].
+We refer to this projectable model as “scVI-projection", and the previous scVI model as “scVI-standard”.
+We built a scVI-projected model trained on human cells and then projected the mouse and macaque data onto it.
+We then compared scVI-projection against the against the previous scVI-standard across all the previously mentioned parameters.
 Using scPOP we first saw that the scVIprojection approach generally performed better than running scVI with all of the data (`r fig_cap(name = 'figBenchmark', display = 'cite')`d).
@@
 study where they used SMART-seq on a retinal bipolar cell enriched population of cells [@shekharComprehensiveClassificationRetinal2016] .
-Even though our ML algorithm was trained only on droplet-based data, our algorithm labels most of this dataset as retinal bipolar cells, with the next most common cell types being Amacrine,  like Shekhkar et al.
-labeled them (`r fig_cap(name = 'supFigShekharSMART', display = 'cite')`).
+Even though our ML algorithm was trained only on droplet-based data, our algorithm labels most of this dataset as retinal bipolar cells, with the next most common cell types being amacrine.
+The same result was found by Shekhkar et al.
+(`r supFig_cap(name = 'supFigShekharSMART', display = 'cite')`).
 To more generally evaluate performance we use the precision recall (PR) curve, which visualizes the ability of the model to precisely label known cells at a given confidence.
 The area under the PR curve (AUC) summarizes the effectiveness of the model across different cell types, with 1 being the highest performance.
-The xgboost model can predict rods, bipolar cells, and Müller glia with near perfect performance (`r fig_cap(name = 'supFig_PR', display = 'cite')`).
-Most of the remaining cell types can be predicted with an AUC over 0.95 `r supTab_cap(name = 'supTab_AUCxgboost', display = 'cite')`.
-The cell types that were the most challenging for the model to predict were the artery, choriocapillaris, and vein.
+The xgboost model can predict rods, bipolar cells, and Müller glia with near perfect performance (`r supFig_cap(name = 'supFig_PR', display = 'cite')`).
+Most of the remaining cell types can be predicted with an AUC over 0.9 `r supTab_cap(name = 'supTab_AUCxgboost', display = 'cite')`.
+Several of the precursor cell types (photoreceptor, AC/HC, and neurogenic) were labelled with lower confidence `r supTab_cap(name = 'supTab_CTmislabels', display = 'cite')`.
+The next most common labels for these cell types were either other precursor cells (e.g.
+100 AC/HC Precursors were labeled as Neurogenic) or the adjacent terminal cell type (e.g.
+254 photoreceptor precursors were labeled as cones).
+Other cell types that were challenging for the model to predict were the artery, choriocapillaris, and vein.
 Artery, choriocapillaris, and vein are constructed from endothelial cells and we find that for all three of these, endothelial was the second most common label.
-The ML model also performs under 0.9 for predicting the neurogenic cell type label; however the most next common label that is predicts are the highly related retinal progenitor populations (RPCs) (`r supTab_cap(name = 'supTab_CTmislabels', display = 'cite')`).
 Overall, our xgboost based ML model shows strong accuracy across the major cell types of the retina (overall AUC of 0.98).
-
-<!-- ## Certain retina cell types have more stable community consensus IDs -->
-
-<!-- With the wide diversity of the scEiaD we can get a sense on how consistent the retina community is at labelling the major cell types of the retina.
-By calculating the AUC against each individual cell type and study, we can see which cell types are consistently labeled correctly across most (`r fig_cap(name = 'supFig_PR_cap', display = 'cite')`).
-Most studies show very similar performance with our ML model.
-The study with the lowest correspondence between published and predicted label is Macoksko et al; this study is notable as it was the first demonstration of the droplet based scRNA technique.
--->
 
@@
 figXGboost_umap_cap <- fig_cap(name='figXGboost_umap', caption =  'a.
-Top genes that are differentially expressed across the major cell types of the retina.
+Top genes that are differentially expressed across the major cell types of the retina (PR is short for Photoreceptor).
 Genes are colored by which cell type they are differentially expressed in.
@@
 b.
-2D UMAP projection of scEiaD, colored by cell type.
-The Tabula Muris data is colored all in gray.
+2D UMAP projection of scEiaD, colored by cell type (Tabula Muris data is gray).
+Arrows are scvelo RNA velocity.
+Longer arrows are cells with higher velocity (relatively more unspliced transcripts).
 c.
@@
                                    cluster_labels,
-                                   cat_to_color_df
+                                   cat_to_color_df,
+                                   velocity = TRUE
 )
@@
 After ML projection of cell type labels from the original `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` labels onto a total of `r meta_filter %>% filter(!is.na(CellType_predict) | !is.na(TabulaMurisCellType_predict)) %>% nrow()` cells we have substantially improved the number of studies per cell type.
-For example, we went from `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(2)` human studies with labeled Muller Glia to `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(5)` after labeling (`r supTab_cap(name = 'supTab_SAcounts', display = 'cite')`).
+For example, we went from `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(2)` human studies with labeled Müller Glia to `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(5)` after labeling (`r supTab_cap(name = 'supTab_SAcounts', display = 'cite')`).
 Overall we go from an average of `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(2) %>% mean() %>% round()` studies per human cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(5) %>% mean() %>% round()` and `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(4) %>% mean() %>% round()` studies per mouse cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(7) %>% mean() %>% round()` after transferring the cell type labels.
@@
 As another check on the quality of the cell type assignments, we ran the cell and cluster independent haystack gene search and pairwise differential expression tests between the predicted cell types (see methods for further details).
-We show the five most consistently differentially expressed genes for each of the major retina cell types (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a).
+We show the five most differentially expressed genes for each of the major retina cell types are consistent with known retinal cell markers (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a).
 As a simple metric to identify known and unknown genes relating to the cell type specific expression we search PubMed for the number of publications with two searches per gene.
@@
 "PDE6H AND Retina").
-Of the 50 genes in (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a), `r 50 - top_markers %>% filter(Gene %in% dplot_markers) %>% filter(c1 == 0) %>% nrow` had one or more citations in the gene - cell type search (`r supTab_cap(name = 'supTab_topMarkers', display = 'cite')`) and `r 50 - top_markers %>% filter(Gene %in% dplot_markers) %>% filter(c1 == 0, c2 == 0) %>% nrow` had one or more citation in the inclusive search.
+Of the 50 genes in (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a), `r 50 - top_markers %>% filter(Gene %in% dplot_markers) %>% filter(c1 == 0) %>% nrow` had one or more citations in the gene - cell type search (Supplementary File celltype_markers.tsv) and `r 50 - top_markers %>% filter(Gene %in% dplot_markers) %>% filter(c1 == 0, c2 == 0) %>% nrow` had one or more citation in the inclusive search.
 The 50 genes had a mean of `r top_markers %>% filter(Gene %in% dplot_markers) %>% pull(c2) %>% mean() %>% round()` studies (with the inclusive gene by "retina" search).
-In contrast, 100 randomly chosen genes had a mean of `r pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% pull(value) %>% mean() %>% round` (wilcox test p < 2.5 x 10^-12).
+In contrast, 100 randomly chosen genes had a mean of `r pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% pull(value) %>% mean() %>% round` (wilcox test p < 1.44 x 10^-17).
 
@@
 
-## The scVI-based scEiaD UMAP projection blends batches while separating cell types
+## The scVI-based scEiaD UMAP projection blends batches and species while separating cell types
 
@@
 We see the neurogenic and progenitor populations from which the retinal cell types are derived near the center of the UMAP visualization.
-The photoreceptor precursors are adjacent to the neurogenic population and flow into the rods and cones.
+The photoreceptor precursors are adjacent to the neurogenic population and, as demonstrated by the RNA velocity dynamics, flow into the rods and cones.
 The amacrine and horizontal precursors (AC/HC) likewise flow from the neurogenic center into the mature amacrine and horizontal cells.
@@
 The photoreceptors (cones and rods) of the retina which are responsible for color and low-light vision, respectively, are near each other in the UMAP space.
-The major remaining retina cell types, by proportion in the mammalian eye are the muller glia, which are a glial cell type which help support the neurons of the retina.
+The major remaining retina cell types, by proportion in the mammalian eye are the Müller Glia, which are a glial cell type which help support the neurons of the retina.
 Next we have the neural cell types which transmit and help interpret the signals from the photoreceptors before they leave the retina via the optic stalk: the amacrine cells, retinal ganglia, horizontal, and bipolar cells.
-All of these cells are in well defined spaces in the UMAP.
-Finally we see even across species that the major cell types still group near each other (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`d).
+All of these cells are in well separated spaces in the UMAP.
+Finally we see across species that the major cell types overlap each other (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`d).
 The macaque retina cells are not present in the precursor/neurogenic center of the UMAP as expected because only fully developed tissues were sampled in these data.
@@
 
+While by eye the UMAP 2D projection generally blends together the three different species (macaque, mouse, human) in the UMAP visualization, we more rigorously tested this by tweaking the inputs to our xgboost cell type predictor machine learning system.
+Briefly, we trained two new models: one with only human data and one with only mouse data.
+We then applied this model to the other two species to see whether the model trained on one species was generally transferable to another species.
+Again, both models (human only and mouse only) both proved to be highly accurate at predicting cell types in the other species with the terminal cell types (`r supFig_cap(name = 'supFigML_CT_transfer', display = 'cite')`.
 ## Projection of outside data onto scEiaD demonstrates similarities and differences of iPSC and organoids to primary cells
@@
 downloading our 13 megabyte scVI model, and 3.
-following the Jupyter notebook on Google colab that we provide as a live demo at http://colab.google.
+following the Jupyter notebook on Google colab that we provide as a live demo available from https://github.com/davemcg/scEiaD.
 We demonstrate the power of this approach in two ways.
@@
 In the Google colab notebook we show how a subset of the reads from Kallman et al.'s SRR12130660 can be processed from the raw reads to a UMAP visualization in under 10 minutes.
-Indeed we see how organoid-based retinal cell types can be detected overlapping primary cells for RPCs, photoreceptors, amacrine cells, and retinal ganglion cells (`r supFig_cap(name = 'projection_fig', display = 'cite')`d).
+Indeed we see how organoid-based retinal cell types can be detected overlapping primary cells for RPCs, photoreceptors, amacrine cells, and retinal ganglion cells (`r supFig_cap(name = 'projection_fig', display = 'cite')`).
 
 Finally, we demonstrate how the RPE derived from the Bharti iPSC differentiation process express canonical RPE markers of TTR and RPE65 but end up in a slightly different position in the UMAP projection`r fig_cap(name = 'RPE_fig', display = 'cite')`a).
-Differential expression analysis identified that vimentin is nearly exclusively expressed in the iPSC-based RPE `r fig_cap(name = 'RPE_fig', display = 'cite')`b).
+Differential expression analysis (see methods) identified that vimentin is nearly exclusively expressed in the iPSC-based RPE `r fig_cap(name = 'RPE_fig', display = 'cite')`b).
 This result is not surprising as Hunt et al.
 demonstrated that cultured and proliferating RPE express high levels of vimentin [@huntAlteredExpressionKeratin1990] and this same trend is seen in bulk RNA-seq from stem cell RPE and primary RPE `r supFig_cap(name = 'eyeIntegration_screenshot', display = 'cite')`) [@swamyEyeDiskEyeIntegration2019].
+
 ```{r RPE_fig, fig.width=14, fig.height=5,  fig.cap=RPE_fig_cap, echo=FALSE, message = FALSE, warning = FALSE}
-RPE_fig_cap <- fig_cap(name='projection_fig', caption =  'a.
+RPE_fig_cap <- fig_cap(name='RPE_fig', caption =  'a.
 RPE distribution colored by study demonstrates how the most RPE are in two locations.
-iPSC-based RPE we provided are located more enriched in cluster 14.
-Tissue RPE more enriched in cluster 46.
-b.
-Violin plot of 3 functional RPE markers (TTR, RPE65, TYRP1) and vimentin')
+iPSC-based RPE we provided are located more enriched in cluster 47.
+Tissue RPE more enriched in cluster 34 b.
+Violin plot of two functional RPE markers (TTR, RPE65) and vimentin (proliferating RPE marker)')
 input <- list()
@@
 The set of Snakemake pipelines that takes in the raw fastq sequence and outputs the scEiaD is at https://github.com/davemcg/scEiad [@kosterSnakemakeScalableBioinformatics2012].
-The publication commit is BLOOP.
-Furthermore, the repository has been deposited at Zenodo under BLOOP.
-We will briefly discuss the pipeline choices, programs and algorithms, and versions below.
+The publication commit is #9d3c5e1 Furthermore, the repository has been deposited at Zenodo under 10.5281/zenodo.4638117 We will briefly discuss the pipeline choices, programs and algorithms, and versions below.
 For the R packages, we provide package versions as the supplementary file "R_session_info.txt"
@@
 A single set of exonic transcript sequences was created for each species for all well techonologies.
-A kallisto quantification index was generated for each of these fastas using  Kallisto index (0.46.2) [@brayNearoptimalProbabilisticRNAseq2016].
+A kallisto quantification index was generated for each of these fastas using kallisto index (0.46.2) [@brayNearoptimalProbabilisticRNAseq2016].
 Well based samples were quantified using kallisto quant.
@@
 
-We tested several gene count normalization approaches as we were not certain which would produce an optimal outcome: standard (default Seurat, library size normalization, then log transform), sqrt (same, but with sqrt normalization), libSize (omit the log or sqrt normalization), scran, SCT from Seurat, and for scVI, no normalization (counts) [@lunStepbystepWorkflowLowlevel2016].
+The following steps (normalization through benchmarking are handled by the SnakePOP file) We tested several gene count normalization approaches as we were not certain which would produce an optimal outcome: standard (default Seurat, library size normalization, then log transform), sqrt (same, but with sqrt normalization), libSize (omit the log or sqrt normalization), scran, SCT from Seurat, and for scVI, no normalization (counts) [@lunStepbystepWorkflowLowlevel2016].
 Our R implementation of the normalization approaches as well as how we constructed the Seurat v3 object can be found in the supplementary file "make_seurat_obj_functions.R"
@@
 We also tried the DESC, liger, and Conos batch corrections methods but were unable to get them to work reliably so they were dropped.
-The batch correction step implementation can be found the supplementary file "merge_methods.R".
+The batch correction step implementation can be found the supplementary file "merge_methods.R" and in the github repo (https://github.com/davemcg/scEiaD/blob/master/src/merge_methods.R).
 
@@
 We also used the leiden algorithm as implemented by PARC with a resolution of 0.6 and 0.8 (higher results in more clusters) [@traagLouvainLeidenGuaranteeing2019; @stassenPARCUltrafastAccurate2020].
-These two resolutions were chosen as they roughly gave the same number of clusters at the Seurat Louvain-Jaccard aprroach with a knn 7.
+These two resolutions were chosen as they roughly gave the same number of clusters at the Seurat Louvain-Jaccard approach with a knn 7.
 
 The UMAP visualization was calculated with the Seurat "RunUMAP" using the uwot R package [@mcinnesUMAPUniformManifold2020].
-We altered the min.dist parameter with 0.001 and 0.1 and tried n.neighbors across 15, 30, 100, and 500.
+We tried min.dist parameters of 0.001, 0.1, and 0.3 and tried n.neighbors across 15, 30, 100, and 500.
 A smaller min.dist value gives "tighter" groupings while a higher number of n.neighbors uses a larger number of near cells to calculate the global positioning.
 
-## Benchmarking
+## Benchmarking and scPOP
 
 We wrote the scPOP R package to unify the LISI and Silhouette metrics from Harmony and kBet, respectively, along with NMI and ARI [@buttnerTestMetricAssessing2019; @korsunskyFastSensitiveAccurate2019].
-LISI and Silhouette require a dense matrix, which is a problem for our data as a `r meta_filter %>% nrow()` cell by 8 latent dimension dense matrix cannot fit in a 1.5 TB memory node.
+LISI and Silhouette require a dense matrix, which is a problem for our data as a `r meta_filter %>% nrow()` cell by 8 latent dimension dense matrix cannot fit in out largest available memory node (1.5 TB).
 We down-sampled the dataset to ~100,000 cells, taking care to keep all rarer cell types for the LISI and Silhouette benchmarking.
@@
 ## Multi-step doublet removal
-While we removed droplet-based cells with more than 3,000 transcripts as these are highly likely to be doublets (more than one cell per droplet) there were likely many more doublets.
-We ran DoubletDetect and scrublet and calculated the distribution of DoubletDetect and scrublet scores across all clusters and removed clusters with a score in both metrics greater than 4 standard deviations above the mean [@adamgayosoJonathanShorDoubletDetectionDoubletdetection2020; @wolockScrubletComputationalIdentification2019].
+To identify probable doublets (more than one cell in a droplet) we ran DoubletDetect and scrublet and calculated the distribution of DoubletDetect and scrublet scores across all clusters and removed clusters with a score in both metrics greater than 4 standard deviations above the mean [@adamgayosoJonathanShorDoubletDetectionDoubletdetection2020; @wolockScrubletComputationalIdentification2019].
 This removed another `r format((umapRef %>% nrow() %>% as.integer()) - (meta_filter %>% nrow() %>% as.integer()), big.mark=",", scientific=FALSE)` cells, leaving `r format((meta_filter %>% nrow()) %>% as.integer(), big.mark=",", scientific=FALSE)` in total.
@@
 In order to identify cell types for the `r format(meta_filter %>% filter(is.na(CellType), is.na(TabulaMurisCellType)) %>% nrow(), big.mark=",", scientific=FALSE)` unlabelledcells we designed a custom xgboost based cell type classifier.
-We took labeled data and split it into training and test sets, stratified by cell type.
+We took labeled data and split it into training (2/3) and test (1/3) sets, stratified by cell type.
 The input features used to train the model are the scVI latent dimensions, the total number of reads in each cell, the number genes detected in each cell, and the percent mitochondrial gene expression of each cell.
@@
 The trained model had an overall macro and micro AUC score of 0.98 and 0.99, respectively.
-The AUC scores ranged from.
 This model was then used to identify labels for each all cell.
-Unlabelled data was pre-processed identically to training data and fed into model to generate a vector of label probabilities for each cell.
+Unlabeled data was pre-processed identically to training data and fed into model to generate a vector of label probabilities for each cell.
 We selected the highest label probability for each cell, and required a minimum probability of 0.5 to assign a label to a cell.
+
+For the organism specific xgboost ML we followed the above procedure, except that we combined Early RPCs, Late RPCs, and RPCs into one category and did not attempt to predict the non-retina cell types (e.g.
+fibroblasts) as there were very few labelled cells across all three organisms.
 
@@
 ## Calculation of RNA velocity
-All RNA velocity calculations were done through the the velocriaptor implementation of the scVelo python library [@bergenGeneralizingRNAVelocity2020].
+
+RNA velocity calculations were with the velocriaptor wrapping of the scVelo python library [@bergenGeneralizingRNAVelocity2020].
 From the anndata objects generated by our Snakemake pipeline we calculated velocity across all genes.
@@
 Another limitation is that discrepancies between cell type labels between different labs makes certain transitioning cell type labels a bit imprecise.
-One example is how the rods and photoreceptor precursor labels overlap somewhat.
+One example is how the rods and photoreceptor precursor labels partially overlap.
 Though we attempted to ameliorate the issue by removing cell type labels in large disagreement with the consensus, some disagreements could propagate into our machine labeled cells type assignments.
+These issues may reflect labeling continuous processes with discrete labels.
 
@@
 Further downstream processing can largely be done a computers with 64+ Gb of memory and a few hundred GB of disk space.
-We believe that our efforts can be replicated in any other tissue / system with a large number of independent studies by an individual lab or small number of scientists.
+We believe that our efforts can be replicated in any other tissue / system with a large number of independent studies by a small number of computational scientists following our general approach.
 We provide the completed analysis as both Scanpy (h5ad) and Seurat objects at https://github.com/davemcg/scEiaD.
@@
 
-We first hand curated over 400,000 published cell type labels.
+We first hand curated over 350,000 published cell type labels across 24 publications.
 With a xgboost algorithm using the latent dimensions, cell age classification (developing or matured), and the UMAP coordinates, we can very accurately label the remaining cells.
@@
 we provide a Google colab/Jupyter notebook which exactly lays out how to use scVI to project (or query) outside data onto our scEiaD with minimum compute resources.
-Finally we also provide our xgboost model so outside groups can label their own datasets with our robust retina cell type labeling system.
+We demonstrate concretely how this can work by showing how iPSC RPE can be queried onto the reference dataset to demonstrate both similarites and dissimilarities in their transcriptomes.
 
@@
 ## Tables
-```{r supTab1_CTcounts, fig.width=8, fig.height=14, out.width=300, echo = F, results='asis', warning = F, fig.cap = supTab1_CTcounts_cap, fig.pos = 'H'}
-supTab1_CTcounts_cap <- supTab_cap(name='supTab1_CTcounts_cap', caption =  'Counts for cell type labels.
+```{r supTab1_CTcounts, fig.width=4, fig.height=14, out.width=300, echo = F, results='asis', warning = F, fig.cap = supTab1_CTcounts_cap, fig.pos = 'H'}
+supTab1_CTcounts_cap <- supTab_cap(name='supTab1_CTcounts', caption =  'Counts for cell type labels.
 Published are the author created labels from the published datasets.
@@
 cat("<div id=\"tbl:supTab1_CTcounts\">")
+source('figs_and_tables/sup_fig__xgboost_PR.R')
 source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
-fontsize(ctTable, size = 8, part = 'all')
+fontsize(ctTable  %>% padding(padding = 0, part = "body") %>% align(align = 'center', part = 'all') %>% autofit(), size = 8, part = 'all')
 #Output Caption (* * for italics)
 cat(paste0("\n*",supTab1_CTcounts_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
+
+```{r supTab_AUCxgboost, fig.cap=supTab_AUCxgboost_cap, fig.width=4, echo=FALSE, results='asis', message = FALSE, warning = FALSE}
+supTab_AUCxgboost_cap <- supTab_cap(name='supTab_AUCxgboost', caption =  'Area under the precision recall curve (AUC) for each cell type, split by study.
+The "All" study is the AUC score across all cells within the cell type')
+
+cat("<div id=\"tbl:supTab_AUCxgboost\">")
+fontsize(xgboost_pr_table_full %>% arrange(`Cell Type`, Study) %>% flextable() %>%  padding(padding = 0, part = "body") %>% align(align = 'center', part = 'all') %>% autofit(), size = 8, part = 'all') %>% colformat_double(digits = 2)
+#Output Caption (* * for italics)
+cat(paste0("\n*",supTab_AUCxgboost_cap,"*\n"))
 # Output </div>
@@
            mutate(Ratio  = round(Count / sum(Count),3 )) %>% 
-           filter(Ratio > 0.05) %>% 
+           filter(Ratio > 0.02) %>% 
            filter(!grepl('Doubl|Periocu', CellType)) %>% 
@@
            flextable()  %>% 
-           autofit(), 
+           padding(padding = 0, part = "body") %>% align(align = 'center', part = 'all') %>% autofit(), 
          size = 8, 
@@
 cat("<div id=\"tbl:supTab_SAcounts\">")
-supTab_SAcounts_cap <- supTab_cap(name='supTab_SAcounts_cap', caption =  'Counts for number of studies with cell types labels before and after cell type label transfer')
+supTab_SAcounts_cap <- supTab_cap(name='supTab_SAcounts', caption =  'Counts for number of studies with cell types labels before and after cell type label transfer')
 
-source('figs_and_tables/sup_fig__xgboost_PR.R')
-fontsize(saTable, size = 8, part = 'all')
+fontsize(saTable  %>% padding(padding = 0, part = "body") %>% align(align = 'center', part = 'all') %>% autofit(), size = 8, part = 'all') 
 #Output Caption (* * for italics)
@@
 
-```{r supTab_AUCxgboost, fig.cap=supTab_AUCxgboost_cap, echo=FALSE, results='asis', message = FALSE, warning = FALSE}
-supTab_AUCxgboost_cap <- supTab_cap(name='supTab_AUCxgboost', caption =  'Area under the precision recall curve (AUC) for each cell type, split by study.
-The "All" study is the AUC score across all cells within the cell type')
+## Figures
 
-cat("<div id=\"tbl:supTab_AUCxgboost\">")
-#source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
-fontsize(xgboost_pr_table_full %>% arrange(`Cell Type`, Study) %>% flextable() %>% width(width = 3), size = 8, part = 'all') %>% colformat_double(digits = 2)
-#Output Caption (* * for italics)
-cat(paste0("\n*",supTab_AUCxgboost_cap,"*\n"))
-# Output </div>
-cat("</div>")
+```{r supFig1, fig.width=4, fig.height=6, out.width=300, fig.cap=supFig1_cap, echo=FALSE, message = FALSE, warning = FALSE}
+supFig1_cap <- supFig_cap(name='supFig1', caption =  'The left bar delineates the number of cells for each organism - technology combination.
+The right bar specifies the number of each cells in each post QC category.
+In silico doublets were identified with scrublet and DoubletDetector.')
+
+source('figs_and_tables/sup_fig1__qc_failure.R')
+supFig1_plot
 ```
-
-```{r supTab_topMarkers, fig.cap=supTab_topMarkers_cap, echo=FALSE, results='asis', message = FALSE, warning = FALSE}
-supTab_topMarkers_cap <- supTab_cap(name='supTab_topMarkers', caption =  'Top differentially expressed genes for each retinal cell type.
-C1 is the number of citations for the gene <-> cell type (e.g.
-Rho and Rods) and C2 is the number of citations for gene and retina (e.g.
-Rho and Retina).')
-
-cat("<div id=\"tbl:supTab_topMarkers\">")
-fontsize(top_markers %>% flextable() %>% width(width = 3), size = 8, part = 'all') %>% colformat_double(digits = 2)
-#Output Caption (* * for italics)
-cat(paste0("\n*",supTab_topMarkers_cap,"*\n"))
-# Output </div>
-cat("</div>")
-```
-
-## Figures
 
@@
   nrow = 2)
+```
+
+```{r supFigShekharSMART, fig.width=3, fig.height=3,  fig.cap=supFigShekharSMART_cap, echo=FALSE, message = FALSE, warning = FALSE}
+supFigShekharSMART_cap <- supFig_cap(name='supFigShekharSMART', caption =  'Our xgboost ML properly labels this Shekhar et al. RBC FAC sorted population as enriched in RBC')
+screenshot_paper <- ggdraw() + draw_image("figs_and_tables/shekhar_screenshot.png", scale = 1.2)
+
+shekhar_counts <- meta_filter %>% 
+  filter(study_accession == 'SRP073242', !is.na(CellType_predict)) %>% 
+  group_by(CellType_predict) %>% summarise(Count = n()) %>% 
+  ggplot(aes(x=CellType_predict, y = Count, label = Count)) + geom_bar(stat = 'identity') + 
+  geom_text(color = 'gray', size = 5) + 
+  cowplot::theme_cowplot(margin(t = 5, r = 5, b = 5, l = 5, unit = "pt")) + 
+  coord_flip()
+cowplot::plot_grid(screenshot_paper, shekhar_counts, labels = c('a','b'), ncol =1)
+```
+
+```{r supFig_PR, fig.width=12, fig.height=10,  fig.cap=supFig_PR_cap, echo=FALSE, message = FALSE, warning = FALSE}
+supFig_PR_cap <- supFig_cap(name='supFig_PR', caption =  'Precision recall curves for our xgboost cell type predictor model across each cell type predicted')
+
+pr_split_curves
 ```
@@
 ```
-```{r eyeIntegration_screenshot, fig.width = 6, fig.height = 4}
-projection_fig_cap <- supFig_cap(name='projection_fig', caption =  'Screenshot of eyeIntegration bulk RNA-seq meta-analysis of vimentin expression in different RPE tissue sources')
 
-ggdraw() + draw_image("figs_and_tables/rpe_eyeintegration.png", scale = 1)
+```{r supFigML_CT_transfer, fig.width=10, fig.height=20,  fig.cap=supFigML_CT_transfer_cap, echo=FALSE, message = FALSE, warning = FALSE}
+supFigML_CT_transfer_cap <- supFig_cap(name='supFigML_CT_transfer', caption =  'F1 scores (1 is perfect) of cell type prediction when using a human or mouse based xgboost cell type prediction model on other organisms (human, macaque, mouse).')
+
+bind_rows(mouse_model__scoring %>% mutate(model = 'Mouse'), human_model__scoring %>% mutate(model = 'Human')) %>% filter(name != 'AC/HC_Precurs') %>% ggplot(aes(x=name,y=value, fill = model, label = round(value, 3))) + geom_bar(stat='identity', position = position_dodge2()) + facet_wrap(~organism, ncol = 1) + coord_flip() + geom_text(position = position_dodge2(width = 1)) + cowplot::theme_cowplot() + ylab('F1 Score') + xlab('Cell Type')
 ```
@@
 
-```{r supFig1, fig.width=4, fig.height=6, out.width=300, fig.cap=supFig1_cap, echo=FALSE, message = FALSE, warning = FALSE}
-supFig1_cap <- supFig_cap(name='supFig1_cap', caption =  'The left bar delineates the number of cells for each organism - technology combination.
-The right bar specifies the number of each cells in each post QC category.
-In silico doublets were identified with scrublet and DoubletDetector.')
+```{r eyeIntegration_screenshot, fig.width=12, fig.height=10,  fig.cap=eyeIntegration_screenshot_cap, echo=FALSE, message = FALSE, warning = FALSE}
+eyeIntegration_screenshot_cap <- supFig_cap(name='eyeIntegration_screenshot', caption =  'Screenshot of eyeIntegration bulk RNA-seq meta-analysis of vimentin expression in different RPE tissue sources')
 
-source('figs_and_tables/sup_fig1__qc_failure.R')
-supFig1_plot
-```
-
-```{r supFig_PR, fig.width=12, fig.height=10,  fig.cap=supFig_PR_cap, echo=FALSE, message = FALSE, warning = FALSE}
-supFig_PR_cap <- supFig_cap(name='supFig_PR', caption =  'Precision recall curves for our xgboost cell type predictor model across each cell type predicted')
-
-pr_split_curves
-```
-
-```{r supFigShekharSMART, fig.width=3, fig.height=3,  fig.cap=supFigShekharSMART_cap, echo=FALSE, message = FALSE, warning = FALSE}
-supFigShekharSMART_cap <- supFig_cap(name='figShekharSMART', caption =  'Our xgboost ML properly labels this RBC enriched population as enriched in RBC')
-screenshot_paper <- ggdraw() + draw_image("figs_and_tables/shekhar_screenshot.png", scale = 1.2)
-
-shekhar_counts <- meta_filter %>% 
-  filter(study_accession == 'SRP073242', !is.na(CellType_predict)) %>% 
-  group_by(CellType_predict) %>% summarise(Count = n()) %>% 
-  ggplot(aes(x=CellType_predict, y = Count, label = Count)) + geom_bar(stat = 'identity') + 
-  geom_text(color = 'gray', size = 5) + 
-  cowplot::theme_cowplot(margin(t = 5, r = 5, b = 5, l = 5, unit = "pt")) + 
-  coord_flip()
-cowplot::plot_grid(screenshot_paper, shekhar_counts, labels = c('a','b'), ncol =1)
+ggdraw() + draw_image("figs_and_tables/rpe_eyeintegration.png", scale = 1)
 ```

### 2021-03-25 figs_and_tables/fig2__integrationPerf.R

@@ -117,7 +117,7 @@ zscore_droplet_scVI_optimize <- perf_scVI %>%
   filter(grepl('scVI', method), epochs == 5) %>% 
   pivot_wider(names_from = c('Score','Group'), values_from = Value) %>% 
   left_join(accuracy %>% group_by(norm, nf, dims, method, knn, epochs) %>% summarise(xgboost_score = mean(score))) %>% 
-  mutate(sumZScale = 
+  mutate( sumZScale = 
            -scale(LISI_CellType)[,1] +
            scale(Silhouette_CellType)[,1] +
            scale(`NMI_CellType-Cluster`)[,1] +
@@ -137,7 +137,10 @@ zscore_droplet_scVI_optimize <- perf_scVI %>%
   #filter(knn == 0.6, method == 'scVI') %>% 
   filter(!is.na(clusterN)) %>% 
   mutate(nf = as.factor(nf),
-         `scVI latent dims` = as.factor(dims)) %>% 
+         `scVI latent dims` = as.factor(dims),
+         method = gsub('scVIprojection', 'scVI-projection',method),
+         method = gsub('scVI$', 'scVI-standard', method),
+         method = factor(method, levels = c('scVI-standard','scVI-projection'))) %>% 
   ggplot(aes(x=sumZScale, y = nf, color = `scVI latent dims`, fill = method)) + 
   #ggbeeswarm::geom_quasirandom(size = 3, groupOnX=FALSE) +
   geom_boxplot() +

### 2021-03-25 figs_and_tables/make_meta_scatter_umap_plot.R

@@ -64,7 +64,8 @@ make_meta_scatter_umap_plot <- function(input, mf, meta_filter,
                                         celltype_labels,
                                         tabulamuris_predict_labels,
                                         cluster_labels,
-                                        cat_to_color_df
+                                        cat_to_color_df,
+                                        velocity = FALSE
 ){
   
   meta_column <- input$meta_column
@@ -185,6 +186,10 @@ make_meta_scatter_umap_plot <- function(input, mf, meta_filter,
   col_size <- {meta_filter[[meta_column]]} %>% n_distinct
   out <- list()
   out$plot <- plot + more
+  if (velocity){
+    out$plot <- plot + geom_segment(data=grid.df, alpha = 0.6, mapping=aes(y=start.1, x=-start.2, 
+                                                             yend=end.1, xend=-end.2), arrow=arrow(length=unit(0.05, "inches"))) + more
+  }
   out$col_size <- col_size
   out
 }

### 2021-03-25 src/organism_celltype_ML.R

@@ -1,0 +1,130 @@
+# run on biowulf2
+# R/4.0.3
+# /data/OGVFB_BG/scEiaD/2021_03_18
+library(glue)
+library(reticulate)
+conda_dir <- '/data/mcgaugheyd/conda/'
+git_dir <- '/home/mcgaugheyd/git/scEiaD/'
+label_id_col <- 'cell_type_id'
+label_name_col <- 'CellType'
+use_python(glue('{conda_dir}/envs/integrate_scRNA/bin/python'), required = TRUE)
+source_python( glue('{git_dir}/src/cell_type_predictor.py'))
+library(Seurat)
+library(tidyverse)
+
+
+load('n_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__preFilter.scEiaD.seuratV3.Rdata')
+
+load('n_features-5000__transform-counts__partition-universe__covariate-batch__method-scVIprojectionSO__dims-8__preFilter.scEiaD__dist-0.1__neighbors-50.umapFilter.predictions.Rdata')
+
+out <- integrated_obj@reductions$scVI@cell.embeddings %>% as_tibble(rownames = 'Barcode') %>% left_join(umap, by = 'Barcode') %>% 
+  filter(study_accession != 'SRP131661')
+
+features <- c(grep('scVI', colnames(integrated_obj@reductions$scVI@cell.embeddings), value = TRUE),
+              'UMAP_1', 'UMAP_2', 'nCount_RNA', 'nFeature_RNA', 'percent.mt')
+
+cluster_outlier <- function(out){
+  groupings <- out %>% 
+    filter(!is.na(CellType)) %>%  
+    group_by(cluster, CellType) %>% 
+    summarise(Count = n()) %>% 
+    mutate(Perc = Count / sum(Count)) %>% 
+    filter(Perc > 0.025)
+  out %>% 
+    left_join(groupings %>% 
+                select(cluster, CellTypeKEEP = CellType)) %>% 
+    filter(CellType == CellTypeKEEP) %>% 
+    select(-CellTypeKEEP)
+}
+
+# remove celltype outliers
+# we will removve 5% furthest from euclidean center
+rm_outlier <- function(out, TM = FALSE){
+  bc_retain <- list()
+  for (i in out$CellType %>% unique){
+    print(i)
[…]

### 2021-07-24 figs_and_tables/fig_umap_benchmarking.R
new data to load, one more color option

@@ -1,9 +1,9 @@
 library(scattermore)
 source('~/git/scEiaD_manuscript/figs_and_tables/fig2__integrationPerf.R')
-load('~/data/scEiaD/n_features-2000__transform-standard__partition-TabulaDroplet__covariate-batch__method-none__dims-30__dist-0.3__neighbors-15__knn-7__umap.Rdata')
+load('~/data/scEiaD/giga_all_methods/gigascience/umap/n_features-2000__transform-standard__partition-universe__covariate-batch__method-none__dims-30__epochs-5__dist-0.1__neighbors-50__knn-20__umap.Rdata')
 umapNone <- umap
 
-load('~/data/scEiaD/n_features-2000__transform-standard__partition-TabulaDroplet__covariate-batch__method-combat__dims-30__dist-0.3__neighbors-15__knn-7__umap.Rdata')
+load('~/data/scEiaD/giga_all_methods/gigascience/umap/n_features-2000__transform-libSize__partition-universe__covariate-batch__method-combat__dims-8__epochs-5__dist-0.1__neighbors-50__knn-20__umap.Rdata')
 umapCombat <- umap
 
 umap_plot_maker <- function(umap, color_against = 'CellType', red = 'UMAP', ptsize = 2){
@@ -50,6 +50,16 @@ umap_plot_maker <- function(umap, color_against = 'CellType', red = 'UMAP', ptsi
       guides(colour = guide_legend(override.aes = list(size=2, alpha = 1))) + 
       theme_void() +
       scale_colour_manual(values = pals::glasbey()) +
+      xlab(paste(red, '1')) + ylab(paste(red, '2')) 
+  } else if (color_against == 'organism'){
+    umapFig %>% 
+      ggplot() + 
+      geom_scattermore(aes(x=umapFig[,paste0(red,'_1')] %>% pull(1), 
+                           y = umapFig[,paste0(red,'_2')] %>% pull(1), 
+                           colour = organism), pointsize = (ptsize/3), alpha = 0.1) + 
+      guides(colour = guide_legend(override.aes = list(size=2, alpha = 1))) + 
+      theme_void() +
+      scale_colour_manual(values = pals::brewer.set3(n=3)) +
       xlab(paste(red, '1')) + ylab(paste(red, '2')) 
   }
 }

### 2021-07-24 figs_and_tables/fig2__integrationPerf.R
new source data, tweak filtering for new knn, add bootstrapping calc

@@ -12,18 +12,16 @@ accuracy <- accuracy %>%
     epochs = str_extract(file, 'epochs-\\d+') %>% gsub('epochs-','',.) %>% as.numeric())
 
 
-load('~/data/scEiaD//metrics_2020_11_28.Rdata')
-perfO <- perf
+load('~/data/scEiaD/giga_all_methods/gigascience/2021_07_22.perf.allmethod.Rdata')
+perfO <- perf %>% filter(dims %in% c(8,30), knn == 20, method != 'scVIprojection')
 
-load('~/data/scEiaD//merged_stats_2021-03-17.Rdata')
+load('~/data/scEiaD/merged_stats_2021-03-17.Rdata')
 
-perf <- perf %>% 
-  mutate(knn = case_when(method == 'scArches' ~ 7, TRUE ~ knn))
+perf <- perf %>% filter(grepl('scVI',method)) 
 cluster_stats <- perf %>% select(clusterN, clusterMedian, dims, nf, knn, method, normalization, set) %>% unique() 
 
 perfO <- perfO %>% 
-  mutate(knn = case_when(method == 'scArches' ~ 7, TRUE ~ knn),
-         normalization = case_when(method == 'scArches' ~ 'standard',
+  mutate(normalization = case_when(method == 'scArches' ~ 'standard',
                                    TRUE ~ normalization))
 
 #cluster_stats <- perf %>% select(clusterN, clusterMedian, dims, nf, knn, method, normalization, set) %>% unique() 
@@ -32,44 +30,44 @@ perfO <- perfO %>%
 #   select(Score, Group, Value, set, dims:normalization) %>% 
 #   filter(Score %in% c('LISI','Silhouette')) 
 
-perf_tabula <- perfO %>% unique() %>% filter(set %in% c('TabulaDroplet')) %>% 
+perf_all_methods <- perfO %>% unique() %>% 
   select(Score, Group, Value, set, dims:normalization) %>% 
   filter(Score %in% c('LISI','Silhouette', 'ARI', 'PCR', 'NMI')) 
 
-perf_scVI <- perf %>% unique() %>% filter(set %in% c('universe')) %>% 
+perf_scVI <- perf %>% unique() %>%
   select(Score, Group, Value, set, dims:normalization) %>% 
   filter(Score %in% c('LISI','Silhouette', 'ARI', 'PCR', 'NMI')) 
 
-lisi <- perf_tabula %>% 
[…]

### 2021-07-24 figs_and_tables/sup_fig__xgboost_PR.R
explictly call the right `rename `

@@ -17,7 +17,7 @@ pr_calculator <- function(test_predictions, partition){
   target_cell_types <-unique(cell_type2id$CellType)
   pr_df <- test_predictions %>% 
     select(-CellTypeID, -CellType) %>% 
-    rename(CellTypeID = true_label_id) %>% 
+    dplyr::rename(CellTypeID = true_label_id) %>% 
     inner_join(cell_type2id) %>% 
     filter(CellType%in% target_cell_types) %>% 
     select(all_of(target_cell_types), CellTypeID, CellType) 

### 2021-07-24 manuscript_01_gigascience_resubmission.Rmd
resubmission paper

--- 2021-03-25 manuscript_00.Rmd
@@
 ---
-title: 'Building the Meta Single Cell Transcriptome Ocular Meta-Atlas'
+title: 'Building the Mega Single Cell Transcriptome Ocular Meta-Atlas'
 author:
-  - Vinay Swamy:
+  - Vinay S Swamy:
       institute: bg
-  - Temesgen Fufa:
+  - Temesgen D Fufa:
       institute: mgog
-  - Robert Hufnagel:
+  - Robert B Hufnagel:
       institute: mgog
-  - David McGaughey:
+  - David M McGaughey:
       institute:
@@
     fig_caption: yes
+    keep_md: yes
     pandoc_args:
@@
 We developed a lightweight R package (scPOP) that brings in batch integration methods and uses a simple heuristic to balance batch merging and celltype/cluster purity.
-We use this package along with a Snakefile based workflow system to demonstrate how to optimally merge 766,615 cells from 34 retina datsets and three species to create a massive ocular single cell transcriptome meta-atlas.
+We use this package along with a Snakefile based workflow system to demonstrate how to optimally merge 766,615 cells from 33 retina datsets and three species to create a massive ocular single cell transcriptome meta-atlas.
 This provides a model how to efficiently create meta-atlases for tissues and cells of interest."
-keywords: "RNA-seq, retina, RPE, ocular, eye, Snakemake, singlecell, scRNA, single-cell, singlecell, scPOP, R"
+keywords: "RNA-seq, retina, RPE, ocular, eye, Snakemake, singlecell, scRNA, single-cell, singlecell, scPOP, R, atlas"
 ---
@@
 The retinal bipolar cells transmit the electrical signal from the photoreceptors to the retinal ganglion cells.
-Horizontal and amacrine regulate and help interpret signals from the photoreceptors.
+Horizontal and amacrine cells regulate and help interpret signals from the photoreceptors.
 The signal is relayed via the retinal ganglion projections through the optic stalk to the brain (for review see [@maslandNeuronalOrganizationRetina2012]).
@@
 The retina was used as the source tissue in one of the earliest works in the high throughput single cell transcriptomics field [@macoskoHighlyParallelGenomewide2015].
-As of late 2020, over a twenty published studies, cumulatively containing over a million cells, have used single cell technology to profile cell type specific gene expression patterns, cell fate trajectory, tissue and cell differentiation, and disease perturbation across multiple mammalian species [@buenaventuraIdentificationGenesEnriched2019; @clarkComprehensiveAnalysisRetinal2018; @collinCRXExpressionPluripotent2019; @collinDeconstructingRetinalOrganoids2019; @cowanCellTypesHuman2020; @diacouSix3Six6Are2018; @eldredThyroidHormoneSignaling2018; @hengHypoxiaToleranceNorrindeficient2019; @huDissectingTranscriptomeLandscape2019; @logiudiceSinglecellTranscriptionalLogic2019; @lukowskiSinglecellTranscriptomeAtlas2019; @luSingleCellAnalysisHuman2020;  @menonSinglecellTranscriptomicAtlas2019; @okorenMicroglialFunctionDistinct2019; @pengMolecularClassificationComparative2019; @phillipsNovelApproachSingle2018; @rheaumeSingleCellTranscriptome2018; @ruedaHippoPathwayBlocks2019; @shekharComprehensiveClassificationRetinal2016; @sridharSingleCellTranscriptomicComparison2020; @tranSingleCellProfilesRetinal2019; @voigtMolecularCharacterizationFoveal2019; @voigtSinglecellTranscriptomicsHuman2019; @yanCellAtlasHuman2020].
+As of late 2020, over twenty published studies, cumulatively containing over a million cells, have used single cell technology to profile cell type specific gene expression patterns, cell fate trajectory, tissue and cell differentiation, and disease perturbation across multiple mammalian species [@buenaventuraIdentificationGenesEnriched2019; @clarkSingleCellRNASeqAnalysis2019; @cowanCellTypesHuman2020; @dharmatEpigeneticAdaptationProlongs2019; @fadlOptimizedProtocolRetina2020; @huDissectingTranscriptomeLandscape2019; @lehmannSinglecellProfilingReveals2020; @logiudiceSinglecellTranscriptionalLogic2019; @logiudiceSinglecellTranscriptionalLogic2019; @lukowskiSinglecellTranscriptomeAtlas2019; @luSingleCellAnalysisHuman2020; @macoskoHighlyParallelGenomewide2015; @menonSinglecellTranscriptomicAtlas2019; @okorenMicroglialFunctionDistinct2019; @okorenMicroglialFunctionDistinct2019; @pengMolecularClassificationComparative2019; @shekharComprehensiveClassificationRetinal2016; @sridharSingleCellTranscriptomicComparison2020; @tranSingleCellProfilesRetinal2019; @voigtBulkSinglecellGene2020; @voigtMolecularCharacterizationFoveal2019; @voigtSingleCellRNASequencing2020; @voigtSinglecellTranscriptomicsHuman2019; @yanCellAtlasHuman2020; @yanMouseRetinalCell2020].
 
@@
 The single cell community has recognized that removal of these technical (also referred to as batch) effects is a critical issue and have independently developed many tools, though it remains unclear which tools and parameters are optimal for a particular dataset [@butlerIntegratingSinglecellTranscriptomic2018; @haghverdiBatchEffectsSinglecell2018; @hieEfficientIntegrationHeterogeneous2019; @johnsonAdjustingBatchEffects2007; @korsunskyFastSensitiveAccurate2019; @liuJointlyDefiningCell2020; @lopezDeepGenerativeModeling2018; @polanskiBBKNNFastBatch2020; @QueryReferenceSinglecell; @simonINSCTIntegratingMillions2020; @stuartComprehensiveIntegrationSingleCell2019; @vandijkRecoveringGeneInteractions2018].
+Reinforcing this point, a couple groups have quantified performance of multiple methods in a consistent framework across several test datasets, finding no consistently best approach [@lueckenBenchmarkingAtlaslevelData2020; @tranBenchmarkBatcheffectCorrection2020].
 
@@
 First, the ocular community who can both search scEiaD for gene expression across many dimensions (e.g.
-cluster, cell type, study) and project their own single cell data onto scEiaD for comparison and rich automatic cell labelling.
+cluster, cell type, study) and project their own single cell data onto scEiaD for comparison and rich automatic cell labeling.
 Second, the computational community can use this very large, well-curated dataset to test algorithms for compute efficiency and performance in a diverse environment.
@@
 A wide variety of methods have been written for scRNA-seq integration.
-As we were uncertain which would perform the best, we ran `r methods %>% length()` tools with a commonly used set of key parameters like number of hyper variable genes (HVG), size of latent dimensional space, and the number of nearest neighbors for the louvain clustering algorithm.
+As we were uncertain which would perform the best, we ran `r methods %>% length()` tools with a commonly used set of key parameters like number of highly variable genes (HVG), size of latent dimensional space, and the number of nearest neighbors for the louvain clustering algorithm.
 The Snakemake system was used to automate the running of the wide variety of tools.
@@
 With our scEiaD meta-atlas insct, trVAE, and magic do not perform particularly well in batch mixing or cluster purity.
-Across the 11 integration methods we tested, scVI achieved the highest sumZScale for out scEiaD meta-atlas.
+For further qualitative investigation of the integration performance we provide the UMAP visualizations of each integration method, normalization, and latent dimensions colored by cell type, study accession, or organism as a supplementary file.
+
+As the LISI and silhouette metrics provide independent cluster and cell type ("purity") and batch ("mixing") scores we looked to see whether the sumZScale scoring is highly influenced by changing the weight placed on purity or mixing by multiplying either by a multiplier of three (`r supFig_cap(name = 'supFig_benchmark_weighting', display = 'cite')`).
+While scVI still performs well, no matter the weight chosen, there were some changes when more weight was placed on batch mixing.
+Most notably the 8 latent dims with fastMNN receives a higher score when a higher weight was placed on batch mixing and CCA's performance with 30 latent dimensions jumps to the top.
+We also bootrap random weights and get similar results (`r supFig_cap(name = 'supFig_benchmark_weighting_ridges', display = 'cite')`).
+Across the 11 integration methods we tested, scVI achieved the highest sumZScale for our scEiaD meta-atlas and furthermore is a desirable choice because of its very short run time (scVI can complete in less than an hour while fastMNN takes over 12 hours and CCA takes more than 24 hours).
 
@@
 Library size (libSize) normalization omits the sqrt or log transformation.
-Finally some methods, like scVI, directly use the raw counts data for modeling the data.
+Finally some methods, like scVI, directly use the raw counts for modeling the data.
 
@@
 sumZScale (higher is better) for each method across a variety of data normalizations.
-All methods shown here use 2000 HVG, louvain clustering, and 8 latent dimensions.
+All methods shown here use 2000 HVG, louvain clustering with knn 20, and 8 or 30 latent dimensions.
 Each color is a different method.
@@
 Most of the remaining cell types can be predicted with an AUC over 0.9 `r supTab_cap(name = 'supTab_AUCxgboost', display = 'cite')`.
-Several of the precursor cell types (photoreceptor, AC/HC, and neurogenic) were labelled with lower confidence `r supTab_cap(name = 'supTab_CTmislabels', display = 'cite')`.
+Several of the precursor cell types (photoreceptor, AC/HC, and neurogenic) were labeled with lower confidence `r supTab_cap(name = 'supTab_CTmislabels', display = 'cite')`.
 The next most common labels for these cell types were either other precursor cells (e.g.
@@
 Confusion matrix of cell type prediction performance of our xgboost labeller between predicted (x axis) and known (y axis) using data withheld from the machine learner.
-Most of the cell types are indeed labelled as their true type.
+Most of the cell types are indeed labeled as their true type.
 e.
@@
 For example, we went from `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(2)` human studies with labeled Müller Glia to `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(5)` after labeling (`r supTab_cap(name = 'supTab_SAcounts', display = 'cite')`).
-Overall we go from an average of `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(2) %>% mean() %>% round()` studies per human cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(5) %>% mean() %>% round()` and `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(4) %>% mean() %>% round()` studies per mouse cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(7) %>% mean() %>% round()` after transferring the cell type labels.
+Overall we go from an average of `r joinedSA %>% filter(CellType != 'Unlabeled') %>% pull(2) %>% mean() %>% round()` studies per human cell type to `r joinedSA %>% filter(CellType != 'Unlabeled') %>% pull(5) %>% mean() %>% round()` and `r joinedSA %>% filter(CellType != 'Unlabeled') %>% pull(4) %>% mean() %>% round()` studies per mouse cell type to `r joinedSA %>% filter(CellType != 'Unlabeled') %>% pull(7) %>% mean() %>% round()` after transferring the cell type labels.
 
-As another check on the quality of the cell type assignments, we ran the cell and cluster independent haystack gene search and pairwise differential expression tests between the predicted cell types (see methods for further details).
+As another check on the quality of the cell type assignments, we ran the cell and cluster independent haystack gene search and pairwise differential expression tests between the predicted cell types (see methods for further details) [@vandenbonClusteringindependentMethodFinding2020].
 We show the five most differentially expressed genes for each of the major retina cell types are consistent with known retinal cell markers (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a).
@@
 "PDE6H AND Retina").
-Of the 50 genes in (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a), `r 50 - top_markers %>% filter(Gene %in% dplot_markers) %>% filter(c1 == 0) %>% nrow` had one or more citations in the gene - cell type search (Supplementary File celltype_markers.tsv) and `r 50 - top_markers %>% filter(Gene %in% dplot_markers) %>% filter(c1 == 0, c2 == 0) %>% nrow` had one or more citation in the inclusive search.
+Of the 50 genes in (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a), `r 50 - top_markers %>% filter(Gene %in% dplot_markers) %>% filter(c1 == 0) %>% nrow` had one or more citations in the gene - cell type search (Supplemental File celltype_markers.tsv) and `r 50 - top_markers %>% filter(Gene %in% dplot_markers) %>% filter(c1 == 0, c2 == 0) %>% nrow` had one or more citation in the inclusive search.
 The 50 genes had a mean of `r top_markers %>% filter(Gene %in% dplot_markers) %>% pull(c2) %>% mean() %>% round()` studies (with the inclusive gene by "retina" search).
-In contrast, 100 randomly chosen genes had a mean of `r pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% pull(value) %>% mean() %>% round` (wilcox test p < 1.44 x 10^-17).
+In contrast, 100 randomly chosen genes had a mean of `r pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% pull(value) %>% mean() %>% round` (wilcox test p < 1.44 x 10**-17).
 
@@
 
-While by eye the UMAP 2D projection generally blends together the three different species (macaque, mouse, human) in the UMAP visualization, we more rigorously tested this by tweaking the inputs to our xgboost cell type predictor machine learning system.
-Briefly, we trained two new models: one with only human data and one with only mouse data.
+While by eye the UMAP 2D projection generally blends together the three different species (macaque, mouse, human) in the UMAP visualization, we more rigorously tested this by changing the inputs to our xgboost cell type predictor machine learning system.
+We trained two new models: one with only human data and one with only mouse data.
 We then applied this model to the other two species to see whether the model trained on one species was generally transferable to another species.
 Again, both models (human only and mouse only) both proved to be highly accurate at predicting cell types in the other species with the terminal cell types (`r supFig_cap(name = 'supFigML_CT_transfer', display = 'cite')`.
+
 ## Projection of outside data onto scEiaD demonstrates similarities and differences of iPSC and organoids to primary cells
@@
 The hard work of creating this resource can be leveraged and extended by the wider community with a few relatively simple steps and modest compute requirements.
-Very briefly, if outside groups quantify their mouse or human scRNA with kallisto and the same references (see methods), they can overlay their data on top of scEiaD by 1.
+Very briefly, if outside groups quantify their mouse or human scRNA with kallisto (bustools) and the same references (see methods), they can overlay their data on top of scEiaD by 1.
 installing scVI (version 0.9.0 or higher), 2.
@@
 
-## Reproduciblity
+## Reproduciblity and data availability
 
 The set of Snakemake pipelines that takes in the raw fastq sequence and outputs the scEiaD is at https://github.com/davemcg/scEiad [@kosterSnakemakeScalableBioinformatics2012].
-The publication commit is #9d3c5e1 Furthermore, the repository has been deposited at Zenodo under 10.5281/zenodo.4638117 We will briefly discuss the pipeline choices, programs and algorithms, and versions below.
+The publication commit is #ffdf738.
+Furthermore, the repository has been deposited at Zenodo under 10.5281/zenodo.4638117 We will briefly discuss the pipeline choices, programs and algorithms, and versions below.
 For the R packages, we provide package versions as the supplementary file "R_session_info.txt"
+
+We have deposited the 10x (v2) fastq files for our iPSC-based RPE data in GEO at accession GSE180662.
+We give download links for the count matrices, seurat and anndata objects, cell level metadata, and our codebases in `supTab_cap(name = 'supTab_links', display = 'cite')`.
 
@@
 A single set of exonic transcript sequences was created for each species for all well techonologies.
-A kallisto quantification index was generated for each of these fastas using kallisto index (0.46.2) [@brayNearoptimalProbabilisticRNAseq2016].
+A kallisto quantification index was generated for each of these fastqs using kallisto index (0.46.2) [@brayNearoptimalProbabilisticRNAseq2016].
 Well based samples were quantified using kallisto quant.
@@
 Otherwise the “–bias” flag was added.
-For the droplet-based samples, we adapted the bustools workflow for generating spliced and unspliced count matricies.
+For the droplet-based samples, we adapted the bustools workflow for generating spliced and unspliced count matrices.
 (0.39.4) [@melstedBarcodeUMISet2019].
@@
 We then removed cells with percent mitochondrial reads of >10%.
-After merging the individual count matrices into one sparse matrix, we created a Seurat version 3 object and removed cells with fewer than 200 detected unique genes, and for the droplet data, more than 3000 detected genes (these are likely to be doublets) [@butlerIntegratingSinglecellTranscriptomic2018].
+After merging the individual count matrices into one sparse matrix, we created a Seurat version 3 object and removed cells with fewer than 200 detected unique genes, and for the droplet data, more than 3000 detected genes (these are more likely to be doublets) [@butlerIntegratingSinglecellTranscriptomic2018].
 
@@
 
-The following steps (normalization through benchmarking are handled by the SnakePOP file) We tested several gene count normalization approaches as we were not certain which would produce an optimal outcome: standard (default Seurat, library size normalization, then log transform), sqrt (same, but with sqrt normalization), libSize (omit the log or sqrt normalization), scran, SCT from Seurat, and for scVI, no normalization (counts) [@lunStepbystepWorkflowLowlevel2016].
+The following steps (normalization through benchmarking are handled by the SnakePOP pipeline) We tested several gene count normalization approaches as we were not certain which would produce an optimal outcome: standard (default Seurat, library size normalization, then log transform), sqrt (same, but with sqrt normalization), libSize (omit the log or sqrt normalization), scran, SCT from Seurat, and for scVI, no normalization (counts) [@lunStepbystepWorkflowLowlevel2016].
 Our R implementation of the normalization approaches as well as how we constructed the Seurat v3 object can be found in the supplementary file "make_seurat_obj_functions.R"
@@
 We tested `r methods` against 2000 HVGs, the different gene count normalization procedures discussed above, and both 8 and 30 outputted batch corrected latent dimensions.
-The latent dimensions are the input for clustering, the 2D UMAP visualization, and the xgboost machine learning to transfer cell type labels to unlabelled cells.
+The latent dimensions are the input for clustering, the 2D UMAP visualization, and the xgboost machine learning to transfer cell type labels to unlabeled cells.
 We were unable to run every method successfully with every normalization method.
-Magic could not complete with the standard or libSize normalization.
-CCA could not complete with the libSize normalization.
+For example, magic could not complete with the standard or libSize normalization.
+Insct was only able to complete the scran normalization.
 We also tried the DESC, liger, and Conos batch corrections methods but were unable to get them to work reliably so they were dropped.
-The batch correction step implementation can be found the supplementary file "merge_methods.R" and in the github repo (https://github.com/davemcg/scEiaD/blob/master/src/merge_methods.R).
+While we attempted to use "default" parameters wherever possible, we had to deviate from this to get Seurat's CCA procedure to complete.
+CCA reciprocally tries to integrate all batches and even with a subset of cells we regularly got "long vector" errors.
+We got around this issue by setting the human batches as the "reference." The batch correction steps implementation can be found the supplementary file "merge_methods.R" and in the github repo (https://github.com/davemcg/scEiaD/blob/master/src/merge_methods.R).
 
@@
 Louvain-Jaccard clustering against the batch corrected latent dimensions used the Seurat implementation [@blondelFastUnfoldingCommunities2008].
-We tried the k-nearest neighbors (knn) parameters 5 and 7 (where 5 gives more clusters than 7).
+For the all methods benchmarking we used k-nearest neighbors (knn) of 20.
+For the scVI-only tuning reduced the knn parameters to 5 and 7 (where 5 gives more clusters than 7) to increase the cluster number.
 We also used the leiden algorithm as implemented by PARC with a resolution of 0.6 and 0.8 (higher results in more clusters) [@traagLouvainLeidenGuaranteeing2019; @stassenPARCUltrafastAccurate2020].
@@
 We wrote the scPOP R package to unify the LISI and Silhouette metrics from Harmony and kBet, respectively, along with NMI and ARI [@buttnerTestMetricAssessing2019; @korsunskyFastSensitiveAccurate2019].
-LISI and Silhouette require a dense matrix, which is a problem for our data as a `r meta_filter %>% nrow()` cell by 8 latent dimension dense matrix cannot fit in out largest available memory node (1.5 TB).
+LISI and Silhouette require a dense matrix, which is a problem for our data as a `r meta_filter %>% nrow()` cell by 8 latent dimension dense matrix cannot fit in out largest available compute node (1.5 TB).
 We down-sampled the dataset to ~100,000 cells, taking care to keep all rarer cell types for the LISI and Silhouette benchmarking.
@@
 
-In order to identify cell types for the `r format(meta_filter %>% filter(is.na(CellType), is.na(TabulaMurisCellType)) %>% nrow(), big.mark=",", scientific=FALSE)` unlabelledcells we designed a custom xgboost based cell type classifier.
+In order to identify cell types for the `r format(meta_filter %>% filter(is.na(CellType), is.na(TabulaMurisCellType)) %>% nrow(), big.mark=",", scientific=FALSE)` unlabeled cells we designed a custom xgboost based cell type classifier.
 We took labeled data and split it into training (2/3) and test (1/3) sets, stratified by cell type.
@@
 The trained model had an overall macro and micro AUC score of 0.98 and 0.99, respectively.
-This model was then used to identify labels for each all cell.
+This model was then used to identify labels for all cells.
 Unlabeled data was pre-processed identically to training data and fed into model to generate a vector of label probabilities for each cell.
@@
 For the organism specific xgboost ML we followed the above procedure, except that we combined Early RPCs, Late RPCs, and RPCs into one category and did not attempt to predict the non-retina cell types (e.g.
-fibroblasts) as there were very few labelled cells across all three organisms.
+fibroblasts) as there were very few labeled cells across all three organisms.
 
@@
 
-To identify marker genes across the CellType (predict) and cluster groups, we used the scran findmarkers (wilcox test) along with the singleCellHaystack algorithm.
+To identify marker genes across the CellType (predict) and cluster groups, we used the scran findmarkers (wilcox test) along with the singleCellHaystack algorithm [@vandenbonClusteringindependentMethodFinding2020].
 The scran findmarkers test runs a wilcox test in a pairwise manner (e.g.
@@
 
-# Limitations
+## Limitations
 
@@
 
-While scEiaD distinguishes the major cell types very well, some of the cell types contain many "sub types" - notably the Amacrine cells have a huge variety in morphology with a recent study identifying over 60 different types of Amacrine cells in mouse.
+While scEiaD distinguishes the major cell types very well, some of the cell types contain many "sub types" - notably the amacrine cells have a huge variety in morphology with a single cell based study identifying over sixty different types of amacrine cells in mouse [@yanMouseRetinalCell2020].
 At this time our batch corrected pan retina cell space does not precisely resolve these sub cell types with high resolution.
@@
 After adding more integration methods we first attempted "hand-assess" the integration results by using the UMAP 2D projection view.
-This proved to scale poorly and this led use to curate some of the more useful benchmarking algorithms (NMI, ARI, silhouette, and LISI) that roughly matched our "hand-assessed" results into the scPOP R package.
+This proved to scale poorly and this led us to curate some of the more useful benchmarking algorithms (NMI, ARI, silhouette, and LISI) that roughly matched our "hand-assessed" results into the scPOP R package.
 While we chose the scVI algorithm, we strongly suggest any other groups attempting a similar meta-atlas construction chose a quantifiable set of criteria so optimal methods and parameters can be picked.
@@
 Many other cell type labeling algorithms and systems exist for those groups less willing or able to tune a machine learning algorithm.
-For example, the developers of scVI also have a cell type label projection algorithm called scANVI.
+For example, the developers of scVI also have a cell type label projection algorithm called scANVI [@ProbabilisticHarmonizationAnnotation2021].
 Whichever approach you use, taking a smaller number of high quality labels and projecting them onto the remaining cells is a powerful way to leverage community knowledge across a huge diverse dataset.
@@
 our data is available via download in several forms at https://github.com/davemcg/scEiaD, and crucially 3.
-we provide a Google colab/Jupyter notebook which exactly lays out how to use scVI to project (or query) outside data onto our scEiaD with minimum compute resources.
-We demonstrate concretely how this can work by showing how iPSC RPE can be queried onto the reference dataset to demonstrate both similarites and dissimilarities in their transcriptomes.
+we provide a Google colab/Jupyter notebook which step-by-step demonstrates out how to use scVI to project (or query) outside data onto our scEiaD with minimum compute resources.
+We demonstrate concretely how this can work by showing how iPSC RPE can be queried onto the reference dataset to demonstrate both similarities and dissimilarities in their transcriptomes.
 
@@
 supTab_CTmislabels_cap <- supTab_cap(name='supTab_CTmislabels', caption =  'Counts of cell type labels with our xgboost machine learning system (PredCellType) and the published cell type labels (TrueCellType).
-Ratio is calculated as CellType that were labelled as CellType_predict.
+Ratio is calculated as CellType that were labeled as CellType_predict.
 Ratio < 0.05 were filtered out from view in the table.
@@
 
-```{r supTab_SAcounts, fig.width=8, fig.height=14, out.width=300, fig.cap=supTab_SAcounts_cap, echo=FALSE, results='asis', message = FALSE, warning = FALSE}
+```{r supTab_SAcounts, fig.width=5, fig.height=14, out.width=300, fig.cap=supTab_SAcounts_cap, echo=FALSE, results='asis', message = FALSE, warning = FALSE}
 cat("<div id=\"tbl:supTab_SAcounts\">")
@@
 cat(paste0("\n*",supTab_SAcounts_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
+
+```{r supTab_links, fig.width=2, fig.height=6, out.width=300, echo = F, results='asis', warning = F, fig.cap = supTab_links_cap, fig.pos = 'H'}
+supTab_links_cap <- supTab_cap(name='supTab_links', caption =  'Links to code and resources')
+cat("<div id=\"tbl:supTab_links\">")
+
+linksTable <- tibble::tribble(
+                                           ~`Resource or File`,                                                                         ~Link,  ~Accession,
+                                             "Seurat Object", "http://hpc.nih.gov/~mcgaugheyd/scEiaD/2021_03_17/scEiaD_all_seurat_v3.Rdata",          NA,
+                                            "Anndata Object",    "http://hpc.nih.gov/~mcgaugheyd/scEiaD/2021_03_17/scEiaD_all_anndata.h5ad",          NA,
+                                   "Counts, R sparse matrix",               "http://hpc.nih.gov/~mcgaugheyd/scEiaD/2021_03_17/counts.Rdata",          NA,
+                                       "Cell level metadata",     "http://hpc.nih.gov/~mcgaugheyd/scEiaD/2021_03_17/metadata_filter.tsv.gz",          NA,
+                              "Codebase for scEiaD creation",                                           "https://github.com/davemcg/scEiaD",   "ffdf738",
+                            "Codebase for scEiad manuscript",                                "https://github.com/davemcg/scEiaD_manuscript",          NA,
+                "Zenodo deposit of all above files and code",                                                      "10.5281/zenodo.4638117",   "4638117",
+                            "iPSC RPE scRNA Raw Fastq Files",                "https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE180662", "GSE180662"
+                )
+fontsize(flextable(linksTable)  %>% padding(padding = 0, part = "body") %>% align(align = 'left', part = 'all') %>% autofit(), size = 4, part = 'all')
+#Output Caption (* * for italics)
+cat(paste0("\n*",supTab_links_cap,"*\n"))
 # Output </div>
@@
 
-```{r supFig_benchmark, fig.width=12, fig.height=8, out.width=300, fig.cap=supFig_benchmark_cap, echo=FALSE, message = FALSE, warning = FALSE}
+```{r supFig_benchmark, fig.width=14, fig.height=12, out.width=300, fig.cap=supFig_benchmark_cap, echo=FALSE, message = FALSE, warning = FALSE}
 supFig_benchmark_cap <- supFig_cap(name='supFig_benchmark', caption =  'Performance of the various batch correction tools across various benchmarking metrics.
@@
   nrow = 2)
+```
+
+```{r supFig_benchmark_weighting, fig.width=9, fig.height=5, out.width=300, fig.cap=supFig_benchmark_weighting_cap, echo=FALSE, message = FALSE, warning = FALSE}
+supFig_benchmark_weighting_cap <- supFig_cap(name='supFig_benchmark_weighting', caption =  'We take the sumZscoring from earlier, and apply two different weighting schemes to demonstrate how different priorities (much higher cluster purity or much more cluster mixing) can influence the scoring.
+In A we give a 3x multiplier to cell and cluster purity relative to batch mixing.
+In B we give a 3x multipler to batch mixing and we see that fastMNN with 8 dims has a higher score and the Seurat CCA metric ranks better.
+scVI performance with 8 latent dimension is consistently high across all weighting schemes.')
+
+cowplot::plot_grid(
+  zscore_allMethods_highPurity_plot + theme(legend.position="none"),
+  zscore_allMethods_highMixing_plot,
+  labels = c('a','b'))
+```
+
+```{r supFig_benchmark_weighting_ridges, fig.width=5, fig.height=15, out.width=300, fig.cap=supFig_benchmark_weighting_ridges_cap, echo=FALSE, message = FALSE, warning = FALSE}
+supFig_benchmark_weighting_ridges_cap <- supFig_cap(name='supFig_benchmark_weighting_ridges', caption =  'The sumZScale is composed of 10 metrics.
+We randomly weighed each zscaled metric by multiplying by a value randomly chosen between 0.1 and 10.
+This is bootstrapped 1000 times.
+The sumZScale is then computed and we extract the rank (by highest sumZScale) for 1000 bootstraps and plot the distribution as a density plot (lower rank is better integration performance).
+The y axis is ordered, top to bottom, by mean sumZScale Rank (higher on the y axis is better).')
+
+zscore_ridges_plot
 ```

### 2021-07-24 figs_and_tables/sup_fig_umap_all_methods.R
sup file umap plots

@@ -1,0 +1,65 @@
+umap_files <- list.files('~/data/scEiaD/giga_all_methods/gigascience/umap/', full.names = TRUE)
+
+umap_files <- umap_files %>% enframe() %>%  mutate(method = str_extract(value, 'method-[a-zA-Z]+') %>% gsub('method-','',.)) %>% arrange(method) %>% pull(value)
+
+umap_data <- list()
+for (i in umap_files){
+  print(i)
+  dims = str_extract(i, 'dims-\\d+') %>% gsub('dims-','',.) %>% as.integer()
+  norm = str_extract(i, 'transform-[a-zA-Z]+') %>% gsub('transform-','',.)
+  load(i)
+  umap$dims <- dims
+  umap$norm <- norm
+  umap_data[[i]] <- umap
+}
+
+
+# plot by celltype
+plots_celltype <- list()
+for (i in umap_files){
+  plots_celltype[[i]] <- umap_plot_maker(umap_data[[i]], color_against <- 'CellType', red = 'UMAP') + 
+    theme(legend.position = 'none') +
+    ggtitle(paste(umap_data[[i]]$Method %>% unique(), 'dims:', umap_data[[i]]$dims%>% unique(), 'normalization:', umap_data[[i]]$norm %>% unique()))
+}
+legend <- get_legend(
+  umap_plot_maker(umap_data[[1]], color_against <- 'CellType', red = 'UMAP') + 
+    ggtitle(paste(umap_data[[1]]$Method %>% unique(), 'dims:', umap_data[[1]]$dims%>% unique(), 'normalization:', umap_data[[1]]$norm %>% unique())) +
+    theme(legend.box.margin = margin(0, 0, 0, 12))
+)
+
+pdf('all_methods_celltype.pdf', width = 12, height = 50)
+plot_grid(plot_grid(plotlist = plots_celltype, ncol = 3), plot_grid(legend), ncol = 1, rel_heights = c(20,4))
+dev.off()
+
+# plot by study accession
+plots_batch <- list()
+for (i in umap_files){
+  plots_batch[[i]] <- umap_plot_maker(umap_data[[i]], color_against <- 'study_accession', red = 'UMAP') + 
+    theme(legend.position = 'none') +
+    ggtitle(paste(umap_data[[i]]$Method %>% unique(), 'dims:', umap_data[[i]]$dims%>% unique(), 'normalization:', umap_data[[i]]$norm %>% unique()))
+}
[…]

### 2021-07-25 manuscript_01_gigascience_resubmission.Rmd
table tweaks, text tweaks

@@
 The publication commit is #ffdf738.
-Furthermore, the repository has been deposited at Zenodo under 10.5281/zenodo.4638117 We will briefly discuss the pipeline choices, programs and algorithms, and versions below.
+Furthermore, the repository has been deposited at Zenodo under 10.5281/zenodo.5129265 We will briefly discuss the pipeline choices, programs and algorithms, and versions below.
 For the R packages, we provide package versions as the supplementary file "R_session_info.txt"
@@
 We have deposited the 10x (v2) fastq files for our iPSC-based RPE data in GEO at accession GSE180662.
-We give download links for the count matrices, seurat and anndata objects, cell level metadata, and our codebases in `supTab_cap(name = 'supTab_links', display = 'cite')`.
+We give download links for the count matrices, seurat and anndata objects, cell level metadata, and our codebases in `r supTab_cap(name = 'supTab_links', display = 'cite')`.
 
@@
 cat("<div id=\"tbl:supTab1_CTcounts\">")
-source('figs_and_tables/sup_fig__xgboost_PR.R')
-source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
-fontsize(ctTable  %>% padding(padding = 0, part = "body") %>% align(align = 'center', part = 'all') %>% autofit(), size = 8, part = 'all')
+suppressMessages( source('figs_and_tables/sup_fig__xgboost_PR.R'))
+suppressMessages( source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R'))
+fontsize(ctTable  %>% padding(padding = 0, part = "body") %>% align(align = 'center', part = 'all') %>% autofit(), size = 8, part = 'all') %>% fit_to_width(7)
 #Output Caption (* * for italics)
@@
 
-```{r supTab_SAcounts, fig.width=5, fig.height=14, out.width=300, fig.cap=supTab_SAcounts_cap, echo=FALSE, results='asis', message = FALSE, warning = FALSE}
+```{r supTab_SAcounts, fig.width=6, fig.height=14, out.width=300, fig.cap=supTab_SAcounts_cap, echo=FALSE, results='asis', message = FALSE, warning = FALSE}
 cat("<div id=\"tbl:supTab_SAcounts\">")
@@
 
-fontsize(saTable  %>% padding(padding = 0, part = "body") %>% align(align = 'center', part = 'all') %>% autofit(), size = 8, part = 'all') 
+fontsize(saTable  %>% padding(padding = 5, part = "body") %>% align(align = 'center', part = 'all'), size = 8, part = 'all') 
 #Output Caption (* * for italics)
@@
 
-```{r supTab_links, fig.width=2, fig.height=6, out.width=300, echo = F, results='asis', warning = F, fig.cap = supTab_links_cap, fig.pos = 'H'}
+```{r supTab_links, fig.width=3, fig.height=6, out.width=600, echo = F, results='asis', warning = F, fig.cap = supTab_links_cap, fig.pos = 'H'}
 supTab_links_cap <- supTab_cap(name='supTab_links', caption =  'Links to code and resources')
@@
                             "Codebase for scEiad manuscript",                                "https://github.com/davemcg/scEiaD_manuscript",          NA,
-                "Zenodo deposit of all above files and code",                                                      "10.5281/zenodo.4638117",   "4638117",
+                "Zenodo deposit of all above files and code",                                                      "https://zenodo.org/record/5129265",   "5129265",
                             "iPSC RPE scRNA Raw Fastq Files",                "https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE180662", "GSE180662"
                 )
-fontsize(flextable(linksTable)  %>% padding(padding = 0, part = "body") %>% align(align = 'left', part = 'all') %>% autofit(), size = 4, part = 'all')
+fontsize(flextable(linksTable)  %>% padding(padding = 5, part = "all") %>% align(align = 'left', part = 'all'), size = 8, part = 'all') %>% autofit() %>% fit_to_width(7)
 #Output Caption (* * for italics)

### 2021-07-27 manuscript_01_gigascience_resubmission.Rmd
text updates, added sup files to git

@@
 To provide a non-ocular reference we also downloaded the raw sequence data from the Tabula Muris project for re-processing.
-In cases where the fastq file from the SRA was not processed properly, we acquired the bam files (SRA or personal correspondence) and re-extracted the fastq.
+In cases where the fastq file from from the SRA was not processed properly (always 10x v2 or v3), we acquired the 10x bam files (via SRA or personal correspondence) and re-extracted the fastq.
 After downloading all the data we had 11 TB across `r scEiaD_2020_v01 %>% tbl('metadata_filter') %>% pull(sample_accession) %>% unique() %>% length()` fastq file sets.
@@
 After empty droplet removal, we had `r format(qc %>% nrow() %>% as.integer() , big.mark=",", scientific=FALSE)` cells.
-We then removed cells which had more than 10% mitochondrial reads across all gene counts or fewer than 200 unique genes quantified.
-After these standard quality control steps we were left with `r format(umapRef %>% nrow, big.mark = ",", scientific=FALSE)` cells (`r supFig_cap(name = 'supFig1', display = 'cite')`).
+We then removed cells which had more than 10% mitochondrial reads across all gene counts, fewer than 200 unique genes quantified, or were identified as an *in silico* doublet (see methods).
+After these quality control (for review see [@lueckenCurrentBestPractices2019] ) steps we were left with `r format(meta_filter %>% nrow, big.mark = ",", scientific=FALSE)` cells (`r supFig_cap(name = 'supFig1', display = 'cite')`).
 
@@
 We studied the metadata and methods of each study to identify the unique biological samples.
-Within the current scEiaD data set we identified `r umapRef %>% pull(batch) %>% unique() %>% length()` batches across `r umapRef %>% pull(study_accession) %>% unique() %>% length()` deposited datasets in `r umapRef %>% left_join(gse) %>% pull(PMID) %>% unique() %>% length()` published papers.
+Within the current scEiaD data set we identified `r umapRef %>% pull(batch) %>% unique() %>% length()` batches across `r umapRef %>% pull(study_accession) %>% unique() %>% length() - 1` deposited datasets in `r umapRef %>% left_join(gse) %>% pull(PMID) %>% unique() %>% length()` published papers.
 
 A wide variety of methods have been written for scRNA-seq integration.
-As we were uncertain which would perform the best, we ran `r methods %>% length()` tools with a commonly used set of key parameters like number of highly variable genes (HVG), size of latent dimensional space, and the number of nearest neighbors for the louvain clustering algorithm.
+As we were uncertain which would perform the best, we ran `r methods %>% length()` tools with a commonly used set of key parameters like number of highly variable genes (HVG), number of latent dimensional space, and the number of nearest neighbors for the louvain clustering algorithm.
 The Snakemake system was used to automate the running of the wide variety of tools.
@@
 In the same manner we plot the silhouette metric (`r supFig_cap(name = 'supFig_benchmark', display = 'cite')` b).
-To merge the different scores we define $sumZScale = \sum scale(m)$ where m is a metric (LISI by batch, LISI by cluster, LISI by cell type, silhouette by batch, silhouette by cluster, silhouette by celltype, NMI, and ARI).
+To merge the different scores we define $sumZScale = \sum scale(m * weight)$ where m is a metric (LISI by batch, LISI by cluster, LISI by cell type, silhouette by batch, silhouette by cluster, silhouette by celltype, NMI, and ARI).
+Weight is set to one, but can be either explicitly set or randomly chosen should it be desired to change the influence of certain metrics.
 
@@
 The other extreme is not using any batch integration method, where you see very distinct groups of cells, but also each nearly study is has a distinct region in the UMAP (`r fig_cap(name = 'figBenchmark', display = 'cite')`b).
-In our scEiaD dataset we see that like ComBat, Harmony and CCA are weighed more towards batch mixing then cluster and cell type purity.
-Scanorama, fastMNN, and bbknn prioritize cleanly separating the clusters.
+In our scEiaD dataset we see that methods like ComBat, Harmony, scArches, CCA are weighed more towards batch mixing then cluster and cell type purity.
+Scanorama and bbknn prioritize cleanly separating the clusters.
 With our scEiaD meta-atlas insct, trVAE, and magic do not perform particularly well in batch mixing or cluster purity.
-For further qualitative investigation of the integration performance we provide the UMAP visualizations of each integration method, normalization, and latent dimensions colored by cell type, study accession, or organism as a supplementary file.
+Overall, CCA, fastMNN, and scVI perform best on this particular dataset.
+For further qualitative investigation of the integration performance we provide the UMAP visualizations of each integration method, normalization, and latent dimensions colored by cell type, study accession, or organism as supplementary files.
 
 As the LISI and silhouette metrics provide independent cluster and cell type ("purity") and batch ("mixing") scores we looked to see whether the sumZScale scoring is highly influenced by changing the weight placed on purity or mixing by multiplying either by a multiplier of three (`r supFig_cap(name = 'supFig_benchmark_weighting', display = 'cite')`).
-While scVI still performs well, no matter the weight chosen, there were some changes when more weight was placed on batch mixing.
-Most notably the 8 latent dims with fastMNN receives a higher score when a higher weight was placed on batch mixing and CCA's performance with 30 latent dimensions jumps to the top.
-We also bootrap random weights and get similar results (`r supFig_cap(name = 'supFig_benchmark_weighting_ridges', display = 'cite')`).
+While scVI still performed well, no matter the weight chosen, there were some changes when more weight was placed on batch mixing.
+Most notably fastMNN with 8 latent dims performed better than fastMNN with 30 latent dims.
+In a reveral of the fastMNN results, CCA with 30 latent dimensions performed better than CCA with 8 latent dimensions.
+We also bootstrap random weights across each individual metric and get similar results (`r supFig_cap(name = 'supFig_benchmark_weighting_ridges', display = 'cite')`).
 Across the 11 integration methods we tested, scVI achieved the highest sumZScale for our scEiaD meta-atlas and furthermore is a desirable choice because of its very short run time (scVI can complete in less than an hour while fastMNN takes over 12 hours and CCA takes more than 24 hours).
@@
 There are several normalization approaches that have been used or published.
-The "standard" approach that the popular analysis packages Seurat and anndata use by default is to, per cell, divide the counts by the sum counts for the cell, multiply by a scaling factor, then log transform.
+The "standard" approach that the popular analysis packages Seurat and scanpy use by default is to, per cell, divide the counts by the sum counts for the cell, multiply by a scaling factor, then log transform.
 This helps make the count distribution more normal, which is an assumption that many algorithms require.
@@
   plot_grid(
-    umap_plot_maker(umapCombat), 
-    umap_plot_maker(umapNone, color_against = 'study_accession'),
+    umap_plot_maker(umapCombat) + guides(color=guide_legend(title="cell type", 
+                                                            override.aes = list(size=7, alpha = 1))), 
+    umap_plot_maker(umapNone, color_against = 'study_accession') + guides(color=guide_legend(title="study accession",
+                                                                                             override.aes = list(size=7, alpha = 1))),
     rel_widths = c(1.3, 1),
@@
 
-As a brief case study of our ML performance, we look at the cell type labels assigned to a Shekhar et al.
+As a brief case study of our ML performance, we look at the cell type labels assigned to the Shekhar et al.
 study where they used SMART-seq on a retinal bipolar cell enriched population of cells [@shekharComprehensiveClassificationRetinal2016] .
@@
 Well based samples were quantified using kallisto quant.
-For the relatively few single ended samples, the params “–single -l 200 -s 30” were used for kallisto quant.
-Otherwise the “–bias” flag was added.
+For the relatively few single ended samples, the params “--single -l 200 -s 30” were used for kallisto quant.
+Otherwise the “--bias” flag was added.
 For the droplet-based samples, we adapted the bustools workflow for generating spliced and unspliced count matrices.
@@
 The rapid of advancement of algorithms to batch correct and process data continue to reduce the computational requirements to handle huge numbers of cells.
-The scVI batch correction step on around one million cells runs within a few hours on a GPU and 150GB of memory.
+The scVI batch correction step with around one million cells runs within an hour on a GPU and 150GB of memory.
 This places this crucial step within the capabilities of a moderately powerful computer or a cloud compute node.
-Further downstream processing can largely be done a computers with 64+ Gb of memory and a few hundred GB of disk space.
+Further downstream processing can largely be done a computers with 64+ GB of memory and a few hundred GB of disk space.
 We believe that our efforts can be replicated in any other tissue / system with a large number of independent studies by a small number of computational scientists following our general approach.
-We provide the completed analysis as both Scanpy (h5ad) and Seurat objects at https://github.com/davemcg/scEiaD.
+We provide the completed analysis as both Scanpy (h5ad) and Seurat objects (`r supTab_cap(name = 'supTab_links', display = 'cite')`).
 
@@
 While we chose the scVI algorithm, we strongly suggest any other groups attempting a similar meta-atlas construction chose a quantifiable set of criteria so optimal methods and parameters can be picked.
+We found in our analysis that substantial differences in the integration performance can come from altering the number of HVGs and the number of latent dimensions.
 
@@
 our dataset/atlas is several times larger than any other published set, 2.
-our data is available via download in several forms at https://github.com/davemcg/scEiaD, and crucially 3.
+our data is available via download in several forms at https://github.com/davemcg/scEiaD and Zenodo accession 5129265, and crucially 3.
 we provide a Google colab/Jupyter notebook which step-by-step demonstrates out how to use scVI to project (or query) outside data onto our scEiaD with minimum compute resources.
-We demonstrate concretely how this can work by showing how iPSC RPE can be queried onto the reference dataset to demonstrate both similarities and dissimilarities in their transcriptomes.
+We demonstrate concretely how this can work by showing how iPSC-based RPE can be queried onto the reference dataset to demonstrate both similarities and dissimilarities in their transcriptomes.
 

### 2021-07-27 manuscript_01_gigascience_resubmission.Rmd
acknowlegemnt added, revision2 docx added

@@
 
+# Acknowledgments
+
+We would like the thank the many groups who provided the raw data required to create this project.
+We keep a updated list of citations for the projects we pulled data from at https://plae.nei.nih.gov.
+We would also like to thank Adam Gayoso, who provided many useful comments on scVI parameter behavior and the Google colab implementation of scVI.
+Finally, this work utilized the computational resources of the NIH HPC Biowulf cluster (http://hpc.nih.gov).
+
 # Works Cited
